---
title: "BCB744 Task B"
format: 
  html:
    fig-format: svg
    fig_retina: 2
    fig-dpi: 400
params: 
  hide_answers: true
execute:
  fig-format: png
  dev: ragg_png
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

```{r code-knitr-opts-chunk-set, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.asp = 0.65,
  out.width = "100%",
  fig.align = "center"
)
library(tidyverse)
library(ggpubr)
library(ggthemes)
```

# [[Assessment Sheet](BCB744_Task_B_Surname.xlsx)]{.my-highlight} {#sec-assessment}

# 5. R and RStudio (Continue)

```{r code-crops-readxl-read-excel-here, echo=FALSE, output=FALSE}
# Read the Excel file into R
crops <- readxl::read_excel(here::here("data", "crops.xlsx"))
```

## Question 18

Explain in words what the pipe operator `%>%` does in R. How does it make your code more readable? **(/3)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

- ✓ The pipe operator `%>%` (or `|>`) in R is used to chain together multiple functions or operations in a sequence. It takes the output of one function and passes it as the first argument to the next function, allowing you to create a pipeline of operations.
- ✓ The pipe operator makes your code more readable by breaking down complex operations into a series of simpler steps. It helps in avoiding nested function calls, improves code clarity, and reduces the need for intermediate variables.
- ✓ In this way you can write code in a more linear and intuitive way, following the flow of data transformations from one step to the next. This makes it easier to understand the logic of the code, the sequence of operations being performed.

`r if (params$hide_answers) ":::"`

## Question 19

Using the various **tidyverse** functions, calculate the mean ± SD for the crop mass within each combination of `block` and `fertiliser` of the `crops` dataset. **(/5)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r code-library-tidyverse}
# Load the tidyverse package
library(tidyverse)

# Calculate the mean ± SD for crop mass within each combination of block and fertiliser
crops %>%
  group_by(block, fertilizer) %>%
  summarise(mean_mass = mean(mass), sd_mass = sd(mass))
```

`r if (params$hide_answers) ":::"`

# 6. Faceting Figures

## Question 1

Create a scatterplot of `bill_length_mm` against `bill_depth_mm` for `Adelie` penguins on `Biscoe` island. **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-library-palmerpenguins, fig.width=5, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Adelie bill length vs. bill depth on Biscoe Island."
library(palmerpenguins) # ✓ 
library(tidyverse) # ✓ 
data(penguins)
head(penguins)

penguins %>% # ✓ 
  filter(island == "Biscoe" & species == "Adelie") %>%  # ✓ 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +  # ✓ 
  geom_point() +  # ✓ 
  labs(title = "Adelie Penguins on Biscoe Island",  # ✓ 
         x = "Bill Length (mm)", # ✓ 
         y = "Bill Depth (mm)") # ✓ 
```


`r if (params$hide_answers) ":::"`

## Question 2

Create histograms of `bill_length_mm` for `Adelie` penguins on all three islands (one figure per island). Save each figure as a separate R object which you can later reuse. Again for `Adelie` penguins, create a boxplot for `bill_length_mm` showing all the data on one plot. Save it too as an R object. Combine the four saved figures into one figure using `ggarrange()`. **(/25)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-library-ggpubr, fig.width=8, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Adelie bill length histograms by island plus overall boxplot."
library(ggpubr) # ✓

# Create histograms
adelie_biscoe <- penguins %>% # ✓ x 5
  filter(island == "Biscoe" & species == "Adelie") %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram() + 
  labs(title = "Adelie Penguins on Biscoe Island", 
       x = "Bill Length (mm)", 
       y = "Frequency")

adelie_dream <- penguins %>% # ✓ x 5
  filter(island == "Dream" & species == "Adelie") %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram() + 
  labs(title = "Adelie Penguins on Dream Island", 
       x = "Bill Length (mm)", 
       y = "Frequency")

adelie_torgersen <- penguins %>% # ✓ x 5
  filter(island == "Torgersen" & species == "Adelie") %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram() + 
  labs(title = "Adelie Penguins on Torgersen Island", 
       x = "Bill Length (mm)", 
       y = "Frequency")

# Create boxplot # ✓ x 5
adelie_boxplot <- penguins %>% 
  filter(species == "Adelie") %>% 
  ggplot(aes(x = island, y = bill_length_mm)) + 
  geom_boxplot() + 
  labs(title = "Adelie Penguins Bill Length Boxplot", 
       x = "Island", 
       y = "Bill Length (mm)")

# Combine figures # ✓ x 1
ggarrange(adelie_biscoe, adelie_dream, adelie_torgersen, adelie_boxplot, 
          ncol = 2, nrow = 2)
```


`r if (params$hide_answers) ":::"`

## Question 3

Create a scatter plot of `flipper_length_mm` against `body_mass_g` and use `facet_wrap()` to create separate panels for each island (combine all species). Also indicate the effect of species. Add a best-fit straight line with 95% confidence intervals through the points, ignoring the effect of species. Take into account which variable best belongs on `x` and `y`. Describe your findings. **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-penguins-x, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Flipper length vs. body mass faceted by island and coloured by species."
penguins %>% # ✓ x 6
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point(aes(colour = species)) + 
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~island) + 
  labs(title = "Flipper Length vs Body Mass", 
       x = "Body Mass (g)",
       y = "Flipper Length (mm)")
```

- The `body_mass_g` variable is best suited to the `x`-axis as it is the independent variable. The `flipper_length_mm` variable is best suited to the `y`-axis as it is the dependent variable.
- ✓ For all penguin species, the `flipper_length_mm` and `body_mass_g` variables show a positive correlation, with larger penguins having longer flippers, higher body masses.
- ✓ The `Adelie` penguins on `Biscoe` island have the shortest flippers and lowest body masses, while `Gentoo` penguins have the longest flippers and highest body masses.
- ✓ `Chinstrap` and `Adelie` penguins are present on `Dream` island; these species' body masses and flipper lengths are difficult to distinguish from one-another.
- ✓ Only `Adelie` penguins are present on `Torgersen` island. The `Adelie` penguins appear to have the same flipper length vs body mass relationship across all three islands.

`r if (params$hide_answers) ":::"`

## Question 4

Create a scatter plot of `bill_length_mm` and `body_mass_g` and use `facet_grid()` to create separate panels for each species and island. **(/6)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-grid-plt-penguins-x, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Bill length vs. body mass faceted by species and island."
grid_plt <-  penguins %>% # ✓ x 6
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) + 
  geom_point() + 
  facet_grid(species ~ island) + 
  labs(title = "Bill Length vs Body Mass", 
       x = "Body Mass (g)",
       y = "Bill Length (mm)")
grid_plt
```


`r if (params$hide_answers) ":::"`

## Question 5

Using the figure created in point 4, also show the effect of `sex` and add a best-fit straight line. Explain the findings. **(/9)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r code-grid-plt-x, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
grid_plt +  # ✓ x 6
  geom_point(aes(col = sex)) +
  geom_smooth(method = "lm", se = TRUE, colour = "black")
```

- ✓ The `bill_length_mm` and `body_mass_g` variables show a positive correlation, with larger penguins having longer bills, higher body masses.
- ✓ The `sex` variable appear to have an effect on the relationship between `bill_length_mm` and `body_mass_g`, with male penguins tending to be heavier with longer bill lengths.
- ✓ There also appears to be differences in the relationship between `bill_length_mm` and `body_mass_g` between the different species and islands.

`r if (params$hide_answers) ":::"`

## Question 6

What are the benefits of using faceting in data visualisation? **(/3)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

- ✓ Faceting allows for the visualisation of multiple relationships in a single plot, making it easier to compare relationships between different groups.
- ✓ Faceting can help to identify patterns and trends in the data that may not be immediately obvious when looking at the data as a whole.
- ✓ Faceting can help to identify differences in relationships between different groups, such as species, islands or allowing for more detailed analysis of the data.

`r if (params$hide_answers) ":::"`
