---
title: "BCB744 Bonus Task"
format: 
  html:
    fig-format: png
    fig_retina: 2
    fig-dpi: 400
execute:
  dev: ragg_png
  fig-format: png
params: 
  hide_answers: true
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

# [[Assessment Sheet](BCB744_Task_Bonus_Surname.xlsx)]{.my-highlight} {#sec-assessment}

# 12--14. Tidy Data

## Question 1

What are the key principles of tidy data? **(/3)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

- ✓ Each variable forms a column.
- ✓ Each observation forms a row.
- ✓ Each type of observational unit forms a table.

```{r code-chunk, echo=FALSE, eval=FALSE, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="80%", fig.align='center'}

```

`r if (params$hide_answers) ":::"`

## Question 2

Using the untidy data (`SACTN2`) and the tidy data (`SACTN2_tidy`), create line graphs, one for each of DEA, SAWS, and KZNSB, showing a time series of temperature. Ensure you have a column of three figures (ncol = 1). Use the fewest number of lines of code possible. You should end up with two graphs, each with three panels. **(/13)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-library-tidyverse, echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="80%", fig.align='center'}
#| fig-cap: "SACTN2 temperature time series faceted by source."
library(tidyverse)

load(here::here("data", "SACTN_mangled.RData")) # ✓

SACTN2_tidy <- pivot_longer(SACTN2, cols = c("DEA", "KZNSB", "SAWS"),
                            names_to = "src",
                            values_to = "temp") # ✓

# Starting with SACTN2: one could be sneaky and cheat by using
# 'pivot_wider()' in the pipeline
SACTN2 |>  # ✓ x 6
  pivot_longer(cols = c("DEA", "KZNSB", "SAWS"),
               names_to = "src",
               values_to = "temp") |> 
  ggplot(aes(x = date, y = temp)) +
  geom_line(aes(col = site, linetype = type)) +
  facet_wrap(~ src, ncol = 1) +
  labs(title = "Untidy Data", # ✓
       x =  "Date", y = "Temperature (°C)")

# For the untidy data, above, I will also allocate marks if you insisted in 
# creating more work for yourself by doing it the long way.... ( # ✓ x 7)

# Starting with SACTN2_tidy
# (creates an identical plot)
ggplot(data = SACTN2_tidy, aes(x = date, y = temp)) +  # ✓ x 4
  geom_line(aes(col = site, linetype = type)) +
  facet_wrap(~ src, ncol = 1) +
  labs(title = "Tidy Data", # ✓
       x =  "Date", y = "Temperature (°C)")
```

`r if (params$hide_answers) ":::"`
