---
title: "BCB744 Task B"
format: 
  html:
    fig-format: svg
    fig_retina: 2
    fig-dpi: 300
    number-sections: false
params: 
  hide_answers: true
execute:
  fig-format: png
  dev: ragg_png
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

```{r code-knitr-opts-chunk-set, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.asp = 0.65,
  out.width = "100%",
  fig.align = "center"
)
library(tidyverse)
library(ggpubr)
library(ggthemes)
```

# [[Assessment Sheet](BCB744_Task_B_Surname.xlsx)]{.my-highlight} {#sec-assessment}

# 6-8. Graphics With ggplot2, Faceting Figures, and Brewing Colours

In these sets of tasks you will generate several figures. For every figure generated, also please provide a narrative explanation of the patterns shown within the figures (sometimes this is mentioned explicitely, but not always; describe your figure(s) in EVERY question).

## Question 1

Create a scatterplot of `bill_length_mm` against `bill_depth_mm` for `Adelie` penguins on `Biscoe` island. **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-library-palmerpenguins, fig.width=5, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Adelie bill length vs. bill depth on Biscoe Island."
library(palmerpenguins) # ✓ 
library(tidyverse) # ✓ 
data(penguins)
head(penguins)

penguins %>% # ✓ 
  filter(island == "Biscoe" & species == "Adelie") %>%  # ✓ 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm)) +  # ✓ 
  geom_point() +  # ✓ 
  labs(title = "Adelie Penguins on Biscoe Island",  # ✓ 
         x = "Bill Length (mm)", # ✓ 
         y = "Bill Depth (mm)") # ✓ 
```


`r if (params$hide_answers) ":::"`

## Question 2

Create histograms of `bill_length_mm` for `Adelie` penguins on all three islands (one figure per island). Save each figure as a separate R object which you can later reuse. Again for `Adelie` penguins, create a boxplot for `bill_length_mm` showing all the data on one plot. Save it too as an R object. Combine the four saved figures into one figure using `ggarrange()`. **(/25)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-library-ggpubr, fig.width=8, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Adelie bill length histograms by island plus overall boxplot."
library(ggpubr) # ✓

# Create histograms
adelie_biscoe <- penguins %>% # ✓ x 5
  filter(island == "Biscoe" & species == "Adelie") %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram() + 
  labs(title = "Adelie Penguins on Biscoe Island", 
       x = "Bill Length (mm)", 
       y = "Frequency")

adelie_dream <- penguins %>% # ✓ x 5
  filter(island == "Dream" & species == "Adelie") %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram() + 
  labs(title = "Adelie Penguins on Dream Island", 
       x = "Bill Length (mm)", 
       y = "Frequency")

adelie_torgersen <- penguins %>% # ✓ x 5
  filter(island == "Torgersen" & species == "Adelie") %>% 
  ggplot(aes(x = bill_length_mm)) + 
  geom_histogram() + 
  labs(title = "Adelie Penguins on Torgersen Island", 
       x = "Bill Length (mm)", 
       y = "Frequency")

# Create boxplot # ✓ x 5
adelie_boxplot <- penguins %>% 
  filter(species == "Adelie") %>% 
  ggplot(aes(x = island, y = bill_length_mm)) + 
  geom_boxplot() + 
  labs(title = "Adelie Penguins Bill Length Boxplot", 
       x = "Island", 
       y = "Bill Length (mm)")

# Combine figures # ✓ x 1
ggarrange(adelie_biscoe, adelie_dream, adelie_torgersen, adelie_boxplot, 
          ncol = 2, nrow = 2)
```


`r if (params$hide_answers) ":::"`

## Question 3

Create a scatter plot of `flipper_length_mm` against `body_mass_g` and use `facet_wrap()` to create separate panels for each island (combine all species). Plot the three species as **distinct point shapes**, and map a **continuous colour scale** to `bill_length_mm`. Add a best‑fit straight line with 95% confidence intervals through the points, ignoring the effect of species. Take into account which variable best belongs on `x` and `y`. Describe your findings. **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-penguins-x, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Flipper length vs. body mass faceted by island with species shapes and bill-length colour scale."
penguins %>% # ✓ x 7
  ggplot(aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point(aes(shape = species, colour = bill_length_mm)) + 
  scale_colour_viridis_c() +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~island) + 
  labs(title = "Flipper Length vs Body Mass", 
       x = "Body Mass (g)",
       y = "Flipper Length (mm)",
       colour = "Bill length (mm)")
```

- The `body_mass_g` variable is best suited to the `x`-axis as it is the independent variable. The `flipper_length_mm` variable is best suited to the `y`-axis as it is the dependent variable.
- ✓ For all penguin species, the `flipper_length_mm` and `body_mass_g` variables show a positive correlation, with larger penguins having longer flippers, higher body masses.
- ✓ The colour scale shows that bill length tends to increase with body mass, especially for Gentoo penguins.
- ✓ Species are clearly separated by shape, and Gentoo penguins occupy the largest body‑mass and flipper‑length ranges.

`r if (params$hide_answers) ":::"`

## Question 4

Create a scatter plot of `bill_length_mm` and `body_mass_g` and use `facet_grid()` to create separate panels for each species and island. Map `bill_length_mm` to a **continuous colour scale** that you customise yourself (do not use the default palette). **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-grid-plt-penguins-x, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Bill length vs. body mass faceted by species and island with a custom colour scale."
grid_plt <-  penguins %>% # ✓ x 7
  ggplot(aes(x = body_mass_g, y = bill_length_mm, colour = bill_length_mm)) + 
  geom_point() + 
  scale_colour_gradient(low = "lightyellow", high = "darkred") +
  facet_grid(species ~ island) + 
  labs(title = "Bill Length vs Body Mass", 
       x = "Body Mass (g)",
       y = "Bill Length (mm)",
       colour = "Bill length (mm)")
grid_plt
```


`r if (params$hide_answers) ":::"`

## Question 5

Using the figure created in point 4, also show the effect of `sex` and add a best-fit straight line. Explain the findings. **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r code-grid-plt-x, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
grid_plt +  # ✓ x 7
  geom_point(aes(shape = sex)) +
  geom_smooth(method = "lm", se = TRUE, colour = "black")
```

- ✓ The `bill_length_mm` and `body_mass_g` variables show a positive correlation, with larger penguins having longer bills, higher body masses.
- ✓ The `sex` variable appears to have an effect on the relationship between `bill_length_mm` and `body_mass_g`, with male penguins tending to be heavier with longer bill lengths.
- ✓ There also appears to be differences in the relationship between `bill_length_mm` and `body_mass_g` between the different species and islands.

`r if (params$hide_answers) ":::"`

## Question 6

What are the benefits of using faceting in data visualisation? **(/3)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

- ✓ Faceting allows for the visualisation of multiple relationships in a single plot, making it easier to compare relationships between different groups.
- ✓ Faceting can help to identify patterns and trends in the data that may not be immediately obvious when looking at the data as a whole.
- ✓ Faceting can help to identify differences in relationships between different groups, such as species, islands or allowing for more detailed analysis of the data.

`r if (params$hide_answers) ":::"`

## Question 7

Use the built-in `ToothGrowth` dataset (guinea pig tooth length) to create a scatter plot of `len` against `dose`, coloured by `supp`, and faceted by `supp`. Add a best‑fit straight line with a 95% confidence interval. **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-toothgrowth-scatter, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Tooth length vs. dose, coloured and faceted by supplement."
data(ToothGrowth) # ✓

ToothGrowth %>% # ✓ x 6
  ggplot(aes(x = dose, y = len, colour = supp)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = TRUE) + 
  facet_wrap(~supp) + 
  labs(title = "Tooth Length vs Dose by Supplement",
       x = "Dose (mg/day)",
       y = "Tooth length")
```

- ✓ `dose` is the explanatory variable, so it belongs on the x‑axis; `len` is the response variable.
- ✓ Tooth length increases with dose for both supplements.

`r if (params$hide_answers) ":::"`

## Question 8

Create histograms of `len` for each `dose` (all supplements together) using `facet_wrap()`. **(/6)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-toothgrowth-hist, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Tooth length distributions by dose."
ToothGrowth %>% # ✓ x 4
  ggplot(aes(x = len)) + 
  geom_histogram(bins = 8) + 
  facet_wrap(~dose) + 
  labs(title = "Tooth Length by Dose",
       x = "Tooth length",
       y = "Count")
```

`r if (params$hide_answers) ":::"`

## Question 9

Create boxplots of `len` by `dose` and facet by `supp` (one panel per supplement). **(/8)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-toothgrowth-box, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Tooth length by dose, separated by supplement."
ToothGrowth %>% # ✓ x 5
  ggplot(aes(x = factor(dose), y = len)) + 
  geom_boxplot() + 
  facet_grid(supp ~ .) + 
  labs(title = "Tooth Length by Dose and Supplement",
       x = "Dose (mg/day)",
       y = "Tooth length")
```

`r if (params$hide_answers) ":::"`

## Question 10

Calculate the mean ± SD of `len` for each combination of `dose` and `supp`, then plot the means with error bars. **(/10)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-toothgrowth-mean-sd, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Mean tooth length ± SD by dose and supplement."
tg_summary <- ToothGrowth %>% # ✓ x 7
  group_by(dose, supp) %>% 
  summarise(mean_len = mean(len),
            sd_len = sd(len),
            .groups = "drop")

tg_summary %>% 
  ggplot(aes(x = factor(dose), y = mean_len, colour = supp, group = supp)) + 
  geom_point(size = 2) + 
  geom_line() + 
  geom_errorbar(aes(ymin = mean_len - sd_len, ymax = mean_len + sd_len), width = 0.15) + 
  labs(title = "Mean Tooth Length ± SD",
       x = "Dose (mg/day)",
       y = "Mean tooth length")
```

`r if (params$hide_answers) ":::"`

## Question 11

Create a violin plot of `len` by `dose`, filled by `supp`, and facet by `supp`. **(/8)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r fig-toothgrowth-violin, warning=FALSE, message=FALSE, fig.width=6, fig.asp=0.65, out.width="60%", fig.align='center'}
#| fig-cap: "Tooth length distributions by dose and supplement."
ToothGrowth %>% # ✓ x 5
  ggplot(aes(x = factor(dose), y = len, fill = supp)) + 
  geom_violin(trim = FALSE) + 
  facet_wrap(~supp) + 
  labs(title = "Tooth Length Distributions by Dose",
       x = "Dose (mg/day)",
       y = "Tooth length")
```

`r if (params$hide_answers) ":::"`

## Question 12

Create a small summary table showing the number of observations for each combination of `dose` and `supp`. **(/6)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r code-toothgrowth-count}
ToothGrowth %>% # ✓ x 3
  count(dose, supp)
```

`r if (params$hide_answers) ":::"`

## Question 13

Briefly describe two patterns you observe in any of the figures above. **(/4)**

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

- ✓ Tooth length increases as dose increases for both supplements.
- ✓ At the same dose, the `OJ` supplement tends to have higher tooth lengths than `VC`, especially at lower doses.

`r if (params$hide_answers) ":::"`
