---
date: "2021-01-01"
title: "16. Recap"
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

> "*Everyone should have their mind blown once a day.*"
>
> --- Neil deGrasse Tyson
  
> "*Somewhere, something incredible is waiting to be known.*"
>
> --- Carl Sagan



# Workshop Recap, Assessment Alignment, and What Comes Next

Over the course of this workshop, you have learned not just *how* to use R, but *how to think* in a tidy, reproducible, and analytical way. These skills are foundational for all subsequent assessments and for the Biostatistics component that follows.

# Alignment with Assessments

The workshop has been deliberately structured to map directly onto your assessed work. Each assessment assumes that you can independently apply the following skills:

* **Assessment readiness**

  * Import, inspect, and tidy real datasets
  * Apply a coherent workflow from raw data to final output
  * Write readable, reproducible R code using tidyverse principles
  * Produce publication-quality figures using `ggplot2`
  * Manipulate data confidently using `dplyr` verbs and pipes

* **What assessors will look for**

  * Logical data workflows (not trial-and-error code)
  * Clear transformation steps (`filter()`, `mutate()`, `summarise()`, `group_by()`)
  * Appropriate visualisation choices
  * Evidence that results were *derived*, not manually curated
  * Liberal application of comments to document your workflow, *i.e.*, code that is understandable by someone else (including your future self)

If you can reproduce the analyses and figures from this workshop *without following along line-by-line*, you are well prepared for the assessments.

# Concept Map: How the Chapters Fit Together

This workshop is not a collection of isolated lessons. You should be able to combine lessons learned in each chapter, because they play specific roles in a single, coherent analytical framework/workflow:

1. **R and RStudio**
   *Orientation* — learning the environment, tools, and expectations of working in R.

2. **Working with Data and Code**
   *Foundations* — understanding scripts, objects, and how R thinks.

3. **R Markdown and Quarto**
   *Reproducibility* — integrating code, results, and narrative into a single document.

4. **Data Classes and Structures**
   *Literacy* — knowing what your data *is* before deciding what to do with it.

5. **R Workflows**
   *Discipline* — structuring analyses so they are repeatable and scalable.

6. **Graphics with ggplot2**
   *Visual reasoning* — learning to explore and communicate data visually.

7. **Faceting Figures**
   *Comparison* — revealing patterns across groups and conditions.

8. **Brewing Colours**
   *Clarity and accessibility* — making figures interpretable and professional.

9. **Mapping with ggplot2**
   *Spatial thinking* — extending tidy principles to geographic data.

10. **Mapping with Style**
    *Polish* — producing maps suitable for reports and publications.

11. **Mapping with Natural Earth / Applied Examples**
    *Integration* — combining data sources, projections, and styling.

13. **Tidy Data**
    *Structure* — learning the rules that make analysis possible.

14. **Tidier Data**
    *Transformation* — filtering, mutating, selecting, and summarising.

15. **Tidiest Data**
    *Power* — grouping, pipelines, and complex workflows.

16. **Recap**
    *Synthesis* — seeing the workflow as a single analytical language.

Together, these chapters teach you how to move from **messy reality → structured data → insight → communication**.

# Prelude to Biostatistics

The Biostatistics component builds directly on everything you have learned here. For those of you taking BCB743 as an elective, this work will also be foundational.

In this workshop, you focused on:

* *How to prepare data*
* *How to explore patterns*
* *How to visualise structure and variation*

In Biostatistics, you will now ask:

* *Are these patterns meaningful?*
* *How much uncertainty is there?*
* *What conclusions are supported by the data?*

Conceptually, the transition looks like this:

* **Tidy data** → prerequisite for valid statistics
* **Grouping and summarising** → foundation of statistical models
* **Visual exploration** → guides hypothesis formulation
* **Reproducible workflows** → ensures transparent inference

Statistical tests, models, and confidence intervals only make sense when applied to well-structured, well-understood data. You now have the tools to ensure that this condition is met.

Think of this workshop as learning the *grammar* of data analysis. Biostatistics is where you begin to write *arguments*.

# Final Note

You are not expected to memorise functions. You *are* expected to recognise workflows, patterns, and logic.

Confidence in R comes not from speed, but from clarity.

If your code reads like a story of what you did and why—you are doing it right.


# Session Info
```{r code-installed-packages-names-sessioninfo, echo=TRUE, include=TRUE}
installed.packages()[names(sessionInfo()$otherPkgs), "Version"]
```
