---
title: "Correlations"
subtitle: "Part 2"  
author: 
  - "AJ Smit"
date: '2020/06/28 (updated: `r Sys.Date()`)'
output:
  xaringan::moon_reader:
    css:
      - default
      - "default-fonts.css"
      - ki
      - "uwc_logo.css"
    yolo: false
    self_contained: TRUE  
    lib_dir: libs
    nature:
      highlightStyle: idea
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include = FALSE, echo = FALSE} 
library(tidyverse)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE,
                      comment = "R>",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      fig.retina = 3,
                      fig.width = 6,
                      fig.height = 3.6,
                      fig.align = "center")
```

```{r, echo=FALSE, include=FALSE}
data(iris)
setosa <- iris %>%
  filter(Species == "setosa") %>%
  select(-Species)
```

class: center, middle
# Doing the correlation

---
## What function do we use to do the correlation?

The name of a very basic function for a correlation test is `cor()`.

Let's find some help on the function first:
```{r, eval = FALSE}
?cor # or,
help(cor)
```

---
## How do we do the Pearson Product Moment correlation?

The equation for the Pearson's correlation coefficient is:

$$r = \frac{\sum_{i=1}^{n}(x_{i}-\bar{x})(y_{i}-\bar{y})}{\sqrt{\sum_{i=1}^{n}(x_{i}-\bar{x})^2\sum_{i=1}^{n}(y_{i}-\bar{y})}}$$

Where $\bar{x}$ and $\bar{y}$ are the means for the X and Y variables, respectively.

The default for `cor()` is to fit a Pearson correlation, so we may omit the `method` argument:

```{r}
with(setosa, cor(Sepal.Length, Sepal.Width))
```

## What does the output mean?

The output and interpretation are simple. There are not associated *p*-values. There are no associated hypothesis tests. It simply tells of about the strength of association between the two variables.

---
## A more comprehensive correlation

We can do a more detailed correlation using `cor.test()`:

```{r}
with(setosa, cor.test(x = Sepal.Length, Sepal.Width))
```

---
## What if want to see the association between all the variables in the `setosa` dataset?

For this we can rely on the `cor()` function again:

```{r}
setosa_pearson <- cor(setosa)
setosa_pearson
```

---
## How would we visualise all these associations?

In the Iris dataset, above, we compared associations for each pair of the following columns: `Sepal.Length`, `Sepal.Width`, `Petal.Length`, and `Petal.Width`. This required six pairs of correlations, which would be a pain if we wanted to create a visualisation for each of the six pairs. We can do it quickly, e.g.:

```{r}
ecklonia <- read_csv("../data/ecklonia.csv") %>% 
  select(-species, - site, - ID)
head(ecklonia)
```

---
```{r, fig.width = 7, fig.height = 7}
ecklonia_pearson <- cor(ecklonia)
library(corrplot)
corrplot(ecklonia_pearson, method = "circle", type = "lower",
         number.digits = 2, addCoef.col = "salmon", tl.col = "black")
```

---
```{r, fig.width = 10, fig.height = 7}
pairs(data = ecklonia, ~ stipe_length + stipe_diameter + frond_length + primary_blade_length +
        primary_blade_width + stipe_mass + frond_mass)
```

---
## What if we had ordinal data instead of continuous data?

Ordinal data are ordered categorical factors; in other words, the data are rank ordered. Let us create a test dataset:

```{r}
lungs <- read_tsv("../data/LungCapData.csv") %>%
  mutate(size_class_intervals = as.factor(cut(Height, breaks = 4)),
         size_class = cut(Height, breaks = 4, labels = c("infant", "toddler", "adolecent", "teen"),
                          ordered = TRUE))
head(lungs)
```

For more information about creating categorical data from continuous data, see <https://www.youtube.com/watch?v=EWs1Ordh8nI>.

---
```{r}
is.ordered(lungs$size_class)
```

```{r}
head(as.numeric(lungs$size_class), 111)
```

---
The equation for Spearman's *rho* is:

$$r = \frac{\sum_{i=1}^{n}(x_{i}'-\bar{x})(y_{i}'-\bar{y})}{\sqrt{\sum_{i=1}^{n}(x_{i}'-\bar{x})^2\sum_{i=1}^{n}(y_{i}'-\bar{y})}}$$

Where $x_{i}'$ and $y_{i}'$ are the ranks for each observation in the X and Y variables, respectively.

## How do we apply a Spearman's *rho* correlation by ranks?

```{r}
cor.test(as.numeric(lungs$size_class), lungs$LungCap, method = "spearman")
```

---
.left-column[## Questions]

.right-column[
Find your own data set suitable for a correlation analysis.

- State the null and alternative hypotheses. Do a Pearson's correlation. Explain the findings.
- Do the tests necessary to evaluate the various assumptions for the above analysis.
- Create all the associated figures for the above analysis.
- Transform one of the variables to ordinal data, and do a Spearman's *rho* or Kendall's *tau* correlation. Explain the findings.
]

---
.right-column[
- Do a full correlation analysis on the full `iris` dataset (i.e. all three species):

    - state all null hypotheses;
    
    - test all assumptions;
    
    - create all necessary plots;
    
    - write a few sentences on the findings.
]
