---
date: last-modified
title: "Occupancy modelling with imperfect detection"
author: L. Dlelapantsi & T.S Selae
format:
  html: default
  pdf: default
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

<!-- # Topic 22: Occupancy models with imperfect detection -->

::: callout-tip
## **Material Required for This Chapter**

| Type        | Name                                 | Link                                                                |
| :--- | :--- | :--- |
| **Slides**  | Occupancy lecture slides            | [💾 `Occupancy_model_with_imperfect_detection.pdf`](.../slides/BCB743_occupancy.pdf) |                                                                                |
| **Theory**  | MacKenzie et al. (2002)                           | [Ecology 83:2248–2255](https://doi.org/10.1890/0012-9658(2002)083[2248:ESORWD]2.0.CO;2) |
| **Theory**  | Becker et al. (2022)                              | [Ecological Applications 32:e2502](https://doi.org/10.1002/eap.2502)                 |
| **Package** | `unmarked`                                        | [CRAN](https://cran.r-project.org/web/packages/unmarked/index.html)                  |
| **Data**    | *Capensibufo rosei*     | [💾 `All_spp_search_data.csv`](https://zivahub.uct.ac.za/articles/dataset/Finding_rare_species_and_estimating_the_probability_that_all_occupied_sites_have_been_found_Dataset/14609838/1?file=28048122)             |              |
           |              |
:::

## Overview

In ecology, occupancy models are frequently used to track biodiversity, investigate rare and endangered species, and understand how species distributions change over time. By taking imperfect detection into account, these models offer more accurate estimates of where species occur and can help identify significant habitats and regions in need of conservation. Occupancy models can also be used to investigate the effects of environmental conditions on species occurrence, including pollution, climate, habitat quality, and human disturbance. While multi-species models can be used to examine patterns of biodiversity and community composition, multi-season models enable researchers to examine changes in occupancy through processes like colonisation and local extinction. These models are useful for evaluating conservation interventions, such as the establishment of protected areas, by tracking changes in species occupancy before and after management action.

### Learning Objectives

By the end of this chapter, students should be able to:

- Explain why imperfect detection can bias biodiversity surveys and how occupancy models address this problem.
- Distinguish between occupancy probability (ψ) and detection probability (p), and describe the ecological processes they represent.
- Construct and interpret single-season occupancy models using environmental and detection covariates.
- Evaluate model outputs, including parameter estimates, confidence intervals, goodness-of-fit, and model selection using AIC.
- Understand the principles of dynamic occupancy models, including colonisation (γ) and extinction (ε), and how species distributions change through time.
- Explain how multi-species occupancy models account for imperfect detection while estimating community-level patterns and species co-occurrence.
- Apply occupancy modelling techniques in R using the `unmarked` package to analyse ecological survey data.

## Ecological Background

### Presence-absence data, false absences, and detection probability

Occurrence or presence-absence data are widely used forms of ecological data, providing the basis for metapopulation studies and large-scale monitoring. This type of data is more efficient compared to estimating full abundance, since estimating the proportion of occupied sites is much easier and less expensive than counting every individual. Within this module, presence-absence data are handled using logistic regression as a form of generalised linear model. Although the data are still handled by logistic regression, this carries an assumption that a zero in the dataset means the species was absent from the site. The occupancy model then argues that *non-detection does not imply absence* unless a species is detected with certainty every time it is present. Species may be present at a site but go unrecorded as a result of low activity levels, poor weather, or a short survey period. Factors like cryptic behaviour also make it difficult to differentiate between a site where the species was not detected and one where the species is truly absent.

### Consequences of ignoring imperfect detection

The occupancy model deals with imperfect detection by splitting the process into two different parameters: *occupancy probability* ($\psi$), which represents the actual ecological state (presence or absence), and *detection probability* ($p$), which is the probability of detecting a species during a survey given that it is present. This statistical method was developed by MacKenzie et al. (2002), and it separates the two probabilities by using repeated surveys. This method is flexible, as it allows the inclusion of site- or survey-specific covariates and accounts for missed observations. When the detection process is ignored, it can result in several kinds of error.

First, when a species is not detected it is assumed to be absent, which produces false negatives in the data. Second, these false negatives cause systematic bias in ecological measures such as population size and species richness. Third, comparisons made across habitats and time can yield incorrect conclusions if the chance of detecting the species differs between groups, rather than the species actually being more or less common.

### Established knowledge versus ongoing debate

The issue of imperfect detection in presence-absence surveys has long existed and was recognised well before the 2000s. Work by Geissler and Fuller (1987) and Azuma et al. (1990) identified that non-detection should not be treated as true absence; however, their approaches were not flexible and could not accommodate covariates or missed observations.

### Core assumptions of the standard occupancy model

Before the model structure is introduced below, it is important to address the assumptions associated with the model and what is required to avoid violating them.

1. **Population closure** -- the site's occupancy status must remain the same throughout the repeated surveys used to estimate detection probability. This assumption is necessary to estimate detection probability and cannot be relied on for mobile or seasonal species.
2. **No false positives occur** -- detection is always assumed to be correct, so a species cannot be recorded at a site where it does not occur. This is the most challenged assumption in the literature, because of how misidentification has arisen with automated survey technologies (e.g. acoustic recorders, camera traps).
3. **Independence of detections** -- repeated detections across a site are assumed to be independent.
4. **No unmodelled heterogeneity in occupancy or detection** -- both occupancy probability ($\psi$) and detection probability ($p$) are assumed to be constant across sites, except as explained by measured covariates.
5. **Correct identification of the sampling unit and season** -- the sample unit and the study season must be constant throughout the study and clearly defined. The season should be short enough that species distributions do not shift within it, and the sampling unit should be appropriate to the species being studied.

## Occupancy Models: Theory and Structure

### What is occupancy?

Understanding what occupancy means is essential before we can understand what occupancy *models* are. Occupancy ($\psi$) refers to the probability that a site being surveyed is occupied by a species. If $x$ is the number of occupied sites and $s$ is the number of surveyed sites, then the naive estimator

$$
\hat{\psi} = \frac{\hat{x}}{s}
$$

can be used to estimate occupancy. In the real world, however, this naive estimator is rarely adequate, because species may not always be detected even when they are present, so we need a method that incorporates detection probability -- the probability that a species is detected at a site, given that it is present. If detection probability is not incorporated, occupancy will be systematically underestimated.

An occupancy survey produces two outcomes: the species is detected at a site, or it is not. Non-detection has two possible explanations: the species was truly absent, or it was present and not detected -- a *false absence*. In most ecological surveys it is reasonable to assume detections themselves are true, though more advanced occupancy models exist to relax this assumption and address *false positives* as well as false negatives.

### The MacKenzie et al. (2002) framework

The single-season (closed) occupancy model rests on three assumptions:

1. The system is **closed**: occupancy state does not change between repeat visits within a season.
2. **Detection is imperfect** but can be estimated: $p < 1$ is the rule, not the exception.
3. Detections are **independent** across sites and across visits to the same site.

The model is then written as a pair of linked equations. Let $z_i \in \{0, 1\}$ be the true
(latent) occupancy state of site $i$ and let $y_{ij} \in \{0, 1\}$ be what the observer recorded
on visit $j$. The state model is:

$$
z_i \sim \text{Bernoulli}(\psi_i)
$$

and the observation model is:

$$
y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \cdot p_{ij})
$$

If the site is unoccupied ($z_i = 0$), no detection is possible regardless of $p$. If it is
occupied, each visit yields a detection with probability $p_{ij}$. Both $\psi$ and $p$ can be
modelled as functions of covariates through logit links, making the framework an extension of
logistic regression to a two-level hierarchical model.

### Understanding the Ecological and Observation Processes

Occupancy modelling separates two processes -- ecological and observation -- rather than treating survey data as a single process. The **ecological process** represents the true state of nature: it asks whether the species is truly present at a site, independently of any surveys. A site is either occupied or unoccupied, and that state does not depend on how many times, or how well, it is surveyed. The **observation process** is what happens during a survey: a species may go undetected even though it is present at a site, and the probability of detecting it depends on factors like habitat structure and weather conditions.

The observation process is conditional on the ecological process, which is why the overall model is described as *hierarchical*. Assumption 2 (no false positives) guarantees that a species cannot be detected at a site where it does not truly occur, so every non-detection remains uncertain -- it could be a true absence or a false absence. The repeated-visit design is what lets the occupancy model separate true occupancy from the chance of detecting it.

### Notation

A useful way to keep the notation straight is that $\psi$ and $z$ belong to the ecological process, while $p$ and $y$ belong to the observation process. Keeping this distinction in mind makes it easier to separate what is happening in nature from what is actually observed:

| Symbol | Process | Meaning |
| :--- | :--- | :--- |
| $\psi_i$ | Ecological | Probability that site $i$ is occupied |
| $z_i$ | Ecological | True (latent) occupancy state of site $i$ ($0$ or $1$) |
| $p_{ij}$ | Observation | Probability of detecting the species on visit $j$ to site $i$, given $z_i = 1$ |
| $y_{ij}$ | Observation | Recorded detection ($1$) or non-detection ($0$) on visit $j$ to site $i$ |

### Model Equations and Likelihood

Occupancy models are built on the hierarchical framework introduced above, which separates the ecological process from the observation process. Together, these two processes form the likelihood function that lets the ecologist estimate occupancy and detection probabilities directly from repeated-visit detection histories, rather than relying on the naive proportion of sites with at least one detection.

To make the observation model concrete, consider a site where the species is truly present ($z = 1$) and detection probability is $p = 0.6$. If three surveys are conducted, the probability of observing the detection history $(1, 0, 1)$ is:

$$
P(y = [1, 0, 1] \mid z = 1) = p \cdot (1 - p) \cdot p = 0.6 \times 0.4 \times 0.6 = 0.144
$$

Each site's contribution to the likelihood is built the same way, from its own sequence of visits, and every site's contribution is then combined (multiplied) across sites to obtain the full likelihood function that `occu()` maximises when fitting a model.

### Link of Assumptions to the Formulas

The assumptions introduced earlier are not separate from the mathematics -- they are what license each part of the model.

- **Closure** is why the true occupancy state is written as $z_i$, indexed only by site and not by survey: the model assumes a species does not colonise or go locally extinct between the repeated surveys used to estimate $p$ within a season.
- **No false positives** is encoded directly in $y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \cdot p_{ij})$: multiplying by $z_i$ guarantees that an unoccupied site ($z_i = 0$) always yields $y_{ij} = 0$, regardless of $p_{ij}$.
- **Independence of detections** is why the likelihood is written as a simple product of visit-level probabilities across surveys and sites, rather than as a chain of conditional probabilities.
- **No unmodelled heterogeneity** means $\psi_i$ and $p_{ij}$ are treated as fixed (or as fixed functions of measured covariates) rather than as quantities that fluctuate for unmeasured reasons; if they did fluctuate beyond what the covariates explain, the resulting estimates would be biased.

Repeated surveys matter precisely because they supply the information needed to separate $\psi$ from $p$. The assumptions therefore determine how a detection history should be interpreted, and how reliable the resulting estimates of $\psi$ and $p$ can be trusted to be.

### Key Variations of Occupancy Models

#### Single-Season Models

Single-season occupancy models are the simplest and most widely used foundation of occupancy modelling. Unlike traditional presence-absence models, they account for the possibility that a species may go undetected even where it occurs, using repeated surveys within a single season to separate non-detection from true absence and so obtain an unbiased estimate of occupancy. The key assumption is *closure*: the occupancy state of a site must remain constant across the surveys used to estimate detection, with no migration, immigration, or extinction occurring at the site during that window. Single-season models are best suited to short-term studies and have been widely used for cryptic and rare species.


#### Multi-Season (Dynamic) Models

Dynamic occupancy models survey one species over multiple seasons, so that the study extends over a long period of time -- often several years, with each year's breeding season (or comparable primary period) surveyed multiple times. Repeated within-season visits are still needed to estimate detection reliably, rather than relying on a single visit per season. Crucially, this model relaxes the closure assumption *between* seasons: it explicitly allows occupancy to change from one season to the next through local colonisation ($\gamma$) and local extinction ($\varepsilon$). Dynamic models are particularly useful for long-term monitoring, helping ecologists understand how species distributions shift over time and how they respond to environmental change or management interventions.

#### Multi-Species/Community Models

Multi-species (community) occupancy models extend the single-species framework by modelling multiple species at the same sites simultaneously, sharing information across species through a hierarchical (community-level) structure. This is especially valuable for rare species that have been detected only a handful of times, since information is "borrowed" from more commonly detected species in the same community. A further extension is the multi-species co-occurrence model, which asks whether the occurrence of one species is influenced by the presence of another at the same site -- helping to determine whether species occur together more or less often than would be expected under independence, and offering insight into ecological relationships such as competition or facilitation.

## Relationship to Generalized Linear Models

Occupancy models are closely related to generalised linear models, and specifically to logistic regression for presence-absence data. In a standard logistic regression, the response variable is coded $1$ if a species is detected at a site and $0$ if it is not, and the model estimates the probability of occurrence as a function of site covariates. This approach is widely used, but it carries the assumption that non-detection is automatically recorded as absence. In ecological surveys this assumption is rarely valid, because a species might be present and simply go undetected -- for example, because it is cryptic -- which is exactly the imperfect-detection problem that occupancy models are designed to solve. An occupancy model can be understood as a two-level extension of logistic regression: one logistic regression for the ecological state ($\psi$), and a second, nested logistic regression for the observation process ($p$), linked through the repeated-visit design.

## Survey Design

### Selecting Sampling Sites and Designing the Sampling Strategy

The first step in occupancy modelling is selecting the sites where the study will be conducted. Sites can be selected randomly or systematically, depending on the research question. It is not advisable to preferentially select sites where the species of interest is already known to occur most, since this can bias the resulting estimates. The number of sites should be large enough to produce reliable estimates of $\psi$ and $p$, and the sites should span a wide range of the environmental conditions relevant to the study (e.g. water availability, degree of disturbance). A well-designed sampling strategy produces results that better represent the wider study area, rather than only the conditions at a handful of convenient sites.

### Repeated Surveys and Survey Timing

Occupancy studies require that each site be surveyed repeatedly, because species can be missed even when present. These repeated surveys supply the information needed to estimate detection probability and to distinguish false absence from true absence. The study period must be short enough that the closure assumption remains defensible, and survey timing must suit the biology of the species being studied -- for example, surveys should be conducted when the species is active and thus detectable.

### Factors Affecting Detection and Covariates

The probability of detecting a species can be affected by several factors, even when the species is present at a site. These include weather conditions, the effort invested in the survey, and habitat characteristics. Such factors are grouped as *detection covariates*, which influence the probability of detecting a species during a given visit. By contrast, environmental factors such as temperature, vegetation cover, and human disturbance are usually grouped as *occupancy covariates*, since they influence where a species occurs in the first place. Separating detection covariates from occupancy covariates is important for correctly interpreting a fitted model, since a factor that affects detectability does not necessarily affect where a species actually occurs, and vice versa.

## Advantages and Limitations of Occupancy Models

### Advantages

Occupancy modelling is very useful, but it is not always the right tool for every ecological question, so it is worth understanding both its strengths and its weaknesses. The most important advantage is that it accounts for imperfect detection: by estimating detection probability explicitly, it reduces the bias caused by false absences. The models require only presence-absence data rather than counts of every individual, which makes them well suited to long-term monitoring programmes that contribute to biodiversity management. They also explicitly separate ecological and observation processes, which makes it possible to model the factors affecting each process independently.

### Limitations

The major limitation of occupancy modelling is the repeated-survey requirement: repeated surveys are time-consuming and require more field effort than a single visit. The single-season model is also limited by the closure assumption, which does not always hold in practice -- animals moving in and out of a site during the survey period can bias the results. Detection probability should not be too low, since very low $p$ makes it difficult to reliably detect the species at all, which in turn widens uncertainty around $\hat\psi$. Occupancy models can also produce misidentifications, since the standard model assumes all detections are correct; when this assumption is violated, false positives can bias the results. More generally, occupancy models operate under a specific set of assumptions, and if the study design does not align with those assumptions, the resulting estimates cannot be considered reliable.

::: {.callout-note appearance="simple"}
## A note on zero-inflation

Standard occupancy models already address one form of zero-inflation through the $\psi$/$p$ framework: they explicitly model two sources of zeros in the data (true absence and non-detection at an occupied site), rather than treating all zeros as equivalent. A further extension -- the *zero-inflated* occupancy model of Miller et al. (2011) -- adds a third source of structural zeros for species that are entirely unavailable for detection at a subset of sites (e.g. because the site is outside the species' range for reasons unrelated to habitat suitability). This extension is outside the scope of this chapter, but is worth knowing about if you encounter a system where "occupancy" itself may not be well defined at every surveyed site.
:::

<!-- #: Real-Data Single-Season Occupancy -->

## Worked Example: Single-Season Occupancy with the Crossbill Dataset

::: callout-tip
## **Crossbill Dataset**

| Columns | Content |
| :--- | :--- |
| `id` | site identifier |
| `ele` | elevation (m) - site covariate |
| `forest` | percent forest cover - site covariate |
| `surveys` | number of surveys conducted |
| `det991`-`det993`, `det001`-`det003`,... | 27 columns: crossbill detection/nondetection data during all surveys over the 9 years. |
| `date991`-`date993`, `date001`-`date003`,... | 27 columns: Julian survey date for each corresponding visit. |
| `x`, `y` | coordinates |
:::

The `crossbill` dataset (built into `unmarked`) comes from the Swiss breeding bird survey (Monitoring Häufige Brutvögel, MHB). It covers 267 1-km² survey quadrats across Switzerland, each visited up to 3 times per year between 1999 and 2007, recording detection/non-detection of the European crossbill (*Loxia curvirostra*).

::: {.callout-note appearance="simple"}
## Reflect Before You Start

Before opening `?crossbill`, write down what you expect. Crossbills depend on conifer seed crops, so where would you predict occupancy to be highest: low-elevation farmland, mid-elevation forest, or high alpine terrain? You will check this prediction against the fitted model further down.
:::

### Load the Data and Inspect Its Shape

```{r step1-load}
#| label: step1-load
#| echo: true

library(unmarked)   # occu(), unmarkedFrameOccu()
library(dplyr)       # data wrangling
library(ggplot2)      # plotting
library(AICcmodavg)   # GOF test, model averaging
library(MuMIn)         # dredge() for model selection

# crossbill ships with unmarked -- no external file needed
data(crossbill)

# 267 sites x (id, ele, forest, x, y, then 9 years x 3 visits of
# detections "detYYYY_V" and survey dates "dateYYYY_V")
str(crossbill[, 1:10])
```

A single-season frame needs three ingredients: a detection history matrix, site-level covariates, and (optionally) observation-level covariates. The crossbill object bundles all three inside one wide data frame, so most of the work here is pulling the right columns out rather than reshaping raw field sheets.

### Carve Out One Season (1999)

I will use 1999, the first year surveyed, so the workflow is directly comparable to the three-visit frog design used in the earlier tiers.

```{r step2-subset}
#| label: step2-subset
#| echo: true

# Detection history: three visits in 1999
det_1999 <- crossbill %>%
  dplyr::select(id, det991, det992, det993)

# Site covariates: elevation and forest cover, standardized as in the
# simulated tiers so coefficients are comparable in magnitude
site_cov <- crossbill %>%
  dplyr::select(id, ele, forest) %>%
  mutate(ele_z = as.numeric(scale(ele)),
         forest_z = as.numeric(scale(forest)))

# Observation covariate: Julian survey date, one column per visit,
# needed to answer Question 4 (does detection vary with survey timing?)
date_1999 <- crossbill %>%
  dplyr::select(date991, date992, date993)

head(det_1999)
head(site_cov)
```

::: {.callout-note appearance="simple"}
## Callout: Missing Visits Are Normal, Not an Error

High-elevation sites in the MHB are only surveyed twice per season, so `det993` (and its matching `date993`) will contain `NA` for those rows. `occu()` handles this natively -- a missing visit simply drops out of that site's likelihood term rather than being treated as a non-detection. Do not recode `NA` to `0`; that claims a survey happened and found nothing, which is a different (false) statement.
:::

### The Occupancy Model with `unmarked`

The `unmarked` package implements the MacKenzie et al. (2002) likelihood. The workflow has three
steps: (1) package the data into an `unmarkedFrameOccu` object, (2) fit one or more `occu()`
models, and (3) extract and interpret the parameter estimates.

#### Setting up the `unmarkedFrame`

For a single-season occupancy model, `unmarked` expects:

- A **detection matrix** ($n_{\text{sites}} \times n_{\text{visits}}$): one row per site, one
  column per visit occasion.
- Optional **site covariates** (covariates on $\psi$): one value per site.
- Optional **observation covariates** (covariates on $p$): one value per site-visit combination.

#### Build the `unmarkedFrameOccu`

```{r step3-frame}
#| label: step3-frame
#| echo: true

crossbill_umf <- unmarkedFrameOccu(
  y = as.matrix(det_1999[, c("det991", "det992", "det993")]),
  siteCovs = site_cov[, c("ele_z", "forest_z")],
  obsCovs = list(date = date_1999)
)

summary(crossbill_umf)
```

### Naive Occupancy vs. the Null Model

Before fitting anything, calculate naive occupancy -- the proportion of sites where the crossbill was detected at least once. This is the same quantity the "Ecological Background" section above builds the "why we need p" argument around, and it will again understate true occupancy here.

```{r step4-naive}
#| label: step4-naive
#| echo: true

# This is the quantity that ignores imperfect detection entirely.
naive_occ <- mean(rowSums(det_1999[, c("det991","det992","det993")],
                           na.rm = TRUE) > 0)
naive_occ

# Null (intercept-only) model gives the model-based comparison point
occ_null <- occu(~1 ~1, data = crossbill_umf)
backTransform(occ_null, type = "state")   # model-based psi
backTransform(occ_null, type = "det")     # model-based p
```

::: {.callout-note appearance="simple"}
## Reflect

Compare naive occupancy to the model-based estimate of &psi; from `occ_null`. Which is larger, and does the gap match what you'd expect given the estimated detection probability?
:::

### Add Covariates: Does Detection Vary with Survey Date?

Before asking what drives occupancy, check what drives detection -- if that is ignored, an occupancy covariate could just be soaking up a detection-probability pattern instead.

```{r step5-datemodel}
#| label: step5-datemodel
#| echo: true

occ_date <- occu(~ scale(date) ~1, data = crossbill_umf)
summary(occ_date)
```

::: {.callout-note appearance="simple"}
## Reflect

Is the coefficient on `date` positive, negative, or indistinguishable from zero? A positive effect would mean later-season visits are more likely to record a crossbill that is present; a null effect means survey timing (within the mid-April to late-June window) doesn't matter much for this species -- either is a defensible, informative finding.
:::

- **Positive coefficient**: Detection improves later in the season.
- **Negative coefficient**: Detection worsens later in the season.
- **Near-zero / CI straddling zero**: No detectable seasonal effect on detectability within this survey window.

### Site-level Covariates: Elevation and Forest on Occupancy

```{r step6-covmodel}
#| label: step6-covmodel
#| echo: true

occ_full <- occu(~ scale(date) ~ ele_z + forest_z, data = crossbill_umf)
summary(occ_full)
```

### Model Selection

I fit all subsets of the general model and rank them by AICc rather than presenting a single model as if it were the only reasonable one.

```{r step6-dredge}
#| label: step6-dredge
#| echo: true

occ_dredge <- dredge(occ_full, rank = "AICc")
occ_dredge
```

### Goodness-of-Fit

```{r step7-gof}
#| label: step7-gof
#| echo: true
#| eval: false

# nsim raised well beyond 10 for real inference; kept low here for a fast
# teaching run. Increase to >=1000 before treating the chi-square p-value
# as informative.
crossbill_gof <- mb.gof.test(occ_full, nsim = 25)
crossbill_gof
```

::: {.callout-note appearance="simple"}
## Callout: Why `eval: false`

The bootstrap GOF test can take several minutes even at `nsim = 25` on a full class's laptops. The chunk is shown but not executed automatically when the chapter renders; run it once locally with `nsim` raised to at least 1000 before quoting a p-value in your write-up.
:::

### Predicted Occupancy Along the Elevation Gradient

```{r step8-predict}
#| label: fig-crossbill-elevation
#| fig-cap: "Predicted crossbill occupancy probability across the standardized elevation gradient, forest cover held at its mean. Shaded band is the 95% confidence interval from the top AICc model."
#| code-fold: true

new_ele <- data.frame(
  ele_z = seq(min(site_cov$ele_z, na.rm = TRUE),
              max(site_cov$ele_z, na.rm = TRUE), by = 0.05),
  forest_z = mean(site_cov$forest_z, na.rm = TRUE)
)

pred_ele <- predict(occ_full, newdata = new_ele, type = "state") %>%
  bind_cols(new_ele)

ggplot(pred_ele, aes(x = ele_z, y = Predicted)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80") +
  geom_line(linewidth = 1) +
  labs(x = "Elevation (standardized)", y = "Predicted occupancy probability") +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic()
```

::: {.callout-note appearance="simple"}
## Reflect

Does the shape of this curve match the prediction when we loaded and inspected the data? If not, what ecological explanation would reconcile crossbill biology with what the model shows?
:::

# South African Case Study: *Capensibufo rosei*

## Background

A species can be absent from a survey record for two fundamentally different reasons: it was
genuinely not there, or it was there but the observer failed to detect it. Confusing these two
produces biased estimates of occupancy — the proportion of sites where a species truly occurs. The
classical MacKenzie et al. (2002) occupancy model was designed to untangle this confusion by
decomposing a detection history at a site across repeated visits into two separate processes: the
ecological state process governing whether the site is occupied ($\psi$), and the observation
process governing whether the species is detected given that it is present ($p$). This section
works through that model using a real South African frog dataset that places the
imperfect-detection problem front and centre.


## The study system: *Capensibufo rosei*

Rose's Mountain Toadlet (*Capensibufo rosei*) is a Critically Endangered frog endemic to the Cape
Peninsula of South Africa. It is known from only two breeding localities — both inside Table
Mountain National Park — and has disappeared from at least four historical sites since the 1980s,
apparently because fire suppression has allowed vegetation to encroach on its open fynbos breeding
pools. The toad lacks an advertisement call and breeds in dense aggregations for only a few weeks
per year, making it exceptionally hard to detect.

Becker et al. (2022) asked a question that precedes occupancy modelling in the usual sense: before
we can monitor a species, we need to know *where it is*. Specifically, they asked what the
probability is that all occupied sites on the Cape Peninsula have already been found, and how much
additional search effort would be needed to reduce the residual probability of an undiscovered
site to an acceptable level.

The dataset we use here — `All_spp_search_data.csv` — contains 101 candidate sites (300 × 300 m
grid cells) across the Peninsula that were considered potential *C. rosei* habitat. Because the
toad itself is so rarely encountered, the study also surveyed four ecologically similar proxy frog
species that share its habitat requirements. These proxy species — *Amietia fuscigula* (Cape River
Frog), *Arthroleptella lightfooti* (Cape Peninsula Moss Frog), *Strongylopus bonaespei* (Clicking
Stream Frog), and *Strongylopus grayii* (Clicking Stream Frog) — have higher detectabilities and
their combined presence/absence pattern carries information about habitat quality and search effort
at each site.

This is worth pausing on. The absence of a `PA.ros` (presence/absence of *C. rosei*) column in
the data is not a data-entry oversight; it is the entire scientific problem. The target species is
so rare that you cannot estimate detection probability directly from its own detection history, so
the detection histories of the proxy species must be used instead. This is an occupancy problem
where imperfect detection is the obstacle that stands between the surveyor and knowledge of the
species' true distribution.

## Setup

```{r pkg-load}
library(tidyverse)    # data wrangling and ggplot2
library(unmarked)     # occupancy models
library(here)         # file path management
library(patchwork)    # multi-panel figures
library(AICcmodavg)   # model selection tables
```

## The Data

```{r load-data}
rosei <- read_csv("c://Users//27659//Downloads//All_spp _search_data.csv")
```

```{r glimpse-data}
glimpse(rosei)
```

The dataset has 16 columns. The ones we will work with are:

| Column          | Role                                                                       |
|:----------------|:---------------------------------------------------------------------------|
| `Cell Nr`       | Site identifier (1–101)                                                    |
| `lat`, `lon`    | Centroid coordinates of the 300 × 300 m grid cell                         |
| `shallowWetland`| Presence (1) or absence (0) of shallow wetland covering ≥5% of the cell   |
| `2/3ClosetSpp`  | Whether ≥2 of the 3 most habitat-similar proxy species were detected       |
| `PA.fus`        | Presence/absence of *Amietia fuscigula*                                    |
| `PA.lig`        | Presence/absence of *Arthroleptella lightfooti*                            |
| `PA.bon`        | Presence/absence of *Strongylopus bonaespei*                               |
| `PA.gra`        | Presence/absence of *Strongylopus grayii*                                  |
| `HS.ros.mean.0` | Predicted habitat suitability for *C. rosei* from a species distribution model (SDM) |

```{r clean-data}
# The first 12 rows contain a data dictionary embedded in the last two columns.
# The actual site data are rows where Cell Nr is not NA and Explanation is NA.
dat <- rosei |>
  filter(!is.na(`Cell Nr`), is.na(Explanation)) |>
  rename(
    site      = `Cell Nr`,
    wetland   = shallowWetland,
    proxy23   = `2/3ClosetSpp`,
    pa_fus    = PA.fus,
    pa_lig    = PA.lig,
    pa_bon    = PA.bon,
    pa_gra    = PA.gra,
    hs_ros    = HS.ros.mean.0,
    hs_fus    = HS.fus.mean,
    hs_lig    = HS.lig.mean,
    hs_bon    = HS.bon.mean,
    hs_gra    = HS.gra.mean
  ) |>
  select(site, lat, lon, wetland, proxy23, pa_fus, pa_lig, pa_bon, pa_gra,
         hs_ros, hs_fus, hs_lig, hs_bon, hs_gra)

glimpse(dat)
```

## Naive occupancy

Before fitting any model, it is instructive to calculate the *naïve* occupancy rate: the raw
proportion of sites where each proxy species was detected, ignoring the possibility of false
absences.

```{r naive-occupancy}
dat |>
  summarise(
    across(c(pa_fus, pa_lig, pa_bon, pa_gra), ~ mean(.x, na.rm = TRUE),
           .names = "naive_{.col}")
  )
```

These naive rates underestimate true occupancy whenever $p < 1$. How much they underestimate
depends on both the true occupancy and the detection probability; the occupancy model will
separate the two.

## Visualising the data

Before modelling, it helps to see where the sites are and how the proxy species are distributed
across them.

```{r fig-site-map, fig.cap="The 101 candidate sites on the Cape Peninsula. Point colour shows whether shallow wetland habitat is present at each site."}
#| code-fold: true
ggplot(dat, aes(x = lon, y = lat, colour = factor(wetland))) +
  geom_point(size = 1.8, alpha = 0.8) +
  scale_colour_manual(
    values = c("0" = "grey70", "1" = "#2166ac"),
    labels = c("Absent", "Present"),
    na.value = "grey90"
  ) +
  labs(
    x = "Longitude", y = "Latitude",
    colour = "Shallow wetland"
  ) +
  coord_fixed() +
  theme(legend.position = "bottom")
```

```{r fig-pa-matrix, fig.cap="Detection matrix for the four proxy species across all 101 sites. Black cells = detected; white = not detected; grey = not surveyed (NA)."}
#| code-fold: true
dat |>
  select(site, pa_fus, pa_lig, pa_bon, pa_gra) |>
  pivot_longer(-site, names_to = "species", values_to = "detected") |>
  mutate(
    species = recode(species,
      pa_fus = "A. fuscigula",
      pa_lig = "A. lightfooti",
      pa_bon = "S. bonaespei",
      pa_gra = "S. grayii"
    )
  ) |>
  ggplot(aes(x = species, y = reorder(site, site), fill = factor(detected))) +
  geom_tile(colour = "white", linewidth = 0.1) +
  scale_fill_manual(
    values = c("0" = "white", "1" = "black"),
    labels = c("Not detected", "Detected"),
    na.value = "grey85"
  ) +
  labs(x = NULL, y = "Site", fill = NULL) +
  theme(
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "bottom"
  )
```

::: callout-note
## What the detection matrix tells you already

The grey cells — sites where a species was not surveyed — are not missing data in the usual sense.
They represent the incompleteness of the search: a site was visited but a particular proxy species
was not recorded (perhaps because it was not the focus of that visit, or conditions were
unsuitable). This kind of missing data, where non-detection and absence of a survey visit are
conflated, is precisely what the occupancy model machinery handles through the `NA`-tolerant
likelihood.
:::


Because our data record only a single composite visit per site per proxy species (rather than
repeated visits across time), we treat the detection history of the four proxy species as the
"repeat visits". This is an important conceptual move: we are not asking whether *C. rosei* itself
was detected on multiple occasions, but whether the proxy community shows evidence of suitable,
well-searched habitat. Each proxy species detection is one "look" at the site.

```{r build-y-matrix}
# Detection matrix: rows = sites, columns = proxy species (the four 'visits')
y_mat <- dat |>
  select(pa_fus, pa_lig, pa_bon, pa_gra) |>
  as.matrix()

# Site covariates (standardise continuous predictors for numerical stability)
site_covs <- dat |>
  select(wetland, proxy23, hs_ros) |>
  mutate(
    hs_ros_sc = as.numeric(scale(hs_ros))
  )

# Package everything into an unmarkedFrameOccu
umf <- unmarkedFrameOccu(
  y      = y_mat,
  siteCovs = site_covs
)

summary(umf)
```

The summary confirms 101 sites and 4 observations (proxy species) per site, along with the
proportion of observations that yielded a detection for each proxy.

## Fitting candidate models

We fit a set of models that vary in what predicts occupancy ($\psi$) while holding detection
probability ($p$) constant. This follows the standard practice of first modelling the detection
process and then, once satisfied with it, exploring the occupancy structure.

The `occu()` formula syntax is `~ detection ~ occupancy`. A `1` on either side specifies an
intercept-only (constant) model.

### Model 1: Null model — constant $\psi$ and $p$

```{r fit-null}
m_null <- occu(~ 1 ~ 1, data = umf)
summary(m_null)
```

The coefficients are on the logit scale. We back-transform to probabilities using `plogis()`.

```{r null-backtransform}
# Back-transformed estimates
psi_null <- plogis(coef(m_null, type = "state"))
p_null   <- plogis(coef(m_null, type = "det"))

cat("Estimated occupancy (psi):", round(psi_null, 3), "\n")
cat("Estimated detection (p):  ", round(p_null,   3), "\n")
```

::: callout-note
## Comparing naive occupancy with model-corrected occupancy

Compare the model-corrected $\hat{\psi}$ with the naive occupancy rates you calculated earlier.
The difference is the magnitude of the detection bias that the model corrects for. For species
with low $p$, this correction can be substantial.
:::

### Model 2: Occupancy predicted by shallow wetland habitat

```{r fit-wetland}
m_wetland <- occu(~ 1 ~ wetland, data = umf)
summary(m_wetland)
```

### Model 3: Occupancy predicted by *C. rosei* SDM habitat suitability score

```{r fit-hs}
m_hs <- occu(~ 1 ~ hs_ros_sc, data = umf)
summary(m_hs)
```

### Model 4: Occupancy predicted by proxy community co-occurrence

```{r fit-proxy}
m_proxy <- occu(~ 1 ~ proxy23, data = umf)
summary(m_proxy)
```

### Model 5: Additive model — wetland and SDM score

```{r fit-additive}
m_add <- occu(~ 1 ~ wetland + hs_ros_sc, data = umf)
summary(m_add)
```

## Model selection

We compare models using AIC. The model with the lowest AIC, or equivalently the highest support
in a relative sense, is preferred. Models within $\Delta\text{AIC} < 2$ of the top model have
substantial support and should not be dismissed.

```{r model-selection}
# Create a named list of fitted models
cand_models <- list(
  "Null (psi(.) p(.))"            = m_null,
  "Wetland (psi(wet) p(.))"       = m_wetland,
  "SDM score (psi(HS) p(.))"      = m_hs,
  "Proxy co-occ (psi(prx) p(.))"  = m_proxy,
  "Additive (psi(wet+HS) p(.))"   = m_add
)

aictab(cand_models, second.ord = FALSE)
```

::: callout-note
## Interpreting the AIC table

The `Delta_AIC` column shows the difference between each model's AIC and the top model's AIC.
`AICWt` gives the Akaike weight, which can be read as the approximate probability that the model
is the best-supported model among those considered. `Cum.Wt` is the cumulative weight: if the top
two models together have a cumulative weight close to 1, little evidence supports the remaining
models.
:::

## Interpreting the best-supported model

Extract the top-ranked model for deeper inspection.

```{r best-model-summary}
library(MuMIn)

# Compare candidate models using AIC
aic_table <- data.frame(
  Model = names(cand_models),
  AIC   = sapply(cand_models, AIC)
)

# Rank models from best to worst
aic_table <- aic_table[order(aic_table$AIC), ]

# Calculate Delta AIC and Akaike weights
aic_table$DeltaAIC <- aic_table$AIC - min(aic_table$AIC)
aic_table$Weight <- MuMIn::Weights(aic_table$AIC)

# Display the model selection table
print(aic_table)

# Extract the best-supported model
best_model <- cand_models[[aic_table$Model[1]]]

cat("\nBest-supported model:", aic_table$Model[1], "\n\n")

summary(best_model)
```

Back-transform both parameters from the best model.

`backTransform()` works only when a model component has a single coefficient (i.e. intercept
only). For models with covariates on $\psi$, we use `predict()` evaluated at the mean of
continuous covariates and the modal value of binary ones — the marginal occupancy at a "typical"
site. Detection probability is always intercept-only here, so `backTransform()` is safe for $p$.

```{r best-backtransform}
# Detection probability: intercept-only across all candidate models, so backTransform() is safe
p_est <- backTransform(best_model, type = "det")
cat("Estimated detection probability (p):\n")
p_est

# Occupancy: use predict() so the code is robust regardless of which model is best.
# Evaluate at mean/modal covariate values to get the marginal occupancy estimate.
psi_pred <- predict(
  best_model,
  newdata = data.frame(
    hs_ros_sc = 0,   # mean of a standardised variable is 0
    wetland   = 0,   # modal value (most sites lack shallow wetland)
    proxy23   = 0    # modal value
  ),
  type = "state"
)

cat("\nEstimated occupancy at mean/modal covariate values (psi):\n")
cat("  Estimate:", round(psi_pred$Predicted, 3), "\n")
cat("  95% CI: [", round(psi_pred$lower, 3), ",", round(psi_pred$upper, 3), "]\n")
```

## Visualising the occupancy–covariate relationship

For models with a continuous covariate on $\psi$, the marginal occupancy curve shows how occupancy
probability changes across the range of the predictor.

```{r fig-occ-curve, fig.cap="Predicted occupancy probability as a function of SDM habitat suitability for *C. rosei*. The shaded ribbon is the 95% confidence interval. Points along the x-axis show observed habitat suitability values at surveyed sites."}
#| code-fold: true
# Generate a prediction grid over the range of hs_ros_sc
hs_seq <- seq(min(site_covs$hs_ros_sc, na.rm = TRUE),
              max(site_covs$hs_ros_sc, na.rm = TRUE),
              length.out = 100)

# Prediction data frame: only include the covariate that m_hs was fitted with.
# Passing extra columns that the model does not recognise causes a warning;
# each model should only receive the covariates it was specified with.
pred_df <- data.frame(hs_ros_sc = hs_seq)

# Predict occupancy from the SDM model
pred_psi <- predict(m_hs,
                    newdata = pred_df,
                    type    = "state")

pred_plot <- bind_cols(pred_df, pred_psi)

ggplot(pred_plot, aes(x = hs_seq)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#2166ac", alpha = 0.2) +
  geom_line(aes(y = Predicted), colour = "#2166ac", linewidth = 0.8) +
  geom_rug(
    data = dat,
    aes(x = as.numeric(scale(hs_ros))),
    sides = "b", alpha = 0.4, linewidth = 0.3, inherit.aes = FALSE
  ) +
  labs(
    x = "SDM habitat suitability (standardised)",
    y = expression(paste("Predicted occupancy (", hat(psi), ")"))
  )
```

## Site-level occupancy predictions

In addition to the marginal curve, `unmarked` can produce a corrected occupancy estimate for each
individual site, conditioning on the actual detection history observed there. A site where all
four proxy species were detected should attract a higher posterior occupancy estimate than a site
where none were seen, even if both have identical habitat suitability scores.

```{r site-level-preds}
# ranef() extracts empirical Bayes estimates of the latent state for each site
re <- ranef(best_model)

# Extract the posterior mean probability of occupancy for each site
site_occ <- bup(re, stat = "mean")

# Bind back to site coordinates for plotting
site_pred_df <- dat |>
  mutate(psi_hat = site_occ)
```

```{r fig-site-psi, fig.cap="Site-level posterior occupancy estimates across the Cape Peninsula. Larger, darker points indicate higher estimated probability of *C. rosei* occupancy."}
#| code-fold: true
ggplot(site_pred_df, aes(x = lon, y = lat, colour = psi_hat, size = psi_hat)) +
  geom_point(alpha = 0.75) +
  scale_colour_gradient(low = "grey85", high = "#08306b",
                        name = expression(hat(psi))) +
  scale_size_continuous(range = c(0.5, 3), guide = "none") +
  labs(x = "Longitude", y = "Latitude") +
  coord_fixed() +
  theme(legend.position = "right")
```

::: callout-note
## What the spatial pattern tells us

Sites with the highest posterior occupancy estimates tend to cluster in areas with high habitat
suitability — but they do not always coincide with the known breeding localities. This is the
practical output of the occupancy model: a ranked list of candidate sites where the probability
of hosting an undiscovered *C. rosei* population is highest. Field teams can use this map to
prioritise future search effort, as Becker et al. (2022) did.
:::

## Reporting Occupancy Estimates

A complete occupancy analysis reports both $\hat{\psi}$ and $\hat{p}$ with uncertainty. The
function below collects these into a tidy table for the best model.

```{r report-table}
# Detection (p): always intercept-only, so backTransform() works directly
p_bt  <- backTransform(best_model, type = "det")
p_row <- tibble(
  Parameter = "Detection (p)",
  Estimate  = as.numeric(coef(p_bt)),
  SE        = SE(p_bt),
  Lower_95  = pmax(0, Estimate - 1.96 * SE),
  Upper_95  = pmin(1, Estimate + 1.96 * SE)
)

# Occupancy (ψ): use predict() at mean/modal covariates
# backTransform(type = "state") throws an error when the state submodel has
# more than one coefficient (i.e. any model with a covariate on psi).
# predict() handles all cases uniformly and returns the estimate on the
# probability scale together with its delta-method SE and 95% CI.
psi_row_raw <- predict(
  best_model,
  newdata = data.frame(hs_ros_sc = 0, wetland = 0, proxy23 = 0),
  type    = "state"
)

psi_row <- tibble(
  Parameter = "Occupancy (ψ) at mean covariates",
  Estimate  = psi_row_raw$Predicted,
  SE        = psi_row_raw$SE,
  Lower_95  = psi_row_raw$lower,
  Upper_95  = psi_row_raw$upper
)

# Combine and round
bind_rows(psi_row, p_row) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

## Summary and Ecological Interpretation

The single-season occupancy model separates two sources of information that are hopelessly
conflated in a raw survey record: the true probability that a site supports the species of
interest, and the probability of detecting it during any given visit. For *Capensibufo rosei*, a
species with extremely limited range and high crypticity, failing to make this separation would
produce a dramatically underestimated distribution and misdirected conservation effort.

Several conceptual points from this analysis are worth taking forward:

**Detection probability is not a nuisance parameter.** The estimate of $\hat{p}$ is biologically
meaningful. A low $p$ for *C. rosei*'s proxy community at a given site signals that the site was
either poorly suited to these species or insufficiently searched. Both interpretations matter for
conservation planning.

**Habitat covariates on $\psi$ quantify what drives occupancy.** Whether shallow wetland presence
or SDM-predicted suitability better predicts proxy occupancy tells us something about which aspect
of habitat quality matters most. A model where $\psi$ increases with SDM suitability confirms
that the SDM captures real ecological signal, not just a statistical artefact.

**The corrected occupancy estimate exceeds the naïve rate.** This is mathematically guaranteed
when $p < 1$: the model allocates some of the observed absences to the detection failure category
rather than the true absence category, raising the estimated proportion of occupied sites.

**The proxy species are doing the detection work.** Because *C. rosei* itself has too few records
to estimate its own $p$, the proxy framework is the only route to a defensible occupancy
estimate. This is one of the most honest acknowledgements in recent South African conservation
research: the rarest species are often the ones for which detection-corrected estimates are most
desperately needed and most difficult to obtain.

## AI-Use Declaration
Artificial intelligence (AI) tools were used in the preparation of this chapter. Specifically, Claude (developed by Anthropic) was used to assist with language refinement, literature summarisation, structural editing and code generating. All AI-generated content was reviewed, verified, and edited by the authors. The authors take full responsibility for the accuracy and integrity of the final submitted work. 

## Author contribution
Both authors designed this chapter as part of the BCB743 Quantitative Ecology module. The authors identified the *Capensibufo rosei* dataset (Becker et al., 2022) as a pedagogically appropriate case study for single-season occupancy modelling in a South African context, directed the structure of the analytical workflow, verified all R code for correctness and pedagogical clarity, and wrote the ecological interpretation and synthesis. Claude (Anthropic) was used as an AI writing and coding assistant under the author's direction. The author takes full responsibility for the accuracy and integrity of all content presented in this chapter.

# References

Azuma DL, Baldwin JA, Noon BR (1990) Estimating the occupancy of spotted owl habitat areas by
sampling and adjusting for bias. General Technical Report PSW-124, USDA Forest Service.

Becker FS, Slingsby JA, Measey J, Tolley KA, Altwegg R (2022) Finding rare species and
estimating the probability that all occupied sites have been found. *Ecological Applications*
32:e2502. <https://doi.org/10.1002/eap.2502>

Dorazio RM, Royle JA, Söderström B, Glimskär A (2006) Estimating species richness and
accumulation by modeling species occurrence and detectability. *Ecology* 87(4):842–854.
<https://doi.org/10.1890/0012-9658(2006)87[842:ESRAAB]2.0.CO;2>

Geissler PH, Fuller MR (1987) Estimation of the proportion of area occupied by an animal species.
Proceedings of the Section on Survey Research Methods, American Statistical Association.

Guillera-Arroita G (2011) Impact of sampling with replacement in occupancy studies with spatial
replication. *Methods in Ecology and Evolution* 2(4):401–406.
<https://doi.org/10.1111/j.2041-210X.2011.00089.x>

Kéry M, Royle JA (2016) *Applied Hierarchical Modeling in Ecology: Analysis of Distribution,
Abundance and Species Richness in R and BUGS* (Vol. 1). Academic Press.

MacKenzie DI, Bailey LL (2004) Assessing the fit of site-occupancy models. *Journal of
Agricultural, Biological and Environmental Statistics* 9:300–318.
<https://doi.org/10.1198/108571104X3361>

MacKenzie DI, Nichols JD, Hines JE, Knutson MG, Franklin AB (2003) Estimating site occupancy,
colonization, and local extinction when a species is detected imperfectly. *Ecology* 84(8):2200–2207.
<https://doi.org/10.1890/02-3090>

MacKenzie DI, Nichols JD, Lachman GB, Droege S, Royle JA, Langtimm CA (2002) Estimating site
occupancy rates when detection probabilities are less than one. *Ecology* 83:2248–2255.
<https://doi.org/10.1890/0012-9658(2002)083[2248:ESORWU]2.0.CO;2>

MacKenzie DI, Nichols JD, Royle JA, Pollock KH, Bailey LL, Hines JE (2017) *Occupancy Estimation
and Modeling: Inferring Patterns and Dynamics of Species Occurrence* (2nd ed.). Academic Press.

MacKenzie DI, Royle JA (2005) Designing occupancy studies: general advice and allocating survey
effort. *Journal of Applied Ecology* 42(6):1105–1114.
<https://doi.org/10.1111/j.1365-2664.2005.01098.x>

Miller DA, Nichols JD, McClintock BT, Grant EHC, Bailey LL, Weir LA (2011) Improving occupancy
estimation when two types of observational error occur: non-detection and species
misidentification. *Ecology* 92(7):1422–1428. <https://doi.org/10.1890/10-1396.1>

Royle JA, Kéry M (2007) A Bayesian state-space formulation of dynamic occupancy models. *Ecology*
88(7):1813–1823. <https://doi.org/10.1890/06-0669.1>

Royle JA, Link WA (2006) Generalized site occupancy models allowing for false positive and false
negative errors. *Ecology* 87(4):835–841.
<https://doi.org/10.1890/0012-9658(2006)87[835:GSOMAF]2.0.CO;2>
