---
title: "Plotting the Whale Sightings and Chlorophyll-*a* Concentrations"
subtitle: "A brief demonstration of data summaries"
date: "`r Sys.Date()`"
description: "This vignette demonstrates exploratory data analyses."
bibliography: ../references.bib
csl: ../styles/marine-biology.csl
format:
  html:
    code-fold: false
    toc-title: "On this page"
    standalone: true
    toc-location: right
    page-layout: full
---

## Load packages

```{r}
#| message: false
library(tidyverse)
library(lubridate)
library(gganimate)
library(raster)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(elevatr)
library(RANN) # for the nearest neighbour search
library(ggpubr)
library(gt) # for nice tables

source(here::here("R", "map_theme.R"))
```

```{r}
#| echo: false
#| eval: false
# OpenTopography API
set_opentopo_key("cd52ef9b8b407a17fd417843e571fc27")
```

![The Azores region in the whales sightings
study.](../images/Azores.png){#fig-map}

This analysis uses the Azores chlorophyll-*a* data downloaded in
[Retrieving Chlorophyll-*a* Data from ERDDAP Servers](chl_ERDDAP.qmd).

## Background

The region around the Azores is characterised by relatively higher
nutrient availability in a 'sea' of otherwise oligotrophic conditions,
and hence they are of major interest as biodiversity hotspots. The
enhanced productivity results from high mesoscale activity (often
measured as Eddy Kinetic Energy, EKE) and the presence (and interaction
with) undersea topographic features [@santos2013phytoplankton]. The
enhanced chlorophyll-a biomass in the region results in it being an
important foraging area for whales *en route* to areas further north in
the Atlantic [@gonzalez2018multi].

We expect seasonal variation to provide a strongly signal in the region,
with typical autumn/winter to spring chl-*a* blooms. All satellite ocean
colour products show the same general pattern with the highest pigment
concentrations during spring months and the lowest during summer. SST
also shows a seasonal trend, with highest SST during summer and the
lowest during winter.

## Rui's objectives

Our main objective is to interpret if Azorean waters are a migratory
corridor for the four main baleen whale’s species sighted in the Azores
(blue, *Balaenoptera musculus*; fin, *B. physalus*; sei, *B. borealis*;
humpback whale, *Megaptera novaeangliae*), during their migration from
breeding to feeding areas, and vice-versa. This leads to other main
questions: 

* How long the target species use the study area during migration? -->
    Unlikely achievable without IDs.
* Are they returning in different years? --> Probably not possible without
    IDs of individual whales.
* How does the intensity and timing of the spring bloom influence
the migration? --> Timing question already addressed, but we can 
    probably improve. Question about intensity vs sightings can be done
    using a regression-type approach.

## Questions

1.  Do whales track temporal appearance of chl-*a* max? \[**Yes**\] What
    is the lag? \[**Table 1**\]

    a.  Is there a correlation in time between whale presence and lagged
        co-located chl-*a* conc.?
    b.  Is the correlation between whale presence and chl-*a* conc
        spatially fixed at i) Pico and Faial and ii) São Miguel, or
    c.  do max aggregations at localities located near i) Pico and Faial
        and ii) São Miguel coincide with the chl-*a* max there? I.e. is
        there a spatial association?

2.  Is there some association between chl-*a* biomass and the number of
    whales visiting the region? Look at inter-annual variation.

3.  Show the inverse association between chl-*a* biomass and SST.

4.  @gonzalez2018multi identify Mean Kinetic Energy (MKE), meridional
    and zonal transport components, eddies, bathymetry (depth), slope,
    nett primary productivity, distance from coast and wind at multiple
    spatial and temporal scales as influencing whale sightings.

5.  Is there a difference in habitat suitability between the north of the
    islands, the south, or around the sea mounts? --> Do a multivariate
    analyses of all variables (see @gonzalez2018multi), with data classified
    a priori into the subsets representing the three regions.

## Load the whale sighting data

```{r}
#| message: false
sights <- read_csv(
  here::here("data", "occurences_sampled.Mn.txt"),
  show_col_types = FALSE
)

# make a column with year only
sights <- sights |>
  mutate(year = year(date),
         month = month(date, label = TRUE),
         week = week(date),
         yday = yday(date))
```

## Explore the whale sighting data

The time span of the study:

```{r}
range(sights$date)
```

```{r}
yrs <- max(year(sights$date)) - min(year(sights$date))
```

The longitudinal and latitudinal range:

```{r}
range(sights$lat)
range(sights$lon)
```

```{r}
median(sights$lat)
median(sights$lon)
```

There is one outlier which I will remove:

```{r}
sights <- sights |> 
  filter(lat > min(lat))
```

From here I define the spatial extent for the study region as:

```{r}
# the extent of the full regional map
# a region around the Azores

ymin <- min(sights$lat) - 0.25; ymax <- max(sights$lat) + 0.25
xmin <- min(sights$lon) - 0.25; xmax <- max(sights$lon) + 0.25

sights_bbox <- st_bbox(c(xmin = xmin, xmax = xmax, ymax = ymax, ymin = ymin),
                       crs = CRS)

area_sf <- st_as_sfc(sights_bbox)

# EPSG:4326
# WGS 84 -- WGS84 - World Geodetic System 1984, used in GPS
st_crs(area_sf) = 4326
```

The bounding box for the study region is:

```{r}
sights_bbox
```

Start preparing all the map layers by loading the Natural Earth data for
the continent outlines:

```{r}
# Get countries
world_ne <- ne_countries(
  scale = "large",
  returnclass = "sf"
)
class(world_ne)
```

A first stab plot of the region shows:

```{r}
#| fig-cap: "The Azores Islands."
#| label: fig-basicMap
#| out-width: "90%"
#| fig-width: 5

ggplot(data = world_ne) +
  geom_sf(col = "black", fill = "black", linewidth = 0.4) +
  coord_sf(xlim = c(xmin, xmax),
           ylim = c(ymin, ymax),
           expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_map()
```

Next we get a DEM of the region using **elevatr**:

```{r}
#| eval: false
dem <- elevatr::get_elev_raster(locations = area_sf, z = 7, 
                                src = "srtm15plus",
                                clip = "bbox")

# keep for later and to prevent having to download each time
save(dem, file = here::here("data", "azores.dem"))
```

```{r}
#| cache: false
# re-use previously downloaded dem
load(here::here("data", "azores.dem"))

# make dataframe from DEM raster
dem_df <- as.data.frame(dem, xy = TRUE, na.rm = TRUE)
colnames(dem_df)[3] <- "layer"
```

## Mapping and plotting observations

Let us add all the map layer together:

i.  the bathymetry for the region,
ii. the land area polygons around the above-water regions, and
iii. all the whale sighting data.

```{r}
#| fig-cap: "A view of the aggregated whale sigtings over the period 1989-08-20 to 2022-05-07."
#| label: fig-aggregatedSightings
#| out-width: "90%"
#| fig-width: 5

# make a colourmap
library(cmocean)
cmap <- cmocean("topo")
cols <- cmap(51) # for bathy/topo
cols3 <- rainbow(33) # for observations

# create the layered graph
ggplot() +
  geom_raster(data = dem_df, aes(x = x, y = y, fill = layer)) +
  geom_sf(data = world_ne, col = "white", fill = NA, linewidth = 0.4) +
  scale_fill_gradientn(colours = cols,
                       values = scales::rescale(c(min(dem_df$layer), 0,
                                                  max(dem_df$layer))),
                       breaks = c(-4000, -2000, -1000, 0, 500, 1000, 2000),
                       name = "Elevation /\nDepth (m)") +
  geom_point(data = sights, aes(x = lon, y = lat, colour = year(date)),
             size = 0.5, shape = 4, alpha = 1) +
  scale_color_gradientn(colours = cols3,
                        name = "Year") +
  coord_sf(xlim = c(xmin, xmax),
           ylim = c(ymin, ymax),
           expand = FALSE) +
  labs(x = NULL, y = NULL) +
  theme_map()
```

Are there distribution differences across years? Difficult to see in the
above figure, so I create an animation of pooled annual observations
across years:

```{r}
#| fig-cap: "An animation of whale sigtings over the period 1989-08-20 to 2022-05-07."
#| label: fig-animationSightings
#| eval: false
#| out-width: "90%"
#| fig-width: 5

p <- ggplot() +
  geom_raster(data = dem_df, aes(x = x, y = y, fill = layer)) +
  geom_sf(data = world_ne, col = "white", fill = NA, linewidth = 0.4) +
  scale_fill_gradientn(colours = cols,
                       values = scales::rescale(c(min(dem_df$layer), 0,
                                                  max(dem_df$layer))),
                       breaks = c(-4000, -2000, -1000, 0, 500, 1000, 2000),
                       name = "Elevation /\nDepth (m)") +
  geom_point(data = sights, aes(x = lon, y = lat),
             colour = "red", size = 0.8, shape = 1, alpha = 1) +
  coord_sf(xlim = c(xmin, xmax),
           ylim = c(ymin, ymax),
           expand = FALSE) +
  theme_map() +
  labs(title = 'Year: {floor(frame_time)}', x = NULL, y = NULL) +
  transition_time(year) +
  ease_aes('linear')

gganimate::animate(p, fps = 2, nframes = 1 * yrs, device = "svg")

gganimate::anim_save(here::here("data", "sightings_anim.gif"))
```

![Animation of yearly whale
sightings.](sightings_anim.gif){#fig-animationSightings}

## Annual chl-*a* climatology

```{r}
#| cache: false
#| eval: false
chlDir <- "/Users/ajsmit/Documents/R/R_in_Ocean_Science/_development/ERDDAP/"
load(paste0(chlDir, "MODIS_chl_data.Rdata"))

# Create a column of weeks
chl_data <- chl_data |> 
  mutate(year = year(time),
         month = month(time, label = TRUE),
         week = week(time),
         yday = yday(time))
```

```{r}
#| fig-cap: "Median chl-*a* concentrations over the period 2003-01-01 to 2022-07-27."
#| label: fig-aggregateChlorophyll
#| out-width: "90%"
#| fig-width: 5
#| cache: false
#| eval: false

library(palr)
pal <- chl_pal(palette = TRUE)

chl_data |> 
  group_by(longitude, latitude) |> 
  summarise(med_chlorophyll = median(chlorophyll, na.rm = TRUE),
            .groups = "drop") |> 
  ggplot() + 
  geom_tile(aes(x = longitude, y = latitude, fill = med_chlorophyll)) +
  geom_sf(data = world_ne, col = "white", fill = "black", linewidth = 0.4) +
  scale_fill_gradientn(
    colours = pal$cols,
    trans = "log",
    breaks = c(0.1, 1, 3, 9)
  ) +
  guides(fill = guide_colourbar(title = "Chl-a [mg/L]",
                                title.position = "top",
                                direction = "horizontal",
                                barwidth = 8)) +
  scale_x_continuous(breaks = seq(-30.5, -25, 2.75)) +
  scale_y_continuous(breaks = seq(37, 39.5, 1.25)) +
  coord_sf(xlim = c(xmin, xmax),
           ylim = c(ymin, ymax),
           expand = FALSE) +
  labs(title = "MODIS Aqua annual chlorophyll-a climatology",
       x = NULL, y = NULL) +
  theme_map() +
  theme(legend.position = "bottom")
```

## Monthly chl-*a* climatology

Plots for the full regional extent:

```{r}
#| fig-cap: "Monthly climatological median chl-*a* concentrations over the period 2003-01-01 to 2022-07-27."
#| label: fig-monthlyChlorophyll
#| out-width: "90%"
#| fig-width: 5
#| cache: false
#| eval: false

chl_data |> 
  mutate(month = month(time, label = TRUE)) |> 
  group_by(longitude, latitude, month) |> 
  summarise(med_chlorophyll = median(chlorophyll, na.rm = TRUE),
            .groups = "drop") |> 
  ggplot() + 
  geom_tile(aes(x = longitude, y = latitude, fill = med_chlorophyll)) +
  geom_sf(data = world_ne, col = "white", fill = "grey50", linewidth = 0.4) +
  scale_fill_gradientn(
    colours = pal$cols,
    trans = "log",
    breaks = c(0.05, 0.1, 1),
    limits = c(0.05, 1)
  ) +
  guides(fill = guide_colourbar(title = "Chl-a [mg/m3]",
                                title.position = "top",
                                direction = "horizontal",
                                barwidth = 8)) +
  scale_x_continuous(breaks = c(-30.5, -25)) +
  scale_y_continuous(breaks = c(37, 39.5)) +
  coord_sf(xlim = c(xmin, xmax),
           ylim = c(ymin, ymax),
           expand = FALSE) +
  labs(title = "MODIS Aqua seasonal chlorophyll-a climatology",
       x = NULL, y = NULL) +
  theme_map() +
  theme(legend.position = "bottom") +
  facet_wrap(vars(month), ncol = 3)
```

I also plot the seasonal profile for the central (Ilha do Faial, Ilha do
Pico, São Jorge, Graciosa, Ilha Terceira) and eastern-most (Ilha do São
Miguel, São Pedro) island groups. The respective bounding boxes are:

```{r}
# center group
c_lonmin <- -29
c_lonmax <- -27.5
c_latmin <- 38
c_latmax <- 39

# eastern group
e_lonmin <- -26
e_lonmax <- -25
e_latmin <- 37.5
e_latmax <- 38
```

```{r}
#| fig-cap: "Monthly climatological median chl-*a* concentrations over the period 2003-01-01 to 2022-07-27 for the central island group comprised of Ilha do Faial, Ilha do Pico, São Jorge, Graciosa, and Ilha Terceira."
#| label: fig-monthlyCentralChlorophyll
#| out-width: "90%"
#| fig-width: 5
#| cache: false
#| eval: false

chl_data |>
  filter(between(longitude, c_lonmin, c_lonmax),
         between(latitude, c_latmin, c_latmax)) |>
  mutate(month = month(time, label = TRUE)) |>
  group_by(longitude, latitude, month) |>
  summarise(med_chlorophyll = median(chlorophyll, na.rm = TRUE),
            .groups = "drop") |>
  ggplot() +
  geom_tile(aes(x = longitude, y = latitude, fill = med_chlorophyll)) +
  geom_point(data = sights, aes(x = lon, y = lat),
             colour = "black", shape = 4, size = 0.5) +
  geom_sf(
    data = world_ne,
    col = "white",
    fill = "grey50",
    linewidth = 0.4
  ) +
  scale_fill_gradientn(
    colours = pal$cols,
    trans = "log",
    breaks = c(0.05, 0.1, 1),
    limits = c(0.05, 1)
  ) +
  guides(
    fill = guide_colourbar(
      title = "Chl-a [mg/m3]",
      title.position = "top",
      direction = "horizontal",
      barwidth = 8
    )
  ) +
  scale_x_continuous(breaks = c(-28.6, -27.8)) +
  scale_y_continuous(breaks = c(38.2, 38.8)) +
  coord_sf(
    xlim = c(c_lonmin, c_lonmax),
    ylim = c(c_latmin, c_latmax),
    expand = FALSE
  ) +
  labs(title = "Seasonal chl-a climatology",
       subtitle = "Central group",
       x = NULL, y = NULL) +
  theme_map() +
  theme(legend.position = "bottom") +
  facet_wrap(vars(month), ncol = 3)
```

```{r}
#| fig-cap: "Monthly climatological median chl-*a* concentrations over the period 2003-01-01 to 2022-07-27 for the eastern island group comprised of Ilha do São Miguel and São Pedro."
#| label: fig-monthlyEasternChlorophyll
#| out-width: "90%"
#| fig-width: 5
#| cache: false
#| eval: false

chl_data |>
  filter(between(longitude, e_lonmin, e_lonmax),
         between(latitude, e_latmin, e_latmax)) |>
  mutate(month = month(time, label = TRUE)) |>
  group_by(longitude, latitude, month) |>
  summarise(med_chlorophyll = median(chlorophyll, na.rm = TRUE),
            .groups = "drop") |>
  ggplot() +
  geom_tile(aes(x = longitude, y = latitude, fill = med_chlorophyll)) +
  geom_point(
    data = sights,
    aes(x = lon, y = lat),
    colour = "black",
    shape = 4,
    size = 0.5
  ) +
  geom_sf(
    data = world_ne,
    col = "white",
    fill = "grey50",
    linewidth = 0.4
  ) +
  scale_fill_gradientn(
    colours = pal$cols,
    trans = "log",
    breaks = c(0.05, 0.1, 1),
    limits = c(0.05, 1)
  ) +
  guides(
    fill = guide_colourbar(
      title = "Chl-a [mg/m3]",
      title.position = "top",
      direction = "horizontal",
      barwidth = 8
    )
  ) +
  scale_x_continuous(breaks = c(-25.8, -25.2)) +
  scale_y_continuous(breaks = c(37.6, 37.9)) +
  coord_sf(
    xlim = c(e_lonmin, e_lonmax),
    ylim = c(e_latmin, e_latmax),
    expand = FALSE
  ) +
  labs(
    title = "Seasonal chl-a climatology",
    subtitle = "Eastern group",
    x = NULL,
    y = NULL
  ) +
  theme_map() +
  theme(
    legend.position = "bottom",
  ) +
  facet_wrap(vars(month), ncol = 3)
```

## 'Climatology' of whale sightings

I create a histogram of Julian days (day of the year), which summarises
the times during the year across the observational record when most
sightings are observed for the full region, the central group, and the
eastern group.

I also want to create a plot of cl-*a* concentration with time for the
full extent, the central group, and the eastern group. All of these
plots will then be displayed in an intuitive manner so that the time of
highest chl-*a* concentration can be displayed next to the timing of
whale sightings.

```{r}
#| fig-cap: "The annual course of sightings (D-F) and the corresponding onset of the chlorophyll-*a* maximum (A-C)."
#| label: fig-timingsFigures
#| out-width: "90%"
#| cache: false
#| eval: false

# make labels to use along the x-axis in stead of yday
fig_labels <-
  data.frame(date = seq.Date(
    from = as.Date("2020-01-01"),
    to = as.Date("2020-12-01"),
    by = "2 month"
  ))
fig_labels <- fig_labels |> 
  mutate(yday = yday(date),
         month = month(date, label = TRUE))

# plot the sightings histograms
a <- sights |>
  ggplot(aes(x = yday)) +
  stat_bin(geom = "step",
           binwidth = 14,
           colour = "navy",
           linewidth = 0.75) + 
  scale_x_continuous(breaks = fig_labels$yday,
                     labels = fig_labels$month) +
  labs(x = NULL, y = "Number of sightings",
       title = "Full region")

b <- sights |>
  filter(between(lon, c_lonmin, c_lonmax),
         between(lat, c_latmin, c_latmax)) |>
  ggplot(aes(x = yday)) +
  stat_bin(geom = "step",
           binwidth = 14,
           colour = "darkcyan",
           linewidth = 0.75) + 
  scale_x_continuous(breaks = fig_labels$yday,
                     labels = fig_labels$month) +
  labs(x = NULL, y = "Number of sightings",
       title = "Central group")

c <- sights |> 
  filter(between(lon, e_lonmin, e_lonmax),
         between(lat, e_latmin, e_latmax)) |>
  ggplot(aes(x = yday)) + 
  scale_x_continuous(breaks = fig_labels$yday,
                     labels = fig_labels$month) +
  stat_bin(geom = "step",
           binwidth = 14,
           colour = "indianred3",
           linewidth = 0.75) +
  labs(x = NULL, y = "Number of sightings",
       title = "Eastern group")

# calculate a cyclic rolling mean with a window width of 30 days
# for the chlorophyll-a data
w_width <- 30
smooth_fun <- function(data) {
  chl <- data |>
    group_by(yday) |>
    summarise(med_chl = median(chlorophyll, na.rm = TRUE),
            .groups = "drop")
  
  chl_pad <-
    data.frame(yday = rbind(tail(chl[, 1], w_width/2),
                            chl[, 1],
                            head(chl[, 1], w_width/2)),
               chlorophyll = rbind(tail(chl[, -1], w_width/2),
                                   chl[, -1],
                                   head(chl[, -1], w_width/2)))

  chl_out <- chl_pad |>
    mutate(s_chl = RcppRoll::roll_mean(
      med_chl,
      n = w_width,
      fill = NA,
      align = "center"
    ))
  return(chl_out[(w_width/2 + 1):(nrow(chl_out) - w_width/2), ])
}

# full <- smooth_fun(chl_data)

# center_group <- chl_data |>
#   filter(between(longitude, c_lonmin, c_lonmax),
#          between(latitude, c_latmin, c_latmax)) |> 
#   smooth_fun()

# east_group <- chl_data |>
#   filter(between(longitude, e_lonmin, e_lonmax),
#          between(latitude, e_latmin, e_latmax)) |> 
#   smooth_fun()

# plot the chlorophyll-a data
d <- ggplot(full, aes(x = yday)) +
  geom_line(aes(y = med_chl), colour = "navy") +
  geom_line(
    aes(y = s_chl),
    colour = "yellow",
    alpha = 0.9,
    linewidth = 0.7
  ) + 
  scale_x_continuous(breaks = fig_labels$yday,
                     labels = fig_labels$month) +
  labs(title = "Full region",
       x = NULL,
       y = expression(Chl-a~(mg.m^-3)))

e <- ggplot(center_group, aes(x = yday)) +
  geom_line(aes(y = med_chl), colour = "darkcyan") +
  geom_line(
    aes(y = s_chl),
    colour = "yellow",
    alpha = 0.9,
    linewidth = 0.7
  ) + 
  scale_x_continuous(breaks = fig_labels$yday,
                     labels = fig_labels$month) +
  labs(title = "Central group",
       x = NULL,
       y = expression(Chl-a~(mg.m^-3)))

f <- ggplot(east_group, aes(x = yday)) +
  geom_line(aes(y = med_chl), colour = "indianred3") +
  geom_line(
    aes(y = s_chl),
    colour = "yellow",
    alpha = 0.9,
    linewidth = 0.7
  ) + 
  scale_x_continuous(breaks = fig_labels$yday,
                     labels = fig_labels$month) +
  labs(title = "Eastern group",
       x = NULL,
       y = expression(Chl-a~(mg.m^-3)))
    
ggarrange(d, e, f, a, b, c, align = "hv",
          nrow = 2, ncol = 3,
          labels = "AUTO")
```

What is the timing of the peak sightings and chl-*a* maximum?
@visser2011timing found that peak abundances of the blue *Balaenoptera
musculus*, fin *B. physalus*, humpback *Megaptera novaeangliae* and sei
whale *B. borealis* occurred from April to May, tracking the onset of
the spring bloom by 13 to 16 wk (depending on species). The lag period
accounts for the development of the whales' zooplankton prey, which is
positioned at a trophic position intermediate between phytoplankton and
whales. Similar findings were obtained in this study (Table 1), although
no distinction was made between cetacean species.

```{r}
#| echo: false
#| eval: false
# day on which max sightings attained for full data extent:
a_dat <- layer_data(a)
a_max <- a_dat |> 
  filter(count == max(count)) |> 
  dplyr::select(x, count)

# day on which max sightings attained for central group:
b_dat <- layer_data(b)
b_max <- b_dat |> 
  filter(count == max(count)) |> 
  dplyr::select(x, count)

# day on which max sightings attained for eastern group:
c_dat <- layer_data(c)
c_max <- c_dat |> 
  filter(count == max(count)) |> 
  dplyr::select(x, count)
```

```{r}
#| echo: false
#| eval: false
timings <- data.frame(
  row.names = c("Full extent",
                "Central group",
                "Eastern group"),
  chl_max = c(full[full$s_chl == max(full$s_chl),][, 1],
              center_group[center_group$s_chl == max(center_group$s_chl),][, 1],
              east_group[east_group$s_chl == max(east_group$s_chl),][, 1]),
  sightings = c(a_max[, 1],
                b_max[, 1],
                c_max[, 1]),
  lag = c(a_max[, 1] - full[full$s_chl == max(full$s_chl),][, 1],
          b_max[, 1] - center_group[center_group$s_chl == max(center_group$s_chl),][, 1],
          c_max[, 1] - east_group[east_group$s_chl == max(east_group$s_chl),][, 1])
)
```

```{r}
#| echo: false
#| eval: false
timings_tbl <- gt(timings, rownames_to_stub = TRUE) |> 
  tab_header(
    title = md("Table 1: Timing of the chlorophyll-*a* maximum and the peak day of sightings"),
    subtitle = md("Chlorophyll-*a* maximum and sightings timing are day of the year, and for lag it is the number of days between the chlorophyll-*a* maximum and peak observations")
  ) |> 
  tab_stubhead(label = "Extent") |> 
  cols_label(
    chl_max = html("Chlorophyll-a"),
    sightings = html("Sightings"),
    lag = html("Lag (days)")
  ) |> 
  tab_spanner(
    label = "Day of the year",
    columns = c(chl_max, sightings)
  ) |> 
  cols_align(
    align = "center",
    columns = c(chl_max, sightings, lag)
  )

timings_tbl
```

## Finding the nearest chl-*a* pixels to the whale sightings

This analysis is continued in [Spatial Localisation, Subsetting, and Aggregation of the Chlorophyll-*a* Data](../vignettes/chl_localisation.qmd).

## References

::: {#refs}
:::
