---
title: "BCB744: BCB Honours Biostatistics Exam 2022"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
## Global options
knitr::opts_chunk$set(cache = TRUE)
```


# Instructions

Please note the following instructions. Failing to comply with them in full will result in a loss of marks.

1. **<u>RMARKDOWN</u>** Submit your exam answers as a compiled Rmarkdown document (your choice of html or MS Word).

2. **<u>R SCRIPTS</u>** Produce *full Rmarkdown scripts*, including the meta-information at the top (name, date, purpose, etc.). Provide ample commentary explaining the purpose of the various tests/sections as necessary.

3. **<u>TESTING OF ASSUMPTIONS</u>** For all questions, make sure that when *formal inferential statistics are required, each is preceded by the appropriate tests for the assumptions*, i.e., state the assumptions, state the statistical procedure for testing the assumptions and mention their corresponding $H_{0}$. If a graphical approach is used to test assumptions, explain the principle behind the approach. Explain the findings emerging from the test of assumptions, and justify your selection of the appropriate inferential test (e.g. *t*-test, ANOVA, etc.) that you will use.

4. **<u>STATE HYPOTHESES</u>** When inferential statistics are required, please provide the full $H_{0}$ and $H_{A}$, and conclude the analysis with a statement of which is accepted or rejected.

5. **<u>GRAPHICAL SUPPORT</u>** All descriptive and inferential statistics must be supported by the appropriate visualisation of the results.

6. **<u>FORMATTING</u>** Pay attention to formatting -- 10% will be allocated to the appearance of the script, including considerations of aspects of the tidiness of the file, the use of appropriate headings, and adherence to code conventions (e.g. spacing etc.).

<b style='color:red;'>Submit the Rmarkdown script and associated MS Word or html file wherein you provide answers to Questions 1–7 by no later than 17:00 today. Label the script as follows:

<u>BCB744_\<Name\>_\<Surname\>_Exam.Rmd</u>, e.g. <u>BCB744_AJ_Smit_Exam.Rmd.</u></b>

<b>Send your documents by email to ajsmit@uwc.ac.za.</b>

# Question 1

## Chromosomal effects of mercury-contaminated fish consumption

These data reside in package **coin**, dataset `mercuryfish`. The dataframe contains the mercury level in blood, the proportion of cells with abnormalities, and the proportion of cells with chromosome aberrations in consumers of mercury-contaminated fish and a control group. Please see the dataset's help file for more information.

Analyse the dataset and answer the following questions:

a. Does the presence of methyl-mercury in a diet containing fish result in a higher proportion of cellular abnormalities?

b. Does the concentration of mercury in the blood influence the proportion of cells with abnormalities, and does this differ between the `control` and `exposed` groups?

c. Is there a relationship between the variables `abnormal` and `ccells`? This will have to be determined independently for the `control` and `exposed` groups.

# Question 2

## Malignant glioma pilot study

Package **coin**, dataset `glioma`: A non-randomized pilot study on malignant glioma patients with pretargeted adjuvant radioimmunotherapy using yttrium-90-biotin.

a. Do `sex` and `group` interact to affect survival time (`time`)?

b. Do `age` and `histology` interact to affect survival time (`time`)?

c. Show a full graphical exploration of the data. Are there any other remaining patterns visible in the data that should be explored statistically? If so, select the most promising and insightful question that remains, state the relevant hypothesis, and undertake the necessary analysis in full.

# Question 3

## Risk factors associated with low infant birth weight

Package **MASS**, dataset `birthwt`: This dataframe has 189 rows and 10 columns. The data were collected at Baystate Medical Center, Springfield, Mass. during 1986.

State three hypotheses and test them. Make sure one of the tests makes use of the 95% confidence interval approach rather than a formal inferential methodology.

# Question 4

## The lung capacity data

a. Using the [Lung Capacity](https://github.com/ajsmit/R_courses/raw/main/static/data/LungCapData.csv) data provided, please calculate the 95% CIs for the `LungCap` variable as a function of:

    i. `Gender`
    
    ii. `Smoke`
    
    iii. `Caesarean`
    
b. Create a graph of the mean ± 95% CIs and determine if there are statistical differences in `LungCap` between the levels of `Gender`, `Smoke`, and `Caesarean`. Do the same using a *t*-test. Are your findings the same using these two approaches?

c. Produce all the associated tests for assumptions -- i.e. the assumptions to be met when deciding whether to use a *t*-test or its non-parametric counterpart.

d. Create a combined tidy dataframe (observe tidy principles) with the estimates for the 95% CI for the `LungCap` data (`LungCap` as a function of `Gender`), estimated using both the traditional and bootstrapping approaches. Create a plot comprising two panels (one for the traditional estimates, one for the bootstrapped estimates) of the mean, median, scatter of raw data points, and the upper and lower 95% CI.

e. Undertake a statistical analysis that factors in the effect of `Age` together with one of the categorical variables on `LungCap`. What new insight does this provide?

# Question 5

## Piglet data

Here is some fictitious data for pigs raised on different diets:

```{r}
feed_1 <- c(60.8, 57.0, 65.0, 58.6, 61.7)
feed_2 <- c(68.7, 67.7, 74.0, 66.3, 69.8)
feed_3 <- c(102.6, 102.1, 100.2, 96.5, 110.3)
feed_4 <- c(87.9, 84.2, 83.1, 85.7, 90.3)

bacon <- data.frame(cbind(feed_1, feed_2, feed_3, feed_4))
```

Does feed type have an effect on the mass of pigs at the end of the experiment?

# Question 6

a. For each line of the script, below, write an English explanation for what the code does.

```{r, eval=FALSE}
ggplot(points, aes(x = group, y = count)) +
  geom_boxplot(aes(colour = group), size = 1, outlier.colour = NA) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_grid(group ~ ., scales = "free") +
    labs(x = "", y = "Number of data points") +
  theme(legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_blank())
```

b. Using the `rnorm()` function, generate some fictitious data that can be plotted using the code, above. Make sure to assemble these data into a dataframe suitable for plotting, complete with correct column titles.

c. Apply the code *exactly as stated* to the data to demonstate your understanding of the code and convince the examiner of your understanding of the correct data structure.

# Question 7

For each line of the script, below, write an English explanation for what the code does.

```{r, eval=FALSE}
SACTN_data <- read_csv("data/SACTN_day_1.csv")

SACTN_monthly <- SACTN_data %>% 
  mutate(month = month(date, label = T)) %>% 
  group_by(site, month) %>% 
  summarise(temp = mean(temp, na.rm = T))

ggplot(data = SACTN_monthly, aes(x = month, y = temp)) +
  geom_point(aes(colour = site)) +
  geom_line(aes(colour = site, group = site)) +
  labs(x = "", y = "Temperature (°C)")

ggplot(data = SACTN_monthly, aes(x = month, y = temp)) +
  geom_point(aes(colour = site)) +
  geom_line(aes(colour = site, group = site)) +
  labs(x = "", y = "Temperature (°C)") +
  facet_wrap(~site, ncol = 1)
```

# The end

<b style='color:red;'>Submit the Rmarkdown script and associated MS Word or html file wherein you provide answers to Questions 1–7 by no later than 17:00 today. Label the script as follows:

<u>BCB744_\<Name\>_\<Surname\>_Exam.Rmd</u>, e.g. <u>BCB744_AJ_Smit_Exam.Rmd.</u></b>

<b>Send your documents by email to ajsmit@uwc.ac.za.</b>

Thank you for playing along! Have a happy weekend...
