---
title: "BCB744: BCB Honours Biostatistics Exam 2022 -- ANSWER MEMO"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

**Note: the outcomes of the statistical procedures will in *most* instances have to be the same as seen here. There might, however, be substantial variations in the workflows around data 'wrangling' and these will be marked on a case-by-case basis.**

# Question 1

## Chromosomal effects of mercury-contaminated fish consumption

Load the data and examine them:

```{r}
fish <- coin::mercuryfish
head(fish)
```

### a. Does the presence of methyl-mercury in a diet containing fish result in a higher proportion of cellular abnormalities?

This question requires that two groups are compared. First, let's create a graph to explore patterns and examine normality:

```{r fig.height=3, fig.width=4}
ggplot(fish, aes(x = group, y = abnormal)) +
  geom_boxplot(notch = TRUE) +
  labs(x = "Experimental group", y = "Proportion abnormalities") +
  theme_linedraw()
```

The lower and upper hinges correspond to the first and third quartiles (the 25th and 75th percentiles). The lower and upper whiskers are defined as:

* upper -- value of 75th percentile + 1.5 * IQR

* lower -- value of 25th percentile - 1.5 * IQR

where the IQR (interquartile range) is the 75th percentile minus the 25th percentile.

Values lower or greater than the whiskers are outliers. There are no outliers. The non-overlapping notches suggest that there is a significant difference between the groups in terms of number of cellular abnormalities. We need to test this statistically.

Let's first see of the data are normally distributed. To check the normality of the data we use the Shapiro-Wilk test. The null hypothesis is:

$H_{0}$: there is no difference between the distribution of the sample data and that expected for normally distributed data.

```{r}
fish %>% 
  group_by(group) %>% 
  summarise(test = as.numeric(shapiro.test(abnormal)[2]))
```

The *p*-value is greater than 0.05 for both groups which means that the data are Gaussian, so we may proceed with a two-sided, two-sample *t*-test.

```{r}
t.test(data = fish, abnormal ~ group)
```

The output will be discussed after we have calculated the SD around the mean abnormalities for each of the groups.

```{r}
fish_stats <- fish %>% 
  group_by(group) %>% 
  summarise(mean = round(mean(abnormal), 1),
            SD = round(sd(abnormal), 1))
```


The result confirms our suspicion raised by the boxplot. There is a significant difference in the amount of abnormal cells between the control and exposed groups, with the control having a smaller proportion abnormal cells (`r fish_stats[1,2]` ± `r fish_stats[2,2]`) compared to the exposed group (`r fish_stats[1,3]` ± `r fish_stats[2,3]`) (*t* = -3.18, d.f. = 37.0, *p* < 0.005).

### b. Does the concentration of mercury in the blood influence the proportion of cells with abnormalities, and does this differ between the `control` and `exposed` groups?

We have two continuous covariates (one is a dependent and the other an independent variable) as well as a categorical grouping variable. This calls for an ANCOVA. We already know the `abnormal` column has normally distributed data. We can do the same for the independent variable, `mercury`.

$H_{0}$: (as before) there is no difference between the distribution of the sample data and that expected for normally distributed data. We plot it and test it:

```{r fig.height=3, fig.width=4}
ggplot(fish, aes(x = group, y = mercury)) +
  geom_boxplot(notch = TRUE) +
  labs(x = "Experimental group", y = "Mercury level") +
  theme_linedraw()
```

```{r}
fish %>% 
  group_by(group) %>% 
  summarise(test = as.numeric(shapiro.test(abnormal)[2]))
```

As before, we see that the data are normally distributed. There is one outlier though, and I choose to remove it before we proceed with the ANCOVA. 

The upper outliers are defined as values that exceed the 75th percentile + 1.5 * the IQR. Let's apply this as a filter:

```{r}
quants <- quantile(filter(fish, group == "exposed")$mercury, probs=c(.25, .75), na.rm = FALSE)
iqr <- IQR(filter(fish, group == "exposed")$mercury)

# we can safely apply the outlier filter to all the mercury data as the 
# mercury values associated with the control group is always less than those
# of the exposed group:
fish_filt <- fish %>% 
  filter(mercury < quants[1] + iqr)
```

There are three $H_{0}$ for the ANCOVA:

1. cellular abnormalities do not vary as a function of mercury concentration;

2. cellular abnormalities do not vary as a function of group (already tested as part of (a)!)

3. there is no interaction between the variation in abnormalities as a function of mercury concentration for the two groups.

One model to rule them all:

```{r}
fish_aov <- aov(abnormal ~ group * mercury, data = fish_filt)
summary(fish_aov)
```

```{r fig.height=3, fig.width=4}
ggplot(fish_filt, aes(x = mercury, y = abnormal)) +
  geom_point(aes(colour = group)) +
  geom_smooth(method = "lm", aes(colour = group)) +
  labs(x = "Mercury level", y = "Proportion abnormalities") +
  theme_linedraw()
```

The ANCOVA shows that only the group membership affects the cell abnormalities (d.f. = 1, Sum Sq. =  104.5, *F* = 5.256, *p* < 0.05), and not mercury concentration as a continuous covariate (but noting that the `exposed` group has a higher mercury level overall compared to the `control` group). I.e., studied separately, there is no linear relationship of cell abnormalities with mercury concentration in the `exposed` and `control` groups (*p* > 0.05). There is also no interaction between group membership and mercury concentration (*p* > 0.05).

### c. Is there a relationship between the variables `abnormal` and `ccells`? This will have to be determined independently for the `control` and `exposed` groups.

Start with a quick visual:

```{r fig.height=3, fig.width=4}
ggplot(fish, aes(x = ccells, y = abnormal)) +
  geom_point(aes(colour = group)) +
  geom_smooth(method = "lm", aes(colour = group), se = FALSE) +
  labs(x = "Proportion of C cells", y = "Abnormality proportion") +
  theme_linedraw()
```

There seems to be a linear relationship of the two covariates *within each* of the two groups. The question simply asks if there is a relationship or not (per group), so no need for an ANCOVA. 

We need to do correlations for each of the groups independently as no claim is made that one of the variables *influences* the other.

$H_{0}$: There is no association between `ccells` and `abnormal` for the `control` and `exposed` groups.

I am fitting the model using the **purrr**, **broom** and **tidyr** packages (not taught). The outcome would be the same if done applying an independent correlation using some other means; the output shows separate correlations for the `control` and `exposed` groups:

```{r}
library(purrr)
library(broom)
library(tidyr)

model <- fish %>%
  nest(data = -group) %>%
  mutate(
    cor = map(data, ~ cor.test(.x$abnormal, .x$ccells)),
    tidied = map(cor, tidy) # for fit stats
    ) 

# view all the fit stats and model estimates
model %>% 
  unnest(tidied)
```
The output is for the defualt Pearson's correlation as the data appear normal and without excessive outliers and with more-or-less equal variances (see below). We have *R*-values of 0.6 and 0.9 for the `control` and `exposed` groups, with *p* < 0.05 in both instances. Clearly there is associations between `abnormal` and `ccells` for both groups.

Although the correct way would have been to do correlations, one could have also fit a separate linear model for each group independently. We could also ask if one model is sufficient for both groups' data (do they share the same intercept and slope?), or if a separate model is necessary for the `control` and `exposed` groups. Again, not necessary for the question as asked. Let's keep it simple with straightforward null hypothesis:

$H_{0}$: There is no relationship between `ccells` and `abnormal` for the `control` and `exposed` groups.

I do not remove the outlier as before (there's one  for `exposed` on the top far right) as it sits on the straight line extended from the denser group of samples towards the lower left. We can test this afterwards for more objective insight.

Again, I am fitting the model using the **purrr**, **broom** and **tidyr** packages. The outcome would be the same if done using more manual labour; it shows separate estimates and tests for significance of the slope and intercept for the `control` and `exposed` groups:

```{r fig.height=5, fig.width=7, message=FALSE}
par(mfrow = c(2, 2))

model <- fish %>%
  nest(data = -group) %>%
  mutate(
    fit = map(data, ~ lm(abnormal ~ ccells, data = .x)),
    tidied = map(fit, tidy), # for fit stats
    glanced = map(fit, glance), # for the R2 and overall model fit
    augmented = map(fit, augment), # for the residuals, fitted values, etc.
    plotted = map(fit, plot, which = c(1, 2))
    ) 

# view all the fit stats and model estimates
model %>% 
  unnest(tidied)
```

```{r message=FALSE, warning=FALSE, echo=FALSE}
dev.off()    
```


Above, the first pair of graphs (row 1) is for the `control` and the second pair for the `exposed` groups.

The $R^{2}$ and overall model fit for the separate models can found found like this:


```{r}
model %>% 
  unnest(glanced)
```

We see that the estimate for the slope for both the `control` and `exposed` groups is significant (*t* = 3.0, *p* < 0.05 and *t* = 8.1, *p* < 0.05, respectively). The $R^{2}$ values are 0.4 and 0.8 for the `control` and `exposed` groups, respectively, and the overall *p*-values are both < 0.05. So, yes, the relationship we suspected exists. The model diagnostic plots show that the variances are reasonably equal throughout (Residuals vs Fitted) and that the data are normally distributed (Normal Q-Q), thus the assumptions for the suitability of a linear regression hold.

Compare the fit of the overall model to the separate ones FYI:

```{r}
model_overall <- lm(abnormal ~ ccells, data = fish)
summary(model_overall)
```

# Question 2

## Malignant glioma pilot study

### a. Do `sex` and `group` interact to affect survival time (`time`)?

```{r}
glioma <- coin::glioma

summary(aov(time ~ sex * group, data = glioma))
```

No, `sex` and `group` do not interact. Survival time is only affected by `group` and it is not affected by the `sex` of the patients.

### b. Do `age` and `histology` interact to affect survival time (`time`)?

```{r}
glioma <- coin::glioma

summary(aov(time ~ age * histology, data = glioma))
```

No, `age` and `histology` do not interact to affect survival `time`, but both of these individually affect `time`.

### c. Show a full graphical exploration of the data. Are there any other remaining patterns visible in the data that should be explored statistically? If so, select the most promising and insightful question that remains, state the relevant hypothesis, and undertake the necessary analysis in full.

First, let's see where some of the differences are. We already saw that `group`, `age` and `histology` affect the survival time. Let's plot these to see how:

```{r fig.height=2, message=FALSE}
library(ggpubr)
plt1 <- ggplot(glioma, aes(x = group, y = time)) +
  geom_boxplot(aes(colour = histology)) +
  labs(y = "Survival time (months)", x = "Treatment group") +
  theme_linedraw()
# plt1
plt2 <- ggplot(glioma, aes(x = age, y = time)) +
  geom_point(aes(colour = histology)) +
  geom_smooth(method = "lm", aes(colour = histology)) +
  facet_wrap(~ group) +
  labs(x = "Age (years)", y = "Survival time (months)") +
  theme_linedraw()

ggarrange(plt1, plt2, nrow = 1, common.legend = TRUE)
```

Let's do a quick ANOVA to see how the `event` column influences the data:

```{r}
summary(aov(time ~ event, data = glioma))
```

What does `event` mean? And what is right-censoring? Reading the paper, we find the answer. In this instance, patients who were still alive are right-censored (FALSE), and those who died are not (TRUE).

Let us explore graphically the effect of right-censoring in the context of treatment `group`, `histology`, `sex`, and `age`:

```{r fig.height=6, message=FALSE, warning=FALSE}
plt1 <- ggplot(glioma, aes(x = event, y = time)) +
  geom_boxplot(aes(colour = histology)) +
  facet_wrap(~ group) +
  labs(x = "Right censored", y = "Survival time (months)") +
  theme_linedraw()

plt2 <- ggplot(glioma, aes(x = age, y = time)) +
  geom_point(aes(colour = histology, shape = group)) +
  facet_wrap(~ event) +
  labs(x = "Age (years)", y = "Survival time (months)") +
  theme_linedraw()

ggarrange(plt1, plt2, nrow = 2)
```

So, right-censoring (`event`) seems to at the very least interact with `histology`, but possibly also with `group`.

$H_{0}$: right-censoring does not depend upon the influences of `histology` and/or `group`.

Let us do an ANOVA:

```{r}
summary(aov(time ~ event * histology * group, data = glioma))
```

No, my suspicion from glancing at the graph was wrong. The effect of `event` is entirely consistent within each of `histology` and `group` and no interactions are seen. The $H_{0}$ is not rejected. The main effects are all significant, confirming partly some previous analyses.

Let consider the effect of `event` in influencing `time` as a function of `age`. I ignore for this analysis the effects that `histology` and `group` have:

$H_{0}$: the effect of `event` on `time` is not influenced by the subjects' `age`.

```{r}
summary(aov(time ~ event * age, data = glioma))
```

There is no interaction of `event` with `age`, but the main effects of `event` and `age` are present. The $H_{0}$ is not rejected.

The analyses are statistically incorrect for various reasons, but none of the reasons were actually taught as part of BCB744. There are many other plausible (or not) hypotheses, and I will assess each on a case-by-case basis. I am here not so much interested in the fine details regarding the statistical correctness of the models as it is beyond the scope of the curriculum.

# Question 3

## Risk factors associated with low infant birth weight

### State three hypotheses and test them. Make sure one of the tests makes use of the 95% confidence interval approach rather than a formal inferential methodology.

```{r}
weight <- MASS::birthwt
str(weight)
```

I'll not state every conceivable hypothesis, but will provide three exampes of my own. I'll assess yours on a case-by-case basis.

All but three of the variables are categorical (binary or with three levels), so they create easy options for *t*-tests and ANOVAs. There's also options for correlations or regressions. We may decide to study interactions, but this is not a requirement for the question as it is stated. Why complicate things?

Before we do the analyses, let's create proper factor variables (there are other ways to do this):

```{r}
col_names <- c("low", "race", "smoke", "ptl", "ht", "ui", "ftv")
weight[,col_names] <- lapply(weight[,col_names] , factor)
str(weight)
```

**Hypothesis 1**

The most obvious and insightful analysis is to simply ask if the birth weight of infants is affected by whether or not the mother smokes during pregnancy. This requires comparing two samples, so either a *t*-test or it's non-parametric counterpart is required.

$H_{0}$: Smoking during pregnancy does not affect the birth mass of babies.

Make a graph to see the overall effect:

```{r fig.height=3, fig.width=4}
ggplot(weight, aes(x = smoke, y = bwt)) +
  geom_boxplot(notch = TRUE) +
  labs(x = "Does the mother smoke?", Y = "Birth mass of infant (g)") +
  theme_linedraw()
```

The notches do not overlap so an effect is possible. Are the data normally distributed?

```{r}
weight %>% 
  group_by(smoke) %>% 
  summarise(test = as.numeric(shapiro.test(bwt)[2]))
```

Yes, the data are normally distributed. We may proceed with a normal *t*-test. I choose to ignore the outlier.

```{r}
t.test(bwt ~ smoke, data = weight)
```

```{r}
weight %>% 
  group_by(smoke) %>% 
  summarise(mean = round(mean(bwt), 1),
            sd = round(sd(bwt), 1)) %>% 
  ungroup()
```

So, yes, smoking is a terrible idea during preganancy (in fact, it is always terrible). The infants of mothers who smoke weigh significantly less (2771.9 ± 659.6 g) than those of healthy mothers (3056.0 ± 752.7g) (*t* = 2.73, d.f. = 170.1, *p*-value < 0.05).

**Hypothesis 2**

Does `race` affect infant birth mass? I can't think that it would, but let's see.

$H_{0}$: Race does not affect infant birth mass.

We start the exploration with a graph:

```{r fig.height=3, fig.width=4}
ggplot(weight, aes(x = race, y = bwt)) +
  geom_boxplot(notch = TRUE) +
  labs(x = "Mother's race", Y = "Birth mass of infant (g)") +
  theme_linedraw()
```

It seems that there's no difference, but need to verify. We already know that birth mass is normally distributed, but let's see if this is still so when data are stratified across race groups ('black' seems to show a non-normal distribution):

```{r}
weight %>% 
  group_by(race) %>% 
  summarise(test = as.numeric(shapiro.test(bwt)[2]))
```

Nope, all's good. An ANOVA will work for this simple analysis:

```{r}
aov1 <- aov(data = weight, bwt ~ race)
summary(aov1)
```

```{r}
weight %>% 
  group_by(race) %>% 
  summarise(mean = round(mean(bwt), 1),
            sd = round(sd(bwt), 1)) %>% 
  ungroup()
```

Interesting! There is a difference. Where is the difference?

```{r fig.height=3, fig.width=4}
plot(TukeyHSD(aov1))
```

Hmmm. White babies weigh significantly more than either black or 'other' babies (d.f. = 2, Sum Sq. = 5015725, *F* = 4.913, *p* < 0.05). The mass of white babies at birth is 3102.7	± 727.9 g, whilst that of black and 'other' babies is 2719.7 ± 638.7 g and 2805.3 ± 722.2	g, respectively. Thinking about it now, I can suggest some reasons for why this might be so. Can you?

**Hypothesis 3**

$H_{0}$: The mass of babies at birth is not affected by the mothers' `age`.

Make a graph:

```{r fig.height=3, fig.width=4}
ggplot(weight, aes(x = age, y = bwt)) +
  geom_point() +
  labs(x = "Mother's age", y = "Birth mass of infant (g)") +
  theme_linedraw()
```

I think the birth mass data are normally distributed, but we have only checked when stratified across `smoke` and `race`. What if we look at all together?

```{r}
shapiro.test(weight$bwt)
```

Yeah, all's good! But what about the `age` variable?

```{r}
shapiro.test(weight$age)
```

Hmmm. Not so normal! How so?

```{r fig.height=3, fig.width=4}
hist(weight$age)
```

It is not surprising that people tend to have babies when they are younger. Let's fix the skewed data by transforming it. I tested a few transformations, but settled on the `log10(x)` transformation for moderately skewed data:

```{r fig.height=3, fig.width=4}
hist(log10(weight$age))
```

```{r}
shapiro.test(log10(weight$age))
```

```{r fig.height=3, fig.width=4}
ggplot(weight, aes(x = log10(age), y = bwt)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Mother's age (log10(years))", y = "Birth mass of infant (g)") +
  theme_linedraw()
```

Do a linear regression:

```{r}
mod <- lm(bwt ~ log10(age), data = weight)
summary(mod)
```

```{r fig.height=3, fig.width=7, message=FALSE}
par(mfrow = c(1, 2))
plot(mod, which = c(1, 2))
```

The regression diagnostic plots seem fine (normally distributed data in the Normal Q-Q plot, and equal variances in the Residual vs Fitted plot). 

That the intercept is significantly positive should not be a surprise -- no baby will have zero mass when born! We can safely ignore the intercept, therefore. The $\beta$ coefficient does not differ from 0, so therefore we can conclude that age does not affect the infants' birth mass (F = 0.9, $R^{2}$ = 0.0, *p* > 0.05).

# Question 4

## The lung capacity data

### a. Using the Lung Capacity data provided, please calculate the 95% CIs for the `LungCap` variable as a function of: i. `Gender`, ii. `Smoke` and iii. `Caesarean`.

I calculate the CIs using bootstrapping using the traditional method (any method will do as this was not specified in the question):

```{r}
library(rcompanion)

lungs <- read.csv("../static/data/LungCapData.csv", sep = "")

(lungs_gender <- groupwiseMean(LungCap ~ Gender,
              data = lungs,
              trad = TRUE))

(lungs_smoke <- groupwiseMean(LungCap ~ Smoke,
              data = lungs,
              trad = TRUE))

(lungs_caesarean <- groupwiseMean(LungCap ~ Caesarean,
              data = lungs,
              trad = TRUE))
```

### b. Create a graph of the mean ± 95% CIs and determine if there are statistical differences in `LungCap` between the levels of `Gender`, `Smoke`, and `Caesarean`. Do the same using a *t*-test. Are your findings the same using these two approaches?

```{r fig.height=6, fig.width=7, message=FALSE}
plt1 <- ggplot(lungs_gender, aes(x = Gender, y = Mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = Trad.lower, ymax = Trad.upper)) +
  labs(y = "Lung capacity") +
  theme_linedraw()

plt2 <- ggplot(lungs_smoke, aes(x = Smoke, y = Mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = Trad.lower, ymax = Trad.upper)) +
  labs(y = "Lung capacity") +
  theme_linedraw()

plt3 <- ggplot(lungs_caesarean, aes(x = Caesarean, y = Mean)) +
  geom_point() +
  geom_errorbar(aes(ymin = Trad.lower, ymax = Trad.upper)) +
  labs(y = "Lung capacity") +
  theme_linedraw()

ggarrange(plt1, plt2, plt3, nrow = 3, labels = "AUTO")
```

```{r}
t.test(data = lungs, LungCap ~ Gender)

t.test(data = lungs, LungCap ~ Smoke)

t.test(data = lungs, LungCap ~ Caesarean)
```

Both `Gender` and `Smoke` affect `LungCap` but `Caesarean` does not. This finding is consistent across the plots of confidence intervals and the *t*-tests.

### c. Produce all the associated tests for assumptions -- i.e. the assumptions to be met when deciding whether to use a *t*-test or its non-parametric counterpart.

Are the data normal?

```{r}
lungs %>% 
  group_by(Gender) %>% 
  summarise(test = as.numeric(shapiro.test(LungCap)[2]))

lungs %>% 
  group_by(Smoke) %>% 
  summarise(test = as.numeric(shapiro.test(LungCap)[2]))

lungs %>% 
  group_by(Caesarean) %>% 
  summarise(test = as.numeric(shapiro.test(LungCap)[2]))
```

```{r fig.height=2}
h1 <- ggplot(data = lungs, aes(x = LungCap, fill = Gender)) +
  geom_histogram(position = "dodge", binwidth = 1, alpha = 0.8) +
  geom_density(aes(y = 1*..count.., fill = Gender), colour = NA, alpha = 0.4) +
  labs(x = "Lung capacity", y = "Count") + theme_linedraw()

h2 <- ggplot(data = lungs, aes(x = LungCap, fill = Smoke)) +
  geom_histogram(position = "dodge", binwidth = 1, alpha = 0.8) +
  geom_density(aes(y = 1*..count.., fill = Smoke), colour = NA, alpha = 0.4) +
  labs(x = "Lung capacity", y = "Count") + theme_linedraw()

h3 <- ggplot(data = lungs, aes(x = LungCap, fill = Caesarean)) +
  geom_histogram(position = "dodge", binwidth = 1, alpha = 0.8) +
  geom_density(aes(y = 1*..count.., fill = Caesarean), colour = NA, alpha = 0.4) +
  labs(x = "Lung capacity", y = "Count") + theme_linedraw()

ggarrange(h1, h2, h3, ncol = 3, labels = "AUTO")
```

The data are right skewed for females, non-smokers, and babies produced from non-cesarean births.

Are the variances equal?

```{r}
lungs %>% 
  group_by(Gender) %>% 
  summarise(sample_var = var(LungCap))

lungs %>% 
  group_by(Smoke) %>% 
  summarise(sample_var = var(LungCap))

lungs %>% 
  group_by(Caesarean) %>% 
  summarise(sample_var = var(LungCap))
```

The variances do not differ by more than about four times between the levels in each grouping variable, so it seems to be homoscedastic.

### d. Create a combined tidy dataframe (observe tidy principles) with the estimates for the 95% CI for the `LungCap` data (`LungCap` as a function of `Gender`), estimated using both the traditional and bootstrapping approaches. Create a plot comprising two panels (one for the traditional estimates, one for the bootstrapped estimates) of the mean, median, scatter of raw data points, and the upper and lower 95% CI.

Confidence intervals (CIs) are used to indicate how accurate a calculated statistic is likely to be, and here we calculate the accuracy of the mean.

All the various kinds of CI and mean estimates can be added into one dataframe with one line of code, which can then be tidied:

```{r}
(lungs_gender <- groupwiseMean(LungCap ~ Gender,
              data = lungs,
              boot = TRUE,
              traditional = TRUE,
              norm = TRUE,
              basic = TRUE,
              percentile = TRUE))
```

Now we `gather()` the data (as taught), or we can use the more recent `pivot_longer()` fundtion. I choose the latter:

```{r}
lungs_gender_tidy <- lungs_gender %>% 
  select(-n, -Conf.level) %>%
  pivot_longer(cols = c("Mean", "Boot.mean"),
               names_to = "mean_type",
               values_to = "mean") %>%
  pivot_longer(cols = c(2:9),
               names_to = c("ci_type", "direction"),
               names_sep = "\\.", # or names_sep = [.]
               values_to = "ci")

# . is a special character intended to match any single character. If you want
# to specify that your variable names are separated by the actual . character
# itself, you need to use names_sep="\\." to escape it.

str(lungs_gender_tidy)
```

This dataframe is fully tidy, but it is difficult to plot. In some instances, data being *too* tidy is not ideal, so lets reshape it so the upper and lower CIs are in separate columns. So, I use `pivot_wider()`:

```{r}
lungs_gender_tidy2 <- lungs_gender_tidy %>% 
  pivot_wider(names_from = "direction",
              values_from = "ci")
```

Okay, much easier to plot. Let's make the plot:

```{r fig.height=3, fig.width=4}
ggplot(filter(lungs_gender_tidy2, mean_type == "Mean"),
       aes(x = Gender, y = mean)) +
  geom_point(shape = 2, colour = "red") +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  facet_wrap(~ ci_type) +
  labs(y = "Lung capacity") +
  theme_linedraw()
```

Putting the full plot together (using all four CI estimate methods, just because I can):

```{r fig.height=3, fig.width=4, message=FALSE, fig.cap = "Plot of confidence intervals for the estimate of the mean produced using four approaches; solid red point -- mean calculated using `mean()` function; blue triangle -- median calculated using `median()`; red triangle -- mean calculated by bootstrapping"}
ggplot(data = filter(lungs_gender_tidy2, mean_type == "Mean"), aes(x = Gender)) +
  geom_errorbar(aes(ymin = lower, ymax = upper)) +
  geom_jitter(data = lungs, aes(x = Gender, y = LungCap),
              colour = "grey50", shape = 1, width = 0.1, alpha = 0.5) +
  geom_point(data = lungs, aes(x = Gender, y = mean(LungCap)),
             colour = "red3", shape = 2) +
  geom_point(data = lungs, aes(x = Gender, y = median(LungCap)),
             colour = "blue3", shape = 6) +
  geom_point(data = filter(lungs_gender_tidy2, mean_type == "Mean"),
             aes(x = Gender, y = mean),
             colour = "deeppink2", shape = 16, size = 2) +
  facet_wrap(~ ci_type) +
  labs(y = "Lung capacity") +
  theme_linedraw()
```

### e. Undertake a statistical analysis that factors in the effect of `Age` together with one of the categorical variables on `LungCap`. What new insight does this provide?

$H_{0}$: The mother's age and smoking do not affect the infant's lung capacity.

```{r}
summary(aov(data = lungs, LungCap ~ Smoke * Age))
```

Do not accept the alternative hypothesis, i.e. the two factors do not interact (d.f. = 1, Sum Sq. = 2, *F* = 0.78, *p* > 0.05); however, `Age` and `Smoke` independently have an effect overall (d.f. = 1, Sum Sq. = 3422, *F* = 1491.64, *p* < 0.05 and d.f. = 1, Sum Sq. = 53, *F* = 22.98, *p* < 0.05, respectively).

However, we have seen earlier that the data are somewhat right skewed, so we could transform to normality. Variances are equal(-ish).

# Question 5

## Piglet data

### Does feed type have an effect on the mass of pigs at the end of the experiment?

```{r}
feed_1 <- c(60.8, 57.0, 65.0, 58.6, 61.7)
feed_2 <- c(68.7, 67.7, 74.0, 66.3, 69.8)
feed_3 <- c(102.6, 102.1, 100.2, 96.5, 110.3)
feed_4 <- c(87.9, 84.2, 83.1, 85.7, 90.3)

bacon <- data.frame(cbind(feed_1, feed_2, feed_3, feed_4))
str(bacon)
```

Make tidy data:

```{r}
bacon_df <- bacon %>% 
  pivot_longer(cols = 1:4,
    names_to = "feed",
               values_to = "mass")

head(bacon_df)
```

Are the data normal?

```{r}
bacon_df %>% 
  group_by(feed) %>% 
  summarise(test = as.numeric(shapiro.test(mass)[2]))
```

Yes, data are all normal.

Now a simple ANOVA will do the trick.

$H_{0}$: Feed does not affect the piglets' mass.

```{r}
summary(aov(mass ~ feed, data = bacon_df))
```

Accept the alternative hypothesis. Feed does indeed affect the piglets' mass at the end of the experiment (d.f. = 3, Sum Sq. = 5138, *F* = 132, *p* < 0.05).

Visualise the data:

```{r fig.height=3, fig.width=4}
ggplot(bacon_df, aes(x = feed, y = mass)) +
  geom_boxplot() +
  labs(x = "Feed type", y = "Mass (dg)") +
  theme_linedraw()
```

# Question 6

### a. For each line of the script, below, write an English explanation for what the code does.

```{r, eval=FALSE}
# create a plot using the `points` data, and plot `count` as a function of
# the `group` it belongs to:
ggplot(points, aes(x = group, y = count)) +
# create a boxplot and allow the boxes' colour to vary accrding to the group
# it belongs to; the size of the lines is 1 and the outliers are switched off:
  geom_boxplot(aes(colour = group), size = 1, outlier.colour = NA) +
# adjust the plotting position so that the boxes do not plot on top of each
# other; increase the transparency to 0.2:
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
# make sure that a separate panel is created for the data that belongs in
# each group; set each panel's y-axis scaling to be independent of each other:
  facet_grid(group ~ ., scales = "free") +
# set-up the labels for the x- and y-axes:
  labs(x = "", y = "Number of data points") +
# create a theme which does not display a legend; the strips along the right of
# each panel are set to empty (no colour); also omit the text in the  strips:
  theme(legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_blank())
```

### b. Using the `rnorm()` function, generate some fictitious data that can be plotted using the code, above. Make sure to assemble these data into a dataframe suitable for plotting, complete with correct column titles.

There are many ways to generate that data. I use a vectorised approach, which was not taught. Other correct approaches will work too, and will be marked accordingly.

```{r}
n <- 50
x <- c(3, 23, 13, 8) # thumb-suck some values for the mean
y <- c(3.4, 4.1, 5.0, 2.9) # thumb-suck some values for the sd

fun <- function(x, y) {
  vals <- rnorm(n = n, mean = x, sd = y)
  return(vals)
}

points <- data.frame(group = rep(c("A", "B", "C", "D"), each = n),
                     count = as.vector(mapply(fun, x, y, SIMPLIFY = TRUE)))
```

### c. Apply the code *exactly as stated* to the data to demonstate your understanding of the code and convince the examiner of your understanding of the correct data structure.

```{r fig.height=3, fig.width=4}
ggplot(points, aes(x = group, y = count)) +
  geom_boxplot(aes(colour = group), size = 1, outlier.colour = NA) +
  geom_point(position = position_jitter(width = 0.2), alpha = 0.3) +
  facet_grid(group ~ ., scales = "free") +
    labs(x = "", y = "Number of data points") +
  theme(legend.position = "none",
    strip.background = element_blank(),
    strip.text = element_blank())
```

# Question 7

### For each line of the script, below, write an English explanation for what the code does.

```{r, eval=FALSE}
# load the data:
SACTN_data <- read_csv("data/SACTN_day_1.csv")

# specify the name of the data to process, and assign the output to a
# new df:
SACTN_monthly <- SACTN_data %>% 
# create a new variable called `month` and add the month's name to it:  
  mutate(month = month(date, label = T)) %>% 
# set up grouping variables, so each subsequent processing is done independently
# for each combination of `site` and `month`:
  group_by(site, month) %>% 
# for each group of `site` x `month`, calculate the mean temperature:  
  summarise(temp = mean(temp, na.rm = T))

# tell ggplot to use the new df just created, and inform which the x and y
# variables are for plotting
ggplot(data = SACTN_monthly, aes(x = month, y = temp)) +
# plot `temp` as a function of `month` as points; colour the points by `site`:  
  geom_point(aes(colour = site)) +
# connect the points with a line, independently for each `site`; also make
# sure the colour varies by `site`:
  geom_line(aes(colour = site, group = site)) +
# create the correct titles for the x- and y-axes:  
  labs(x = "", y = "Temperature (°C)")

# the first four lines below are as for the previous ggplot graph:
ggplot(data = SACTN_monthly, aes(x = month, y = temp)) +
  geom_point(aes(colour = site)) +
  geom_line(aes(colour = site, group = site)) +
  labs(x = "", y = "Temperature (°C)") +
# make sure a panel is created for each `site` rather than plotting all
# the lines on one panel as before:
  facet_wrap(~site, ncol = 1)
```
