---
title: "BCB744 Task G"
format: 
  html:
    fig-format: svg
    fig_retina: 2
    fig-dpi: 400
params: 
  hide_answers: true
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.asp = 0.65,
  out.width = "100%",
  fig.align = "center"
)
library(tidyverse)
library(ggpubr)
library(ggthemes)
```

# [[Assessment Sheet](BCB744_Task_H_Surname.xlsx)]{.my-highlight} {#sec-assessment}

# 9. Simple Linear Regressions

## Question 1

a. Examine the content of the regression model object `sparrows.lm` produced in the [linear regression](https://tangledbank.netlify.app/BCB744/basic_stats/09-regressions.html) chapter. Explain the meaning of the components within, and tell us how they relate to the model summary produced by `summary(eruption.lm)`.
b. Using the values inside of the model object, write some R code to show how you can reconstruct the observed values for the dependent variable from the residuals and the fitted values.
c. Fit a linear regression through the model residuals (use `sparrows.lm`). Explain your findings.
d. Similarly, fit a linear regression through the the fitted values. Explain.

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r}

```

- ✓

`r if (params$hide_answers) ":::"`

## Question 2

Find your own **two datasets** and do a full regression analysis on it. Briefly describe the data and the reason for their existence. Start with setting the appropriate hypotheses. Follow with an EDA, make some exploratory figures, fit the linear model, make a figure with the fitted linear model, provide diagnostic plots to test assumptions, and present the output in a Results section suitable for publication.

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r}

```

- ✓

`r if (params$hide_answers) ":::"`

## Question 3

Find your own **two datasets** and do a full correlation analysis on it. Briefly describe the data and the reason for their existence. Start with setting the appropriate hypotheses. Follow with an EDA, make some exploratory figures, fit the correlation, make figures with the fitted correlation line, provide diagnostic plots to test assumptions, and present the output in a Results section suitable for publication.

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r}

```

- ✓

`r if (params$hide_answers) ":::"`
