---
title: "BCB744: Summative End-of-Intro-R Task"
subtile: "Final CA Task and Exam Preparation"
date: "Due: 1 March 2023 at 23:59"
format:
  html:
    code-fold: false
    toc-title: "On this page"
    standalone: true
    toc-location: right
    page-layout: full
    embed-resources: false
number-sections: false
---

## Honesty Pledge

**This assignment requires that you work as an individual and not share
your code, results, or discussion with your peers. Penalties and
disciplinary action will apply if you are found cheating.**

::: callout-note
## Acknowledgement of the Pledge

Copy the statement, below, into your document and replace the
underscores with your name acknowledging adherence to the UWC's Honesty
Pledge.

**I, \_\_\_\_\_\_\_\_\_\_\_\_, hereby state that I have not communicated
with or gained information in any way from my peers and that all work is
my own.**
:::

## Format and mode of submission

This Assignment requires submission as both a Quarto (.qmd) file and the
knitted .html product. You are welcome to copy any text from here to use
as headers or other pieces of informative explanation to use in your
Assignment.

## Style and organisation

As part of the assessment, we will look for a variety of features,
including, but not limited to the following:

-   Content:
    -   Questions answered in order
    -   A written explanation of approach included for each question
    -   Appropriate formatting of text, for example, fonts not larger
        than necessary, headings used properly, etc. Be sensible and
        tasteful.
-   Code formatting:
    -   Use **Tidyverse** code and [style conventions](https://style.tidyverse.org/)
    -   No more than \~80 characters of code per line (pay particular
        attention to the comments)
    -   Application of [R code
        conventions](http://adv-r.had.co.nz/Style.html), e.g. spaces
        around `<-`, after `#`, after `,`, etc.
    -   New line for each `dplyr` function (lines end in `%>%`) or
        `ggplot` layer (lines end in `+`)
    -   Proper indentation of pipes and `ggplot()` layers
    -   All chunks labelled without spaces
    -   No unwanted / commented out code left behind in the document
-   Figures:
    -   Sensible use of themes / colours
    -   Publication quality
    -   Informative and complete titles, axes labels, legends, etc.
    -   No redundant features or aesthetics

## Packages

For this assignment, you will have to install the **nycflights13**
package. The package contains the dataset `flights` and some associated
meta-data, all of which you need to complete the questions below. You
will also need **tidyverse** and **ggpubr**

```{r load-packages, message=FALSE}
#| echo: true
#| eval: true
library(tidyverse)
library(ggpubr)
library(nycflights13)
```

## Questions

### Question 1

Insert **Task G** which can be found
[here](../BCB744/intro_r/11-tidy.html#pivot_longer).

### Question 2

What are the 10 most common destinations for flights from NYC airports
in 2013, and what is the total distance travelled to each of these
airports? Make a 2-panel figure and display these data graphically.

```{r}
#| echo: true
#| eval: true
#| fig-cap: "The figure for Question 2."
tab1 <- flights %>%
  group_by(dest) %>%  
  summarise(n = n(),
            total_distance = sum(distance)) %>% 
  arrange(desc(n)) %>%
  slice(1:10)

plt1 <- ggplot(tab1) +
  geom_col(aes(x = dest, y = n), fill = "grey90", alpha = 0.7,
           colour = "indianred") +
  theme_minimal() +
  labs(x = "Airport code",
       y = "Number of\ndestimations")
plt2 <- ggplot(tab1) +
  geom_col(aes(x = dest, y = total_distance * 0.621371), # convert to km
           fill = "grey90", alpha = 0.7, colour = "indianred") +
  theme_minimal() +
  labs(x = "Airport code",
       y = "Total\ndistance (km)")
ggarrange(plt1, plt2, ncol = 1, labels = "AUTO")
```

### Question 3

Which airlines have the most flights departing from NYC airports in
2013? Make a table that lists these in descending order of frequency and
shows the number of flights for each airline. In your table, list the
names of the airlines as well. *Hint:* you can use the `airlines`
dataset to look up the airline name based on `carrier` code.

```{r}
#| echo: true
#| eval: true
popular_destinations <- flights %>%
  count(carrier) %>%
  arrange(desc(n)) %>%
  inner_join(airlines, by = "carrier") %>% 
  as_tibble()
head(popular_destinations, n = 16)
```

<!-- The carrier with the highest number of flights departing from NYC airports in 2013 is United Airlines, followed by JetBlue Airways and ExpressJet Airlines. -->

### Question 4

Consider only flights that have non-missing arrival delay information.
Your answer should include the name of the carrier in addition to the
carrier code and the values asked.

a.  Which carrier had the highest mean arrival delay?

b.  Which carrier had the lowest mean arrival delay?

Make sure that your answer includes the name of the carrier and the
calculated mean (±SD) delay times, and use a sensible number of decimal
digits.

```{r}
#| echo: true
#| eval: true
flights %>%
  filter(!is.na(arr_delay)) %>%
  group_by(carrier) %>%
  summarise(mean_arr_delay = round(mean(arr_delay), 1),
            sd_arr_delay = round(sd(arr_delay), 1)) %>%
  arrange(desc(mean_arr_delay)) %>%
  inner_join(airlines, by = "carrier") %>%
  slice(c(1, n())) %>% 
  as_tibble()
```

The longest arrival delay was on Frontier Airlines that, on average, arrived 21.9 ± 61.5 (mean ± SD) minutes late. On the contrary, Alaska Airlines typically arrived earlier than anticipated by 9.9 ± 36.5 (mean ± SD) minutes.

### Question 5

What were the mean values for the weather variables at the origin
airport on the top 10 days with the highest departure delays? Contrast
this with a similar view on the 10 days with the lowest departure
delays. Your table(s) should include the names of origin airports, the
dates with the highest (lowest) departure delays, and the mean (±SD)
weather variables on these days.

Can you make any inferences about the effect of weather conditions on
flight delays? Are there any problems with this analysis, and how might
you improve this analysis for a clearer view of the effect of weather
conditions on the ability of flights to depart on time?

```{r}
#| echo: true
#| eval: true
#| warning: false
top_delay <- bind_rows(
    flights %>% slice_max(order_by = dep_delay, n = 10),
    flights %>% slice_min(order_by = dep_delay, n = 10),
) %>% 
  select(carrier, flight, tailnum, dep_delay, month, day, origin) %>%
  inner_join(weather, by = c("origin", "month", "day")) %>% 
  pivot_longer(cols = temp:visib,
               names_to = "weather_var",
               values_to = "value") %>% 
  na.omit() %>% 
  group_by(carrier, flight, tailnum, dep_delay, weather_var) %>% 
  summarise(mean_weather_var = round(mean(value, na.rm = TRUE), ),
            sd_weather_var = round(sd(value, na.rm = TRUE), 1)) %>% 
  unite("mean_sd", mean_weather_var:sd_weather_var, sep = " ± ") %>% 
  arrange(desc(dep_delay)) %>% 
  pivot_wider(names_from = weather_var, values_from = mean_sd) %>% 
  as_tibble()
top_delay
```

Nothing obvious I can see about the effect of weather variables in affecting the departure delay. We would need to do some multivariate stats to assess.

### Question 6

Partition each day into four equal time intervals, e.g. 00:01-06:00,
06:01-12:00, 12:01-18:00, and 18:01-00:00.

a.  At each time interval, what is the proportion of flights delayed at
    departure? Illustrate your finding in a figure.

```{r}
#| echo: true
#| eval: true
# Create time of day variable
flights_tod <- flights %>%
  mutate(time_of_day = case_when(
    sched_dep_time >= 001  & sched_dep_time <= 600  ~ "00:01-06:00",
    sched_dep_time >= 601  & sched_dep_time <= 1200 ~ "06:01-12:00",
    sched_dep_time >= 1201 & sched_dep_time <= 1800 ~ "12:01-18:00",
    sched_dep_time >= 1801                          ~ "18:01-00:00"
  )) %>% 
  mutate(day_of_week = wday(time_hour))

# Find proportion of delayed flights for each time of day
delay_time <- flights_tod %>%
  filter(!is.na(dep_delay)) %>%
  mutate(dep_delayed = ifelse(dep_delay > 0, "delayed", "ontime")) %>%
  count(time_of_day, dep_delayed) %>%
  group_by(time_of_day) %>%
  mutate(prop_delayed = n / sum(n)) %>%
  filter(dep_delayed == "delayed") %>%
  arrange(prop_delayed) %>% 
  as_tibble()
head(delay_time)
```

```{r}
#| fig-cap: "The figure for Question 6a."
#| echo: true
#| eval: true
ggplot(delay_time, aes(x = time_of_day, y = prop_delayed)) +
  geom_col(fill = "grey90", alpha = 0.7, colour = "indianred") +
  theme_minimal() +
  labs(x = "Time of day",
       y = "Proportion of\nflights delayed")
```

About 21% of flights are delayed between midnight and 6:00, 26% are delayed between 6:00-12:00, 46% delays between 12:00-18:00, and 52% delays between 18:00pm and midnight. As the day progresses, the better the chance is of there being a delay.

b.  Based on your analysis, does the chance of being delayed change
    throughout the day?

See answer to Q.6a.

c.  For each weekday (1-7) aggregated over 2013, which of the time
    intervals has the most flights? Create a figure to show your
    finding.

```{r}
#| fig-cap: "The figure for Question 6c."
#| echo: true
#| eval: true
flights_tod %>% 
  group_by(time_of_day, day_of_week) %>% 
  summarise(n_flights = n()) %>% 
  ggplot(aes(x = day_of_week, y = n_flights)) +
    geom_col(fill = "grey90", alpha = 0.7, colour = "indianred") +
    theme_minimal() +
    labs(x = "Time of day",
         y = "Number of\nflights") +
    facet_wrap(~time_of_day)
```

Midnight to 6-am has the fewest flights, regardless of what day of the week we are looking at -- although Sundays (day 1) and Saturday (day 7) have slightly fewer flights. Similarly, Sun?Sat have slightly fewer flights between 6-am to 6-pm, although the total number of flights are much higher (closer to 20000 flights per day). Evenings from 6-pm to midnight the flights decrease in numbers, and Saturdays have fewer flights during this time than other days.

### Question 7

Find the 10 planes that spend the longest time (cumulatively) in the air.

a.  For each model, what are the cumulative and mean flight times? In this table, also mention their type, manufacturer, model, number of engines, and speed.

```{r}
#| echo: true
#| eval: true
cum_flights <- flights %>%
  group_by(tailnum) %>% 
  summarise(cum_air_time = sum(air_time),
            mean_air_time = round(mean(air_time, na.rm = TRUE), 1)) %>% 
  slice_max(order_by = cum_air_time, n = 10) %>% 
  inner_join(planes, by = "tailnum") %>%
  select(tailnum, cum_air_time, mean_air_time, type, manufacturer, model, engines, speed) %>% 
  as_tibble()
head(cum_flights)
```

b.  Create a table that lists, for each air-plane identified in (a.), each flight (and associated destination) that it undertook during 2013. 

```{r}
#| echo: true
#| eval: true
flight_dest <- flights %>% 
  filter(tailnum %in% cum_flights$tailnum) %>% 
  select(tailnum, origin, dest, time_hour) %>% 
  as_tibble()
head(flight_dest)
```

c.  Summarise all the in formation in (b.) on a map of the USA. Use
    lines to connect departure and destination locations (each
    labelled). Different facets in the figure must be used for each of
    the 10 planes. You can use the `alpha` value in **ggplot2** such
    that the colour intensity of overlapping flight lines is
    proportional to the number of flights taken along the path. For
    bonus marks, ensure that the curvature of Earth is indicated in the
    flight lines. *Hint:* such lines would display as curves, not straight
    lines.

```{r}
# to be done in due course...
```


### Question 8

Limit this analysis to only the coldest three winter and warmest three
summer months (show evidence for how this is decided). For each of these
two seasons, create a visualisation to explore if there is a
relationship between the mean daily departure delay and the mean daily
temperature. Be as economical with your code as possible.

Discuss your answer.

```{r ec}
#| echo: true
#| eval: true
seas <- weather %>% 
  group_by(month) %>% 
  summarise(mean_temp = round(mean(temp, na.rm = TRUE), 1)) %>% 
  as_tibble()
seas

ggplot(seas, aes(x = month, y = (mean_temp - 32)*5/9)) +
  geom_col(fill = "grey90", alpha = 0.7, colour = "indianred") +
  scale_x_continuous(breaks = seq(2, 12, by = 2)) +
  theme_minimal() +
  labs(x = "Month",
       y = "Mean temp. (°C)")
```

The coldest months are December, January, and February. The warmest time of year is during June, July, August.

```{r}
#| echo: true
#| eval: true
flights %>% 
  filter(month %in% c(12, 1, 2, 6, 7, 8)) %>% 
  inner_join(weather, by = c("origin", "month", "day")) %>% 
  group_by(month) %>% 
  summarise(mean_dep_delay = mean(dep_delay, na.rm = TRUE),
            mean_temp = round(mean(temp, na.rm = TRUE), 1)) %>% 
  mutate(seas = c("winter", "winter", "summer", "summer", "summer", "winter")) %>% 
  ggplot(aes(x = (mean_temp - 32)*5/9, y = mean_dep_delay)) +
    geom_point(aes(col = seas)) +
  theme_minimal() +
  labs(x = "Mean temp. (°C)", y = "Mean delay (min)")
```

It seems that, in general, shorter delays are experienced during winter months. To fully assess the effect of weather variables on delays, a more detailed statistical analysis will be required.

## Submission instructions

::: callout-important
## Submission instructions

Submit your .qmd and .html files wherein you provide answers to these
Questions by no later than 1 March 2023 at 23:59.

Label the files as follows:

-   `BCB744_<first_name>_<last_name>_Final.qmd`, and

-   `BCB744_<first_name>_<last_name>_Final.html`

(the `<` and `>` must be omitted as they are used in the example as
field indicators only).

Failing to follow these instructions carefully, precisely, and
thoroughly will cause you to lose marks, which could cause a significant
drop in your score as formatting counts for 15% of the final mark (out
of 100%).

Submit your Tasks on the Google Form when ready.
:::
