---
title: "BCB744 Task F"
format: 
  html:
    fig-format: svg
    fig_retina: 2
    fig-dpi: 400
params: 
  hide_answers: true
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.asp = 0.65,
  out.width = "100%",
  fig.align = "center"
)
library(tidyverse)
library(ggpubr)
library(ggthemes)
```

# [[Assessment Sheet](BCB744_Task_F_Surname.xlsx)]{.my-highlight} {#sec-assessment}

# 5. Statistical Inference

## Question 1

Devise an experiment in which you demonstrate the following two principles:

a. the effect of sample size on the estimate of the mean and variance (use SD), and
b. the effect of repeated sampling on the value of the mean and variance. In both cases, apply the correct *t*-test (later in this Chapter) to test whether sample size and repeat sampling has a statistically significant effect. What conclusion do you reach?

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r}

```

- ✓ 
- ✓ 
- ✓ 

`r if (params$hide_answers) ":::"`

## Question 2

Please refer to the [*t*-test in the lecture.](http://localhost:7028/BCB744/basic_stats/07-t_tests.html#one-sided-two-sample-t-test).

a. Repeat the analyses using the old-fashioned `t.test()` function.
b. Repeat the analysis using the even more old-fashioned Equation 2 in the lecture. Show the code and talk us through the step you followed to read the *p*-values off the table of *t*-statistics.

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r}

```

- ✓

`r if (params$hide_answers) ":::"`

## Question 3

Please report back on Task F.3 presented in the lecture.

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r}

```

- ✓

`r if (params$hide_answers) ":::"`

## Question 4

Please report back the analysis and results for Task F.4. in the lecture.

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

```{r}

```

- ✓

`r if (params$hide_answers) ":::"`
