---
date: "2021-01-01"
title: "6. Deep Dive into Gradients"
subtitle: "Marine gradients around South Africa"
---

<!--- # Topic 3: A deeper dive into gradients --->


::: callout-tip
## **Material required for this chapter**

| Type        | Name                                     | Link                                                               |
|-------------|------------------------------------------|--------------------------------------------------------------------|
| **Reading** | Smit et al. (2017)                       | [💾 `Smit_et_al_2017.pdf`](../docs/Smit_et_al_2017.pdf)            |
|             | Smit et al. (2013)                       | [💾 `Smit_et_al_2013.pdf`](../docs/Smit_et_al_2013.pdf)            |
|             | Supp. to Smit et al. (2017)              | [💾 `Smit_the_seaweed_data.pdf`](../docs/Smit_the_seaweed_data.pdf)|
| **Data**    | The seaweed environmental data           | [💾 `SeaweedEnv.RData`](../data/seaweed/SeaweedEnv.RData)          |
|             | The seaweed species data                 | [💾 `dists_mat.RData`](../data/seaweed/dists_mat.RData)            |
|             | The bioregions                           | [💾 `bioregions.csv`](../data/seaweed/bioregions.csv)              |
:::


In the previous Chapter we looked at calculations involving biodiversity
(specifically the dissimilarity matrices made from a species table) and
environmental variables (distances). What can we do with the two forms
of contemporary β-diversity? What do they mean? Can we look to
environmental distances for more insight?

Let's do a deeper analysis and create a figure to demonstrate these
findings. I regress $\beta_{\text{sør}}$ on the spatial distance between
section pairs (see below) and on the environmental distance
$\beta_{\text{E}}$ in each bioregion and used the magnitude of the slope
(per 100 km) of this relationship as a metric of β-diversity or
'distance decay' of dissimilarity.

What these lines of code do is recreate Figure 5 in @smit2017seaweeds.
Please read the paper for an interpretation of this figure as this is
critical for an understanding of the role that gradients play in
structuring patterns of biodiversity.

(To be updated...)

```{r}
## Setting up the analysis environment
library(tidyverse)
library(plyr)
library(vegan)
library(betapart) # for partitioning beta-diversity
library(BiodiversityR) # for calcs of species richness etc.
```

## Load all of the data

```{r}
# load the environmental data...
load("../data/seaweed/SeaweedEnv.RData")
env <- as.data.frame(env)
# keep only some...
env <- env[, c("annMean", "annRange", "annSD", "febMean", "febRange",
               "febSD", "augMean", "augRange", "augSD")]
```

Since the connectivity between sections is constrained by their location
along a shoreline, we calculated the distances between sections not as
'as the crow flies' distances (e.g. Section 1 is not connected in a
straight line to Section 58 because of the intervening land in-between),
but as the great circle geodesic distances between each pair of sections
along a 'route'. Travelling from **1** to **58** therefore requires
visiting **2**, then **3**, and eventually all the way up to **58**. The
total distance between a pair of arbitrary sections is thus the
cumulative sum of the great circle distances between each consecutive
pair of intervening sections along the route. These data are contained
in `dists_mat.RData` (I prepared it earlier):

```{r}
# load the distances matrix...
load("../data/seaweed/dists_mat.RData")
# loaded as dists.mat
dists.mat[1:10, 1:8]
```

Various bioregions have been defined for South African marine biota. I
prefer to use the one made by @bolton2002seaweed:

```{r}
# load the bioregions data...
bioreg <- read.csv("../data/seaweed/bioregions.csv",
                   header = TRUE)
rbind(head(bioreg, 3), tail(bioreg, 3))
```

Make a copy of the original matrix of distances between pairs of sites
to create a full matrix which constrains pairwise comparisons to pairs
within bioregions:

```{r}
bioreg.mat <- dists.mat
bioreg.mat[1:58, 1:58] <- "out"
bioreg.mat[1:16, 1:16] <- "BMP"
bioreg.mat[17:21, 17:21] <- "B-ATZ"
bioreg.mat[22:41, 22:41] <- "AMP"
bioreg.mat[42:58, 42:58] <- "ECTZ"
dim(bioreg.mat)

# see what is inside the matrix...
bioreg.mat[1:3, 1:10] 
bioreg.mat[56:58, 53:58]

# convert to show only the lower left triangle
# requires the gdata package...
bioreg.tri <- gdata::lowerTriangle(bioreg.mat, diag = FALSE) 
```

In `bioreg.mat`, pairs of sites that do not fall within any of the
bioregions is called 'out':

```{r}
# print output below...
bioreg.mat[1:3, 53:58]
```

We extract the slices (groups of rows) of the original species table
into separate dataframes, one for each of the four bioregions:

```{r}
env.BMP <- env[1:16, ]
env.BATZ <- env[17:21, ]
env.AMP <- env[22:41, ]
env.ECTZ <- env[42:58, ]
```

Now we make an environmental dataframe for use with plots of pairwise
correlations etc.:

```{r}
env.df <- data.frame(bio = bioreg$bolton, round(env, 3))
rbind(head(env.df, 3), tail(env.df, 3))
```

```{r}
# load the seaweed data...
spp <- read.csv('../data/seaweed/SeaweedSpp.csv')
spp <- dplyr::select(spp, -1)
```

## Start calculating

Calculate β-diversity using the Sørensen index of dissimilarity. This is
used throughout; binary Bray-Curtis is equivalent to Sørensen in
**vegan**.

```{r}
# ---- Sorensen-index ----
# this is used throughout...
Y <- vegdist(spp, binary = TRUE)
Y.mat <- as.matrix(Y)
# extract the subdiagonal...
Y.diag <- diag(Y.mat[-1, -nrow(Y.mat)]) 
# add a zero in front...
Y.diag <- append(0, Y.diag, after = 1) 
```

Decompose into turnover and nestedness-resultant beta-diversity:

```{r}
# ---- do-betapart ----
## Calculations with betapart...
Y.core <- betapart.core(spp)

# Using the Sørensen index, compute three distance matrices accounting for
# the (i) turnover (replacement), (ii) nestedness-resultant component, and
# (iii) total dissimilarity (i.e. the sum of both components)
# use for pairwise plotting...
Y.pair <- beta.pair(Y.core, index.family = "sor") 
```

Extract the subdiagonal for plotting later on:

```{r}
Y1.mat <- as.matrix(Y.pair$beta.sim)
# extract the subdiagonal...
Y1.diag <- diag(Y1.mat  [-1, -nrow(Y1.mat)]) 
# add a zero in front...
Y1.diag <- append(0, Y1.diag, after = 1) 

Y2.mat <- as.matrix(Y.pair$beta.sne)
Y2.diag <- diag(Y2.mat[-1, -nrow(Y2.mat)])
Y2.diag <- append(0, Y2.diag, after = 1)
```

Create separate matrices for each bioregion:

```{r}
# ---- spp-bioregion ----
spp.BMP <- spp[1:16, ]
Y.BMP <- vegdist(spp.BMP, binary = TRUE)
spp.core.BMP <- betapart.core(spp.BMP)
# use below for pairwise plotting...
Y.pair.BMP <- beta.pair(spp.core.BMP, index.family = "sor")

spp.BATZ <- spp[17:21, ]
Y.BATZ <- vegdist(spp.BATZ, binary = TRUE)
spp.core.BATZ <- betapart.core(spp.BATZ)
# use below for pairwise plotting...
Y.pair.BATZ <- beta.pair(spp.core.BATZ, index.family = "sor")

spp.AMP <- spp[22:41, ]
Y.AMP <- vegdist(spp.AMP, binary = TRUE)
spp.core.AMP <- betapart.core(spp.AMP)
# use below for pairwise plotting...
Y.pair.AMP <- beta.pair(spp.core.AMP, index.family = "sor")

spp.ECTZ <- spp[42:58, ]
Y.ECTZ <- vegdist(spp.ECTZ, binary = TRUE)
spp.core.ECTZ <- betapart.core(spp.ECTZ)
# use below for pairwise plotting...
Y.pair.ECTZ <- beta.pair(spp.core.ECTZ, index.family = "sor")
```

Calculate species richness (alpha-diversity):

```{r}
# ---- do-species-richness ----
spp.richness.site <- diversityresult(spp, index = 'richness',
                                     method = 'each site')
```

Calculate the environmental distances:

```{r}
# ---- environmental-distance ----
# Euclidian distances on temperatures
# first make a copy so we can use untransformed data later on...
env.raw <- env 
# calculate z-scores...
env <- decostand(env, method = "standardize")
```

Using individual thermal variables, calculate Euclidian distances, make
a matrix and extract the subdiagonal. The data have already been
standardised in `env`:

```{r}
# augMean
# to be used in env.rda2...
env4 <- dplyr::select(env, augMean) 
env4 <- vegdist(env4, method = 'euclidian')
env4.mat <- as.matrix(env4)
env4.diag <- diag(env4.mat[-1, -nrow(env4.mat)])
env4.diag <- append(0, env4.diag, after = 1)
```

```{r}
# febRange
# to be used in env.rda2...
env5 <- dplyr::select(env, febRange) 
env5 <- vegdist(env5, method = 'euclidian')
env5.mat <- as.matrix(env5)
env5.diag <- diag(env5.mat[-1, -nrow(env5.mat)])
env5.diag <- append(0, env5.diag, after = 1)
```

```{r}
# febSD
# to be used in env.rda2...
env6 <- dplyr::select(env, febSD) 
env6 <- vegdist(env6, method = 'euclidian')
env6.mat <- as.matrix(env6)
env6.diag <- diag(env6.mat[-1, -nrow(env6.mat)])
env6.diag <- append(0, env6.diag, after = 1)
```

```{r}
# augSD
# to be used in env.rda2...
env7 <- dplyr::select(env, augSD) 
env7 <- vegdist(env7, method = 'euclidian')
env7.mat <- as.matrix(env7)
env7.diag <- diag(env7.mat[-1, -nrow(env7.mat)])
env7.diag <- append(0, env7.diag, after = 1)
```

```{r}
# annMean
# to be used in env.rda2...
env8 <- dplyr::select(env, annMean) 
env8 <- vegdist(env8, method = 'euclidian')
env8.mat <- as.matrix(env8)
env8.diag <- diag(env8.mat[-1, -nrow(env8.mat)])
env8.diag <- append(0, env8.diag, after = 1)
```

```{r}
# combined variables selected with the db-RDA
# these have a far poorer fit...
env.comb <- dplyr::select(env, augMean, febRange, febSD, augSD)
env.comb <- vegdist(env.comb, method = 'euclidian')
env.comb.mat <- as.matrix(env.comb)
env.comb.diag <- diag(env.comb.mat[-1, -nrow(env.comb.mat)])
env.comb.diag <- append(0, env.comb.diag, after = 1)
```

```{r}
# ---- do-figure-5 ----
# Assemble data frame for plotting...
spp.df <- data.frame(dist = as.vector(dists.mat),
                     bio = as.vector(bioreg.mat),
                     augMean = as.vector(env4.mat),
                     febRange = as.vector(env5.mat),
                     febSD = as.vector(env6.mat),
                     augSD = as.vector(env7.mat),
                     annMean = as.vector(env8.mat),
                     Y = as.vector(Y.mat),
                     Y1 = as.vector(Y1.mat),
                     Y2 = as.vector(Y2.mat))
```

Do the various linear regressions of Sørensen dissimilarities
(*β~sor~*), turnover (*β~sim~*) and nestedness-related *β*-diversity
(*β~sne~*) as a function of the various thermal distances. I only
display the results of the linear regression for $Y1$ regressed on
geographical distance, `dist`, but do all the calculations:

```{r}
spp.df2 <- droplevels(subset(spp.df, bio !=  "out"))
head(spp.df2)

# turnover...
Y1.lm1 <- dlply(spp.df2, .(bio), function(x) lm(Y1 ~ dist, data = x))
lapply(Y1.lm1, summary)
Y1.lm2 <- dlply(spp.df2, .(bio), function(x) lm(Y1 ~ augMean , data = x))
# lapply(Y1.lm2, summary)
Y1.lm3 <- dlply(spp.df2, .(bio), function(x) lm(Y1 ~ augSD , data = x))
# lapply(Y1.lm3, summary)
Y1.lm4 <- dlply(spp.df2, .(bio), function(x) lm(Y1 ~ febRange , data = x))
# lapply(Y1.lm4, summary)
Y1.lm5 <- dlply(spp.df2, .(bio), function(x) lm(Y1 ~ febSD , data = x))
# lapply(Y1.lm5, summary)

# nestedness-resultant...
Y2.lm1 <- dlply(spp.df2, .(bio), function(x) lm(Y2 ~ dist, data = x))
# lapply(Y2.lm1, summary)
Y2.lm2 <- dlply(spp.df2, .(bio), function(x) lm(Y2 ~ annMean , data = x))
# lapply(Y2.lm2, summary)
```

## Make the plots

Now assemble **Figure 5.** in @smit2017seaweeds. It is a plot of pairwise (a) Sørensen
dissimilarities (*β~sor~*), (b) turnover (*β~sim~*) and (c)
nestedness-related *β*-diversity (*β~sne~*) [*sensu* @baselga2010partitioning] as a
function of distance between sections. Section pairs falling within
individual bioregions are colour-coded; where the pairs include sections
across different bioregions the symbols are coloured grey and labeled
'out'.

Combine the data in a way that makes for easy plotting:

```{r}
# Plots...
spp.long <- spp.df %>%
  gather(beta, dissim, Y:Y2) %>%
  gather(metric, distance, c(dist, augMean:annMean))
spp.long$metric = factor(spp.long$metric,
                         levels = c('dist', 'augMean', 'febRange',
                                    'febSD', 'augSD', 'annMean'))
```

The repetive portions of code needed to create each of the panels. I was
too lazy to write neater and more concise code:

```{r}
# sim as a function of geographic distance...
plt5a <- spp.long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "dist") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x, alpha = 1.0,
            size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(name = "Bioregion",
                        values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(name = "Bioregion",
                      palette = "Set1") +
  scale_shape_manual(name = "Bioregion",
                     values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(name = "Bioregion",
                    values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(name = "Bioregion",
                     values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste("Distance (km)"))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 1000)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        # legend.position = c(0.2, 0.7),
        # legend.direction = "vertical",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of distance")))
```

```{r}
# sim as a function of augMean...
plt5b <- spp.long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "augMean") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 2)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        # legend.title = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key = element_blank(),
        legend.key.height = unit(.22, "cm"),
        legend.background = element_blank(),
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of augMean")))
```

```{r}
# sim as a function of febRange...
plt5c <- spp.long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "febRange") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 4)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of febRange")))
```

```{r}
# sim as a function of febSD...
plt5d <- spp.long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "febSD") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 3)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of febSD")))
```

```{r}
# sim as a function of augSD...
plt5e <- spp.long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "augSD") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 3)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of augSD")))
```

```{r}
# sne as a function of distance...
plt5f <- spp.long %>%
  dplyr::filter(beta %in% "Y2" & metric %in% "dist") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste("Distance (km)"))) +
  ylab(expression(paste(beta[sne]))) +
  scale_y_continuous(limits = c(0, 0.22)) +
  scale_x_continuous(limits = c(0, 1000)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sne], " as a function of distance")))
```

```{r}
# sne as a function of annMean...
plt5g <- spp.long %>%
  dplyr::filter(beta %in% "Y2" & metric %in% "annMean") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y~x, alpha = 1.0, size = 0.6,
            colour = "black",
            aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted", "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sne]))) +
  scale_y_continuous(limits = c(0, 0.22)) +
  scale_x_continuous(limits = c(0, 2)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sne], " as a function of annMean")))
```

```{r}
plt5h <- ggplot(spp.long, aes(x = distance, y = dissim)) +
  geom_blank() +
  theme(plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

Assemble using the **cowplot** package:

```{r}
library(cowplot)

# turn off warnings...
oldw <- getOption("warn") 
options(warn = -1)

l <- get_legend(plt5a)
# pdf("Fig5.pdf", width = 9, height = 6.5)
ggdraw() +
  draw_plot(plot_grid(plt5a + theme(legend.position = 'none'), plt5b, plt5c,
                      plt5d, plt5e, l,
                      plt5f, plt5g, plt5h,
                      ncol = 3, align = 'hv'),
            width = 1.0)
```

## References

::: {#refs}
:::
