---
date: "2021-01-01"
title: "11a. non-Metric Multidimensional Scaling (nMDS)"
subtitle: ""
bibliography: ../references.bib
csl: ../marine-biology.csl
format:
  html:
    code-fold: false
    toc-title: "On this page"
    standalone: true
    toc-location: right
    page-layout: full
---

<!--- # Topic 11: non-Metric multidimensional scaling (nMDS) --->

::: callout-tip
## **Material required for this chapter**

| Type           | Name                                     | Link                                                                              |
|----------------|------------------------------------------|-----------------------------------------------------------------------------------|
| **Slides**     | nMDS lecture slides                      | [💾 `BCB743_11_nMDS.pdf`](../slides/BCB743_11_nMDS.pdf)                           |
| **Data**       | The Doubs River data                     | [💾 `Doubs.RData`](../data/NEwR-2ed_code_data/NeWR2-Data/Doubs.RData)             |
:::

nMDS is a rank-based *indirect gradient analysis* (i.e. not an eigen-analysis) that uses a distance or dissimilarity matrix as input. Whereas the ordination methods discussed thus far try to maximise the variance or correspondence between sites, nMDS strives to represent pairwise dissimilarities between sites in ordination space. It does not use the distances or dissimilarities directly (hence *indirect*), but these are substituted with their ranks (e.g. dissimilarities between pairs of sites are ordered by rank) (and hence *non-metric* in the name). This results in a loss of insight into the magnitude of difference between site pairs, but we benefit from the technique being more robust and less influenced by deviations from idealised data distributions.

nMDS is the non-metric equivalent to PCoA, the latter sometimes being called metric multi-dimensional scaling.

## Set-up the analysis environment

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(vegan)

# setting up a 'root' file path so I don't have to keep doing it later...
root <- "../data/"
```

## The Doubs River data

We continue to use the species data:

```{r}
load(paste0(root, "NEwR-2ed_code_data/NEwR2-Data/Doubs.RData"))
spe <- dplyr::slice(spe, -8)
```

## Do the nMDS

```{r}
spe_nmds <- metaMDS(spe, distance = "bray")
spe_nmds
```

As always, reading the help file for (accessible as `?metaMDS`) is invaluable (as are the help files for all other ordination techniques).

There's a summary method available, but it is not particularly useful and I don't display the output here:

```{r}
#| eval: false
summary(spe_nmds)
```
Although `summary(spe_nmds)` does not return anything interesting, the species and site scores are nevertheless available directly through the `scores()` command, and they can be plotted as layer in **ggplot2** if need be:

```{r}
scores(spe_nmds)
```

See *Numerical Ecology in R* (pp. 145 to 149) for information about the interpretation of a nMDSand the ordination diagrams shown below.

## Ordination diagrams

We create the ordination diagrammes as before, but new concepts introduced here are **stress**, **Shepard plots**, and **goodness of fit**. The stress indicates the scatter of observed dissimilarities against an expected monotone regression, while a Shepard diagram plots ordination distances against original dissimilarities, and adds a monotone or linear fit line to highlight this relationship. The `stressplot()` function also produces two fit statistics. The goodness-of-fit of the ordination is measured as the $R^{2}$ of either a linear or a non-linear regression of the nMDS distances on the original ones.

```{r}
#| fig-width: 8
#| fig-height: 6
par(mfrow = c(2, 2))
stressplot(spe_nmds, main = "Shepard plot")
ordiplot(spe_nmds, type = "t", cex = 0.6,
         main = paste0("nMDS stress = ", round(spe_nmds$stress, 2)))
gof = goodness(spe_nmds)
plot(spe_nmds, type = "t", main = "Goodness of fit")
points(spe_nmds, display = "sites", cex = gof * 200)
# ...bigger bubbles indicate a worse fit
```

A good rule of thumb: stress <0.05 provides an excellent representation in reduced dimensions, <0.1 is great, <0.2 is so-so, and stress <0.3 provides a poor representation.

We can also build ordination plots from scratch to suit specific needs:

```{r}
#| fig-width: 8
#| fig-height: 6
pl <- ordiplot(spe_nmds, type = "none", main = "nMDS fish abundances ")
points(pl, "sites", pch = 21, cex = 1.75, col = "grey80", bg = "grey80")
points(pl, "species", pch = 21, col = "turquoise", arrows = TRUE)
text(pl, "species", col = "blue4", cex = 0.9)
text(pl, "sites", col = "red4", cex = 0.9)
```

Or we can fit response surfaces using `ordisurf()` and project environmental drivers:

```{r}
#| fig-width: 8
#| fig-height: 6
require('viridis')
palette(viridis(8))
par(mar = c(4, 4, 0.9, 0.5) + .1, mfrow = c(2, 2))
with(spe, tmp <- ordisurf(spe_nmds ~ Satr, bubble = 3,
                          family = quasipoisson, knots = 2, col = 6,
                          display = "sites", main = "Satr"))
abline(h = 0, v = 0, lty = 3)
with(spe, tmp <- ordisurf(spe_nmds ~ Scer, bubble = 3,
                          family = quasipoisson, knots = 2, col = 6,
                          display = "sites", main = "Scer"))
abline(h = 0, v = 0, lty = 3)
with(spe, tmp <- ordisurf(spe_nmds ~ Teso, bubble = 3,
                          family = quasipoisson, knots = 2, col = 6,
                          display = "sites", main = "Teso"))
abline(h = 0, v = 0, lty = 3)
with(spe, tmp <- ordisurf(spe_nmds ~ Cogo, bubble = 3,
                          family = quasipoisson, knots = 2, col = 6,
                          display = "sites", main = "Cogo"))
abline(h = 0, v = 0, lty = 3)

env <- dplyr::slice(env, -8)

(spe_nmds_env <- envfit(spe_nmds, env)) 
plot(spe_nmds_env, col = "grey40")
plot(spe_nmds_env, p.max = 0.05, col = "red")
```

:::{.callout-important}
## Lab 6

1. Using two unconstrained ordination techniques of your choice, analyse the mite data in the **vegan** package. Provide a brief description and discussion of what you have found, and produce the R code.

2. Using two unconstrained ordination techniques of your choice (not already used in 1, above) analyse the dune data in the **vegan** package. Provide a brief description and discussion of what you have found, and produce the R code.
:::

:::{.callout-important}
## Submission instructions

The deadline for this submission is Monday x August 2022.

Provide a **neat and thoroughly annotated** Rmarkdown file which
outlines the graphs and all calculations and which displays the
resultant distance matrix. Use separate tabs for the different
questions.

Please label the Rmarkdown and resulting HTML files as follows:

-   `BCB743_<first_name>_<last_name>_Lab_6.Rmd`, and

-   `BCB743_<first_name>_<last_name>_Lab_6.html`

(the `<` and `>` must be omitted as they are used in the example as
field indicators only).

Failing to follow these instructions carefully, precisely, and
thoroughly will cause you to lose marks, which could cause a significant
drop in your score as formatting counts for 15% of the final mark (out
of 100%).

Submit your Labs on iKamva when ready.
:::


## References

::: {#refs}
:::

