---
date: "2021-01-01"
title: "Correspondence Analysis (CA)"
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

<!-- # Topic 9: Correspondence Analysis (CA) -->

::: callout-tip
## **Material Required for This Chapter**

| Type           | Name                                     | Link                                                                              |
| --- --- --- --- --- -| --- --- --- --- --- --- --- --- --- --- --- --- --- --- | --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --|
| **Theory**     | Numerical Ecology with R                   | See pages 132-140                                                                 |
| **Slides**     | CA lecture slides                        | [💾 `BCB743_09_CA.pdf`](../slides/BCB743_09_CA.pdf)                               |
| **Data**       | The Doubs River data                     | [💾 `Doubs.RData`](../data/BCB743/NEwR-2ed_code_data/NeWR2-Data/Doubs.RData)             |
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

* [Task E 1--4](tasks/Task_E.qmd)
:::

Correspondence Analysis (CA) is an eigenvector-based ordination method that handles nonlinear species responses more effectively than Principal Component Analysis (PCA). PCA relies on linear relationships and maximises variance explained using a covariance or correlation matrix, but CA employs similar regression techniques based on $\chi^2$-standardised data and weights. This makes it more appropriate for species count and presence/absence data.

CA maximises the correspondence between species scores and sample scores by preserving $\chi^2$ distances between sites in a species-by-site matrix instead of Euclidean distances. The $\chi^2$ distance metric is not influenced by [double zeros](https://www.davidzeleny.net/anadat-r/doku.php/en:similarity), making it suitable for situation when many species might be absent from several sites. The process involves performing a Singular Value Decomposition (SVD), or eigenvalue decomposition (two different approaches in linear algebra applied to the analysis of matrices) on the standardised data matrix, and reporting the eigenvalues and eigenvectors.

In CA ordination diagrams, species, and sites are presented as points within a reduced-dimensional space. Their relative positions encode the strength and structure of their associations. The distances between these points approximate the $\chi^2$ distances calculated between the rows (sites) or columns (species) of the original contingency or abundance matrix, and preserve a measure of compositional dissimilarity that is sensitive to the distributional asymmetries characteristic of ecological data. The ordination thus provides a geometric framework for addressing inferential questions of the type: *Which sites have compositional affinities with particular species assemblages?*, or *Which species distributions align with which site characteristics?*

The species scores are derived as weighted averages of site scores or equivalently as linear combinations of the original species data. As such, they are constructed to maximise the dispersion of species configurations along successive ordination axes. So, they capture dominant gradients, and patterns of variation that may reflect underlying ecological processes. Whereas PCAs provide a linear mapping of species onto environmental gradients, CAs better approximate species’ nonlinear, often unimodal, or skewed, responses to latent environmental factors. Because of this nonlinear structure, species points in CA biplots are not represented as vectors radiating from the origin (as they are in PCA, where linear monotonic gradients predominate). Instead, CA are better suited to visualisations involving curved response surfaces, which indicate the fact that species’ occurrence, or abundance peaks at intermediate positions along gradients rather than increasing or decreasing uniformly across the ordination space.

One potential downside of CA is that it assumes the total abundance or presence of species across sites to be constant, which may not always hold true. Additionally, some ecologists argue that CA might be overly influenced by rare species, as their contributions to the $\chi^2$ statistic can be disproportionately large. This issue can be mitigated by applying appropriate transformations or down weighting rare species in the analysis.

Another problem with CA is the 'arch effect.' This is similar to the [horseshoe effect in PCA](PCA.qmd#sec-horseshoe-effect), but less severe. The arch effect can be mitigated by using a [Detrended Correspondence Analysis (DCA)](DCA.qmd), which is a variation of CA that detrends the arch effect by removing the linear trend from the eigenvalues.

CA produces one axis fewer than the minimum of the number of sites (*n*) or the number of species (*p*). Like PCA, CA produces orthogonal axes ranked in decreasing order of importance. However, the variation represented is the total inertia, which is the sum of squares of all values in the $\chi^2$ matrix, rather than the sum of eigenvalues along the diagonal as in PCA. Individual eigenvalues in CA can be greater than 1, indicating that the corresponding axis captures a significant portion of the total variance in the data.

The scaling of ordination plots in CA is similar to that in PCA. Scaling 1 (site scaling) means that sites close together in the plot have similar species relative frequencies, and any site near a species point will have a relatively large abundance of that species. Scaling 2 (species scaling) means that species points close together will have similar abundance patterns across sites, and any species close to a site point is more likely to have a high abundance at that site.

As with all ordination techniques, interpreting CA results should be done with caution, and in conjunction with additional ecological knowledge and statistical tests, as the ordination axes may not always have a clear ecological interpretation. Please supplement your reading by referring to [GUSTA ME](https://sites.google.com/site/mb3gustame/indirect-gradient-analysis/ca)) and David Zelený's writing on the topic in [Analysis of community ecology data in R](https://www.davidzeleny.net/anadat-r/doku.php/en:ca_dca).

## Set-up the Analysis Environment

```{r code-library-tidyverse}
library(tidyverse)
library(vegan)
library(viridis)

# Files will be referenced using here::here() for absolute paths
```

## The Doubs River Data

In the PCA chapter we analysed the environmental data. This time we work with the species data.

```{r code-load-here-here-data}
load(here::here("data", "BCB743", "NEwR-2ed_code_data", "NEwR2-Data", "Doubs.RData"))
head(spe, 8)
```

## Do the CA

The **vegan** function `cca()` can be used for CA and Constrained Correspondence Analysis (CCA). When we do not specify constraints, as we do here, we will do a simple CA:

```{r code-spe-ca-cca-spe, error=TRUE, message=FALSE, warning=FALSE, include=TRUE}
spe_ca <- cca(spe)
```

Okay, so there is a problem. The error message says that at least one of the rows sums to 0. Which one?

```{r code-apply-spe-sum}
apply(spe, 1, sum)
```

We see that the offending row is row 8, so we can omit it. This function will omit any row that sums to zero (or less):

```{r code-spe-spe-rowsums-spe}
spe <- spe[rowSums(spe) > 0, ]
head(spe, 8)
```

Now we are ready for the CA:

```{r code-spe-ca-cca-spe-2}
spe_ca <- cca(spe)
spe_ca
```

The more verbose `summary()` output:

```{r code-summary-spe-ca}
summary(spe_ca)
```

The output looks similar to that of a PCA. The important things to note are the **inertia** (unconstrained and total inertia are the same), the **Eigenvalues for the unconstrained axes**, the **Species scores,**, and the **Site scores**. Their interpretation is the same as before, but we can reiterate. Let us calculate the total inertia:

```{r code-round-sum-spe-ca-ca}
round(sum(spe_ca$CA$eig), 5)
```

The inertia for the first axis (CA1) is:

```{r code-round-spe-ca-ca-eig}
round(spe_ca$CA$eig[1], 5)
```

The inertia of CA1 *and* CA2 is:

```{r code-round-sum-spe-ca-ca-2}
round(sum(spe_ca$CA$eig[1:2]), 5)
```

The fraction of the variance explained by CA1 and CA2 is:

```{r code-round-sum-spe-ca-ca-3}
round(sum(spe_ca$CA$eig[1:2]) / sum(spe_ca$CA$eig) * 100, 2) # result in %
```

Above, the value is the same one as in **Cumulative Proportion** in the `summary(spe_ca)` output under the CA2 column.

```{r fig-ca-screeplot, fig.height=4, fig.width=6}
#| fig.align: center
#| fig.cap: "Scree plot of the Doubs River environmental data PCA."

# make a scree plot using the vegan function:
screeplot(spe_ca, bstick = TRUE, type = "lines")
```

The scree plot (@fig-ca-screeplot) shows the eigenvalues of the CA axes which helps us decide how many axes to retain in the analysis. In this case, we will retain the first two axes, as they explain the most variance in the data.

Species scores are *actual* species scores as they now relate to species data (in the PCA, the environmental variables were in the columns, and so the species scores referred instead to the environment). The most positive and most negative eigenvectors (or loadings) indicate those species that dominate in their influence along particular CA axes. For example, CA1 will be most heavily loaded by the species Cogo and Satr (eigenvectors of 1.50075 and 1.66167, respectively). If there is an environmental gradient, it will be these species that show the strongest response. At the very least, we can say that the contributions of these species are having an overriding influence on the community differences seen between sites.

Site scores are also as seen earlier in PCA. The highest positive or negative loadings indicate sites that are dispersed far apart on the biplot (in ordination space). They will have large differences in fish community composition.

Please see *Numerical Ecology with R* (pp. 133 to 140). There you will find explanations for how to interpret the ordinations and the ordination diagrams shown below.

## Ordination Diagrams

The biplots for the above ordination are given in @fig-ca-biplots.

```{r fig-ca-biplots}
#| fig-width: 10
#| fig-height: 5.625 
#| fig.align: center
#| fig.cap: "CA ordination plot of the Doubs River species data showing site scaling (left) and species scaling (right)."

opar <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot(spe_ca, scaling = 1, main = "CA fish abundances - biplot scaling 1")
plot(spe_ca, scaling = 2, main = "CA fish abundances - biplot scaling 2")
par(opar)
```

**Scaling 1:** This is site scaling, which *emphasises relationships between rows* accurately in low-dimensional ordination space. Distances among objects (samples, or sites) in the biplot are approximations of their $\chi^{2}$ distances in multidimensional space. Objects found near a point representing a species are likely to contain a high contribution of that species. *Site scaling means that sites close together in the plot have similar species relative frequencies, and any site near a species point will have a relatively large abundance of that species.*

**Scaling 2:** Species scaling. This *emphasises relationships between columns* accurately in low-dimensional ordination space. Distances among objects (samples or sites) in the biplot are not approximations of their $\chi^{2}$ distances in multidimensional space, but the distances among species are. *Species scaling means that species points close together will have similar abundance patterns across sites, and any species close to a site point is more likely to have a high abundance at that site.*

Below I provide biplots with site and species scores for four selected species (@fig-ca-ordisurf2). The point size of the site scores scales with species scores: the larger the point, the greater the species score. Here, the species score is seen as a centre of abundance; it represents the species' maximum abundance, which decreases in every direction from the centroid. The plots are augmented with response surfaces created using the `ordisurf()` function. This function fits models to predict the abundance of the species *Salmo trutta fario* (Brown Trout), *Scardinius erythrophthalmus* (Rudd), *Telestes souffia* (Souffia, or Western Vairone), and *Cottus gobio* (Bullhead) using a Generalised Additive Model (GAM) of the Correspondence Analysis (CA) site scores on axes 1, and 2 as the predictor variables. The response surfaces illustrate where the species are most abundant and the direction of their response.

Additionally, I used the `envfit()` function to project biplot arrows for the continuous environmental variables into the ordination space. Each arrow points in the direction of the maximum increase of the variable. The length of the arrow is proportional to the correlation between the variable and the ordination axes. The significance of the correlation is tested by permutation, with significant vectors shown in red. The environmental variables are the same as those used in the PCA.

```{r fig-ca-ordisurf2}
#| fig-width: 10
#| fig-height: 10.25 
#| fig.align: center
#| fig.cap: "CA ordination plots with species response surfaces of the Doubs River species data emphasising four species of fish: A) Satr, B) Scer, C) Teso, and D) Cogo. D) additionally has the environmental vectors projected on the plot, with the significant vectors shown in red."

palette(viridis(8))
opar <- par(no.readonly = TRUE)
par(mar = c(4, 4, 0.9, 0.5) + .1, mfrow = c(2, 2))

invisible(ordisurf(spe_ca ~ Satr, data = spe, bubble = 3,
                   family = quasipoisson, knots = 2, col = 6,
                   display = "sites", main = "Salmo trutta fario"))
abline(h = 0, v = 0, lty = 3)

invisible(ordisurf(spe_ca ~ Scer, data = spe, bubble = 3,
                   family = quasipoisson, knots = 2, col = 6,
                   display = "sites", main = "Scardinius erythrophthalmus"))
abline(h = 0, v = 0, lty = 3)

invisible(ordisurf(spe_ca ~ Teso, data = spe, bubble = 3,
                   family = quasipoisson, knots = 2, col = 6,
                   display = "sites", main = "Telestes souffia"))
abline(h = 0, v = 0, lty = 3)

invisible(ordisurf(spe_ca ~ Cogo, data = spe, bubble = 3,
                   family = quasipoisson, knots = 2, col = 6,
                   display = "sites", main = "Cottus gobio"))
abline(h = 0, v = 0, lty = 3)

env <- env[-8, ] # because we removed the eighth site in the spp data

# A posteriori projection of environmental variables in a CA
# The last plot produced (CA scaling 2) must be active
spe_ca_env <- envfit(spe_ca, env, scaling = 2) # Scaling 2 is default
plot(spe_ca_env)

# Plot significant variables with a different colour
plot(spe_ca_env, p.max = 0.05, col = "red")
par(opar)
```

The species response surfaces in @fig-ca-ordisurf2 show the change of species abundance across the ordination space and the vectors indicate how the species distribution and abundance relate to the predominant environmental gradients. Seen in this way, it quickly becomes evident that the biplot is a simplification of [coenospaces](../BDC334/Lec-03-gradients.qmd).

## References

::: {#refs}
:::
