---
date: "2021-01-01"
title: "Deep Dive into Gradients"
format:
  html: default
---

<!-- # Topic 3: A deeper dive into gradients -->


::: callout-tip
## **Material Required for This Chapter**

| Type        | Name                                     | Link                                                                       |
| --- --- --- --- -| --- --- --- --- --- --- --- --- --- --- --- --- --- --- | --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- --- -|
| **Reading** | Smit et al. (2017)                       | [💾 `Smit_et_al_2017.pdf`](../docs/Smit_et_al_2017.pdf)                    |
|             | Smit et al. (2013)                       | [💾 `Smit_et_al_2013.pdf`](../docs/Smit_et_al_2013.pdf)                    |
|             | Supp. to Smit et al. (2017)              | [💾 `Smit_the_seaweed_data.pdf`](../docs/Smit_the_seaweed_data.pdf)        |
 **Related**  | Appendices to Smit et al. (2017)         | [💾 `Appendices`](two_oceans_appendices.qmd)  |
| **Data**    | The seaweed environmental data           | [💾 `SeaweedEnv.RData`](../data/seaweed/SeaweedEnv.RData)                  |
|             | The seaweed species data                 | [💾 `dists_mat.RData`](../data/seaweed/dists_mat.RData)                    |
|             | The bioregions                           | [💾 `bioregions.csv`](../data/seaweed/bioregions.csv)                      |
:::


In the previous chapter we looked at calculations involving biodiversity (specifically the dissimilarity matrices made from a species table) and environmental variables (distances) from the paper by @smit2017seaweeds. What can we do with the two forms of contemporary β-diversity? What do they mean? Can we look to environmental distances for more insight?

Let us do a deeper analysis and create a figure to demonstrate these findings. I regress $\beta_{\text{sør}}$ on the spatial distance between section pairs (see below) and on the environmental distance $\beta_{\text{E}}$ in each bioregion and used the magnitude of the slope (per 100 km) of this relationship as a metric of $\beta$-diversity or 'distance decay' of dissimilarity.

What these lines of code do is recreate Figure 5 in @smit2017seaweeds. Please read the paper for an interpretation of this figure as this is critical for an understanding of the role that gradients play in structuring patterns of biodiversity.

(To be updated...)

```{r code-library-tidyverse}
## Setting up the analysis environment
library(tidyverse)
library(plyr)
library(vegan)
library(betapart) # for partitioning beta-diversity
```

## Load and Prepare All the Data

### The environmental data

```{r code-load-here-here-data}
# load the environmental data...
load(here::here("data", "seaweed", "SeaweedEnv.RData"))
env <- as.data.frame(env)
# keep only some...
env <- env[, c("annMean", "annRange", "annSD", "febMean", "febRange",
               "febSD", "augMean", "augRange", "augSD")]
```

### The bioregional classification

Various bioregions have been defined for South African marine biota. I prefer to use the one made by @bolton2002seaweed:

```{r code-bioreg-read-csv-here}
# load the bioregions data...
bioreg <- read.csv(here::here("data", "seaweed", "bioregions.csv"),
                   header = TRUE)
rbind(head(bioreg, 3), tail(bioreg, 3))
```

### The geographic distances

Since the connectivity between sections is constrained by their location along a shoreline, we calculated the distances between sections not as 'as the crow flies' distances (e.g. Section 1 is not connected in a straight line to Section 58 because of the intervening land in-between), but as the great circle geodesic distances between each pair of sections along a 'route'. Travelling from **1** to **58** therefore requires visiting **2**, then **3**, and eventually all the way up to **58**. The total distance between a pair of arbitrary sections is thus the cumulative sum of the great circle distances between each consecutive pair of intervening sections along the route. These data are contained in `dists_mat.RData` (I prepared it earlier):

```{r code-load-here-here-data-2}
# load the distances matrix...
load(here::here("data", "seaweed", "dists_mat.RData"))
# loaded as dists_mat
dists.mat[1:10, 1:8]
```

Make a copy of the original matrix of distances between pairs of sites to create a full matrix which constrains pairwise comparisons to pairs within bioregions:

```{r code-bioreg-mat-dists-mat}
bioreg_mat <- dists.mat
bioreg_mat[1:58, 1:58] <- "out"
bioreg_mat[1:16, 1:16] <- "BMP"
bioreg_mat[17:21, 17:21] <- "B-ATZ"
bioreg_mat[22:41, 22:41] <- "AMP"
bioreg_mat[42:58, 42:58] <- "ECTZ"
dim(bioreg_mat)

# see what is inside the matrix...
bioreg_mat[1:3, 1:10] 
bioreg_mat[56:58, 53:58]

# convert to show only the lower left triangle (not used later)
# requires the gdata package...
bioreg_tri <- gdata::lowerTriangle(bioreg_mat, diag = FALSE) 
```

In `bioreg_._mat`, pairs of sites that do not fall within any of the bioregions are labelled 'out':

```{r code-bioreg-mat}
# print output below...
bioreg_mat[1:3, 53:58]
```

We extract the slices (groups of rows) of the original species table into separate dataframes, one for each of the four bioregions:

```{r code-env-bmp-env}
env_BMP <- env[1:16, ]
env_BATZ <- env[17:21, ]
env_AMP <- env[22:41, ]
env_ECTZ <- env[42:58, ]
```

Now we make an environmental dataframe for use with plots of pairwise correlations etc.:

```{r code-env-df-data-frame-bio}
env_df <- data.frame(bio = bioreg$bolton, round(env, 3))
rbind(head(env_df, 3), tail(env_df, 3))
```

### The seaweed species data

```{r code-spp-read-csv-here}
# load the seaweed data...
spp <- read.csv(here::here("data", "seaweed", "SeaweedSpp.csv"))
spp <- dplyr::select(spp, -1)
spp[1:10, 1:10]
```

## Calculate $\beta$-Diversity Indices

Calculate $\beta$-diversity using the Sørensen index of dissimilarity. This is used throughout; binary Bray-Curtis is equivalent to Sørensen in **vegan**. I then extract the subdiagonal from this matrix of species dissimilarities. The subdiagonal refers to the elements immediately below the main diagonal. For a matrix $Y$ with elements $y_{ij}$, the subdiagonal elements are $y_{i, i+1}$.

```{r code-y-vegdist-spp-binary}
# ---- Sorensen-index ----
## this is used throughout...
Y <- vegdist(spp, binary = TRUE)
Y_mat <- as.matrix(Y)
# extract the subdiagonal...
Y_diag <- diag(Y_mat[-1, -nrow(Y_mat)]) 
# add a zero in front...
Y_diag <- append(0, Y_diag, after = 1) 
```

Decompose into turnover and nestedness-resultant beta-diversity:

```{r code-y-core-betapart-core}
# ---- do-betapart ----
## Calculations with betapart...
Y.core <- betapart.core(spp)

# Using the Sørensen index, compute three distance matrices accounting for
# the (i) turnover (replacement), (ii) nestedness-resultant component, and
# (iii) total dissimilarity (i.e. the sum of both components)
# use for pairwise plotting...
Y.pair <- beta.pair(Y.core, index.family = "sor") 
```

Extract the subdiagonal for plotting later on:

```{r code-y1-mat-as-matrix-y}
# Y1 will be the turnover component
Y1_mat <- as.matrix(Y.pair$beta.sim)
# extract the subdiagonal...
Y1_diag <- diag(Y1_mat  [-1, -nrow(Y1_mat)]) 
# add a zero in front...
Y1_diag <- append(0, Y1_diag, after = 1) 

# Y2 will be the nestedness-resultant component
Y2_mat <- as.matrix(Y.pair$beta.sne)
Y2_diag <- diag(Y2_mat[-1, -nrow(Y2_mat)])
Y2_diag <- append(0, Y2_diag, after = 1)
```

Create separate matrices for each bioregion:

```{r code-spp-bmp-spp}
# ---- spp-bioregion ----
spp.BMP <- spp[1:16, ]
Y.BMP <- vegdist(spp.BMP, binary = TRUE)
spp.core.BMP <- betapart.core(spp.BMP)
# use below for pairwise plotting...
Y.pair.BMP <- beta.pair(spp.core.BMP, index.family = "sor")

spp.BATZ <- spp[17:21, ]
Y.BATZ <- vegdist(spp.BATZ, binary = TRUE)
spp.core.BATZ <- betapart.core(spp.BATZ)
# use below for pairwise plotting...
Y.pair.BATZ <- beta.pair(spp.core.BATZ, index.family = "sor")

spp.AMP <- spp[22:41, ]
Y.AMP <- vegdist(spp.AMP, binary = TRUE)
spp.core.AMP <- betapart.core(spp.AMP)
# use below for pairwise plotting...
Y.pair.AMP <- beta.pair(spp.core.AMP, index.family = "sor")

spp.ECTZ <- spp[42:58, ]
Y.ECTZ <- vegdist(spp.ECTZ, binary = TRUE)
spp.core.ECTZ <- betapart.core(spp.ECTZ)
# use below for pairwise plotting...
Y.pair.ECTZ <- beta.pair(spp.core.ECTZ, index.family = "sor")
```

Calculate species richness ($\alpha$-diversity):

```{r code-spp-richness-site-specnumber}
# ---- do-species-richness ----
spp.richness.site <- specnumber(spp)
```

Calculate the environmental distances:

```{r code-env-raw-env}
# ---- environmental-distance ----
# Euclidean distances on temperatures
# first make a copy so we can use untransformed data later on...
env_raw <- env 
# calculate z-scores...
env <- decostand(env, method = "standardize")
```

Using individual thermal variables, calculate Euclidean distances, make a matrix, and extract the subdiagonal. The data have already been standardised in `env`:

```{r code-env4-mat-env}
# augMean
# to be used in env_rda2...
env4_mat <- env |> 
  dplyr::select(augMean) |> 
  vegdist(method = 'euclidian') |> 
  as.matrix()

env4_diag <- diag(env4_mat[-1, -nrow(env4_mat)])
env4_diag <- append(0, env4_diag, after = 1)
```

```{r code-env5-mat-env}
# febRange
# to be used in env_rda2...
env5_mat <- env |> 
  dplyr::select(febRange) |> 
  vegdist(method = 'euclidian') |> 
  as.matrix()

env5_diag <- diag(env5_mat[-1, -nrow(env5_mat)])
env5_diag <- append(0, env5_diag, after = 1)
```

```{r code-env6-mat-env}
# febSD
# to be used in env_rda2...
env6_mat <- env |> 
  dplyr::select(febSD) |> 
  vegdist(method = 'euclidian') |> 
  as.matrix()

env6_diag <- diag(env6_mat[-1, -nrow(env6_mat)])
env6_diag <- append(0, env6_diag, after = 1)
```

```{r code-env7-mat-env}
# augSD
# to be used in env_rda2...
env7_mat <- env |> 
  dplyr::select(augSD) |> 
  vegdist(method = 'euclidian') |> 
  as.matrix()

env7_diag <- diag(env7_mat[-1, -nrow(env7_mat)])
env7_diag <- append(0, env7_diag, after = 1)
```

```{r code-env8-mat-env}
# annMean
# to be used in env_rda2...
env8_mat <- env |> 
  dplyr::select(annMean) |> 
  vegdist(method = 'euclidian') |> 
  as.matrix()

env8_diag <- diag(env8_mat[-1, -nrow(env8_mat)])
env8_diag <- append(0, env8_diag, after = 1)
```

```{r code-env-comb-mat-env}
# combined variables selected with the db-RDA
# these have a far poorer fit...
env_comb_mat <- env |> 
  dplyr::select(augMean, febRange, febSD, augSD) |> 
  vegdist(method = 'euclidian') |> 
  as.matrix()

env_comb_diag <- diag(env_comb_mat[-1, -nrow(env_comb_mat)])
env_comb_diag <- append(0, env_comb_diag, after = 1)
```

```{r code-spp-df-data-frame-dist}
# ---- do-figure-5 ----
# assemble data frame for plotting...
spp_df <- data.frame(dist = as.vector(dists.mat),
                     bio = as.vector(bioreg_mat),
                     augMean = as.vector(env4_mat),
                     febRange = as.vector(env5_mat),
                     febSD = as.vector(env6_mat),
                     augSD = as.vector(env7_mat),
                     annMean = as.vector(env8_mat),
                     Y = as.vector(Y_mat),
                     Y1 = as.vector(Y1_mat),
                     Y2 = as.vector(Y2_mat))

# include only site pairs that fall within bioregions...
spp_df2 <- droplevels(subset(spp_df, bio !=  "out"))
rbind(head(spp_df2, 3), tail(spp_df2, 3))
```

I will save this file with the combined data for use later in the [Multiple Regression Chapter](multiple_regression.Rmd):

```{r code-write-csv-spp-df2-file}
#| eval: false
write.csv(spp_df2, file = here::here("data", "seaweed", "spp_df2.csv"))
```


Do the various linear regressions of Sørensen dissimilarities ($\beta_\text{sør}$), turnover ($\beta_\text{sim}$) and nestedness-related $\beta$-diversity ($\beta_\text{sne}$) as a function of the various thermal distances. I only display the results of the linear regression for $Y1$ regressed on geographical distance, `dist`, but do all the calculations:

```{r fn-y1-lm1-dlply-spp-df2-bio}
# turnover...
Y1_lm1 <- dlply(spp_df2, .(bio), function(x) lm(Y1 ~ dist, data = x))
lapply(Y1_lm1, summary)
Y1_lm2 <- dlply(spp_df2, .(bio), function(x) lm(Y1 ~ augMean , data = x))
# lapply(Y1_lm2, summary)
Y1_lm3 <- dlply(spp_df2, .(bio), function(x) lm(Y1 ~ augSD , data = x))
# lapply(Y1_lm3, summary)
Y1_lm4 <- dlply(spp_df2, .(bio), function(x) lm(Y1 ~ febRange , data = x))
# lapply(Y1_lm4, summary)
Y1_lm5 <- dlply(spp_df2, .(bio), function(x) lm(Y1 ~ febSD , data = x))
# lapply(Y1_lm5, summary)

# nestedness-resultant...
Y2_lm1 <- dlply(spp_df2, .(bio), function(x) lm(Y2 ~ dist, data = x))
# lapply(Y2_lm1, summary)
Y2_lm2 <- dlply(spp_df2, .(bio), function(x) lm(Y2 ~ annMean , data = x))
# lapply(Y2_lm2, summary)
```

## Make the Plots

Now assemble **Figure 5.** in @smit2017seaweeds. It is a plot of pairwise (a) Sørensen dissimilarities ($\beta_\text{sør}$), (b) turnover (*β~sim~*) and (c) nestedness-related *β*-diversity (*β~sne~*) [*sensu* @baselga2010partitioning] as a function of distance between sections. Section pairs falling within individual bioregions are colour-coded; where the pairs include sections across different bioregions the symbols are coloured grey and labelled 'out'.

Combine the data in a way that makes for easy plotting:

```{r code-spp-long-spp-df}
# Plots...
spp_long <- spp_df %>%
  gather(beta, dissim, Y:Y2) %>%
  gather(metric, distance, c(dist, augMean:annMean))
spp_long$metric = factor(spp_long$metric,
                         levels = c('dist', 'augMean', 'febRange',
                                    'febSD', 'augSD', 'annMean'))
```

The repetitive portions of code needed to create each of the panels. I was too lazy to write neater and more concise code:

```{r fig-plt5a-spp-long}
# sim as a function of geographic distance...
plt5a <- spp_long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "dist") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x, alpha = 1.0,
            size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(name = "Bioregion",
                        values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(name = "Bioregion",
                      palette = "Set1") +
  scale_shape_manual(name = "Bioregion",
                     values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(name = "Bioregion",
                    values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(name = "Bioregion",
                     values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste("Distance (km)"))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 1000)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        # legend.position = c(0.2, 0.7),
        # legend.direction = "vertical",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of distance")))
```

```{r fig-plt5b-spp-long}
# sim as a function of augMean...
plt5b <- spp_long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "augMean") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 2)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        # legend.title = element_blank(),
        legend.title = element_text(size = 8),
        legend.text = element_text(size = 8),
        legend.key = element_blank(),
        legend.key.height = unit(.22, "cm"),
        legend.background = element_blank(),
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of augMean")))
```

```{r fig-plt5c-spp-long}
# sim as a function of febRange...
plt5c <- spp_long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "febRange") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 4)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of febRange")))
```

```{r fig-plt5d-spp-long}
# sim as a function of febSD...
plt5d <- spp_long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "febSD") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 3)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of febSD")))
```

```{r fig-plt5e-spp-long}
# sim as a function of augSD...
plt5e <- spp_long %>%
  dplyr::filter(beta %in% "Y1" & metric %in% "augSD") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sim]))) +
  scale_y_continuous(limits = c(0, 0.75)) +
  scale_x_continuous(limits = c(0, 3)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sim], " as a function of augSD")))
```

```{r fig-plt5f-spp-long}
# sne as a function of distance...
plt5f <- spp_long %>%
  dplyr::filter(beta %in% "Y2" & metric %in% "dist") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y ~ x,
            alpha = 1.0, size = 0.6, colour = "black", aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted",
                                   "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste("Distance (km)"))) +
  ylab(expression(paste(beta[sne]))) +
  scale_y_continuous(limits = c(0, 0.22)) +
  scale_x_continuous(limits = c(0, 1000)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sne], " as a function of distance")))
```

```{r fig-plt5g-spp-long}
# sne as a function of annMean...
plt5g <- spp_long %>%
  dplyr::filter(beta %in% "Y2" & metric %in% "annMean") %>%
  ggplot(aes(x = distance, y = dissim, group = bio)) +
  geom_point(aes(colour = bio, shape = bio), size = 1.2, alpha = 0.8) +
  # geom_point(aes(colour = bio, size = bio, alpha = bio, shape = bio)) +
  geom_line(stat = "smooth", method = "lm", formula = y~x, alpha = 1.0, size = 0.6,
            colour = "black",
            aes(linetype = bio)) +
  scale_linetype_manual(values = c("dashed", "solid", "dotted", "longdash", "blank")) +
  scale_colour_brewer(palette = "Set1") +
  scale_shape_manual(values = c(0, 19, 2, 5, 46)) +
  scale_size_manual(values = c(1.0, 1.2, 1.0, 1.0, 0.6)) +
  scale_alpha_manual(values = c(0.85, 1.0, 0.85, 0.85, 0.1)) +
  xlab(expression(paste(d[E]))) +
  ylab(expression(paste(beta[sne]))) +
  scale_y_continuous(limits = c(0, 0.22)) +
  scale_x_continuous(limits = c(0, 2)) +
  theme_grey() +
  theme(panel.grid.minor = element_line(colour = NA),
        plot.title = element_text(hjust = 0, size = 10),
        legend.position = "none",
        aspect.ratio = 0.6) +
  ggtitle(expression(paste(beta[sne], " as a function of annMean")))
```

```{r fig-plt5h-ggplot-spp-long-aes}
plt5h <- ggplot(spp_long, aes(x = distance, y = dissim)) +
  geom_blank() +
  theme(plot.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_blank(),
        panel.background = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank())
```

Assemble using the **cowplot** package:

```{r code-library-cowplot}
#| fig-width: 16
#| fig-height: 10
#| fig.align: center
#| column: screen-inset-right

library(cowplot)

# turn off warnings...
oldw <- getOption("warn") 
options(warn = -1)

l <- get_legend(plt5a)
# pdf("Fig5.pdf", width = 9, height = 6.5)
ggdraw() +
  draw_plot(plot_grid(plt5a + theme(legend.position = 'none'), plt5b, plt5c,
                      plt5d, plt5e, l,
                      plt5f, plt5g, plt5h,
                      ncol = 3, align = 'hv'),
            width = 1.0)
```

## References

::: {#refs}
:::
