---
date: last-modified
title: "10: Principal Coordinate Analysis (PCoA)"
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

<!-- # Topic 10: Principal Coordinate Analysis (PCoA) -->

::: callout-tip
## **Material Required for This Chapter**

| Type           | Name                                     | Link                                                                              |
| :--- | :--- | :--- |
| **Theory**     | Numerical Ecology with R                   | See pages 140-145                                                                 |
| **Slides**     | CA lecture slides                        | [💾 `BCB743_10_PCoA.pdf`](../slides/BCB743_10_PCoA.pdf)                           |
| **Data**       | The Doubs River data                     | [💾 `Doubs.RData`](../data/BCB743/NEwR-2ed_code_data/NeWR2-Data/Doubs.RData)             |
|                | The seaweed environmental data           | [💾 `SeaweedEnv.RData`](../data/BCB743/seaweed/SeaweedEnv.RData)                         |
|                | The seaweed bioregion classification     | [💾 `bioregions.csv`](../data/BCB743/seaweed/bioregions.csv).                            |
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

* [Task F 1--6](tasks/Task_F.qmd)
:::

Suppose two river sites share not a single species. Ecologically they could hardly be more different, and a dissimilarity measure such as Bray-Curtis records that difference exactly. PCA cannot ordinate a Bray-Curtis matrix: it is built on Euclidean distance, just as CA is built on $\chi^2$ distance, so each fixes the geometry in advance. Principal Coordinates Analysis (PCoA) removes that constraint. It takes a dissimilarity matrix, computed with whatever measure is ecologically appropriate, and finds the low-dimensional arrangement of sites that best preserves those dissimilarities. So, I shift from ordinating *variables* to ordinating *distances*, and that is the principle behind this chapter.

PCoA is also known as Classical Multidimensional Scaling (MDS). It is an ordination technique used to analyse and represent multivariate data based on a (dis)similarity matrix. I use PCoA when it is necessary to specify dissimilarity measures other than Euclidean distance or $\chi^2$ distance, as in PCA, and CA, respectively.

Unlike PCA and CA, which operate directly on the raw data matrix, PCoA takes a (dis)similarity matrix as input. This matrix can be calculated using various dissimilarity measures available in **vegan**'s `vegdist()` function, which may be necessary when my dataset includes quantitative, semi-quantitative, qualitative, and mixed variables. If the dissimilarities are Euclidean distances among centred observations, PCoA results are equivalent to those obtained from PCA, up to ordinary differences in scaling, signs, and rotation.

That flexibility is what places PCoA in the wider ordination landscape. The choice of method follows from the kind of difference between sites that is ecologically relevant:

| Situation | Suitable method |
| :--- | :--- |
| Continuous environmental variables | PCA |
| Species abundances under $\chi^2$ geometry | CA |
| Any ecological distance, such as Bray-Curtis | PCoA |
| Rank-order dissimilarities | nMDS |

The workflow is always the same, so I build a dissimilarity matrix from the data, then ordinate the matrix (@fig-pcoa-workflow). The first step is where the ecology enters, since that is where you choose how two sites are to be compared.

```{mermaid}
%%| label: fig-pcoa-workflow
%%| echo: false
%%| fig-cap: "The PCoA workflow. The ecological decision is the choice of dissimilarity at the second step; everything after it is mechanical."
flowchart TD
  A["Species table<br/>(sites × species)"] --> B["Choose a dissimilarity<br/>(e.g. Bray-Curtis)"]
  B --> C["Distance matrix<br/>(site × site)"]
  C --> D["PCoA"]
  D --> E["Ordination plot"]
```

PCoA performs a dimensionality reduction on the (dis)similarity matrix, scaling the dissimilarities, and returning a set of points in a lower-dimensional space (typically 2D, or 3D). When plotted, the Euclidean distances between these points approximate the original dissimilarities, effectively representing the dissimilarities between objects as Euclidean distances in a lower-dimensional space. This representation can be helpful in visualising, and interpreting complex relationships in a more interpretable form.

Conceptually, PCoA is similar to PCA and CA in that it aims to represent the relationships between objects in a lower-dimensional space. However, it differs in its approach to preserving distances. While PCA preserves Euclidean distances between objects and CA preserves $\chi^2$ distances, PCoA starts from whichever (dis)similarity measure is provided as input. If that dissimilarity is Euclidean, the full set of real axes can reproduce it exactly; if it is not Euclidean, PCoA gives the best real-coordinate representation and may report negative eigenvalues for the part that cannot be embedded.

The cleanest way to hold the relationship in mind is that PCoA generalises the geometry of PCA. PCA is the special case of PCoA that you obtain when the chosen dissimilarity is Euclidean:

| PCA | PCoA |
| :--- | :--- |
| starts with variables | starts with distances |
| Euclidean distance only | any dissimilarity measure |
| needs the raw variables | a distance matrix is enough |
| a special case of distance ordination | the general distance ordination |

In PCoA, the eigenvalues represent the extent to which each principal coordinate axis captures the variability in the original dissimilarity matrix. The proportion of a given positive eigenvalue to the sum of positive eigenvalues indicates the relative importance of each real axis, but this proportion must be read carefully when the dissimilarity is non-Euclidean and negative eigenvalues are present. Higher positive eigenvalues represent axes that capture more of the embeddable structure in the data, helping identify the strongest gradients in the dataset.

When `capscale()` is given a raw community table and a distance method, it can add species scores as weighted averages of the site scores (wascores). If I instead supply only a precomputed dissimilarity matrix, that species information is no longer available from the input and the species scores are absent.

## Set-up the Analysis Environment

```{r code-library-tidyverse}
library(tidyverse)
library(vegan)
library(ggrepel) # for tidy biplot labels

# Files will be referenced using here::here() for absolute paths
```

## The Doubs River Data

I continue to use the species data:

```{r code-load-here-here-data}
load(here::here(
  "data",
  "BCB743",
  "NEwR-2ed_code_data",
  "NEwR2-Data",
  "Doubs.RData"
))
# remove the 8th row because it sums to zero
spe <- dplyr::slice(spe, -8)
```

## Calculate a Suitable Dissimilarity Matrix

You may or may not want to calculate a dissimilarity index upfront (see below). Here I calculate the Bray-Curtis dissimilarity which is appropriate for abundance data:

```{r code-spe-bray-vegdist-spe}
spe_bray <- vegdist(spe)
```

## Do the PCoA

<!-- The **vegan** function `cmdscale()` can be used for PCoA: -->

The book *Numerical Ecology with R* uses a built-in function `cmdscale()` or the function `pcoa()` in **ape** for its PCoA calculation. The **vegan** function `capscale()` can also be used for PCoA, and this is the approach I take here. The 'CAP' in `capscale()` stands for 'Canonical Analysis of Principal Coordinates'. `capscale()` works differently from `rda()` or `cca()` in that I specify the input through a formula interface (see `?capscale`). To run a PCoA without constraints I put `1` on the right-hand side of the formula (the community data go on the left). The same interface also accepts predictors on the right, which is how `capscale()` performs *constrained* analysis (db-RDA).

The cleanest way to run it is to hand `capscale()` the **raw species table** and let it compute the dissimilarities. Supplying the raw table lets the function add species scores (as weighted averages of the site scores, or wascores), so the ordination places species as well as sites; a precomputed matrix carries no species information, and the scores are then missing. I built `spe_bray` above to make the dissimilarity-first logic clear, but `capscale()` can construct it internally, which is what I do here:

```{r code-spe-pcoa-capscale-spe-distance}
spe_pcoa <- capscale(spe ~ 1, distance = "bray")
spe_pcoa
```

::: {.callout-note appearance="simple"}
## Why Are There Negative ("Imaginary") Eigenvalues?

The printout reports a small **Imaginary** component (here about $-0.30$, against a real total of about $7.06$). This is expected, and is not a sign of error. Bray-Curtis, and most ecological dissimilarities, are not Euclidean, so they cannot be embedded perfectly in real coordinate space; PCoA represents the part that will not embed as negative eigenvalues. When the negative component is small relative to the leading positive axes, the ordination on the first two or three axes is usually still useful, but the proportions of variation need to be read with that limitation in mind.

If the negative component is large enough to affect interpretation, there are standard corrections. A square-root transformation of the dissimilarities can often make semi-metric distances closer to Euclidean. Cailliez and Lingoes corrections add constants so the corrected matrix has no negative eigenvalues; these are available through `ape::pcoa(correction = "cailliez")` or `ape::pcoa(correction = "lingoes")`, and base R's `cmdscale(add = TRUE)` applies a Cailliez-style additive correction. The correction should be reported because it changes the geometry being analysed. One practical consequence of leaving negative eigenvalues in place is that the broken-stick test (`bstick()`) cannot be computed, which is why the scree plot below is drawn by hand.
:::

The `summary()` now produces output that should look familiar from PCA and CA, with the **Species scores** present because the raw table was supplied:

```{r code-summary-spe-pcoa-2}
summary(spe_pcoa)
```

### How to Read This PCoA

The detailed plotting comes later; the result itself comes first. @fig-pcoa-interp redraws the ordination with the sites coloured from source to mouth and the most distinctive species labelled.

```{r fig-pcoa-interp}
#| fig-cap: "An annotated PCoA (Bray-Curtis) of the Doubs fish data. Points are sites, coloured from the source (dark) to the mouth (yellow); green crosses are species, labelled where they sit away from the crowded centre. MDS1 orders the sites along the river, as the arrow shows."
#| fig-width: 6.5
#| fig-height: 5
#| code-fold: true
pos_eig <- spe_pcoa$CA$eig[spe_pcoa$CA$eig > 0]
pcoa_pct <- round(100 * pos_eig / sum(pos_eig), 1)

pcoa_sites <- as.data.frame(scores(spe_pcoa, display = "sites", choices = 1:2))
pcoa_sites$site <- as.integer(rownames(pcoa_sites))
pcoa_spp <- as.data.frame(scores(spe_pcoa, display = "species", choices = 1:2))
pcoa_spp$lab <- rownames(pcoa_spp)
pcoa_spp$d <- sqrt(pcoa_spp$MDS1^2 + pcoa_spp$MDS2^2)
pcoa_lab <- pcoa_spp[pcoa_spp$d > quantile(pcoa_spp$d, 0.45), ]

ggplot(pcoa_sites, aes(MDS1, MDS2)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(aes(colour = site), size = 2) +
  scale_colour_viridis_c(name = "Site\n(1 = source)") +
  geom_point(
    data = pcoa_spp,
    aes(MDS1, MDS2),
    colour = "seagreen4",
    shape = 3,
    size = 0.8
  ) +
  geom_text_repel(
    data = pcoa_lab,
    aes(MDS1, MDS2, label = lab),
    colour = "seagreen4",
    size = 2.5,
    max.overlaps = Inf,
    segment.colour = "grey80"
  ) +
  annotate(
    "segment",
    x = -1.15,
    y = -1.5,
    xend = 1.0,
    yend = -1.5,
    arrow = arrow(length = unit(2.5, "mm")),
    colour = "firebrick"
  ) +
  annotate(
    "text",
    x = -0.1,
    y = -1.7,
    label = "upstream to downstream (MDS1)",
    colour = "firebrick",
    size = 2.7
  ) +
  labs(
    x = paste0("MDS1 (", pcoa_pct[1], "%)"),
    y = paste0("MDS2 (", pcoa_pct[2], "%)")
  ) +
  coord_equal()
```

The ecological interpretation is the gradient I have met throughout the module:

- **MDS1 has about 52% of the positive inertia**, so it is the dominant pattern.
- **The sites fall along MDS1 in river order**, from the source at one end to the mouth at the other. The Bray-Curtis ordination recovers the river sequence from species composition alone.
- **The two faunas separate along MDS1.** The upstream end holds cool-water trout-zone species, brown trout (`Satr`) and minnow (`Phph`) among them; the downstream end holds lowland species such as bleak (`Alal`), roach (`Ruru`), and ruffe (`Gyce`).
- **MDS1 is therefore the upstream-to-downstream river gradient**, the same gradient recovered by the [PCA of the environmental data](PCA.qmd) and the [CA of the fish](CA.qmd). Fitting the environmental vectors (below) confirms it: dissolved oxygen points towards the trout end, and distance from source, nitrate, and organic load towards the lowland end.

The ecological story is the point of the analysis. Two other ways to call the function, and how to recover species scores when a precomputed matrix leaves them out, are given in a note further down.

### Eigenvalues and How Many Axes to Retain

I can unpack what is inside the results, and there I can see that I can access the eigenvalues as I did for PCA, and CA:

```{r code-str-spe-pcoa-not-shown, eval=FALSE}
str(spe_pcoa) # not shown due to length of output
```

The percentage inertia explained by the first three axes is therefore:

```{r code-round-sum-spe-pcoa-ca}
round(sum(spe_pcoa$CA$eig[1:3]) / sum(spe_pcoa$CA$eig) * 100, 2)
```

```{r fig-pcoa-screeplot, fig.height=4, fig.width=6}
#| fig.align: center
#| fig.cap: "Scree plot of the Doubs River fish species PCoA."
# The `bstick()` function is not compatible with PCoA
# when negative eigenvalues are present
# Plot the scree plot without the broken stick model

# Extract eigenvalues
eigenvalues <- spe_pcoa$CA$eig

# Calculate the proportion of positive ordination variance explained
pos_eig <- eigenvalues[eigenvalues > 0]
variance_explained <- pos_eig / sum(pos_eig)

# Create a scree plot
plot(
  variance_explained,
  type = "b",
  main = "Scree Plot",
  xlab = "Principal Components",
  ylab = "Prop. Var. Explained"
)
```

The scree plot (@fig-pcoa-screeplot) shows the proportion of variation explained by the PC axes. In this case, I will still only retain the first two axes. I selected these because after the 2nd PC, the proportion of variance explained by each additional PC is less than 10%, and the plot starts levelling off.

See *Numerical Ecology with R* (pp. 140 to 145) for information about the interpretation of a PCoA and the ordination diagrams shown below.

## Ordination Diagrams

I create the ordination diagrammes as before:

```{r fig-pcoa-biplots}
#| fig-width: 10
#| fig-height: 5.625 
#| fig.align: center
#| fig.cap: "PCoA ordination plot of the Doubs River species data showing site scaling (left) and species scaling (right)."
opar <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
plot(spe_pcoa, scaling = 1, main = "PCoA fish abundances - biplot scaling 1")
plot(spe_pcoa, scaling = 2, main = "PCoA fish abundances - biplot scaling 2")
par(opar)
```

Scaling 1 and scaling 2 is the same as in PCA and CA.

The plots above work okay, but I can improve them. Note that you can also apply these improvements to PCA, and CA ordinations. Let me build plots from scratch:

```{r code-pcoa-ordiplots}
#| fig-width: 10
#| fig-height: 5.625 
#| fig.align: center
#| fig.cap: "PCoA ordination plot made with `ordiplot()` of the Doubs River species data showing site scaling (left) and species scaling (right)."
opar <- par(no.readonly = TRUE)
par(mfrow = c(1, 2))
pl1 <- ordiplot(
  spe_pcoa,
  type = "none",
  scaling = 1,
  main = "PCoA fish abundances - biplot scaling 1"
)
points(pl1, "sites", pch = 21, cex = 1.75, col = "grey80", bg = "grey80")
points(pl1, "species", pch = 21, col = "turquoise", arrows = TRUE)
text(pl1, "species", col = "blue4", cex = 0.9)
text(pl1, "sites", col = "red4", cex = 0.9)

pl2 <- ordiplot(
  spe_pcoa,
  type = "none",
  scaling = 2,
  main = "PCoA fish abundances - biplot scaling 2"
)
points(pl2, "sites", pch = 21, cex = 1.75, col = "grey80", bg = "grey80")
points(pl2, "species", pch = 21, col = "turquoise", arrows = TRUE)
text(pl2, "species", col = "blue4", cex = 0.9)
text(pl2, "sites", col = "red4", cex = 0.9)
par(opar)
```

I can also fit response surfaces using `ordisurf()`:

```{r fig-pcoa-ordisurf}
#| fig-width: 10
#| fig-height: 10
#| fig.align: center
#| fig.cap: "PCoA ordination plots with species response surfaces of the Doubs River species data emphasising four species of fish: A) Satr, B) Scer, C) Teso, and D) Cogo. D) additionally has the environmental vectors projected on the plot, with the significant vectors shown in red."
require('viridis')
palette(viridis(8))

opar <- par(no.readonly = TRUE)
par(mar = c(4, 4, 0.9, 0.5) + .1, mfrow = c(2, 2))
with(
  spe,
  tmp <- ordisurf(
    spe_pcoa ~ Satr,
    bubble = 3,
    family = quasipoisson,
    knots = 2,
    col = 6,
    display = "sites",
    main = "Salmo trutta fario"
  )
)
abline(h = 0, v = 0, lty = 3)
with(
  spe,
  tmp <- ordisurf(
    spe_pcoa ~ Scer,
    bubble = 3,
    family = quasipoisson,
    knots = 2,
    col = 6,
    display = "sites",
    main = "Scardinius erythrophthalmus"
  )
)
abline(h = 0, v = 0, lty = 3)
with(
  spe,
  tmp <- ordisurf(
    spe_pcoa ~ Teso,
    bubble = 3,
    family = quasipoisson,
    knots = 2,
    col = 6,
    display = "sites",
    main = "Telestes souffia"
  )
)
abline(h = 0, v = 0, lty = 3)
with(
  spe,
  tmp <- ordisurf(
    spe_pcoa ~ Cogo,
    bubble = 3,
    family = quasipoisson,
    knots = 2,
    col = 6,
    display = "sites",
    main = "Cottus gobio"
  )
)
abline(h = 0, v = 0, lty = 3)

env <- dplyr::slice(env, -8)

(spe_pcoa_env <- envfit(spe_pcoa, env, scaling = 2))
plot(spe_pcoa_env, col = "grey40")
plot(spe_pcoa_env, p.max = 0.05, col = "red")
par(opar)
```

::: {.callout-note collapse="true" appearance="simple"}
## Other Ways to Run a PCoA, and the Species-Score Issue

Above I gave `capscale()` the raw species table, which is the approach I recommend because it keeps the species scores. Two alternatives are worth knowing.

**Supply a precomputed dissimilarity matrix.** If I hand `capscale()` the Bray-Curtis matrix I built earlier, the ordination of the sites is identical, but the **Species scores** are missing, since the function sees only the site-by-site distances and never the species themselves:

```{r code-spe-pcoa-capscale-spe-bray}
# spe_pcoa <- cmdscale(spe_bray, k = nrow(spe) - 1, eig = TRUE)
spe_pcoa <- capscale(spe_bray ~ 1)
spe_pcoa
```

```{r code-summary-spe-pcoa}
summary(spe_pcoa)
```

**Add the species back with `sppscores()`.** When a precomputed matrix is unavoidable, e.g. a Gower matrix from another package (see [Handling Mixed Variable Types](#handling-mixed-variable-types), below), the species scores can be restored afterwards:

```{r code-spe-pcoa-capscale-spe-bray-2}
spe_pcoa <- capscale(spe_bray ~ 1)
sppscores(spe_pcoa) <- spe

summary(spe_pcoa)
```
:::

## Handling Mixed Variable Types

The simplest way to handle mixed variable types is to plot the factor variable of interest as a differently shaped or coloured symbol on the ordination diagram. I will show this with the seaweed dataset.

First, I construct an environmental dataset that contains some mixed variables by column binding a dataset of seawater temperatures, and a bioregional classification of the 58 coastal sections:

```{r code-bioreg-read-csv-here}
bioreg <- read.csv(
  here::here("data", "BCB743", "seaweed", "bioregions.csv"),
  header = TRUE
)
load(here::here("data", "BCB743", "seaweed", "SeaweedEnv.RData"))
E <- cbind(bioreg, env) %>%
  mutate(
    spal.prov = factor(spal.prov),
    spal.ecoreg = factor(spal.ecoreg),
    lombard = factor(lombard),
    bolton = factor(bolton)
  )
head(E)
str(E)
```

In Smit et al. (2017) I used forward selection and the assessment of VIF to find only the non-collinear variables, which included `augMean`, `febRange`, `febSD` and `augSD` as the most parsimonious descriptors. I will do a PCoA of the seaweed environmental data (only the subset indicated above) and colour the sites by the bioregion:

```{r fig-pcoa-seaweed-env}
#| fig-width: 5.625
#| fig-height: 5.625 
#| fig.align: center
#| fig.cap: "PCoA ordination plot of the seaweed environmental data with sites coloured by bioregion."
E_pcoa <- capscale(env[, c("augMean", "febRange", "febSD", "augSD")] ~ 1)

col <- c("firebrick1", "seagreen4", "blue2", "goldenrod2")
pch <- c(17, 19)

opar <- par(no.readonly = TRUE)
ordiplot(
  E_pcoa,
  type = "n",
  scaling = 1,
  xlim = c(-1.8, 2),
  ylim = c(-2.8, 1.2),
  main = "PCoA of seaweed env. data"
)
points(E_pcoa, "sites", pch = 21, cex = 1.75, col = col[E$bolton], bg = "white")
text(E_pcoa, "sites", col = col[E$bolton], cex = 0.5)
```

The arrangement of sites in the ordination diagram (@fig-pcoa-seaweed-env) is still only affected by `augMean`, `febRange`, `febSD`, and `augSD` and colour is used only to identify the sites by bioregion. Colour does not affect the outcome of the ordination, and yet I can already see that sites show a clear bioregional grouping. Please see the analysis of the [Mayombo diatom dataset](nMDS_diatoms.qmd) for additional approaches to deal with categorical variables that are presumed to be  influential. To formally account for bioregion in the ordination, I must take a different approach.

The second option is to use a suitable dissimilarity metric. Numerical, nominal, ordinal, and binary variables can all be accommodated with the Gower distance. I do not use **vegan** for this, but rather the `daisy()` function in **cluster**.

I calculate the Gower distances and proceed with the PCoA as before:

```{r code-library-cluster}
library(cluster)

# cannot use mixed var  types
# E_gower <- vegdist(E, method = "gower")

# can handle mixed var types... use instead of vegdist() gower dissimilarity
E_gower <- daisy(E, metric = "gower")

summary(E_gower)

E_mat <- as.matrix(E_gower)
E_mat[1:5, 1:5]

E_pcoa <- capscale(E_mat ~ 1)

# sadly this means that the names in the Species scores are now missing
summary(E_pcoa)
```

I can extract the various kinds of scores for manual plotting.

## Three Methods, Three Geometries

PCA, CA, and PCoA all reduce dimensionality, and they differ in the geometry they preserve. PCA preserves Euclidean distances, CA preserves $\chi^2$ distances, and PCoA preserves whichever dissimilarity the analyst supplies. Because a Euclidean dissimilarity reduces PCoA to PCA, PCoA is the general method and the other two are particular choices of distance. This freedom to choose the distance on ecological grounds is what makes PCoA one of the most widely used ordination methods in community ecology, and it is the foundation for the distance-based methods that follow. The next chapter takes up [non-metric Multidimensional Scaling (nMDS)](nMDS.qmd), which starts from the same kind of dissimilarity matrix but abandons the eigenvalue machinery in favour of preserving the rank order of the distances.

## References

::: {#refs}
:::
