---
title: "BCB744 (BioStatistics): Final Assessment (Rewrite) 2024"
output: html_document
date: "2024-07-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Honesty Pledge

**This assignment requires that you work as an individual and not share
your code, results, or discussion with your peers. Penalties and
disciplinary action will apply if you are found cheating.**

::: callout-note
## Acknowledgement of the Pledge

Copy the statement, below, into your document and replace the
underscores with your name acknowledging adherence to the UWC's Honesty
Pledge.

**I, \_\_\_\_\_\_\_\_\_\_\_\_, hereby state that I have not communicated
with or gained information in any way from my peers and that all work is
my own.**
:::

```{r}
#| echo: false
library(tidyverse)
library(ggpubr)
```


## Instructions

Please note the following instructions. Failing to comply with them in full will result in a loss of marks.

* **<u>QUARTO --> HTML</u>** Submit your assessment answers as an .html file compiled from your Quarto document. Produce *fully annotated reports*, including the meta-information at the top (name, date, purpose, etc.). Provide ample commentary explaining the purpose of the various tests/sections as necessary.

* **<u>TESTING OF ASSUMPTIONS</u>** For all questions, make sure that when *formal inferential statistics are required, each is preceded by the appropriate tests for the assumptions*, i.e., state the assumptions, state the statistical procedure for testing the assumptions and mention their corresponding $H_{0}$. If a graphical approach is used to test assumptions, explain the principle behind the approach. Explain the findings emerging from the test of assumptions, and justify your selection of the appropriate inferential test (e.g. *t*-test, ANOVA, etc.) that you will use.

* **<u>STATE HYPOTHESES</u>** When inferential statistics are required, please provide the full $H_{0}$ and $H_{A}$, and conclude the analysis with a statement of which is accepted or rejected.

* **<u>GRAPHICAL SUPPORT</u>** All descriptive and inferential statistics must be supported by the appropriate figures of the results.

* **<u>STATEMENT OF RESULTS</u>** Make sure that the textual statement of the final result is written exactly as required for it to be published in a journal article. Please consult a journal if you don't know how.

* **<u>FORMATTING</u>** Pay attention to formatting. Some marks will be allocated to the appearance of the script, including considerations of aspects of the tidiness of the file, the use of the appropriate headings, and adherence to code conventions (e.g. spacing etc.).

* **<u>MARK ALLOCATION</u>** Please see the [Introduction Page](https://tangledbank.netlify.app/bcb744/bcb744_index#summative-tasks) for an explanation of the assessment approach that will be applied to these questions.

<b style='color:#CD5C5B;'>Submit the .html file wherein you provide answers to Questions 1–7 by no later than 19:00 today. Label the script as follows:

<u>BCB744_\<Name\>_\<Surname\>_Final_Assessment.html</u>, e.g.

<u>BCB744_AJ_Smit_Final_Assessment.html.</u></b>


## Question 1

### The Effect of Vitamin C on Tooth Growth in Guinea Pigs

Package **datasets**, dataset `ToothGrowth`: The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as `VC`).

a. Undertake the analysis, describe the findings, and create an appropriate graphical summary of your findings.
b. Provide a justification for your choice of statistical method.

## Question 2

### The wheat yield data

Please see the file '**fertiliser_crop_data.csv**' for this dataset. The data represent an experiment designed to test whether or not fertiliser type and the density of planting have an effect on the yield of wheat. The dataset contains the following variables:

- Final yield (kg per acre)
- Type of fertiliser (fertiliser type A, B, or C)
- Planting density (1 = low density, 2 = high density)
- Block in the field (north, east, south, west)

a. Do fertiliser type and planting density affect the yield of wheat? If so, which is the best density to plant wheat at, and which fertiliser produces the best yield?
b. Does it matter if the wheat is planted in portions of the experimental fields that face north, east, south, or west?
c. Provide a justification for why you chose the statistical tests you used to answer the question.

## Question 3

### The shells data

See the '**shells.csv**' file. This dataset contains measurements of shell widths and lengths of the left and right valves of two species of mussels, *Aulacomya* sp. and *Choromytilus* sp. Length and width measurements are presented in mm.

a. Which species of mussel is the i) widest and ii) longest?
b. Within each species of mussel, are the four different measurements correlated with each other?
c. Considering *Aulacomya* sp. only, use a linear regression to predict the length of the left valve when the width of the left valve is 15 and 17 mm. 
d. Provide a justification for why you chose the statistical tests you used to answer the question.

## Question 4

### The health data

These data are in '**health.csv**'. Inside the file are several columns, but the ones that are relevant to this question are:

- 'Sex', which is the gender of the individuals assessed
- 'Substance', indicating the kind of drug abused by the individuals in question
- 'Mental\_score', which is the outcome of a test designed to test the cognitive ability of individuals

a. Do males and females suffer the same cognitive impairments if they abuse cocaine, alcohol, or heroin?
b. Which drug is worst in terms of affecting the user's mental health?
c. Provide a justification for why you chose the statistical tests you used to answer the question.

## Question 5

## The crickets data

The file '**crickets.csv**' contains data for some crickets whose chirp rate was measured at several temperatures. The temperature was measured in °F, but please make sure you do all the calculations using °C instead.

a. Does the chirp rate of the crickets depend on the temperature?
b. Provide an equation that quantifies this relationship.
c. Provide a justification for why you chose the statistical tests you used to answer the question.

## Question 6

### The SST data

The file '**SST.csv**' contains sea surface temperatures for Port Nolloth and Muizenberg in °C. The data are from 1 January 2010 to 31 December 2011.

a. Do the temperatures differ between the two places?
b. For each of the two sites, which month has the i) lowest and ii) highest temperature?
c. For each of the two sites, is the winter temperature colder than the summer temperature?
d. Same as (c), but use 95% confidence intervals to approach this problem (and provide the supporting graphs).

Hint: The **lubridate** package (and others) offers convenient ways to work with time series (i.e. in this case coding a variable for month).

## Question 7

### Oudekraal's seawater temperature data

The file '**oudekraal.csv**' contains seawater temperature data for Oudekraal, Cape Town, South Africa. The temperature data were collected every minute from 4 March 2016 to 23 March 2016

a. What is the average daily temperature of the seawater at Oudekraal? Present your findings as a graph.
b. Are nighttime temperatures colder than daytime temperatures?
c. Are the water temperatures of Sundays different from those of Wednesdays?
d. When was the warmest and coldest days during the period of data collection?
e. Provide a justification for why you chose the statistical tests you used to answer the question.


## The end

<b style='color:#CD5C5B;'>Submit the .html file wherein you provide answers to Questions 1–7 by no later than 19:00 today. Label the script as follows:

<u>BCB744_\<Name\>_\<Surname\>_Summative_Task_2.html</u>, e.g.

<u>BCB744_AJ_Smit_Summative_Task_2.html.</u></b>

<b>Upload your .html files onto [Google Forms](https://docs.google.com/forms/d/e/1FAIpQLSfBO9a42ESrL3E3ytIFIqMGvnei19ynWisPpJHGU8P9DXvyow/viewform?usp=sf_link).</b>

