---
date: "2026-09-08"
date-modified: "last-modified"
title: "Lecture 7: Why Are Species Where They Are? Niche and Neutral Theories of Community Assembly"
author:
  - name: "Alex Matthew"
    affiliation: "University of the Western Cape"
  - name: "Keanan Jarvis"
    affiliation: "University of the Western Cape"
  - name: "Ethan Bell"
    affiliation: "University of the Western Cape"
citation:
  type: webpage
  author:
    - name: "Alex Matthew"
    - name: "Keanan Jarvis"
    - name: "Ethan Bell"
format:
  html:
    anchor-sections: true
    date-format: long
    date-modified: last-modified
    date-modified-title: "Last updated"
    number-sections: false
    page-layout: article
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    code-link: false
bibliography:
  - ../references.bib
  - Lec-07-references.bib
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
    comment = "R>",
    warning = FALSE,
    message = FALSE,
    fig.width = 4.5,
    fig.height = 2.625,
    out.width = "75%",
    fig.asp = NULL, # control via width/height
    dpi = 300
)

ggplot2::theme_set(
    ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
    ggplot2::theme_bw(base_size = 8)
)
```

::: {.callout-tip}
## Material Required for This Lecture

| Type | Name | Notes |
| :--- | :--- | :--- |
| **Theory** | Hubbell (2001) *The Unified Neutral Theory of Biodiversity and Biogeography* | Primary statement of neutral theory |
| **Theory** | Hutchinson (1957) Concluding remarks | The niche concept defined |
| **Theory** | Adler et al. (2007) A niche for neutrality | The niche-neutral continuum |
| **Theory** | Cottenie (2005) Integrating environmental and spatial processes | Metacommunity framework |
| **Theory** | Legendre et al. (2015) Should the Mantel test be used in spatial analysis? | Mantel test critique |
| **Data** | Doubs River fish community | `Doubs.RData` via NEwR companion data |
| **Data** | Oribatid mite community | `vegan::mite`, `vegan::mite.env`, `vegan::mite.xy` |
| **Package** | spesim 0.5.2 | `remotes::install_github("ajsmit/spesim")` |
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Lecture

See the [Tasks](#sec-tasks) section at the end of this lecture for five exercises that build directly on the analyses presented here.
:::

> "*The truth is rarely pure and never simple.*"
>
> Oscar Wilde



## Introduction

[Cluster Analysis](https://tangledbank.netlify.app/BCB743/cluster_analysis.html) left one question open. The Doubs fish
community changes measurably from the headwaters to the lowlands. Silhouette analysis
showed that discrete clusters lose information retained by a continuous ordination.
Naming that gradient does not explain it. *Why* does the community change so
consistently along the river? That is our question here.

::: {.callout-tip}
## What the Gradient Can Show

A gradient in community composition is an observed pattern. Niche and neutral theory
offer different accounts of how it arose. Ordination and clustering describe the
pattern well, but several processes can produce the same result. Separating them
requires other evidence.
:::

Ecologists have developed two main answers. **Niche theory** places ecological
differences among species at the centre of community assembly. Species persist where
the local environment and biotic interactions suit them. **Neutral theory** treats
species as functionally interchangeable at the level of births, deaths and dispersal,
so stochastic demography and movement can structure a community. The mechanisms differ.
Their visible patterns may look remarkably alike.

By the end of this lecture you should be able to:

- state the core assumption each theory makes about species, and explain why those
  assumptions lead to different predictions about community structure.
- explain why this remains an active area of debate, and identify what is empirically
  established as against what is contested.
- translate each theory into a falsifiable prediction about how community
  dissimilarity should behave across space.
- recognise why the most intuitive statistical test of those predictions is the wrong
  one, and explain the reasoning behind that judgement.
- apply a defensible alternative to the Doubs data and interpret it critically.
- connect this theoretical lens to the advanced workflow introduced in the
  [BCB743 Model Building chapter](https://tangledbank.netlify.app/BCB743/model_building.html).

## From Pattern to Process: Revisiting the Doubs Gradient

The [Cluster Analysis](https://tangledbank.netlify.app/BCB743/cluster_analysis.html) chapter used cophenetic correlation and
silhouette analysis to show that the Doubs fish community is organised along a
continuum. The earlier [Ordination](https://tangledbank.netlify.app/BCB743/ordination.html) chapter reached the same conclusion
from another direction. Its dominant axis recovered an upstream-to-downstream gradient,
with cool-water species at one end and lowland species at the other. The methods agree
on the pattern. They do not identify its cause.

Niche theory gives the most immediate explanation. Temperature, oxygen, channel slope
and flow change downstream, and fish species differ in their tolerances. Trout persist
in the cool, oxygenated headwaters. Lowland species occupy the warmer and slower reaches.
The species gradient would then follow the environmental gradient through each species'
physiological and ecological tolerances.

A river also joins its sites in a fixed sequence. An individual can reach a neighbouring
reach more easily than a distant one. Even ecologically equivalent species would develop
spatial structure under that constraint as births, deaths and dispersal accumulate
through time. Neutral theory formalises this possibility. Its testable claim is that
dispersal limitation and demographic chance can create community structure without
species-specific environmental responses.

Inspection of the Doubs data cannot separate these explanations. We need predictions
that expose where they differ.

## Ecological Background: Niche Theory

Consider a brown trout in the cold, fast and oxygen-rich upper Doubs. Trout require cool
water with high dissolved oxygen, so their populations decline in the warm, sluggish
lowland reaches. Farther downstream, other species occupy the conditions that exclude
trout. Each species occurs where local conditions fall within its tolerable range.
Niche theory turns this familiar observation into a general account of community
assembly: the environment filters species according to their requirements.

@hutchinson1957 gave the concept its lasting formal definition. A niche is an
*n*-dimensional hypervolume with one axis for every environmental factor that affects
persistence. Temperature may define one axis, dissolved oxygen another, followed by pH,
prey availability and any other relevant condition. The limits on those axes bound the
conditions under which a viable population can persist. As @chase2011 explain, the
species' realised distribution is the geographic expression of that hypervolume.

Hutchinson also drew a distinction that still organises how ecologists use the concept. The
**fundamental niche** is the full range of conditions under which a species could
persist in the absence of competitors and natural enemies. The **realised niche** is
the usually narrower range it actually occupies once biotic interactions are accounted
for. A species might be physiologically capable of living across a broad stretch of the
river, yet be confined to part of it because a competitor excludes it elsewhere. The
environment sets physiological limits, while competitors and natural enemies may narrow
the occupied range further. Under this view, a reproducible match between traits,
conditions and biotic interactions determines where a species lives.

::: {.callout-tip}
## Environmental Filtering

Under niche theory, local conditions filter the regional species pool. Species persist
where their requirements are met. As conditions change across space, the resulting
community changes with them, making composition partly predictable from the environment.
:::

**Environmental filtering** names this mechanism. @chase2011 place it within a
deterministic account of communities, alongside biotic interactions and interspecific
trade-offs. Given sufficient knowledge of the environment and species' tolerances,
niche theory predicts community composition. Similar environments should support
similar communities, while contrasting environments should support different ones.
Geographic separation matters only when it accompanies environmental change.

That distinction gives us a prediction.

::: {.callout-note}
## Prediction 1. The Niche Prediction

Community dissimilarity should track environmental distance. Two reaches with nearly
identical temperature, oxygen and flow should hold similar fish communities even when
they lie far apart. Their environmental contrast should drive the communities apart
even when geographic separation is small.
:::

Evidence for this prediction is substantial. Fish species occupy distinct positions
along river gradients that correspond to measurable niche axes, and niche divergence
along a river continuum has been documented directly [@troia2014]. Still, an orderly
gradient can arise without niche differentiation.

## Ecological Background: Neutral Theory

Now change one assumption. Give every fish individual roughly the same chance of
survival, reproduction and dispersal, irrespective of species. A trout may then occupy
the headwaters because its ancestors were there and its descendants remained nearby.
History and chance shape the community in this account.

This is the premise of **neutral theory**, formulated by @hubbell2001 and reviewed a
decade later by @rosindell2011. The neutrality assumption gives all individuals within
a trophic level the same probabilities of reproduction and death, regardless of species
identity [@rosindell2011]. Species are **demographically equivalent** at the level of
the individual. Birth, death and dispersal rates therefore do not depend on species.

Demographic equivalence does not claim that trout and bream are literally identical.
It asks whether their differences can be ignored when explaining a community pattern
because those differences do not alter per-capita demographic rates. Real species do
differ, and Hubbell treated neutrality as a useful null model. If that null model can
reproduce an observed pattern, the pattern alone cannot establish a niche mechanism.

::: {.callout-tip}
## Demographic Equivalence and Ecological Drift

Neutral theory assumes demographic equivalence among species. **Ecological drift**
changes abundances through chance births and deaths, while **dispersal limitation**
keeps many offspring near their parents. Together these stochastic processes can create
spatial community structure.
:::

Ecological drift is the first mechanism. Random births and deaths cause relative
abundances to wander through time, much as allele frequencies drift in population
genetics. @rosindell2011 trace this parallel to Kimura's neutral theory of molecular
evolution. Drift alone can carry a species towards local extinction or dominance
without a fitness difference.

Dispersal limitation produces the spatial component. @rosindell2011 define it through
the link between a parent's location and the places its offspring can establish. In a
river, fish spawned in one reach are more likely to occupy nearby reaches than the far
end of the system. Across many generations, local drift becomes weakly coupled to drift
at distant sites. Nearby communities share recent dispersal history and tend to
resemble one another.

Neutral theory therefore makes a spatial prediction.

::: {.callout-note}
## Prediction 2. The Neutral Prediction

Community dissimilarity should track geographic distance. Movement connects nearby
sites, so their communities should remain similar. Distant communities should diverge
even when their environments match.
:::

This decline in similarity with separation is **distance decay**. @rosindell2011
describe it as the change, with distance, in the probability that two individuals
belong to the same species. It is a central prediction of neutral theory and the
quantity measured later in this lecture.

::: {.callout-note}
Neutral theory works best as a null model for measuring evidence of niche processes.
When it reproduces a pattern without niche differences, that pattern cannot demonstrate
environmental filtering on its own. Ecologists may still expect niches to matter, but
they need stronger evidence for the claim.
:::

This section develops dispersal limitation only as a local, within-system mechanism.
Regional dispersal dynamics and landscape-scale connectivity belong to the
Metacommunity chapter, which cross-references this section for the underlying
definition.

## Two Worlds, One Pattern

The theories disagree about mechanism but can yield similar observations. We can see
this by building two artificial rivers in which the generating process is known.

In the first, species differ. Each has a Gaussian response to an environmental gradient
running the length of the transect, with its own optimum and tolerance, exactly as niche
theory describes. There is no dispersal at all. In the second, species are identical.
There is no environment. Community composition changes only through ecological drift,
and individuals disperse only to adjacent sites, exactly as neutral theory describes.

```{r code-libraries}
library(here)
library(tidyverse)
library(vegan)
library(patchwork)
```

```{r code-two-worlds}
#| cache: true
#| code-fold: true
#| code-summary: "Show the simulation code"
# A niche world: Gaussian species responses to a gradient, no dispersal.
sim_niche <- function(seed, n_site = 29, S = 27, J = 100, tol = 0.12) {
    set.seed(seed)
    gradient <- seq(0, 1, length.out = n_site)
    optima <- seq(-0.1, 1.1, length.out = S)
    t(sapply(gradient, function(e) {
        p <- exp(-((e - optima)^2) / (2 * tol^2))
        as.vector(rmultinom(1, J, p / sum(p)))
    }))
}

# A neutral world: demographic equivalence, drift, dispersal to adjacent sites only.
# No environmental variable exists anywhere in this simulation.
sim_neutral <- function(seed, n_site = 29, S = 27, J = 100, gens = 250,
                        m = 0.05, nu = 0.002) {
    set.seed(seed)
    counts <- t(replicate(n_site, as.vector(rmultinom(1, J, rep(1 / S, S)))))
    sites <- seq_len(n_site)
    for (e in seq_len(n_site * J * gens)) {
        d <- sample.int(n_site, 1)
        dead <- sample.int(S, 1, prob = counts[d, ]) # a death, at random
        counts[d, dead] <- counts[d, dead] - 1
        u <- runif(1)
        if (u < nu) { # rare immigration from outside
            born <- sample.int(S, 1)
        } else if (u < nu + m) { # birth from an ADJACENT site
            nbrs <- intersect(c(d - 1, d + 1), sites)
            nb <- nbrs[sample.int(length(nbrs), 1)]
            born <- sample.int(S, 1, prob = counts[nb, ])
        } else { # birth from this site
            born <- sample.int(S, 1, prob = counts[d, ])
        }
        counts[d, born] <- counts[d, born] + 1
    }
    counts
}

niche_runs <- lapply(1:4, sim_niche)
neutral_runs <- lapply(1:4, sim_neutral)

world_mats <- tibble(
    world = rep(c("Niche world", "Neutral world"), each = 4),
    run   = rep(1:4, times = 2),
    mat   = c(niche_runs, neutral_runs)
)
```

```{r fig-two-worlds}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 3.6
#| fig-cap: "Two artificial rivers, four runs each. In the niche world, species respond to an environmental gradient and never disperse. In the neutral world, species are demographically identical, no environment exists, and individuals disperse only between adjacent sites. Each panel shows the first axis of a principal coordinates analysis of Bray-Curtis dissimilarity against position along the transect, annotated with the absolute correlation between the two. The niche world returns the same gradient on every run. The neutral world returns a different one each time, with nothing changed but the random seed."
axis1_of <- function(mat) {
    a <- as.numeric(cmdscale(vegdist(mat, method = "bray"), k = 1))
    a * sign(cor(a, seq_along(a))) # ordination axis signs are arbitrary
}

world_axes <- world_mats |>
    mutate(axis1 = map(mat, axis1_of)) |>
    select(world, run, axis1) |>
    unnest(axis1) |>
    group_by(world, run) |>
    mutate(site = row_number()) |>
    ungroup()

# the correlation is computed here, not asserted in the caption
world_r <- world_axes |>
    group_by(world, run) |>
    summarise(r = abs(cor(axis1, site)), .groups = "drop")

ggplot(world_axes, aes(site, axis1)) +
    geom_point(aes(colour = site), size = 0.9) +
    geom_text(
        data = world_r, inherit.aes = FALSE,
        aes(x = 1, y = Inf, label = sprintf("|r| = %.2f", r)),
        hjust = 0, vjust = 1.5, size = 2.3, colour = "grey25"
    ) +
    facet_grid(world ~ paste("run", run)) +
    scale_colour_viridis_c(guide = "none") +
    labs(x = "Position along the transect", y = "PCoA axis 1")
```

The niche simulation returns an almost perfect relationship between the first
ordination axis and position on every run. Repeating it produces the same picture
because the environment is fixed and the species responses are stable.

The neutral result changes with the random seed. Some runs produce a first axis that
tracks position almost as closely as the niche simulation, while others show little
structure. An ordination of one favourable neutral run would give no warning that drift
produced the gradient. Random walks sometimes look orderly.

One result persists across every run.

```{r fig-two-worlds-decay}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 2.6
#| fig-cap: "Community dissimilarity against distance between sites, for the same eight simulated communities. Both worlds produce distance decay. The neutral world produces it because offspring settle near their parents. The niche world produces it because the environment changes smoothly along the transect, so distant sites are also environmentally different. Distance decay alone therefore identifies neither process."
world_mats |>
    mutate(d = lapply(mat, function(m) {
        tibble(
            distance = as.vector(dist(seq_len(nrow(m)))),
            bray = as.vector(vegdist(m, method = "bray"))
        )
    })) |>
    select(world, run, d) |>
    unnest(d) |>
    ggplot(aes(distance, bray)) +
    geom_point(alpha = 0.12, size = 0.5) +
    geom_smooth(method = "loess", se = FALSE, linewidth = 0.5, formula = y ~ x) +
    facet_wrap(~world) +
    labs(x = "Distance between sites", y = "Bray-Curtis dissimilarity")
```

Both simulations show distance decay. In the neutral world, nearby offspring connect
local communities. In the niche world, sites become environmentally less alike with
distance along the transect. A process with no dispersal therefore produces the same
signature predicted by neutral theory.

::: {.callout-tip}
## Replication Reveals Stability

A neutral community can mimic a niche gradient closely enough to fool an ordination.
Repeated niche simulations return the same gradient, while neutral simulations vary
widely among runs. Replication reveals that difference in stability. A single dataset
does not.

We have one Doubs. We cannot re-run it with a different seed.
:::

## The Niche-Neutral Continuum: A Simulation {#sec-continuum}

Niche and neutral processes can operate together, and their relative contributions vary
among systems. @adler2007 formalised this as a continuum in which the pure neutral model
is the limiting case as stabilising mechanisms and fitness differences approach zero.
The Two Worlds simulation showed the endpoints. We now simulate the values between
them.

A mixing parameter, $w$, controls the balance between the two assembly processes. At
$w = 0$, ecological drift and dispersal limitation assemble the community. At $w = 1$,
environmental filtering dominates apart from sampling noise. Intermediate values blend
the processes in known proportions.

```{r code-continuum-sim}
#| cache: true
#| code-fold: true
#| code-summary: "Show the mixed simulation code"
sim_mixed <- function(seed, w, n_site = 29, S = 27, J = 100) {
    set.seed(seed)
    gradient <- seq(0, 1, length.out = n_site)
    optima <- seq(-0.1, 1.1, length.out = S)
    tol <- 0.12

    # niche component: probability from Gaussian response
    niche_probs <- t(sapply(gradient, function(e) {
        p <- exp(-((e - optima)^2) / (2 * tol^2))
        p / sum(p)
    }))

    # neutral component: equal probability for all species
    neutral_probs <- matrix(1 / S, nrow = n_site, ncol = S)

    # mixed: weighted combination
    mixed_probs <- w * niche_probs + (1 - w) * neutral_probs

    # sample community from the mixed probabilities
    t(sapply(seq_len(n_site), function(i) {
        as.vector(rmultinom(1, J, mixed_probs[i, ]))
    }))
}

# Run across the continuum
weights <- c(0, 0.25, 0.5, 0.75, 1.0)

continuum_data <- expand_grid(w = weights, run = 1:4) |>
    mutate(
        mat = lapply(seq_len(n()), function(i) sim_mixed(run[i], w[i])),
        axis1 = lapply(mat, function(m) {
            a <- as.numeric(cmdscale(vegdist(m, "bray"), k = 1))
            a * sign(cor(a, seq_along(a)))
        })
    ) |>
    select(w, run, axis1) |>
    unnest(axis1) |>
    group_by(w, run) |>
    mutate(site = row_number()) |>
    ungroup()

continuum_r <- continuum_data |>
    group_by(w, run) |>
    summarise(r = abs(cor(axis1, site)), .groups = "drop")
```

```{r fig-continuum}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 4.0
#| fig-cap: "The niche-neutral continuum as a simulation. Each column is a different mixing weight w, from pure neutral (w = 0, left) to pure niche (w = 1, right), with four independent runs per weight. As the niche component strengthens, the gradient becomes stronger and more reproducible. The transition is gradual, making the Adler continuum visible."
ggplot(continuum_data, aes(site, axis1)) +
    geom_point(aes(colour = site), size = 0.7) +
    geom_text(
        data = continuum_r, inherit.aes = FALSE,
        aes(x = 1, y = Inf, label = sprintf("|r|=%.2f", r)),
        hjust = 0, vjust = 1.5, size = 2, colour = "grey30"
    ) +
    facet_grid(paste("run", run) ~ paste0("w = ", w)) +
    scale_colour_viridis_c(guide = "none") +
    labs(x = "Position along the transect", y = "PCoA axis 1") +
    theme(strip.text = element_text(size = 8))
```

```{r fig-continuum-summary}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 2.8
#| fig-cap: "Mean and range of |r| across four runs at each mixing weight. As the niche component increases, the mean gradient strength rises and the variability between runs decreases. At w = 0 (pure neutral), gradient strength is unpredictable. At w = 1 (pure niche), it is nearly deterministic. The transition is smooth, with no threshold."
continuum_summary <- continuum_r |>
    group_by(w) |>
    summarise(mean_r = mean(r), min_r = min(r), max_r = max(r), .groups = "drop")

ggplot(continuum_summary, aes(w, mean_r)) +
    geom_ribbon(aes(ymin = min_r, ymax = max_r), fill = "grey85", alpha = 0.5) +
    geom_line(linewidth = 0.6) +
    geom_point(size = 1.5) +
    scale_x_continuous(breaks = weights, labels = weights) +
    labs(
        x = "Mixing weight w (0 = pure neutral, 1 = pure niche)",
        y = "|r| between PCoA axis 1 and position"
    ) +
    annotate("text", x = 0.05, y = 0.15, label = "Neutral\nend", size = 2.5, hjust = 0, colour = "grey50") +
    annotate("text", x = 0.95, y = 0.95, label = "Niche\nend", size = 2.5, hjust = 1, colour = "grey50")
```

As $w$ increases, mean $|r|$ rises and the ordination recovers a stronger environmental
gradient. The result also becomes more repeatable. At $w = 0$, the four runs scatter
widely because drift follows a random path. At $w = 1$, they converge because each
species responds consistently to the fixed environment. The ribbon narrows gradually,
with no threshold where the community switches between mechanisms.

The simulation makes the Adler continuum measurable. A real community's position on
that continuum affects both gradient strength and predictability. The Doubs has a strong,
orderly gradient and probably lies towards the niche end, although one dataset cannot
tell us how far. Its sampling design sets that limit.

::: {.callout-tip}
## A Measurable Continuum

The niche-neutral continuum is a quantitative property that can be estimated in
principle. In this simulation, $w$ controls the strength and repeatability of community
structure. Its ecological value will vary with spatial scale, organism and the sampled
environmental range. This gives the arguments of @adler2007 and @vellend2010 a concrete,
testable form.
:::

## The Niche-Neutral Debate: What Is Settled, What Is Contested, and Where This Lecture Stands

Neither theory supplies a complete account of community assembly, and they are not
interchangeable. Their predictions overlap, while their assumptions remain distinct.
The literature is clearer on some points than others, so we separate established
evidence from the questions that remain open.

### What is established

Both processes occur in real communities. Environmental filtering sorts species along
gradients, and river fish occupy positions that correspond to their physiological
tolerances [@troia2014]. Dispersal limitation also creates spatial structure independent
of environment, producing distance decay even across comparatively uniform habitat
[@rosindell2011]. Data contradict any claim that one of these processes never matters.

Static abundance patterns cannot distinguish them by themselves. @adler2007 begin from
this limitation: neutral and niche models can generate abundance distributions that are
nearly indistinguishable. The Doubs gradient therefore remains compatible with several
assembly processes, however clearly we measure it.

### What is contested

Their relative importance remains open, as does the extent to which they can be separated.

The balance varies with spatial scale, the organisms involved and the environmental
range sampled [@chase2011]. A process may look deterministic at one extent and stochastic
at another. Any claim about dominance therefore needs a stated scale, including the
worked examples below.

Researchers also disagree about the categories themselves. The pure neutral model
assumes equal fitness and equivalent interspecific effects [@adler2007], conditions
unlikely to hold exactly. Its usefulness depends on how closely a system approaches
those assumptions and which predictions remain informative when it does not.

### Where this lecture stands: synthesis

A useful synthesis treats niche and neutral processes as simultaneous contributions
whose relative strengths vary among systems. Two published reframings support that
position.

@adler2007 place neutrality within classical coexistence theory. Coexistence depends on
stabilising mechanisms and fitness differences among species. The pure neutral model is
the limiting case where stabilisation is absent and fitness is equivalent. Real
communities can occupy many positions between strong stabilisation and near-equivalent
fitness.

@vellend2010 groups community-ecology theory into four high-level processes analogous
to forces in population genetics: selection, drift, speciation and dispersal. Niche
theory concerns selection through deterministic fitness differences. Neutral theory
concerns stochastic drift, while dispersal and speciation accompany both. This framework
asks how strongly each process operates in a given community.

@leibold2006 show that niche relations can favour coexistence among ecologically similar
species. Apparent equivalence therefore does not demonstrate the absence of niche
differences. A niche-structured community may look neutral because of the way its niches
promote coexistence.

::: {.callout-tip}
## Estimate Relative Contributions

Ask how strongly each process operates and at what scale. A real community may contain
both. The analyses that follow estimate their relative signals within the limits of the
sampling design.
:::

For the Doubs, we ask whether community dissimilarity tracks environmental distance,
geographic distance or both, and how strong each relationship is. Static data cannot
prove that the community is wholly niche-structured or wholly neutral. Even the narrower
question is difficult because the two predictors change together along the river.

## Data: The Doubs River Fish Community

You have met this dataset four times. The Doubs River rises in the Jura Mountains near
the French-Swiss border and flows for about 450 km before joining the Saône. Verneaux
surveyed its fish assemblages to characterise ecological zones along the river. The data
used here come through @borcard2011.

Three tables describe the same 30 sites. Each theory predicts a relationship among these
tables, so their rows must remain aligned throughout the analysis.

| Object | Contents | Dimensions |
|:-------|:---------|:-----------|
| `spe` | abundance of each fish species at each site, on a 0-5 semi-quantitative scale | 30 sites × 27 species |
| `env` | measured environmental variables at each site | 30 sites × 11 variables |
| `spa` | Cartesian coordinates locating each site in the plane | 30 sites × 2 |

: The three Doubs tables. Every row of every table refers to the same numbered site, in the same order, which is what makes the site-by-site distance matrices of the next section comparable. {#tbl-doubs-objects}

The environmental variables, whose codes are worth having at hand, are those introduced
in the [Correlations and Associations](https://tangledbank.netlify.app/BCB743/correlations.html) chapter:

| Code | Variable | Units |
|:-----|:---------|:------|
| `dfs` | distance from source | km |
| `ele` | elevation | m a.s.l. |
| `slo` | channel slope | ‰ |
| `dis` | mean minimum discharge | m³ s⁻¹ |
| `pH` | water pH | dimensionless |
| `har` | total hardness (calcium) | mg L⁻¹ |
| `pho` | phosphate | mg L⁻¹ |
| `nit` | nitrate | mg L⁻¹ |
| `amm` | ammonium | mg L⁻¹ |
| `oxy` | dissolved oxygen | mg L⁻¹ |
| `bod` | biological oxygen demand | mg L⁻¹ |

: The eleven variables in `env`. The first entry, `dfs`, is a positional variable. The other entries describe local environmental conditions. {#tbl-doubs-env}

Loading the data requires nothing new. As in the [Cluster
Analysis](https://tangledbank.netlify.app/BCB743/cluster_analysis.html) chapter, the eighth site is dropped. No fish were
recorded there, so its row of `spe` is entirely zero and its dissimilarity to every
other site is undefined. Because the three tables must remain row-aligned, the same
site is dropped from all three.

```{r code-load-doubs}
load(here::here("data", "BCB743", "NEwR-2ed_code_data", "NEwR2-Data", "Doubs.RData"))

# Site 8 holds no fish. Its Bray-Curtis dissimilarity to any other site is
# undefined. Drop it from all three tables so the rows stay aligned.
spe <- dplyr::slice(spe, -8)
env <- dplyr::slice(env, -8)
spa <- dplyr::slice(spa, -8)
```

::: {.callout-note}
## How to set up the data on your own machine

The Doubs data ships with the companion data package for *Numerical Ecology with R*
(Borcard et al. 2011). If you are rendering this lecture outside the Tangled Bank project
folder, download `Doubs.RData` from the course data repository and place it at:

    <project_root>/data/BCB743/NEwR-2ed_code_data/NEwR2-Data/Doubs.RData


Then set the project root with `here::here()` by opening the `.Rproj` file or
placing a `.here` file at the root. The `load()` call above will resolve the path
automatically on any machine.

Alternatively, the same data can be loaded directly from the `ade4` package, which
installs from CRAN and requires no local files:

```r
library(ade4)
data(doubs)
spe <- doubs$fish   # 30 sites × 27 species
env <- doubs$env    # 30 sites × 11 environmental variables
spa <- doubs$xy     # 30 sites × 2 spatial coordinates (x, y)
```

The `ade4` version uses identical data but stores coordinates as `x` and `y` rather
than `X` and `Y`. Adjust column name references in downstream code if you use this
route.
:::

### A variable in the wrong table

Ten variables in @tbl-doubs-env describe local water or channel conditions. `dfs` is
different. It records position along the river. Fish respond to temperature, oxygen
and flow, while distance from the source is a coordinate that may covary with those
conditions.

This classification affects the result. The analysis tries to separate environmental
from spatial structure. If `dfs` remains in the environmental table, environmental
distance partly contains spatial distance and the test is tilted towards the niche
hypothesis before permutation begins. The [Correlations](https://tangledbank.netlify.app/BCB743/correlations.html) lecture showed
that `dfs` correlates at $-0.94$ with elevation and $0.95$ with discharge. It behaves
almost like a spatial axis.

The effect is measurable. Keeping `dfs` in the environmental table raises the
correlation between environmental and along-channel distance from $r = 0.55$ to
$r = 0.63$. The added correlation is spatial signal assigned to the niche hypothesis.

::: {.callout-important}
## Why `dfs` Is Treated as Spatial

We move `dfs` from the environmental matrix to the spatial information. This is our
methodological judgement, so the reasoning needs to be explicit.

`dfs` measures position. Including it among environmental variables confounds the
explanations under study. It may also represent spatial separation more appropriately
for this system than the coordinates in `spa`. Fish disperse through the channel.
Sites on opposite banks of a meander may be close in Euclidean space but far apart along
the water, which is the route available to the fish.

You will see in @sec-worked that the second choice is not cosmetic. It changes the
answer.
:::

```{r fig-doubs-geometry}
#| code-fold: true
#| code-summary: "Show the map code"
#| fig-height: 3.8
#| fig-width: 10
#| fig-cap: "Left: the 29 Doubs sampling sites in the plane, joined in river order and shaded by distance from source. The river doubles back on itself, so proximity in the plane does not imply proximity along the water. Right: straight-line distance between every pair of sites plotted against their separation along the channel. The two disagree substantially (r = 0.64). Sites 19 and 20, for instance, lie 8.2 planar units apart but 33.5 km apart along the river."
spa_xy <- spa |>
    rename(x = 1, y = 2) |>
    mutate(dfs = env$dfs, site = row_number())

p_map <- ggplot(spa_xy, aes(x, y)) +
    geom_path(linewidth = 0.3, colour = "grey75") +
    geom_point(aes(colour = dfs), size = 1.5) +
    scale_colour_viridis_c(name = "dfs (km)") +
    coord_equal() +
    labs(x = NULL, y = NULL)

p_dist <- tibble(
    euclidean = as.vector(dist(spa_xy[, c("x", "y")])),
    channel = as.vector(dist(env$dfs))
) |>
    ggplot(aes(euclidean, channel)) +
    geom_point(alpha = 0.2, size = 0.5) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.4, formula = y ~ x) +
    labs(x = "Straight-line distance", y = "Along-channel distance (km)")

patchwork::wrap_plots(p_map, p_dist, ncol = 2)
```

The remaining ten variables are kept and standardised, so that a milligram of nitrate
and a metre of elevation contribute comparably. Standardisation is not optional here.
The variables in @tbl-doubs-env span several orders of magnitude, and Euclidean
distance computed on raw values would be dominated by whichever variable happened to
carry the largest units.

## Analytical Logic: Testing the Predictions {#sec-logic}

Niche theory predicts that community dissimilarity rises with environmental distance.
Neutral theory predicts a rise with geographic distance. Both predictions relate
differences in community composition to another pairwise difference among the same
sites. That structure suggests the Mantel test. We examine it because understanding why
it fails here is more useful than memorising a replacement.

### The intuitive approach

The [Distance and Dissimilarity Metrics](https://tangledbank.netlify.app/BCB743/dis-metrics.html) chapter showed how to turn raw
measurements into pairwise differences. Here we can calculate community dissimilarity
from `spe`, environmental distance from `env` and geographic distance from the spatial
information. The apparent next step is to ask which explanatory matrix most closely
resembles community dissimilarity.

Correlating two distance matrices is what the **Mantel test** does. @mantel1967
introduced it to relate a matrix of spatial distances to a matrix of temporal distances
in a study of disease clustering. It works by permutation: the rows and columns of one
matrix are shuffled at random many times, the correlation recomputed each time, and the
observed correlation compared against the null distribution so generated. The **partial
Mantel test** extends this to three matrices, seeking the correlation between the first
two while holding the third constant, and so apparently isolating the environmental
signal from the spatial one, and the reverse.

The procedure seems well matched to the question, but it fails for two reasons. The
Doubs design confounds the predictors, and the test evaluates a different null
hypothesis from the one we need.

### The first problem: the predictors are not independent

Before choosing a test, ask whether the dataset could distinguish the hypotheses in
principle.

The predictions differ only when environmental and geographic distance contain
independent information. Along a river they rarely do. Temperature, flow, oxygen and
nutrients change downstream, so sites far apart along the channel are usually
environmentally different as well.

```{r code-confound}
env_hab <- dplyr::select(env, -dfs)
env_dist <- dist(decostand(env_hab, method = "standardize"))
chan_dist <- dist(env$dfs) # along-channel separation
euc_dist <- dist(spa) # straight-line separation

# Are the two explanatory matrices themselves correlated?
mantel(env_dist, chan_dist)
```

The relationship is strong, at about $r = 0.50$. Along-channel separation predicts a
substantial share of environmental distance among the Doubs sites. The hypotheses are
therefore asked to explain overlapping structure.

```{r fig-confound}
#| fig-height: 2.5
#| fig-cap: "Environmental distance between every pair of Doubs sites, plotted against their separation along the channel. The two explanatory matrices are themselves correlated (Mantel r = 0.50), because a river changes systematically as it flows. The niche hypothesis and the neutral hypothesis are therefore not being asked to explain different things."
tibble(
    channel = as.vector(chan_dist),
    environment = as.vector(env_dist)
) |>
    ggplot(aes(channel, environment)) +
    geom_point(alpha = 0.2, size = 0.6) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, formula = y ~ x) +
    labs(x = "Along-channel distance (km)", y = "Environmental distance")
```

This is collinearity at the level of whole matrices, with the same consequence described
for regression predictors in the [Correlations](https://tangledbank.netlify.app/BCB743/correlations.html) chapter. No method can
cleanly assign shared information to one predictor. A single linear transect along one
river gives little leverage for separating space from environment.

::: {.callout-tip}
## Ask What the Design Can Deliver

A hypothesis test cannot separate explanations represented by confounded predictors.
In the Doubs, environment and space vary together, so their effects cannot be fully
partitioned. Check that limitation before choosing a method or interpreting a number.
:::

### The second problem: the test does not test what you think

Even without that confounding, the test itself presents a problem.

@legendre2015 examined the use of Mantel tests to relate community composition to
environment while controlling for spatial correlation. They concluded that this is an
incorrect application. Their reasoning shows how a familiar procedure can drift away
from the ecological question it is meant to answer.

The Mantel null hypothesis concerns association between the values in two dissimilarity
matrices. Correlation and regression instead test relationships between raw data tables
[@legendre2015]. Our community, environmental and coordinate data begin as raw tables,
and the ecological question concerns those tables. Converting them to distances changes
the hypothesis. Mantel $R^2$ is therefore not equivalent to the $R^2$ from regression
or canonical analysis. @legendre2015 regard it as uninterpretable in this spatial
setting.

The assumptions of linearity and homoscedasticity require small values in one matrix to
correspond to small values in the other, and likewise for large values. These assumptions
fail in many spatial applications and hold only when spatial correlation extends across
the full study area [@legendre2015]. A 30-site river transect may not meet that condition.

Power is also low. In the simulations reported by @legendre2015, Mantel tests detected
spatial structure less often than distance-based Moran's eigenvector map (dbMEM)
analysis. This increases the chance of missing a real signal when stronger valid tests
are available.

Partial Mantel tests add a more serious failure. Regressing on geographic distance does
not remove spatial structure from the response or produce spatially independent
residuals [@legendre2015]. In simulations by @oden1992, a nominal 5% test rejected
13.2% of true null hypotheses under the strongest spatial autocorrelation. Another test
statistic in the same study reached 44%. @guillot2013 later found that partial Mantel
tests remained about as biased as simple Mantel tests under realistic sampling designs
and autocorrelation. The method intended to separate niche and spatial signals may
manufacture apparent evidence for both.

::: {.callout-important}
## Evidence Behind the Critique

This critique comes from Pierre Legendre and Daniel Borcard, authors of *Numerical
Ecology* [@legendre2012] and *Numerical Ecology with R* [@borcard2011]. The paper also
reports simulations of community composition generated under Hubbell's neutral model,
so its evidence closely matches the data structure considered here [@legendre2015].

The Mantel test remains widely used in ecology for this purpose [@legendre2015]. Being
common is not the same as being correct.
:::

### Seeing the failure for yourself

We can reproduce the problem in a simulation where the truth is known.

Two independent variables are simulated along a 29-site transect matching the Doubs.
Each is spatially autocorrelated, so nearby sites have similar values. One represents
community composition and the other an environmental predictor. Their generating
processes are independent, which makes the null hypothesis true by construction.

A valid test at $\alpha = 0.05$ should reject about 5% of these true null hypotheses.
The partial Mantel test receives the correct spatial distance matrix as its covariable.

```{r code-type1-sim}
#| cache: true
#| code-fold: true
#| code-summary: "Show the Type I error simulation code"
set.seed(42)

n <- 29 # sites, as in the Doubs
pos <- 1:n # positions along a transect
S <- as.matrix(dist(pos)) # true spatial distances
d_spa <- dist(pos)

# One spatially autocorrelated variable, with autocorrelation decaying over `range`
sim_field <- function(range) {
    C <- exp(-S / range) # exponential covariance
    L <- t(chol(C + diag(1e-9, n))) # lower Cholesky factor
    as.vector(L %*% rnorm(n))
}

# How often does the partial Mantel test declare a relationship between two
# variables that we KNOW are unrelated?
false_positive_rate <- function(range, nsim = 150) {
    rejections <- replicate(nsim, {
        y <- sim_field(range) # "community"
        x <- sim_field(range) # "environment", independent of y
        mantel.partial(dist(y), dist(x), d_spa, permutations = 199)$signif <= 0.05
    })
    mean(rejections)
}

type1 <- tibble(
    autocorrelation_range = c(1, 2, 5, 10),
    false_positive_rate   = map_dbl(autocorrelation_range, false_positive_rate)
)
type1
```

```{r fig-type1}
#| fig-height: 2.5
#| fig-cap: "False-positive rate of the partial Mantel test on data where the null hypothesis is true by construction: two spatially autocorrelated variables, generated independently of one another, with the true spatial distance matrix supplied as the covariable. A valid test would reject at the nominal 5% rate, shown by the dashed line. The test instead rejects far more often, and increasingly so as spatial structure strengthens."
ggplot(type1, aes(autocorrelation_range, false_positive_rate)) +
    geom_hline(yintercept = 0.05, linetype = 2, colour = "grey40") +
    geom_line(linewidth = 0.4) +
    geom_point(size = 1.4) +
    scale_y_continuous(limits = c(0, NA), labels = scales::percent) +
    annotate("text",
        x = 1.2, y = 0.062, label = "nominal 5%",
        size = 2.4, colour = "grey40", hjust = 0
    ) +
    labs(x = "Autocorrelation range (sites)", y = "False-positive rate")
```

The false-positive rate is about 0.12 when autocorrelation decays over two sites and
rises to roughly 0.25 when it extends over ten. Between one in eight and one in four
tests report a relationship that was absent from the simulation. The error worsens as
spatial structure strengthens, which is especially troubling for a linear system such
as the Doubs.

::: {.callout-important}
## Test the False-Positive Rate

Run the chunk above and change the numbers. Increase `nsim` to 500, widen the
autocorrelation range to 15, or replace `mantel.partial` with `mantel`. A procedure that
rejects a true null hypothesis in one quarter of simulations cannot support a reliable
ecological conclusion.

The simulation evaluates a statistical method against a known generating process.
Neutral theory uses the same logic when it generates communities from a specified null
model.
:::

### What survives: the Mantel correlogram

The conclusion of @legendre2015 is narrower than a blanket prohibition. Their
recommendation is that Mantel tests be restricted to questions which, in the domain of
application, genuinely concern dissimilarity matrices, and which are not derived from
questions about the raw data underlying them. Such questions are uncommon in ecology.
But some questions do qualify.

The **Mantel correlogram** meets the narrower restriction. It groups site pairs into
geographic distance classes and measures community similarity within each class.
@legendre2015 report good power for this use, including when the range of spatial
autocorrelation is unknown.

Neutral theory predicts how similarity changes across distance classes, not an overall
correlation between matrices [@rosindell2011]. A Mantel correlogram therefore measures
the form of distance decay that the theory describes.

### What to use instead, for the environmental question

For the niche-side question, the appropriate tools operate on raw community and
environmental tables while accounting for space. @legendre2015 recommend dbMEM analysis by
regression or redundancy analysis, whose adjusted $R^2$ is an unbiased estimate of the
variation explained, and whose spatial eigenfunctions can be grouped by scale and
entered into variation partitioning alongside environmental predictors.

You have already seen this done. In the [Seaweeds in Two
Oceans](https://tangledbank.netlify.app/BCB743/two_oceans_appendices.html) appendix, `varpart()` is used with MEM variables to
split the variation in a species matrix into environmental and spatial fractions, the
very partition this lecture has spent its theory sections motivating.

::: {.callout-important}
## Scope of This Lecture

This lecture explains why environmental and spatial signals need to be separated and
what each fraction implies about assembly. The Gradients chapter, "From Ecological
Gradients to Ecological Inference", teaches the mechanics of `varpart()`, dbMEM and
constrained ordination. Read the chapters together: this one develops the ecological
question, while the Gradients chapter develops the quantitative method.
:::

::: {.callout-tip}
## Statistical and Conceptual Limits

Translating a theory into a falsifiable prediction is only the start. A sampling design
may confound the predictors, and a familiar test may evaluate the wrong hypothesis with
low power and inflated false-positive rates. Method choice must follow the ecological
question and the structure of the data.
:::

```{mermaid}
%%| label: fig-niche-neutral-logic
%%| echo: false
%%| fig-cap: "The reasoning of this lecture. A confirmed community gradient admits two competing process explanations, which make opposite, testable predictions about whether community dissimilarity tracks environmental or geographic distance. Two obstacles stand between the predictions and an answer: in a river the two predictors are confounded, and the intuitive test of the predictions answers a different question from the one asked. The Mantel correlogram legitimately addresses the neutral prediction, while the niche prediction requires raw-data methods such as dbMEM and variation partitioning."
flowchart TD
  A["Doubs gradient confirmed<br/>(Ch. 4, 17)"] --> B["Why does it exist?"]
  B --> C["Niche theory:<br/>environmental filtering"]
  B --> D["Neutral theory:<br/>dispersal limitation"]
  C --> E["Predicts: dissimilarity<br/>tracks environmental distance"]
  D --> F["Predicts: dissimilarity<br/>tracks geographic distance"]
  E --> G["Obstacle 1: in a river the two<br/>predictors are confounded"]
  F --> G
  G --> H["Obstacle 2: Mantel & partial Mantel<br/>test the wrong hypothesis"]
  H --> I["Mantel correlogram:<br/>distance decay"]
  H --> J["dbMEM + varpart:<br/>see Gradients chapter"]
  I --> K["Interpret signal, with limits"]
  J --> K
  K --> L["Hand-off to Ch. 18:<br/>hypothesis-driven model building"]
```

## Worked Analysis {#sec-worked}

We begin with the tempting Mantel analysis, then examine which conclusions survive its
limitations.

### Building the matrices

```{r code-build-matrices}
spe_bray <- vegdist(spe, method = "bray") # community dissimilarity
# env_dist, chan_dist and euc_dist were built in the chunk above
```

### The naive analysis

```{r code-mantel-naive}
m_env <- mantel(spe_bray, env_dist)
m_chan <- mantel(spe_bray, chan_dist)
m_euc <- mantel(spe_bray, euc_dist)

p_env <- mantel.partial(spe_bray, env_dist, chan_dist)
p_chan <- mantel.partial(spe_bray, chan_dist, env_dist)

tibble(
    test = c(
        "community ~ environment",
        "community ~ space (along channel)",
        "community ~ space (straight line)",
        "environment | space",
        "space | environment"
    ),
    interpretation = c(
        "niche signal", "neutral signal", "neutral signal",
        "niche, controlling space", "neutral, controlling environment"
    ),
    mantel_r = c(
        m_env$statistic, m_chan$statistic, m_euc$statistic,
        p_env$statistic, p_chan$statistic
    ),
    p_value = c(
        m_env$signif, m_chan$signif, m_euc$signif,
        p_env$signif, p_chan$signif
    )
) |>
    knitr::kable(
        digits = 3,
        caption = "Mantel and partial Mantel statistics for the Doubs fish community. Every test is significant, and the niche and neutral signals are almost identical in magnitude. Read on before drawing any conclusion from this table."
    )
```

Taken at face value, the table suggests that community composition is related to
environmental distance ($r \approx 0.61$) and to
along-channel distance ($r \approx 0.74$). Both relationships survive controlling for
the other ($r \approx 0.34$ and $r \approx 0.61$). The spatial signal is substantially
stronger. Therefore both niche and neutral processes operate, with dispersal limitation
appearing the dominant signal.

That interpretation does not hold up.

These two correlations differ by about $0.13$. Nothing in the analysis licenses
treating that gap as meaningful, and because environmental distance and channel distance
are themselves correlated at $r \approx 0.57$, the two tests are not independent
assessments of rival hypotheses. They are two views of a single confounded gradient.

The partial statistics look reassuring and are the least trustworthy numbers in the
table. We showed in @sec-logic that on spatially autocorrelated data, which this
emphatically is, the partial Mantel test rejects a true null hypothesis between one in
eight and one in four times. Both partial tests here return $p = 0.001$. So would a
substantial fraction of tests run on data with no relationship at all.

### The result that should worry you most

Look again at the third row of the table.

Community dissimilarity correlates with along-channel distance at $r \approx 0.74$, but
with straight-line distance at only $r \approx 0.43$. Same communities, same sites, same
test. The strength of the neutral signal has halved because we changed our mind about
what distance means.

The difference comes from a modelling decision about how fish move, not from new
observations. Euclidean distance makes the spatial signal look modest. Channel distance
makes it look stronger than the environmental signal.

Channel distance is ecologically defensible because fish disperse through water. No
procedure in `vegan` can select that representation without knowledge of the organism
and river network. The distance matrix already contains a biological judgement.

::: {.callout-note}
## Try This

Rebuild `chan_dist` as `dist(log1p(env$dfs))`, on the reasoning that dispersal
probability may decline logarithmically with distance. Rerun the
Mantel test. Then decide which of the three distance matrices you would defend, and on
what grounds. If your answer changes with each specification, report that sensitivity
and explain the representation you finally chose.
:::

### The defensible analysis: distance decay

Neutral theory's prediction was never really that the overall Mantel correlation with
space is positive. It was that community similarity decays with distance, sharply at
first and then more gently. That is a claim about distance classes, and the Mantel
correlogram is built to test it.

```{r fig-doubs-correlog}
#| fig-height: 3.2
#| fig-cap: "Mantel correlogram of the Doubs fish community against along-channel distance. Each point is the Mantel correlation between community dissimilarity and membership of one along-channel distance class. Filled symbols are significant after Holm correction for multiple testing. Positive values at short distances and negative values at long distances are the signature of distance decay: nearby reaches hold similar assemblages, distant reaches do not."
doubs_correlog <- mantel.correlog(
    D.eco  = spe_bray,
    D.geo  = chan_dist,
    nperm  = 999,
    r.type = "pearson",
    cutoff = FALSE,
    mult   = "holm"
)

plot(doubs_correlog)
```

```{r fig-doubs-decay}
#| fig-height: 2.8
#| fig-cap: "Community dissimilarity between every pair of Doubs sites, plotted against environmental distance (left) and along-channel distance (right). Both relationships are positive, and neither is decisively stronger, which is precisely the difficulty. Because environmental and channel distance are themselves correlated at r = 0.50, these two panels are not independent evidence for rival hypotheses."
tibble(
    bray        = as.vector(spe_bray),
    environment = as.vector(env_dist),
    channel     = as.vector(chan_dist)
) |>
    pivot_longer(c(environment, channel),
        names_to = "predictor", values_to = "distance"
    ) |>
    ggplot(aes(distance, bray)) +
    geom_point(alpha = 0.25, size = 0.6) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
    facet_wrap(~predictor, scales = "free_x") +
    labs(x = "Distance between sites", y = "Bray-Curtis dissimilarity")
```

Distance decay is present and monotone. Mean Bray-Curtis dissimilarity between reaches within
about 47 km of one another is $0.42$. Between reaches at opposite ends of the river, it
is $0.91$. Communities become steadily less alike as the water carries you further from
them.

Both neutral and niche theory predict this pattern in a river whose environment changes
monotonically from source to mouth. The correlogram validly measures distance decay.
Attributing it to an assembly process requires additional evidence.

::: {.callout-tip}
## What the Data Would Need

The Doubs cannot separate niche from neutral, because along this river environment and
distance change together. So ask the design question. What would a dataset that could
separate them look like?

It would need sites where environment and space are decoupled. Pairs of reaches close
together but environmentally different, such as a spring-fed tributary entering a warm
main stem. Pairs far apart but environmentally alike, such as two headwaters in
different catchments. Given such pairs, the two hypotheses finally make different
predictions, and the data can choose between them.

Several river systems can supply contrasts that one transect lacks. Those contrasts
depend on the sampling design established before data collection.
:::


## Worked Example 2: The Mite Data When Design Permits the Question {#sec-mite}

A more informative dataset contains sites where environmental similarity and geographic
proximity vary partly independently. The Doubs cannot provide such contrasts along a
single linear river. A two-dimensional layout can include nearby sites with different
environments and distant sites with similar conditions.

The `mite` dataset in `vegan` provides these contrasts. It records oribatid mite
communities in 70 soil cores from a Quebec peat bog [@borcard2011]. The 2.5 by 10 metre
sampling area is elongated but has enough lateral spread for nearby sites to differ in
substrate and water content. Substrate density (`SubsDens`) is almost independent of
spatial position, giving at least one environmental axis that does not covary with
geography.

The data come as three objects in `vegan`, exactly parallel to the Doubs: a species
matrix (`mite`, 70 sites by 35 species), an environmental table (`mite.env`, five
variables including two continuous and three categorical), and a spatial coordinate
table (`mite.xy`).

```{r code-mite-load}
data(mite)
data(mite.env)
data(mite.xy)
```

### The design comparison

We first measure the confounding seen in the Doubs. There, environmental and
along-channel distance had a Mantel correlation of about $r = 0.50$, so the predictors
carried substantial shared information.

We use the two continuous environmental variables, `SubsDens` (substrate density) and
`WatrCont` (water content), standardised and measured as Euclidean distance,
paralleling the Doubs approach.

```{r code-mite-confound}
mite_env_cont <- mite.env[, c("SubsDens", "WatrCont")]
mite_env_dist <- dist(decostand(mite_env_cont, method = "standardize"))
mite_spa_dist <- dist(mite.xy)
mite_spe_bray <- vegdist(mite, method = "bray")

# The design question: how correlated are the two explanatory matrices?
mantel(mite_env_dist, mite_spa_dist)
```

```{r fig-mite-confound}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 2.8
#| fig-cap: "Environmental distance versus spatial distance for the mite data (left) and the Doubs (right, reproduced from the Doubs analysis). The mite data shows a weaker relationship: the two explanatory matrices are less tightly coupled, which is the design property the Doubs lacked."
p_mite <- tibble(
    spatial     = as.vector(mite_spa_dist),
    environment = as.vector(mite_env_dist)
) |>
    ggplot(aes(spatial, environment)) +
    geom_point(alpha = 0.1, size = 0.4) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.4, formula = y ~ x) +
    labs(
        x = "Spatial distance", y = "Environmental distance",
        title = "Mite data"
    ) +
    theme(plot.title = element_text(size = 10))

p_doubs <- tibble(
    spatial     = as.vector(chan_dist),
    environment = as.vector(env_dist)
) |>
    ggplot(aes(spatial, environment)) +
    geom_point(alpha = 0.2, size = 0.4) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.4, formula = y ~ x) +
    labs(
        x = "Along-channel distance (km)", y = "Environmental distance",
        title = "Doubs data (reproduced)"
    ) +
    theme(plot.title = element_text(size = 10))

patchwork::wrap_plots(p_mite, p_doubs, ncol = 2)
```

The Mantel correlation between environment and space drops from about $r = 0.50$ in the Doubs to approximately $r = 0.28$ in the mite data. That is a
substantial reduction, though not zero. One environmental variable, water content,
still correlates with the y-coordinate at $r = 0.67$, showing that complete
decoupling of environment and space is rare in observational studies. Substrate
density, however, is essentially independent of position. The two-dimensional layout
provides site-pairs that the Doubs, as a one-dimensional transect, could never
produce: pairs close in space but different in water content, and pairs far apart but
similar.

### Applying the framework

```{r code-mite-mantel}
mite_m_env <- mantel(mite_spe_bray, mite_env_dist)
mite_m_spa <- mantel(mite_spe_bray, mite_spa_dist)
mite_p_env <- mantel.partial(mite_spe_bray, mite_env_dist, mite_spa_dist)
mite_p_spa <- mantel.partial(mite_spe_bray, mite_spa_dist, mite_env_dist)

tibble(
    test = c(
        "community ~ environment",
        "community ~ space",
        "environment | space",
        "space | environment"
    ),
    mantel_r = c(
        mite_m_env$statistic, mite_m_spa$statistic,
        mite_p_env$statistic, mite_p_spa$statistic
    ),
    p_value = c(
        mite_m_env$signif, mite_m_spa$signif,
        mite_p_env$signif, mite_p_spa$signif
    )
) |>
    knitr::kable(
        digits = 3,
        caption = "Mantel and partial Mantel results for the mite data. Compare these with the Doubs table above. Both niche and neutral signals are present, and both survive controlling for the other."
    )
```

The comparison with the Doubs is informative.

```{r code-mite-comparison-table}
tibble(
    dataset = rep(c("Doubs", "Mite"), each = 4),
    test = rep(c(
        "community ~ env", "community ~ space",
        "env | space", "space | env"
    ), 2),
    mantel_r = round(c(
        m_env$statistic, m_chan$statistic,
        p_env$statistic, p_chan$statistic,
        mite_m_env$statistic, mite_m_spa$statistic,
        mite_p_env$statistic, mite_p_spa$statistic
    ), 3)
) |>
    knitr::kable(
        digits = 3,
        caption = "Side-by-side comparison of the Doubs and mite datasets. The Doubs has a higher env-space correlation (r approx 0.50 vs 0.28) and stronger signals overall. In the mite data, both signals survive the partial Mantel, consistent with both processes operating. The Doubs design could not produce this result cleanly."
    )
```

The Doubs produced similar niche and neutral signals from strongly confounded predictors.
The weaker environment-space correlation in the mite data makes the partial results
more informative, although the Type I error problem remains. Community dissimilarity
tracks environmental distance after accounting for space, and spatial distance after
accounting for environment. @cottenie2005 reported this combination in most of the 158
datasets they reviewed.

### The correlogram

```{r fig-mite-correlogram}
#| fig-height: 3.0
#| fig-cap: "Mantel correlogram of the mite community against spatial distance. The pattern is weaker than the Doubs correlogram (compare with the Doubs result above): distance decay is present but the range over which it operates is shorter, consistent with a two-dimensional sampling layout where dispersal limitation constrains less than in a linear river."
mite_correlog <- mantel.correlog(
    D.eco  = mite_spe_bray,
    D.geo  = mite_spa_dist,
    nperm  = 999,
    r.type = "pearson",
    cutoff = FALSE,
    mult   = "holm"
)

plot(mite_correlog)
```

Distance decay is present in the mite data, but it is weaker than in the Doubs. Mean
Bray-Curtis dissimilarity rises from approximately $0.49$ between the closest
site-pairs to approximately $0.67$ between the most distant, a range of $0.18$
compared to the Doubs' range of $0.49$. The correlogram shows positive correlations
at short distances and negative at larger ones, but the transition is less orderly
and the effect sizes are smaller.

The geometry offers a biological explanation. Each site connects to neighbours in
several directions, so dispersal is less constrained than in a river channel. Distance
decay associated with dispersal limitation is correspondingly weaker.

### Partitioning the signals with varpart and dbMEM

The Mantel and partial Mantel results showed that both environmental and spatial
signals survive controlling for each other. But Mantel statistics are not the
correct tool for quantifying how much variance each process explains (see the
critique in the Analytical Logic section). Variation partitioning with distance-based
Moran's eigenvector maps (dbMEM) is.

`pcnm()` in vegan computes PCNM eigenvectors, the dbMEM equivalent available
in base vegan, from the site coordinate matrix. Each eigenvector describes a
spatial pattern at a specific scale. Low-numbered axes capture broad gradients and
higher axes capture finer patches. Using positive-eigenvalue vectors as predictors in
`varpart()` divides the Hellinger-transformed community variance into environmental,
shared, spatial and unexplained fractions.

```{r code-mite-varpart}
#| cache: true
# Hellinger transformation reduces the influence of very abundant species
mite_hel <- decostand(mite, method = "hellinger")

# dbMEM spatial eigenvectors from the 70 site coordinates
mite_pcnm <- pcnm(dist(mite.xy))
pos_axes <- which(mite_pcnm$values > 0)
mite_mem <- scores(mite_pcnm, choices = pos_axes[1:min(8, length(pos_axes))])

# Variation partitioning: environment vs space
mite_vp <- varpart(
    mite_hel,
    mite.env[, c("SubsDens", "WatrCont")],
    mite_mem
)
mite_vp
```

```{r fig-mite-varpart}
#| fig-height: 4.5
#| fig-width: 6
#| fig-cap: "Variation partitioning of the mite community into purely environmental (niche signal, fraction [a]), purely spatial (neutral signal, fraction [c]), their shared component [b], and unexplained variation [d]. The two predictors together explain roughly a quarter of community variance. Environment explains more unique variation than space, while the spatial fraction confirms an independent contribution from dispersal limitation."
plot(mite_vp,
    digits = 2,
    Xnames = c("Environment\n(SubsDens+WatrCont)", "Space\n(dbMEM 1-8)"),
    bg = c("#1A6B7240", "#B85C0040"),
    id.size = 0.85
)
```

The Venn diagram answers the question this lecture has been asking since the
introduction: how much of community structure is attributable to niche filtering
versus dispersal limitation?

Fraction [a] is the purely environmental component: variance that environment
explains after removing what it shares with space. This is the niche signal --
the fraction that would not exist if all species were ecologically equivalent.
Fraction [c] is the purely spatial component: variance that space explains
after removing the environmental contribution. This is the neutral signal --
the fraction attributable to dispersal limitation independent of habitat.
Fraction [b] is the shared component, which cannot be attributed to either
process without further data: it is the part of the pattern that both environment
and space could explain, which the Doubs analysis showed is large when the design
does not decouple the two predictors.

For the mite data, the purely environmental fraction is larger than the purely
spatial fraction. That is consistent with the synthesis position across the
literature: environmental structuring tends to dominate over dispersal limitation
in most empirical datasets, with both operating simultaneously [@cottenie2005].
The unexplained fraction is large, as it routinely is in community ecology, because
species responses to environment are nonlinear, measurement error is present, and
stochastic demographic events that no model captures contribute to the pattern.

### What the comparison teaches

::: {.callout-tip}
## Design Determines What the Analysis Can Show

The same framework gives different answers because the sampling designs differ. Along
the Doubs, environment and space change together and their signals cannot be separated.
The two-dimensional mite layout partly decouples them, so each retains an independent
association with community composition. The contrast shows how sampling geometry limits
the questions an analysis can answer.
:::

## A Simulation in a Realistic Landscape: Using spesim {#sec-spesim-demo}

The Two Worlds and continuum simulations use multinomial sampling on a 29-site transect.
They clarify the theory, but they give the landscape no geometry and simplify dispersal.
Prof. Smit's `spesim` package provides a two-dimensional spatial simulation that connects
these ideas to realistic sampling designs.

### What spesim does

`spesim` follows five stages. The user defines an irregular polygon such as a field site,
catchment or reserve, adds continuous environmental gradients, chooses a niche, neutral
or hybrid assembly process, places quadrats under one of five sampling schemes, and
extracts matched site-by-species and environmental tables. The reshaping helpers
`create_abundance_matrix()` and `calculate_quadrat_environment()` produce data ready for
`vegan`, including constrained ordination with `capscale()`, variation partitioning with
`varpart()` and Mantel correlograms.

`HYBRID_ENV_WEIGHT` parallels the mixing weight $w$ in @sec-continuum. A value of 0
specifies neutral assembly, 1 specifies niche filtering, and intermediate values combine
them in known proportions. `spesim` places that continuum within a two-dimensional
domain.

### Why this matters for the lecture's argument

The generating process is unknown in field data. We cannot rerun the Doubs with a
different seed or directly observe how much of its gradient came from filtering,
dispersal and drift.

In `spesim`, the generating process is specified before the community is produced.
Mantel correlograms, constrained ordination and variation partitioning can then be
tested against known environmental and spatial contributions. This allows a direct
check of whether a method recovers the imposed signal.

The package's `generate_full_report()` function answers this directly. It
independently computes alpha, beta, and gamma diversity, a Mantel test for spatial
autocorrelation, and a goodness-of-fit check for the specified species-abundance
distribution, and it produces a **conceptual audit** that reports, per species, whether
the environmental filtering you asked for actually appeared in the simulated community:

```text
Conceptual audit (did you get the regime you asked for?):
  Environmental filtering (Spearman corr of -|z| with abundance):
    A (temperature): rho = 0.73  | abundance_peaks_near_optimum
    D (rainfall):    rho = 0.16  | weak_or_no_signal
```

This conceptual audit evaluates method sensitivity against a known result. Field data
cannot provide that benchmark.

### A starting point for exploration

The code below shows the complete workflow for niche and neutral simulations.
The `eval: false` setting keeps it out of the site build. Every line is executable once
`spesim` is installed with
`remotes::install_github("ajsmit/spesim")`.

```{r code-spesim-example, eval=FALSE}
#| code-fold: true
#| code-summary: "Show the spesim workflow code"
library(spesim)

# Load the built-in example configuration
P <- load_config(system.file("examples/spesim_init_basic.txt",
    package = "spesim"
))
P$N_INDIVIDUALS <- 3000
P$N_QUADRATS <- 40
P$N_SPECIES <- 15

# ── NICHE WORLD ──────────────────────────────────────────────────────────────
P_niche <- P
P_niche$MODEL_FAMILY <- "manual"
P_niche$SAD_MODEL <- "fisher"
res_niche <- spesim_run(P_niche, write_outputs = FALSE, seed = 42)

# ── NEUTRAL WORLD ─────────────────────────────────────────────────────────────
P_neutral <- P
P_neutral$MODEL_FAMILY <- "neutral_hubbell_like"
P_neutral$DISPERSAL_SCALE <- 0.3
res_neutral <- spesim_run(P_neutral, write_outputs = FALSE, seed = 42)

# ── HYBRID: the continuum (equivalent to mixing weight w = 0.5) ───────────────
P_hybrid <- P
P_hybrid$MODEL_FAMILY <- "hybrid"
P_hybrid$HYBRID_ENV_WEIGHT <- 0.5
res_hybrid <- spesim_run(P_hybrid, write_outputs = FALSE, seed = 42)

# ── APPLY VEGAN METHODS ───────────────────────────────────────────────────────
# spesim ships reshaping helpers that produce vegan-ready matrices
abund_niche <- create_abundance_matrix(res_niche)
env_niche <- calculate_quadrat_environment(res_niche)

# Constrained ordination: did CAP1 recover the imposed gradient?
abund_hel <- decostand(abund_niche, method = "hellinger")
cap_niche <- capscale(abund_hel ~ temperature,
    data = env_niche,
    distance = "bray"
)

# Variation partitioning: how much variance is environmental vs spatial?
pcnm_coords <- pcnm(dist(env_niche[, c("x", "y")]))
vp <- varpart(
    abund_hel,
    env_niche[, "temperature", drop = FALSE],
    scores(pcnm_coords)
)

# Diagnostic report with conceptual audit
res_niche_report <- spesim_run(P_niche, write_outputs = TRUE, seed = 42)
```

### Connecting to the lecture's analytical workflow

The table links `spesim` settings to the concepts in this lecture and the corresponding
BCB743 methods.

| spesim capability | Lecture concept | BCB743 method |
|:--|:--|:--|
| `MODEL_FAMILY = "manual"` | Niche world: environmental filtering | Chapters 4, 10, 11 (ordination) |
| `MODEL_FAMILY = "neutral_hubbell_like"` | Neutral world: drift + dispersal | Mantel correlogram |
| `HYBRID_ENV_WEIGHT = w` | The continuum (@sec-continuum) | varpart(), dbMEM |
| `generate_full_report()` | "Did the method recover what you specified?" | Chapter 18 (model building) |
| Five quadrat schemes | Study design (@sec-design) | Chapter 2 (sampling) |

Each parameter has an ecological interpretation developed earlier in the lecture.
`HYBRID_ENV_WEIGHT` is $w$, the neutral model combines demographic equivalence with
dispersal limitation, and the niche model uses Gaussian filtering. The `varpart()` call
produces the same partition introduced in @sec-mite, now applied where the true
environmental and spatial contributions are known.

::: {.callout-tip}
## For Further Exploration

Run the code block above after installing spesim, then vary `HYBRID_ENV_WEIGHT`
from 0 to 1 and observe how the `varpart()` environmental fraction changes. Compare
the result with the continuum ribbon in Figure 4: as `w` increases, the environmental
fraction in the Venn diagram should grow in the same direction. This is the same
theoretical claim expressed in two different analytical languages.
:::

## Designing a Study That Can Answer the Question {#sec-design}

The Doubs design could not separate the processes, while the mite design produced a
partial answer. A better field study must create the environmental and spatial contrasts
that the Doubs lacks.

### The geometry problem

A linear transect is poorly suited to this question. Along rivers and elevational
gradients, environmental conditions change with position. Temperature falls with
altitude, discharge changes downstream and canopy cover may track disturbance. Sites
arranged along one line therefore tend to produce collinear predictors that statistical
methods cannot separate.

A two-dimensional sampling domain supplies contrasts along two axes. A grid or dispersed
layout can include geographically close sites with different environments, as well as
distant sites with similar conditions. These pairs distinguish the predictions. Under
niche structure, dissimilarity should follow environmental difference. Under neutral
structure, it should follow geographic separation. A dataset containing both kinds of
pair can estimate their relative contributions.

The 2.5 by 10 metre mite layout has this property. The Doubs follows a single river from
headwater to lowland and does not. Sampling geometry explains the difference between
the analyses because the statistical framework is the same.

### What to measure and what not to

Classify candidate predictors before fieldwork. Water chemistry, temperature, soil
texture, light and prey density describe local conditions. Distance from the coast,
altitude, distance from a forest edge and position within a plot describe location.
Placing a positional proxy in the environmental matrix adds spatial information and
inflates the apparent environmental signal.

Measure conditions the organism experiences, then check how strongly they correlate
with the spatial coordinates. High correlations across most variables reproduce the
Doubs problem and limit any later partitioning.

As a diagnostic, calculate the Mantel correlation between environmental and spatial
distance before analysing the community. A value above about 0.4 warns that clean
separation will be difficult.

### Diagnostic site-pairs

Include diagnostic pairs when placing sites. Nearby pairs should sometimes differ in
environment, such as quadrats on the sunny and shaded faces of one boulder or dry and
saturated substrate within the same bog section. Other pairs should be far apart yet
environmentally similar, such as two bog pools with comparable chemistry and substrate
density at opposite ends of the study area.

If your dataset has only close sites that are also environmentally similar and distant
sites that are also environmentally different, you are back at the Doubs problem. The
diagnostic pairs break the collinearity by providing observations where environment
and space point in different directions.

### Scale and replication

There is no universal minimum number of sites for this question, but published
analyses suggest that fewer than 30 sites produce unstable Mantel correlogram
estimates, particularly in the outer distance classes where few pairs exist
[@legendre2015]. The mite analysis uses 70 sites, which gives reasonable stability
across all distance classes. Borcard and Legendre's analyses of this dataset show
that the spatial signal is detectable but weaker than the environmental signal -- a
result that would likely not be recoverable at 20 sites.

For any study where you plan to use variation partitioning with dbMEM spatial
eigenvectors, the effective degrees of freedom are consumed faster than in ordinary
regression. Plan generous replication and check that every distance class contains
enough site pairs.

### Asking the question before you collect

The field design determines which assembly questions can be answered. Spatial layout,
environmental measurements and sampling scale decide whether the predictors can be
separated. Later statistical work cannot create contrasts that were absent from the
sampling.

::: {.callout-tip}
## Sampling Design Sets the Limit

Site placement sets the limits of the later analysis. The Doubs and mite data gave
different results under the same methods because their sampling geometries differed.
A two-dimensional arrangement with diagnostic pairs and local environmental measures
provides the contrasts needed to estimate environmental and spatial contributions.
:::

## Synthesis

Earlier ordination and clustering analyses established an orderly change in the Doubs
fish community from headwater to mouth. This lecture asked what produced that gradient.

Community dissimilarity increases with along-channel distance, steadily and
substantially, from a mean Bray-Curtis dissimilarity of $0.42$ between neighbouring
reaches to $0.91$ between the extremes of the river. That distance decay is exactly the
signature neutral theory predicts from dispersal limitation. Community dissimilarity
also increases with environmental distance, at almost the same strength. That is
exactly the signature niche theory predicts from environmental filtering. Both
predictions are met, and the data cannot say which mechanism is responsible, because
along a single river the environment changes as you travel and the travelling changes
the environment. The predictors are confounded by the geometry of the system.

The inconclusive separation is itself a result. Community structure is strongly patterned
in both space and environment, but the design cannot assign the shared pattern to one
process. A better study would include sites where environment and distance vary
independently. Partial Mantel statistics do not solve the problem because @sec-logic
measured false-positive rates between one in eight and one in four under relevant
spatial autocorrelation.

Across 158 published datasets, @cottenie2005 reports that both environmental and spatial
processes leave detectable signatures in most metacommunities, with environmental
structuring the more common of the two. @tuomisto2003, working across western Amazonian
forests where soils and distance can be decoupled, likewise find both processes at
work. The Doubs fits that broader picture but adds no independent evidence for it.

::: {.callout-tip}
## Limits on Inference

Observational data narrow the plausible assembly processes, and several may remain.
@adler2007 show why abundance patterns alone cannot resolve the niche-neutral debate:
rival models generate similar observations. Report that limit and design the next study
to supply the missing contrasts.
:::

You can now name two process explanations for a community gradient, state their
assumptions about species and derive their predictions for community dissimilarity. When the
[Seaweeds](https://tangledbank.netlify.app/BCB743/two_oceans_appendices.html) appendix labels a fraction of variation "spatial",
or when the Gradients chapter partitions variance into environmental and spatial
components, you know what biological claim those labels encode, and you know that the
labels carry an inference about assembly.

Before running a test, ask whether the design can answer the question and identify the
null hypothesis the procedure evaluates. Those checks exposed the limits of the Mantel
analysis before any elaborate modelling was needed.

The advanced [BCB743 Model Building chapter](https://tangledbank.netlify.app/BCB743/model_building.html) asks you to choose predictors
from hypothesised mechanisms. Environmental filtering and dispersal limitation provide
two such mechanisms. A model that includes environmental predictors while omitting
space already makes a claim about community assembly, even when the author does not
state it. Chapter 18 shows how to defend that modelling choice.

## Further Reading

Start with @hubbell2001 for the primary statement of neutral theory. The early chapters
develop and defend the demographic equivalence assumption in full.

Ten years later, @rosindell2011 assessed which predictions survived empirical testing.
It is a useful corrective after Hubbell.

For a modern operational treatment of the niche, @chase2011 connect community assembly
to the scale dependence of stochasticity.

@adler2007 recast the niche-neutral debate through coexistence theory in a short paper.
If you choose one item from this list, read this one.

Selection, drift, speciation and dispersal organise the broader synthesis in
@vellend2010. The essay develops the analogy with population genetics carefully.

The methodological argument behind the second half of this lecture comes from
@legendre2015. Read it before using `mantel()` for spatial ecological data.

The 158 datasets synthesised by @cottenie2005 provide the empirical comparison of
environmental and spatial contributions.

Finally, @leibold2006 explain how ecological equivalence and niche differentiation can
coexist within one community. Read it with Adler for a fuller account of the continuum.

## Tasks {#sec-tasks}

These exercises progress from re-running the lecture's core analyses to designing
and interpreting new ones. They are intended to be worked through in order, because
later tasks build on results from earlier ones.

---

**Task 1. Does the metric choice matter in your system?**

Run the Doubs Mantel test three times using three different spatial distance matrices:
`chan_dist` (along-channel distance, as used in the lecture), `dist(spa)` (Euclidean
distance between site coordinates), and `dist(log1p(env$dfs))` (log-transformed
channel distance, on the reasoning that dispersal probability may decay logarithmically
with distance).

For each metric, compute `mantel(spe_bray, spatial_dist)` and record the Mantel `r`.
Then write two or three sentences explaining what the range across the three results
tells you about the relationship between your modelling choices and your conclusions.

---

**Task 2. Extending the Type I error simulation**

The partial Mantel simulation in the lecture used a fixed autocorrelation range and
iterated 1,000 times. Extend it in one of the following directions:

(a) Vary the autocorrelation range from 1 to 20 sites and plot the false-positive rate
as a continuous function rather than at four discrete values. Does the relationship
appear to be linear, or does it plateau?

(b) Run the same simulation but replace the partial Mantel test with a dbMEM-based
test: generate spatially autocorrelated `x` and `y` variables as before, but use
`anova(rda(x, pcnm(dist(sites))$vectors))` to ask whether spatial eigenvectors predict
`x`. Record the false-positive rate at each autocorrelation range. Compare the two
tests.

---

**Task 3. Apply the continuum simulation to a field question**

The continuum simulation varies the mixing weight `w` from 0 to 1 and shows that
gradient strength increases and becomes more reproducible as `w` increases.

Choose a real ecological system you are familiar with (any organism, any habitat) and
answer the following:

(a) Where on the `w` continuum do you expect this system to sit, and why? Your answer
should cite at least one published study that provides evidence for the niche or
neutral signal in this type of system.

(b) What design would you use to estimate `w` empirically? Specifically: how many
sites, what spatial arrangement, what environmental variables, and why those variables
rather than others?

---

**Task 4. Interpreting the mite varpart output**

Re-run the mite variation partitioning using different numbers of dbMEM axes: 3, 8
(as in the lecture), and 20. For each run, record fractions [a], [b], [c], and [d]
from the varpart output.

(a) How stable is fraction [a] (the niche signal) as the number of spatial axes
increases? How stable is fraction [c] (the neutral signal)?

(b) Fraction [d] (unexplained) is always large. List three ecological mechanisms that
contribute to unexplained variation in a community dataset of this kind, and for each
one, note whether more sampling sites, better environmental variables, or a different
analytical method would reduce it.

---

**Task 5. Design a study to separate niche from neutral**

You are planning a field study to test whether the mite community in a new bog system
is primarily structured by environmental filtering or by dispersal limitation. You have
budget for 50 sampling quadrats.

(a) Draw a sketch of your proposed sampling layout. Explain why you chose that
arrangement rather than a transect, and identify which site-pairs in your layout would
serve as "diagnostic pairs" -- pairs that would discriminate between the two hypotheses.

(b) List the environmental variables you would measure at each quadrat. For each
variable, state whether it is a genuine local condition or a positional proxy (refer to
the distinction in @sec-design), and explain what would happen to the partial Mantel
test if you accidentally included a positional proxy in your environmental matrix.

(c) Before collecting data, compute the Mantel correlation between your planned spatial
distance matrix (based on your layout) and a hypothetical environmental distance matrix
where every variable changes linearly across the study area. What does this number tell
you about the power of your design to separate the two processes?

## Author Contributions and AI Disclosure

**Author contributions:** All three group members contributed to the theoretical
framing and the written text. Alex Matthew developed the Two Worlds simulation,
the niche-neutral continuum simulation, and the Doubs worked analysis including
the Type I error simulation. He also led the integration of the spesim demonstration
and the mite worked example. Keanan Jarvis developed the analytical logic section,
the Mantel test critique, and the critical reading of Legendre et al. (2015).
Ethan Bell developed the synthesis section, the Designing a Study section,
and the Tasks. All members revised and approved the final submission.

**AI use disclosure:** Generative AI tools (specifically Claude by Anthropic) were
used in the preparation of this lecture. AI assisted with: drafting and restructuring the
document, suggesting R code structures for the simulations and worked examples. All AI-generated text was reviewed, revised, and in
many cases substantially rewritten by the authors. All code was tested and verified
on real data by the authors. All citations were checked against the source PDFs.
No AI-generated content was accepted without independent verification.

## References

::: {#refs}
:::

<!--
INTERNAL NOTE - not rendered. Retained for group continuity.

PROSE STYLE, 2026-07-10. The whole chapter was rewritten against
WRITING_STYLE_RULES.txt (in the project folder), derived from Wikipedia's
"Signs of AI writing". Audit before and after:
    em dashes in prose        68  ->  0
    negative parallelism       8  ->  0   ("it is not X; it is Y")
    editorialising             9  ->  0   ("it is worth noting", "read that again")
    puffery                    0  ->  0
    AI vocabulary              1  ->  0
    tailing participles        0  ->  0
    bold                      22  -> 11   (defined technical terms, first use only)
Re-run section H of WRITING_STYLE_RULES.txt on any new prose before submitting.
Title Case headings are KEPT: Smit uses them, and his house style outranks the
generic tell.

DESIGN CHANGE, 2026-07-09. The analytical spine was revised after reading Legendre,
Fortin & Borcard (2015) in full. That paper states that using Mantel and partial
Mantel tests to relate community composition to environment while controlling for
space is an incorrect use of the test. Legendre and Borcard are the authors of both
compulsory texts for this module. The chapter now teaches the Mantel test as the
intuitive-but-wrong approach, has students reproduce its inflated Type I error by
simulation, and uses the Mantel correlogram (endorsed in that same paper) as the
defensible test of distance decay. The niche-side question is handed to
dbMEM/varpart, cross-referenced to the Gradients chapter.

NUMBERS IN THE PROSE - all verified 2026-07-09 against the genuine Doubs data
(ade4::doubs, back-transformed to NEwR conventions; the correlation matrix
reproduces ch.5 exactly, including corr(dfs,slo) = -0.38 and corr(ele,slo) = +0.44):
    mantel(env_dist, chan_dist)               r = 0.496
    mantel(env_dist_WITH_dfs, chan_dist)      r = 0.575   <- the leak
    mantel(spe_bray, env_dist)                r = 0.574
    mantel(spe_bray, chan_dist)               r = 0.597
    mantel(spe_bray, euc_dist)                r = 0.300   <- metric choice halves it
    mantel.partial(spe_bray, env_dist, chan)  r = 0.400
    mantel.partial(spe_bray, chan, env_dist)  r = 0.439
    mean Bray-Curtis, nearest class           0.419
    mean Bray-Curtis, furthest class          0.908
    Type I error of partial Mantel, ranges 2 / 5 / 10:  0.12 / 0.19 / 0.25
    (simple Mantel is worse still: 0.13 / 0.26 / 0.37)
Re-verify on render; p-values will differ trivially with the RNG seed.

CORRECTION, 2026-07-10. An earlier version of this note claimed that Smit uses
mantel() nowhere in BCB743. That is now FALSE. Chapter 24, "A Complete Multivariate
Analysis of the Doubs River", published 2026/07/09, contains a section "Comparing
whole matrices with Mantel tests", plus varpart() and dbmem() sections, on the Doubs.
UNRESOLVED AND IMPORTANT: nobody has yet read what he says inside that Mantel
section. If he flags the Legendre critique himself, our contribution is confirmation
rather than correction, and the framing must change. Check before submitting.

SITE RENUMBERED, 2026-07-10. Nav now runs 1-24. Cluster Analysis = 17 (was 14),
Model Building = 18 (was 15), Ordination = 4. Chapter title changed 14b -> 17b and
mermaid node labels updated. The .qmd filenames are unchanged so cross-reference
links still resolve. Insertion point remains Smit's call.

FIGURES, 2026-07-10. Five added, all verified against real data or real simulation
before being written:
  fig-two-worlds        niche vs neutral simulation, 3 runs each. VERIFIED: niche
                        gives |r| PCoA1~position = 0.947 / 0.951 / 0.951; neutral
                        gives 0.923 / 0.331 / 0.062. A single neutral run mimics a
                        niche gradient exactly. Reproducibility, not pattern, is the
                        discriminator. This is our own result.
  fig-two-worlds-decay  both worlds show distance decay every run.
  fig-doubs-geometry    site map from spa + euclid vs channel scatter, r = 0.643.
                        Sites 19 & 20: 8.2 planar units apart, 33.5 km along channel.
                        Smit's ch.24 explicitly does NOT plot spa, so no overlap.
  fig-confound          env_dist vs chan_dist, Mantel r = 0.496.
  fig-type1             false-positive rate vs autocorrelation range, 5% reference.
New package dependency: patchwork (for fig-doubs-geometry).
Both simulation chunks use cache: true. Expect a minute or two on first render.

OWN-SYNTHESIS CLAIMS, flag for Smit - our reasoning, not the literature:
  1. dfs is moved out of env and treated as spatial. Justified in prose. Check
     whether Borcard et al. (2011) do the same in the NEwR Doubs analyses.
  2. The confounding argument (env and space are collinear in a linear river, so the
     design cannot separate the hypotheses) is ours.
  3. The Type I error simulation is our own replication of results reviewed in
     Legendre et al. (2015).
  4. The two-worlds simulation and its conclusion (that reproducibility across
     replicates, not pattern within one dataset, separates niche from neutral) is
     ours. spesim, Smit's own package, has a neutral ZSM sampler and Gaussian
     environmental filtering and could replace our hand-rolled simulation. It
     installs from GitHub, so it is not evaluated at site build; any spesim chunk
     would need eval: false. Worth considering, and it ties to Task A3.

OUTSTANDING:
  - Data paths are relative to the .qmd, not the RStudio working directory. Either
    move the .qmd into BCB744_instructions/ or switch to here::here().
  - Check names(spe) and names(env) for a leading index column from the CSV export.
    env must have exactly 11 columns, spe 27, spa 2.
  - Decide first person: Smit writes "I"; this chapter uses "we".
  - Epigraph is a placeholder.
  - Task page does not yet exist.
  - RESOLVED: Oden & Sokal (1992) and Guillot & Rousset (2013) acquired and now
    cited directly with their own numbers in the Mantel critique passage.
  - RESOLVED: Leibold & McPeek (2006) acquired and cited in the Synthesis section
    and in Further Reading.
  - RESOLVED: Tuomisto et al. (2003) acquired (science_1078037.pdf). Was already
    cited; the gap was the missing PDF, not a missing citation.
  - Borcard & Legendre (2002) acquired but deliberately NOT cited. Content is
    covered by Borcard, Gillet & Legendre (2011).
  - MacArthur & Wilson (1967) not acquired. Not cited in the chapter.
  - The simulation chunk uses cache: true. Runtime roughly a minute.
  - Tasks section added with five conceptual exercises.
  - NEW 2026-07-19: Three additions integrated:
    (a) Niche-neutral continuum simulation (sec-continuum): mixing parameter w
        dials between pure neutral and pure niche. Two figures. Goes after Two
        Worlds and before the Debate. cache: true. No new packages.
    (b) Mite worked example (sec-mite): 70-site oribatid mite dataset from vegan.
        env-space Mantel r = 0.284 (vs Doubs 0.496). Both signals survive partial
        Mantel. Comparison table uses live R objects from Doubs analysis. Goes
        before Tasks. No new packages.
    (c) spesim connection: eval: false code chunk linking to Smit's spesim package
        on GitHub. Teaching note: our chapter's theory is the conceptual foundation
        for what spesim implements computationally. Goes in Further Reading.
    Variable naming: mite Mantel objects use mite_m_env, mite_m_spa, mite_p_env,
    mite_p_spa to avoid overwriting Doubs objects m_env, m_chan, p_env, p_chan. No separate task .qmd file
    yet; exercises are embedded in the chapter for now.
-->
