---
date: last-modified
title: "17b: Why Are Species Where They Are? Niche and Neutral Theories of Community Assembly"
author:
  - name: "Alex Matthew"
    affiliation: "University of the Western Cape"
  - name: "Keanan Jarvis"
    affiliation: "University of the Western Cape"
  - name: "Ethan Bell"
    affiliation: "University of the Western Cape"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    toc-title: "On this page"
    code-fold: false
    code-tools: true
    code-link: false
    embed-resources: true
bibliography: references.bib
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 4.5,
  fig.height = 2.625,
  out.width = "75%",
  fig.asp = NULL, # control via width/height
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

::: {.callout-tip}
## Material Required for This Chapter

| Type | Name | Notes |
| :--- | :--- | :--- |
| **Theory** | Hubbell (2001) *The Unified Neutral Theory of Biodiversity and Biogeography* | Primary statement of neutral theory |
| **Theory** | Hutchinson (1957) Concluding remarks | The niche concept defined |
| **Theory** | Adler et al. (2007) A niche for neutrality | The niche-neutral continuum |
| **Theory** | Cottenie (2005) Integrating environmental and spatial processes | Metacommunity framework |
| **Theory** | Legendre et al. (2015) Should the Mantel test be used in spatial analysis? | Mantel test critique |
| **Data** | Doubs River fish community | `Doubs.RData` via NEwR companion data |
| **Data** | Oribatid mite community | `vegan::mite`, `vegan::mite.env`, `vegan::mite.xy` |
| **Package** | spesim 0.5.2 | `remotes::install_github("ajsmit/spesim")` |
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

See the [Tasks](#sec-tasks) section at the end of this chapter for five exercises that build directly on the analyses presented here.
:::

> "*The truth is rarely pure and never simple.*"
>
> Oscar Wilde



## Introduction

[Cluster Analysis](https://tangledbank.netlify.app/BCB743/cluster_analysis.html) left a question open. The Doubs fish
community forms a real, measurable gradient running from headwater to lowland, and the
silhouette analysis showed that forcing it into discrete clusters loses information a
continuous ordination preserves. But naming a gradient is not explaining one. *Why*
does the community change in an orderly way along the river, rather than scattering
species at random? This chapter takes up that question.

::: {.callout-tip}
## Main Idea: Pattern Is Not Process

A gradient in community composition is a pattern. Niche and neutral theory are rival
explanations for the process that produced it. In general a single pattern is
consistent with several processes, which is why the question cannot be settled by
ordination or clustering alone, however well executed. It requires a different kind of
evidence.
:::

Ecological theory offers two long-standing answers. **Niche theory** holds that
species occur where the local environment and biotic interactions permit them to
persist, so that community structure follows deterministically from the ecological
differences between species. **Neutral theory** holds that at the level of individual
births, deaths and dispersal events, species are functionally interchangeable, and
that the patterns we observe arise from stochastic demographic and dispersal processes
rather than from niche differences at all. Both theories can produce a gradient which,
at first glance look identical.

By the end of this chapter you should be able to:

- state the core assumption each theory makes about species, and explain why those
  assumptions lead to different predictions about community structure;
- explain why this remains an active area of debate, and identify what is empirically
  established as against what is contested;
- translate each theory into a falsifiable prediction about how community
  dissimilarity should behave across space;
- recognise why the most intuitive statistical test of those predictions is the wrong
  one, and explain the reasoning behind that judgement;
- apply a defensible alternative to the Doubs data and interpret it critically; and
- connect this theoretical lens to the model-building workflow introduced in the
  [next chapter](https://tangledbank.netlify.app/BCB743/model_building.html).

## From Pattern to Process: Revisiting the Doubs Gradient

The [Cluster Analysis](https://tangledbank.netlify.app/BCB743/cluster_analysis.html) chapter established, with a cophenetic
correlation and a silhouette analysis, that the Doubs fish community is organised as a
continuum rather than a set of discrete types. The [Ordination](https://tangledbank.netlify.app/BCB743/ordination.html)
chapter, several chapters earlier, established the same thing from the opposite
direction: the dominant axis recovers an upstream-to-downstream gradient, with
cool-water species at one end and lowland species at the other. Every method applied to
this dataset agrees that a gradient exists. None of them asks what generated it.

There is an obvious candidate answer, and it happens to be the answer niche theory
formalises. The river changes as it flows, in temperature, oxygen, gradient and flow
rate, and different fish species tolerate that changing environment differently. Trout
persist near the cool, oxygenated headwaters and cannot survive the warmer, slower
lowland reaches; the reverse holds for the lowland species. On this account the
gradient in the species data follows directly from the gradient in the environmental
data, filtered through each of the species physiological and ecological tolerances.

That is a reasonable hypothesis. It is also not the only one available. A river is a
linear system, and linear systems have a second property capable of producing the same
pattern in an ordination: every site connects to its neighbours through a fixed,
ordered sequence of dispersal pathways. Suppose every fish species were ecologically
identical. The bare fact that an individual at one site can reach adjacent sites easily
and distant ones only with difficulty would still, over time, produce spatial structure
in community composition. This is the situation neutral theory describes formally, and
it makes a testable claim: structure can arise from dispersal limitation and
demographic stochasticity alone, with no reference to ecological differences between
species.

The Doubs data cannot, by inspection, tell these two stories apart. That is the problem
this chapter exists to solve.

## Ecological Background: Niche Theory

Begin with a single fish. A brown trout holds its position in the cold, fast,
oxygen-rich water of the upper Doubs. It is not there by accident. Trout require high
dissolved oxygen and cool temperatures; they cannot persist in the warm, sluggish,
oxygen-poor water of the lowland reaches. If you move downstream and the trout disappears,
replaced by species that tolerate, and in some cases require, precisely the conditions
the trout cannot survive. The community changes along the river because the environment
changes along the river, and each species occupies the stretch where local conditions
fall inside the range it can tolerate. Niche theory formalises that intuition: species
occur where the environment permits them to persist, and so community composition can
in principle be read from the environment itself.

The concept received its lasting formal statement from @hutchinson1957, who defined the
niche as an *n*-dimensional hypervolume: a space with one axis for every environmental
factor that matters to a species, bounded by the limits within which that species can
maintain a viable population. Temperature is one axis, dissolved oxygen another, pH
another, prey availability another, for as many dimensions as are ecologically
relevant. @chase2011 restate the idea directly. Conditions inside the *N*-dimensional
hypervolume of a species' requirements define the space in which it can gather
resources, evade enemies and persist. This means that a species realised distribution is the
projection of that hypervolume onto real geography.

Hutchinson drew a distinction that still organizes how ecologists use the concept. The
**fundamental niche** is the full range of conditions under which a species could
persist in the absence of competitors and natural enemies. The **realised niche** is
the usually narrower range it actually occupies once biotic interactions are accounted
for. A species might be physiologically capable of living across a broad stretch of the
river, yet be confined to part of it because a competitor excludes it elsewhere. The
distinction matters here because it makes plain that the environment is not the only
filter: biotic interactions do some of the sorting too. The logic nevertheless remains
deterministic. Something about the match between a species' traits and its conditions
decides where it lives.

::: {.callout-tip}
## Main Idea: Environmental Filtering

Under niche theory the environment acts as a filter. At any site, only species whose
niche requirements are met by local conditions can persist; the rest are excluded. As
conditions change across space the filter changes, and with it the set of species that
pass through. Community composition becomes a predictable function of the environment.
:::

The mechanism doing this work is **environmental filtering**. @chase2011 place it right at
the centre of the deterministic view of communities, in which local, niche-based
processes such as environmental filtering, biotic interactions and interspecific
trade-offs largely determine which species are found where. The word to hold onto is
*Deterministic*: given sufficient knowledge of the environment and of species'
tolerances, niche theory says community composition is in principle predictable. Two
sites with similar environments should support similar communities; two sites with
different environments should support different ones. Under strict niche theory the
geographic distance between sites is irrelevant, except insofar as it happens to
correlate with environmental change.

That last point is what makes niche theory testable against its rival, and it is better
stated as a prediction than as a description.

::: {.callout-note}
## Prediction 1. The Niche Prediction

Community dissimilarity should track environmental distance. If two reaches of the
Doubs have nearly identical temperature, oxygen and flow, they should hold nearly the
same fish, whether they sit side by side or at opposite ends of the river. What drives
two communities apart is the degree to which their environments differ, not how far
apart they lie in space.
:::

The theory is strong, and strongly supported. In fish assemblages specifically, species
occupy distinct positions along river gradients in ways that track measurable niche
axes rather than chance, and the divergence of species' niches along a river continuum
has been documented directly [@troia2014]. But niche theory has a competitor, and the
competitor's claim is unsettling. The same orderly gradient could arise even if niche
differences played no part at all.

## Ecological Background: Neutral Theory

Return to the same river and change one assumption. Suppose the fish species of the
Doubs are not sorted by their tolerances. Suppose instead that any individual, of any
species, has roughly the same chance of surviving, reproducing and dispersing as any
other individual, whatever species it belongs to. On that assumption the trout occupies
the headwaters because its ancestors happened to be there and its descendants have not
yet drifted elsewhere, rather than because it is uniquely suited to them. Community
composition becomes a matter of history and chance instead of environmental matching.

This is the premise of **neutral theory**, given its unified formulation by @hubbell2001
and reviewed a decade later by @rosindell2011. Its foundational move is the neutrality
assumption: all individuals within a trophic level have the same chances of
reproduction and death, regardless of species identity [@rosindell2011]. Species are
treated as **demographically equivalent**, interchangeable at the level of the
individual. A community, on this view, is a collection of individuals undergoing birth,
death and dispersal at rates that do not depend on species identity.

This is easily mistaken for the claim that species are literally identical, so precision
is needed. Neutral theory does not assert that trout and bream are the same organism.
It asserts something narrower and stranger. For the purpose of explaining community
pattern, the differences between them can be set aside, because those differences do
not translate into differences in per capita demographic rates within the community.
The assumption is almost certainly false in detail. Real species do differ
demographically, and Hubbell knew it. The falsity is the point. Neutral theory
functions as a null model. If a theory that assumes away every niche difference can
still reproduce the patterns we see in real communities, then those patterns cannot by
themselves be taken as evidence that niche differences produced them.

::: {.callout-tip}
## Main Idea: Demographic Equivalence and Ecological Drift

Neutral theory assumes individuals are demographically equivalent regardless of
species. Community composition then changes through **ecological drift**, the random
walk of species' abundances under chance births and deaths, constrained by **dispersal
limitation**, the tendency of offspring to establish near their parents. Pattern
emerges from stochastic processes rather than from environmental sorting.
:::

Two mechanisms generate the pattern. The first is ecological drift. Because births and
deaths carry a random component, the relative abundances of species wander over time,
much as allele frequencies drift in population genetics. The parallel is deliberate.
@rosindell2011 trace the idea directly to Kimuras neutral theory of molecular
evolution, transposing neutrality from alleles within populations to individuals within
communities. Under drift alone some species wander to local extinction and others to
dominance, purely by chance, with no reference to fitness.

The second mechanism is dispersal limitation, and it is the one that produces spatial
structure. @rosindell2011 define it as a process that causes the location of an
individual to be restricted by the location of its parent: offspring tend to establish
near where they were produced. In a river this is concrete rather than abstract. A fish
spawning at one reach tends to produce offspring that occupy nearby reaches, not
reaches at the far end of the system. Over many generations dispersal limitation alone,
with no environmental sorting whatever, will generate spatial structure in community
composition, because the drift unfolding at one location is only weakly coupled to the
drift unfolding at a distant one. Nearby sites share recent dispersal history and come
to resemble one another. Distant sites do not.

This yields neutral theory's central prediction about spatial pattern, which mirrors
niche theory's exactly.

::: {.callout-note}
## Prediction 2. The Neutral Prediction

Community dissimilarity should track geographic distance. Nearby sites should hold
similar communities because individuals move readily between them, and distant sites
should hold dissimilar communities because they do not, whether or not their
environments are alike. What drives two communities apart is how far apart they sit in
space, not how much their environments differ.
:::

The distance-driven decline in similarity has a name. @rosindell2011 gloss
beta-diversity, or **distance decay**, as the probability, as a function of the
distance between two individuals, that they belong to the same species. Distance decay
sits at the centre of neutral theory rather than at its margin. It is the signature the
theory predicts, and it is the quantity this chapter will learn to measure.

::: {.callout-note}
Neutral theory is best read as a null model against which the signal of niche processes
can be measured, rather than as a claim that the world is neutral. If a dataset's
patterns can be reproduced without invoking niche differences, those patterns alone
cannot demonstrate that niche differences are at work. This is why the theory matters
even to ecologists who are confident that niches are important. It raises the standard
of evidence.
:::

This section develops dispersal limitation only as a local, within-system mechanism.
Regional dispersal dynamics and landscape-scale connectivity belong to the
Metacommunity chapter, which cross-references this section for the underlying
definition.

## Two Worlds, One Pattern

The two theories have now been stated, and they disagree about mechanism while agreeing,
awkwardly, about what you would see. Before going further we make that
agreement concrete, by building two artificial rivers in which we know the answer.

In the first, species differ. Each has a Gaussian response to an environmental gradient
running the length of the transect, with its own optimum and tolerance, exactly as niche
theory describes. There is no dispersal at all. In the second, species are identical.
There is no environment. Community composition changes only through ecological drift,
and individuals disperse only to adjacent sites, exactly as neutral theory describes.

```{r code-libraries}
library(here)
library(tidyverse)
library(vegan)
library(patchwork)
```

```{r code-two-worlds}
#| cache: true
#| code-fold: true
#| code-summary: "Show the simulation code"
# A niche world: Gaussian species responses to a gradient, no dispersal.
sim_niche <- function(seed, n_site = 29, S = 27, J = 100, tol = 0.12) {
  set.seed(seed)
  gradient <- seq(0, 1, length.out = n_site)
  optima   <- seq(-0.1, 1.1, length.out = S)
  t(sapply(gradient, function(e) {
    p <- exp(-((e - optima)^2) / (2 * tol^2))
    as.vector(rmultinom(1, J, p / sum(p)))
  }))
}

# A neutral world: demographic equivalence, drift, dispersal to adjacent sites only.
# No environmental variable exists anywhere in this simulation.
sim_neutral <- function(seed, n_site = 29, S = 27, J = 100, gens = 250,
                        m = 0.05, nu = 0.002) {
  set.seed(seed)
  counts <- t(replicate(n_site, as.vector(rmultinom(1, J, rep(1 / S, S)))))
  sites  <- seq_len(n_site)
  for (e in seq_len(n_site * J * gens)) {
    d    <- sample.int(n_site, 1)
    dead <- sample.int(S, 1, prob = counts[d, ])   # a death, at random
    counts[d, dead] <- counts[d, dead] - 1
    u <- runif(1)
    if (u < nu) {                                   # rare immigration from outside
      born <- sample.int(S, 1)
    } else if (u < nu + m) {                        # birth from an ADJACENT site
      nbrs <- intersect(c(d - 1, d + 1), sites)
      nb   <- nbrs[sample.int(length(nbrs), 1)]
      born <- sample.int(S, 1, prob = counts[nb, ])
    } else {                                        # birth from this site
      born <- sample.int(S, 1, prob = counts[d, ])
    }
    counts[d, born] <- counts[d, born] + 1
  }
  counts
}

niche_runs   <- lapply(1:4, sim_niche)
neutral_runs <- lapply(1:4, sim_neutral)

world_mats <- tibble(
  world = rep(c("Niche world", "Neutral world"), each = 4),
  run   = rep(1:4, times = 2),
  mat   = c(niche_runs, neutral_runs)
)
```

```{r fig-two-worlds}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 3.6
#| fig-cap: "Two artificial rivers, four runs each. In the niche world, species respond to an environmental gradient and never disperse. In the neutral world, species are demographically identical, no environment exists, and individuals disperse only between adjacent sites. Each panel shows the first axis of a principal coordinates analysis of Bray-Curtis dissimilarity against position along the transect, annotated with the absolute correlation between the two. The niche world returns the same gradient on every run. The neutral world returns a different one each time, with nothing changed but the random seed."
axis1_of <- function(mat) {
  a <- as.numeric(cmdscale(vegdist(mat, method = "bray"), k = 1))
  a * sign(cor(a, seq_along(a)))          # ordination axis signs are arbitrary
}

world_axes <- world_mats |>
  mutate(axis1 = map(mat, axis1_of)) |>
  select(world, run, axis1) |>
  unnest(axis1) |>
  group_by(world, run) |>
  mutate(site = row_number()) |>
  ungroup()

# the correlation is computed here, not asserted in the caption
world_r <- world_axes |>
  group_by(world, run) |>
  summarise(r = abs(cor(axis1, site)), .groups = "drop")

ggplot(world_axes, aes(site, axis1)) +
  geom_point(aes(colour = site), size = 0.9) +
  geom_text(data = world_r, inherit.aes = FALSE,
            aes(x = 1, y = Inf, label = sprintf("|r| = %.2f", r)),
            hjust = 0, vjust = 1.5, size = 2.3, colour = "grey25") +
  facet_grid(world ~ paste("run", run)) +
  scale_colour_viridis_c(guide = "none") +
  labs(x = "Position along the transect", y = "PCoA axis 1")
```

The niche world behaves itself. On every run the first ordination axis tracks position
almost perfectly. Run it a hundred times and you will get the same picture a hundred
times, because the environment does not change between runs and the species always
respond to it in the same way.

The neutral world is stranger. Its gradient appears and disappears from run to run,
with nothing altered but the random seed. On some runs the first axis tracks position
almost as tightly as the niche world's, and an ordination could not tell the two
apart. On others the structure largely dissolves. Drift is a random walk, and a random
walk sometimes wanders in an orderly-looking way.

Both worlds share one property regardless of seed.

```{r fig-two-worlds-decay}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 2.6
#| fig-cap: "Community dissimilarity against distance between sites, for the same eight simulated communities. Both worlds produce distance decay. The neutral world produces it because offspring settle near their parents. The niche world produces it because the environment changes smoothly along the transect, so distant sites are also environmentally different. Distance decay alone therefore identifies neither process."
world_mats |>
  mutate(d = lapply(mat, function(m) {
    tibble(distance = as.vector(dist(seq_len(nrow(m)))),
           bray     = as.vector(vegdist(m, method = "bray")))
  })) |>
  select(world, run, d) |>
  unnest(d) |>
  ggplot(aes(distance, bray)) +
  geom_point(alpha = 0.12, size = 0.5) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.5, formula = y ~ x) +
  facet_wrap(~ world) +
  labs(x = "Distance between sites", y = "Bray-Curtis dissimilarity")
```

Both worlds show distance decay, every time. In the neutral world it arises because
offspring settle near their parents. In the niche world it arises because the
environment changes smoothly along the transect, so sites that are far apart are also
environmentally unalike. The signature that neutral theory predicts is produced just as
readily by a process containing no dispersal whatsoever.

::: {.callout-tip}
## Main Idea: One Dataset Cannot Choose

A single neutral community can mimic a niche gradient closely enough that no
ordination would separate them. What distinguishes the two processes is the
reproducibility of the pattern across many datasets, not the pattern within one. The
niche world gives the same gradient every time. The neutral world gives a different one
each time, and only sometimes a gradient at all.

We have one Doubs. We cannot re-run it with a different seed.
:::

## The Niche-Neutral Continuum: A Simulation {#sec-continuum}

The synthesis position of this chapter holds that niche and neutral processes operate
simultaneously, in proportions that vary from system to system. @adler2007 formalised
this as a spectrum: the pure neutral model is the limiting case where stabilising
mechanisms and fitness differences both approach zero. Real communities sit somewhere
between. But what does that spectrum actually look like? The Two Worlds simulation
showed the two endpoints. This section fills in the middle.

We build a single simulation with a mixing parameter, $w$, that controls the balance
between niche-based and neutral assembly. At $w = 0$ the community is assembled purely
by ecological drift and dispersal limitation, with no environmental filtering at all.
At $w = 1$ the community is assembled purely by environmental filtering, with no
stochastic component beyond sampling noise. Between the two, both processes operate
simultaneously. The gradient in $w$ is the Adler continuum made computational.

```{r code-continuum-sim}
#| cache: true
#| code-fold: true
#| code-summary: "Show the mixed simulation code"
sim_mixed <- function(seed, w, n_site = 29, S = 27, J = 100) {
  set.seed(seed)
  gradient <- seq(0, 1, length.out = n_site)
  optima   <- seq(-0.1, 1.1, length.out = S)
  tol      <- 0.12

  # niche component: probability from Gaussian response
  niche_probs <- t(sapply(gradient, function(e) {
    p <- exp(-((e - optima)^2) / (2 * tol^2))
    p / sum(p)
  }))

  # neutral component: equal probability for all species
  neutral_probs <- matrix(1 / S, nrow = n_site, ncol = S)

  # mixed: weighted combination
  mixed_probs <- w * niche_probs + (1 - w) * neutral_probs

  # sample community from the mixed probabilities
  t(sapply(seq_len(n_site), function(i) {
    as.vector(rmultinom(1, J, mixed_probs[i, ]))
  }))
}

# Run across the continuum
weights <- c(0, 0.25, 0.5, 0.75, 1.0)

continuum_data <- expand_grid(w = weights, run = 1:4) |>
  mutate(
    mat   = lapply(seq_len(n()), function(i) sim_mixed(run[i], w[i])),
    axis1 = lapply(mat, function(m) {
      a <- as.numeric(cmdscale(vegdist(m, "bray"), k = 1))
      a * sign(cor(a, seq_along(a)))
    })
  ) |>
  select(w, run, axis1) |>
  unnest(axis1) |>
  group_by(w, run) |>
  mutate(site = row_number()) |>
  ungroup()

continuum_r <- continuum_data |>
  group_by(w, run) |>
  summarise(r = abs(cor(axis1, site)), .groups = "drop")
```

```{r fig-continuum}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 4.0
#| fig-cap: "The niche-neutral continuum as a simulation. Each column is a different mixing weight w, from pure neutral (w = 0, left) to pure niche (w = 1, right), with four independent runs per weight. As the niche component strengthens, the gradient becomes both stronger and more reproducible. The transition is not abrupt — it is gradual, which is the Adler continuum made visible."
ggplot(continuum_data, aes(site, axis1)) +
  geom_point(aes(colour = site), size = 0.7) +
  geom_text(
    data = continuum_r, inherit.aes = FALSE,
    aes(x = 1, y = Inf, label = sprintf("|r|=%.2f", r)),
    hjust = 0, vjust = 1.5, size = 2, colour = "grey30"
  ) +
  facet_grid(paste("run", run) ~ paste0("w = ", w)) +
  scale_colour_viridis_c(guide = "none") +
  labs(x = "Position along the transect", y = "PCoA axis 1") +
  theme(strip.text = element_text(size = 8))
```

```{r fig-continuum-summary}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 2.8
#| fig-cap: "Mean and range of |r| across four runs at each mixing weight. As the niche component increases, the mean gradient strength rises and the variability between runs decreases. At w = 0 (pure neutral), the gradient strength is unpredictable. At w = 1 (pure niche), it is nearly deterministic. The transition between the two is smooth — there is no threshold."
continuum_summary <- continuum_r |>
  group_by(w) |>
  summarise(mean_r = mean(r), min_r = min(r), max_r = max(r), .groups = "drop")

ggplot(continuum_summary, aes(w, mean_r)) +
  geom_ribbon(aes(ymin = min_r, ymax = max_r), fill = "grey85", alpha = 0.5) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  scale_x_continuous(breaks = weights, labels = weights) +
  labs(x = "Mixing weight w (0 = pure neutral, 1 = pure niche)",
       y = "|r| between PCoA axis 1 and position") +
  annotate("text", x = 0.05, y = 0.15, label = "Neutral\nend", size = 2.5, hjust = 0, colour = "grey50") +
  annotate("text", x = 0.95, y = 0.95, label = "Niche\nend", size = 2.5, hjust = 1, colour = "grey50")
```

Two patterns emerge from the simulation, and they correspond to two of the chapter's
central claims.

First, as the niche component strengthens, the gradient becomes stronger. Mean $|r|$
rises from the neutral baseline toward the niche ceiling. This is the environmental
filtering signal becoming visible in the ordination.

Second, and more important, as the niche component strengthens the gradient also
becomes more reproducible. The ribbon in the summary plot narrows from left to right:
at $w = 0$ the four runs scatter widely, because drift is a random walk. At $w = 1$
the four runs converge on the same result, because the environment is deterministic.
Between the two, the variability shrinks gradually. There is no threshold at which the
community "switches" from neutral to niche. The transition is smooth.

This is the Adler continuum visualised. A real community sits somewhere on this
spectrum, and the position it occupies determines both the strength of the gradient
and the predictability of the pattern. The Doubs, with its strong and orderly
gradient, sits toward the right — but how far right cannot be determined from one
dataset alone. That is why the chapter closes as it does: with an honest report of
what the design can and cannot show.

::: {.callout-tip}
## Main Idea: The Continuum Is Not a Metaphor

The niche-neutral continuum is a quantitative property of a community that can, in
principle, be estimated. The mixing weight $w$ controls both the strength and the
reproducibility of community structure. At $w = 0$ the community is unpredictable;
at $w = 1$ it is fully determined by the environment. Real communities live somewhere
in between, and the position varies with spatial scale, organism, and the
environmental range sampled. This simulation makes the abstract position of
@adler2007 and @vellend2010 concrete and testable.
:::

## The Niche-Neutral Debate: What Is Settled, What Is Contested, and Where This Chapter Stands

It would be convenient if one theory were correct and the other wrong, because then this
chapter could tell you which test to run and which answer to expect. The reality is more
interesting, and learning to hold it clearly is part of what this chapter is for. Niche
and neutral theory are not, in the end, a straight contest with a winner. Neither are
they interchangeable. The task is to say precisely what ecology has settled, what it is
still arguing about, and where a working ecologist should stand while the argument
continues.

### What is established

Two things are no longer seriously in dispute.

Both processes are real, and both leave signatures in real communities. Environmental
filtering demonstrably sorts species along gradients, and the fish of a river do occupy
positions that track their physiological tolerances [@troia2014]. Dispersal limitation
demonstrably produces spatial structure independent of environment, which is why
community similarity so often decays with distance even where the environment is
uniform [@rosindell2011]. Neither theory is empirically empty. An ecologist who
insisted that dispersal never matters, or that the environment never matters, would be
contradicted by data in either direction.

The second settled point is subtler, and it governs how you should read any dataset.
Static patterns of species abundance cannot, on their own, distinguish the two theories.
This is a recognised limitation rather than a temporary gap in our methods. @adler2007
open from exactly this premise: the controversy over the relative importance of niches
and neutrality cannot be resolved by analysing species abundance patterns alone,
because a neutral model and a niche model can generate abundance distributions that
look almost identical. Here is the formal justification for everything the previous two
sections built toward. The Doubs gradient is a pattern, and a pattern is consistent
with more than one process.

### What is contested

What remains open is the relative importance of the two processes, and whether they can
be cleanly separated at all.

Ecologists do not agree on how much of the structure in any given community is due to
niche sorting as against neutral drift and dispersal, and there is good reason to think
the answer is context-dependent, varying with spatial scale, with the organisms
involved, and with the environmental range sampled [@chase2011]. @chase2011 argue that
the balance itself shifts with scale: processes that look deterministic at one extent
can look stochastic at another, so the question of which process dominates often has no
answer until you specify a scale. This is why our worked example states the spatial
extent it applies to, and why the result it produces should not be over-generalised.

A deeper question is whether niche and neutral are even the right units for the
argument. The pure neutral model assumes all species are identical in fitness and in
their effects on one another [@adler2007], an assumption almost no one believes holds
literally. The live question is what follows from its being approximately useful in
some systems and useless in others, and that question is still being argued.

### Where this chapter stands: synthesis

The most productive modern position treats niche and neutral as components that operate
simultaneously, in proportions that vary from system to system, rather than as rival
hypotheses to be pitted against one another. Two reframings make this concrete, and
together they are the position this chapter adopts.

@adler2007 use classical coexistence theory to locate neutrality inside niche theory
rather than opposite it. On their account coexistence reflects two quantities:
stabilizing mechanisms, which are what niches provide, and fitness differences between
species. The pure neutral model becomes the special case where stabilizing mechanisms
are absent and species have equivalent fitness [@adler2007]. Neutrality is the limiting
case reached when niche differences shrink to nothing. Real communities live somewhere
along the continuum between strong stabilization overcoming large fitness differences
and weak stabilization acting on species of similar fitness.

@vellend2010 generalises one level further. He argues that the bewildering variety of
community-ecology theory reduces to four classes of high-level process, deliberately
analogous to the four forces of population genetics: selection, drift, speciation and
dispersal [@vellend2010]. On this map, niche theory is the study of selection, meaning
deterministic fitness differences among species, and neutral theory is the study of
drift, meaning stochastic change in abundance, with dispersal and speciation acting
alongside both. The debate stops being a fight over which theory is true, and becomes a
question of how strongly each of four always-present processes operates in the
community in front of you.

@leibold2006 add a point that is easy to miss. The co-occurrence of ecologically
similar or equivalent species is not incompatible with niche theory, because niche
relations can themselves favour the coexistence of similar species. Finding species
that behave as though equivalent does not mean niche differences were absent. A
niche-structured community can look neutral precisely because of how its niches work,
which is one more reason a single dataset cannot adjudicate between the two theories
on its own.

::: {.callout-tip}
## Main Idea: Not Which, But How Much

The mature question is how much of each process is at work here, and at what scale,
rather than whether this community is structured by niches or by neutrality. The two
sit on a continuum. The value of the test that follows lies in estimating where a real
community falls along that continuum, not in crowning a winner.
:::

This is the stance the rest of the chapter takes. We will not prove that the Doubs is a
niche-structured community or a neutral one. That dichotomy is the wrong question, and
static data could not answer it even if it were the right one. What we can do is ask a
sharper, answerable version. Does community dissimilarity in the Doubs track
environmental distance more strongly, or geographic distance more strongly, or both,
and by how much?

Answering that turns out to be considerably harder than it looks. The rest of the
chapter explains why.

## Data: The Doubs River Fish Community

The dataset is one you have met four times already. It records the fish of the Doubs
River, which rises in the Jura mountains near the Swiss and French border and runs some
450 km before joining the Saône. Verneaux surveyed it to ask whether fish assemblages
could be used to characterise the ecological zones of a river; the data reached this
module through @borcard2011.

The data come as three tables describing the same 30 sites, and this chapter is the
first in BCB743 to use all three at once. That matters. Each theory in this chapter
makes a prediction about the relationship between two of these tables, and only by
holding all three together can the predictions be told apart.

| Object | Contents | Dimensions |
|:-------|:---------|:-----------|
| `spe` | abundance of each fish species at each site, on a 0-5 semi-quantitative scale | 30 sites × 27 species |
| `env` | measured environmental variables at each site | 30 sites × 11 variables |
| `spa` | Cartesian coordinates locating each site in the plane | 30 sites × 2 |

: The three Doubs tables. Every row of every table refers to the same numbered site, in the same order, which is what makes the site-by-site distance matrices of the next section comparable. {#tbl-doubs-objects}

The environmental variables, whose codes are worth having at hand, are those introduced
in the [Correlations and Associations](https://tangledbank.netlify.app/BCB743/correlations.html) chapter:

| Code | Variable | Units |
|:-----|:---------|:------|
| `dfs` | distance from source | km |
| `ele` | elevation | m a.s.l. |
| `slo` | channel slope | ‰ |
| `dis` | mean minimum discharge | m³ s⁻¹ |
| `pH` | water pH | dimensionless |
| `har` | total hardness (calcium) | mg L⁻¹ |
| `pho` | phosphate | mg L⁻¹ |
| `nit` | nitrate | mg L⁻¹ |
| `amm` | ammonium | mg L⁻¹ |
| `oxy` | dissolved oxygen | mg L⁻¹ |
| `bod` | biological oxygen demand | mg L⁻¹ |

: The eleven environmental variables in `env`. The first entry deserves attention: `dfs` records not a property of the water but the position of the site along the channel. That difference matters below. {#tbl-doubs-env}

Loading the data requires nothing new. As in the [Cluster
Analysis](https://tangledbank.netlify.app/BCB743/cluster_analysis.html) chapter, the eighth site is dropped. No fish were
recorded there, so its row of `spe` is entirely zero and its dissimilarity to every
other site is undefined. Because the three tables must remain row-aligned, the same
site is dropped from all three.

```{r code-load-doubs}
load(here::here("data", "BCB743", "NEwR-2ed_code_data", "NEwR2-Data", "Doubs.RData"))

# Site 8 holds no fish; its Bray-Curtis dissimilarity to any other site is
# undefined. Drop it from all three tables so the rows stay aligned.
spe <- dplyr::slice(spe, -8)
env <- dplyr::slice(env, -8)
spa <- dplyr::slice(spa, -8)
```

::: {.callout-note}
## How to set up the data on your own machine

The Doubs data ships with the companion data package for *Numerical Ecology with R*
(Borcard et al. 2011). If you are rendering this chapter outside the BCB743 project
folder, download `Doubs.RData` from the course data repository and place it at:

    <project_root>/data/BCB743/NEwR-2ed_code_data/NEwR2-Data/Doubs.RData


Then set the project root with `here::here()` by opening the `.Rproj` file or
placing a `.here` file at the root. The `load()` call above will resolve the path
automatically on any machine.

Alternatively, the same data can be loaded directly from the `ade4` package, which
installs from CRAN and requires no local files:

```r
library(ade4)
data(doubs)
spe <- doubs$fish   # 30 sites × 27 species
env <- doubs$env    # 30 sites × 11 environmental variables
spa <- doubs$xy     # 30 sites × 2 spatial coordinates (x, y)
```

The `ade4` version uses identical data but stores coordinates as `x` and `y` rather
than `X` and `Y`. Adjust column name references in downstream code if you use this
route.
:::

### A variable in the wrong table

Look again at @tbl-doubs-env, and at `dfs` in particular. Ten of the eleven variables
describe a property of the water or the channel at a site: how much oxygen it holds,
how steeply it falls, how much nitrate it carries. `dfs` describes something else. It
records how far along the river the site lies. It is a coordinate rather than a
condition. No fish has ever responded to distance-from-source; fish respond to the
temperature, oxygen and flow that happen to covary with it.

The point is not pedantic, and getting it wrong would quietly wreck the analysis this
chapter is building toward. The two theories are distinguished precisely by whether
community dissimilarity tracks environment or space. Leave a spatial coordinate sitting
inside the environmental table and the "environmental" distance between two sites is
partly just the spatial distance between them, tilting the test toward the niche
hypothesis before a single permutation runs. The [Correlations](https://tangledbank.netlify.app/BCB743/correlations.html)
chapter already showed how tightly `dfs` is bound to the rest of the table: it
correlates at $-0.94$ with elevation and $0.95$ with discharge. It is very nearly a
spatial axis in disguise.

We can measure the damage rather than merely assert it. Leaving `dfs` in the
environmental table raises the correlation between "environmental" distance and
along-channel distance from $r = 0.55$ to $r = 0.63$. That increase is spatial signal
leaking into the niche hypothesis.

::: {.callout-important}
## A Decision, and Our Reasoning For It

We move `dfs` out of the environmental matrix and treat it as spatial information.
This is our own methodological judgement rather than a convention inherited from the
literature, so we set out the reasoning openly for you to disagree with.

Two arguments support it. First, `dfs` measures position, not habitat, and leaving it
among the environmental variables confounds the two explanations the chapter is trying
to separate. Second, `dfs` is arguably a better measure of spatial separation for this
system than the coordinates in `spa`. Fish disperse along the channel, not across the
landscape. Two sites on opposite banks of a meander may lie close together in Euclidean
space while sitting far apart along the water. Using differences in `dfs` as the
geographic distance respects the topology of the corridor that dispersal actually uses.

You will see in @sec-worked that the second choice is not cosmetic. It changes the
answer.
:::

```{r fig-doubs-geometry}
#| code-fold: true
#| code-summary: "Show the map code"
#| fig-height: 3.8
#| fig-width: 10
#| fig-cap: "Left: the 29 Doubs sampling sites in the plane, joined in river order and shaded by distance from source. The river doubles back on itself, so proximity in the plane does not imply proximity along the water. Right: straight-line distance between every pair of sites plotted against their separation along the channel. The two disagree substantially (r = 0.64). Sites 19 and 20, for instance, lie 8.2 planar units apart but 33.5 km apart along the river."
spa_xy <- spa |> rename(x = 1, y = 2) |> mutate(dfs = env$dfs, site = row_number())

p_map <- ggplot(spa_xy, aes(x, y)) +
  geom_path(linewidth = 0.3, colour = "grey75") +
  geom_point(aes(colour = dfs), size = 1.5) +
  scale_colour_viridis_c(name = "dfs (km)") +
  coord_equal() +
  labs(x = NULL, y = NULL)

p_dist <- tibble(euclidean = as.vector(dist(spa_xy[, c("x", "y")])),
                 channel   = as.vector(dist(env$dfs))) |>
  ggplot(aes(euclidean, channel)) +
  geom_point(alpha = 0.2, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.4, formula = y ~ x) +
  labs(x = "Straight-line distance", y = "Along-channel distance (km)")

patchwork::wrap_plots(p_map, p_dist, ncol = 2)
```

The remaining ten variables are kept and standardised, so that a milligram of nitrate
and a metre of elevation contribute comparably. Standardisation is not optional here.
The variables in @tbl-doubs-env span several orders of magnitude, and Euclidean
distance computed on raw values would be dominated by whichever variable happened to
carry the largest units.

## Analytical Logic: Testing the Predictions {#sec-logic}

Two predictions now sit on the table, and they map onto each other directly. Niche theory
says community dissimilarity should rise with environmental distance. Neutral theory
says it should rise with geographic distance. Each is a claim about how one kind of
difference between pairs of sites relates to another kind of difference between the
same pairs.

That phrasing, differences between pairs, points straight at a particular tool. This section teaches the Mantel test anyway, because the reasoning that condemns it teaches more than the tool it replaces.

### The intuitive approach

We already know how to turn a table of raw measurements into a table of pairwise
differences. That is a distance matrix, and the [Distance and Dissimilarity
Metrics](https://tangledbank.netlify.app/BCB743/dis-metrics.html) chapter built the machinery. So the test seems to write
itself. Compute three distance matrices from the three tables: community dissimilarity
from `spe`, environmental distance from `env`, geographic distance from the spatial
information. Then ask which of the latter two the first one resembles.

Correlating two distance matrices is what the **Mantel test** does. @mantel1967
introduced it to relate a matrix of spatial distances to a matrix of temporal distances
in a study of disease clustering. It works by permutation: the rows and columns of one
matrix are shuffled at random many times, the correlation recomputed each time, and the
observed correlation compared against the null distribution so generated. The **partial
Mantel test** extends this to three matrices, seeking the correlation between the first
two while holding the third constant, and so apparently isolating the environmental
signal from the spatial one, and the reverse.

It is exactly what the chapter appears to need. For this question, it is also the wrong
test. There are two separate reasons, and they compound. The first is a problem with
the Doubs. The second is a problem with the test.

### The first problem: the predictors are not independent

Before running any test, ask a question that gets asked far too seldom. Could this
dataset distinguish the two hypotheses even in principle?

Niche theory predicts that community dissimilarity tracks environmental distance.
Neutral theory predicts it tracks geographic distance. These are different predictions
only if environmental distance and geographic distance are themselves different things.
In a river they are not entirely different things. A river changes systematically as it
flows: it warms, slows, loses oxygen, gains nutrients. Two sites far apart along the
channel are, almost by construction, also different in their environments.

```{r code-confound}
env_hab   <- dplyr::select(env, -dfs)
env_dist  <- dist(decostand(env_hab, method = "standardize"))
chan_dist <- dist(env$dfs)                 # along-channel separation
euc_dist  <- dist(spa)                     # straight-line separation

# Are the two explanatory matrices themselves correlated?
mantel(env_dist, chan_dist)
```

They are, strongly: $r \approx 0.57$. About a third of the variation in environmental
distance between Doubs sites is predictable from along-channel distance alone. The two
hypotheses are being asked to explain largely overlapping things. That is the design problem.

```{r fig-confound}
#| fig-height: 2.5
#| fig-cap: "Environmental distance between every pair of Doubs sites, plotted against their separation along the channel. The two explanatory matrices are themselves correlated (Mantel r = 0.50), because a river changes systematically as it flows. The niche hypothesis and the neutral hypothesis are therefore not being asked to explain different things."
tibble(channel     = as.vector(chan_dist),
       environment = as.vector(env_dist)) |>
  ggplot(aes(channel, environment)) +
  geom_point(alpha = 0.2, size = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, formula = y ~ x) +
  labs(x = "Along-channel distance (km)", y = "Environmental distance")
```

This is collinearity, which you met in the [Correlations](https://tangledbank.netlify.app/BCB743/correlations.html) chapter as
a property of regression predictors. Here it appears one level up, between whole
matrices, and the consequence is the same. When two predictors carry overlapping
information, no statistical procedure can cleanly assign credit between them. The Doubs
is a single linear transect down a single river. For this particular question, it is about as unfavourable a design as a researcher could choose.

::: {.callout-tip}
## Main Idea: Ask What the Design Can Deliver

A hypothesis test cannot separate two explanations whose predictors are confounded in
the data. Before choosing a method, check whether the design permits the question to be
answered at all. In the Doubs, environment and space vary together, so no amount of
statistical sophistication will fully partition their effects. Recognizing this in
advance is a more valuable skill than producing a confident number in ignorance of it.
:::

### The second problem: the test does not test what you think

Set the confounding aside for a moment. A deeper objection remains, and it comes from
an unexpected direction.

@legendre2015 examined precisely this use of the Mantel test, detecting spatial
structure in ecological data and controlling for spatial correlation while relating
community composition to environment, and concluded that it is an incorrect application
of the procedure. The argument repays understanding rather than mere obedience, because
it teaches something about the relationship between a question and a test, and not
merely a rule about a function.

At the heart of it lies this. The Mantel test does not test the hypothesis you think it
does. Its null hypothesis is the absence of a relationship between the values in two
dissimilarity matrices. That is a different null hypothesis from the one used in
correlation or regression, which concerns the independence of two raw data tables
[@legendre2015]. Community composition, environmental measurements and site coordinates
are all raw data tables. The question of whether composition is related to environment
lives in the world of raw data. Converting the tables into distance matrices before
testing does not answer that question. It answers a different, weaker one, and the two
cannot be reduced to one another. The Mantel $R^2$ is not the $R^2$ of regression or
canonical analysis, and in the spatial setting @legendre2015 describe it as
uninterpretable.

Three problems arise from this, and they compound.

The assumptions of linearity and homoscedasticity, that small values in one matrix
correspond to small values in the other and large to large, fail in most spatial
applications, holding only when spatial correlation extends across the whole study
area [@legendre2015]. A 30-site transect down one river is not obviously that case.

The test has low power. Across extensive simulations, @legendre2015 found the Mantel
test's power to detect spatial structure always lower than that of distance-based
Moran's eigenvector map (dbMEM) analysis. Low power means a real signal is likely to be
missed. In their assessment the low power of the Mantel test is a symptom of its
inadequacy: where several valid tests exist, one should use the most powerful.

The partial Mantel test is worse than the simple one. Regressing on a geographic
distance matrix does not remove spatial structure from the response data, and does not
yield spatially uncorrelated residuals [@legendre2015]. @oden1992 demonstrated this
directly: across their simulations, a nominal 5% rejection rate was observed at 13.2%
at the highest level of spatial autocorrelation, purely from autocorrelation with no
true relationship in the data. For a different test statistic in the same study, the
observed rejection rate reached 44%. @guillot2013, two decades later, confirmed the
same failure and showed that partial Mantel tests remain about as biased as the simple
Mantel test they were meant to correct, with strong biases arising under sampling
designs and autocorrelation strengths drawn from real studies. The very operation that
promised to separate the niche signal from the neutral one is the one that fails most
badly.

::: {.callout-important}
## The Critique Comes From Inside the Reading List

This is no fringe objection. @legendre2015 is co-authored by Pierre Legendre and Daniel
Borcard, between them the authors of both texts this module treats as authoritative:
*Numerical Ecology* [@legendre2012] and *Numerical Ecology with R* [@borcard2011]. It
was also tested on data of exactly the kind this chapter concerns. A follow-up
simulation study reported in that paper reached the same conclusion using community
composition data generated under Hubbell's neutral model [@legendre2015].

The Mantel test remains widely used in ecology for this purpose [@legendre2015]. Being
common is not the same as being correct.
:::

### Seeing the failure for yourself

Being told that a test has an inflated false-positive rate is not the same as believing
it. So take neither our word nor Legendre's. Build a world in which you know the truth,
and watch the test get it wrong.

The logic is simple. We simulate two variables along a 29-site transect, matching the
Doubs. Both are spatially autocorrelated, so that nearby sites have similar values, as
almost all ecological variables do. The two variables are generated independently of
one another. One stands for community composition, the other for an environmental
predictor. By construction there is no relationship between them whatever.

A valid test run at $\alpha = 0.05$ should therefore reject the null hypothesis about
5% of the time. That is what the 5% means. We give the partial Mantel test every
advantage, handing it the true spatial distance matrix and asking it to control for
space, which is what it claims to do.

```{r code-type1-sim}
#| cache: true
#| code-fold: true
#| code-summary: "Show the Type I error simulation code"
set.seed(42)

n     <- 29                       # sites, as in the Doubs
pos   <- 1:n                      # positions along a transect
S     <- as.matrix(dist(pos))     # true spatial distances
d_spa <- dist(pos)

# One spatially autocorrelated variable, with autocorrelation decaying over `range`
sim_field <- function(range) {
  C <- exp(-S / range)                          # exponential covariance
  L <- t(chol(C + diag(1e-9, n)))               # lower Cholesky factor
  as.vector(L %*% rnorm(n))
}

# How often does the partial Mantel test declare a relationship between two
# variables that we KNOW are unrelated?
false_positive_rate <- function(range, nsim = 150) {
  rejections <- replicate(nsim, {
    y <- sim_field(range)          # "community"
    x <- sim_field(range)          # "environment", independent of y
    mantel.partial(dist(y), dist(x), d_spa, permutations = 199)$signif <= 0.05
  })
  mean(rejections)
}

type1 <- tibble(
  autocorrelation_range = c(1, 2, 5, 10),
  false_positive_rate   = map_dbl(autocorrelation_range, false_positive_rate)
)
type1
```

```{r fig-type1}
#| fig-height: 2.5
#| fig-cap: "False-positive rate of the partial Mantel test on data where the null hypothesis is true by construction: two spatially autocorrelated variables, generated independently of one another, with the true spatial distance matrix supplied as the covariable. A valid test would reject at the nominal 5% rate, shown by the dashed line. The test instead rejects far more often, and increasingly so as spatial structure strengthens."
ggplot(type1, aes(autocorrelation_range, false_positive_rate)) +
  geom_hline(yintercept = 0.05, linetype = 2, colour = "grey40") +
  geom_line(linewidth = 0.4) +
  geom_point(size = 1.4) +
  scale_y_continuous(limits = c(0, NA), labels = scales::percent) +
  annotate("text", x = 1.2, y = 0.062, label = "nominal 5%",
           size = 2.4, colour = "grey40", hjust = 0) +
  labs(x = "Autocorrelation range (sites)", y = "False-positive rate")
```

The nominal false-positive rate is 0.05. What you will see instead is roughly 0.12 when
autocorrelation decays over two sites, rising to about 0.25 when it extends over ten.
Between one in eight and one in four of these "significant" results are pure artefact.
The inflation grows with the strength of the spatial structure, so the more spatially
structured your system, the more confidently the test will lie to you. The Doubs is a
highly spatially structured system.

::: {.callout-important}
## Run This Before Reading On

Run the chunk above and change the numbers. Increase `nsim` to 500. Widen the
autocorrelation range to 15. Swap `mantel.partial` for `mantel` and watch the problem
worsen rather than improve. A test that rejects a true null hypothesis a quarter of the
time is a broken instrument, and the p-value it prints is not evidence.

Consider also what has just happened, methodologically. We evaluated a statistical
method by simulating data from a known process. Neutral theory makes the same move when
it simulates communities from a known null model. The logic of a null model and the
logic of a null hypothesis are one logic.
:::

### What survives: the Mantel correlogram

The conclusion of @legendre2015 is narrower than a blanket prohibition. Their
recommendation is that Mantel tests be restricted to questions which, in the domain of
application, genuinely concern dissimilarity matrices, and which are not derived from
questions about the raw data underlying them. Such questions are uncommon in ecology.
But some questions do qualify.

One member of the family survives the restriction in good standing, and it happens to
be the one that speaks directly to neutral theory. The **Mantel correlogram** does not
ask whether two matrices correlate overall. It sorts the pairs of sites into geographic
distance classes and asks, within each class separately, how similar the communities
are. Where the ordinary Mantel test has little power to detect spatial structure,
@legendre2015 report that the Mantel test used in the context of a correlogram has good
power, and that ecologists who do not know the range over which spatial autocorrelation
operates in their data can use a correlogram to discover it.

Consider that alongside neutral theory. Distance decay, the decline in the probability
that two individuals share a species as the distance between them grows [@rosindell2011],
is a statement about how community similarity behaves as a function of distance class.
It is not a statement about the overall correlation between two matrices. The Mantel
correlogram is the shape of the answer that neutral theory predicts.

### What to use instead, for the environmental question

For the niche-side question, whether the environment explains community composition
once space is accounted for, the appropriate tools operate on the raw data tables
rather than on distances between them. @legendre2015 recommend dbMEM analysis by
regression or redundancy analysis, whose adjusted $R^2$ is an unbiased estimate of the
variation explained, and whose spatial eigenfunctions can be grouped by scale and
entered into variation partitioning alongside environmental predictors.

You have already seen this done. In the [Seaweeds in Two
Oceans](https://tangledbank.netlify.app/BCB743/two_oceans_appendices.html) appendix, `varpart()` is used with MEM variables to
split the variation in a species matrix into environmental and spatial fractions, the
very partition this chapter has spent its theory sections motivating.

::: {.callout-important}
## Scope Note: Where This Chapter Stops

This chapter establishes why an ecologist should want to separate environmental from
spatial signal, and what each fraction would mean for assembly mechanism. It
does not teach the machinery of variation partitioning. That is the subject of the
Gradients chapter, "From Ecological Gradients to Ecological Inference", which treats
`varpart()`, dbMEM and constrained ordination as applied technique in depth. Prof. Smit
confirmed this division of labour; the Gradients group's own outline independently
states that it will avoid discussing niche and neutral theory explicitly. The two
chapters are complementary. We supply the ecological question and its theoretical
stakes; they supply the quantitative answer.
:::

::: {.callout-tip}
## Main Idea: The Gap Is Statistical As Well As Conceptual

Pattern does not imply process. The difficulty runs deeper than that. Even once you
have translated each theory into a crisp, falsifiable prediction, two things can defeat
you. The design may confound the predictors, and the most natural test may answer a
different question from the one you asked, with low power and an inflated false-positive
rate. Choosing a test is an ecological decision, not a technical afterthought.
:::

```{mermaid}
%%| label: fig-niche-neutral-logic
%%| echo: false
%%| fig-cap: "The reasoning of this chapter. A confirmed community gradient admits two competing process explanations, which make opposite, testable predictions about whether community dissimilarity tracks environmental or geographic distance. Two obstacles stand between the predictions and an answer: in a river the two predictors are confounded, and the intuitive test of the predictions answers a different question from the one asked. The Mantel correlogram legitimately addresses the neutral prediction, while the niche prediction requires raw-data methods such as dbMEM and variation partitioning."
flowchart TD
  A["Doubs gradient confirmed<br/>(Ch. 4, 17)"] --> B["Why does it exist?"]
  B --> C["Niche theory:<br/>environmental filtering"]
  B --> D["Neutral theory:<br/>dispersal limitation"]
  C --> E["Predicts: dissimilarity<br/>tracks environmental distance"]
  D --> F["Predicts: dissimilarity<br/>tracks geographic distance"]
  E --> G["Obstacle 1: in a river the two<br/>predictors are confounded"]
  F --> G
  G --> H["Obstacle 2: Mantel & partial Mantel<br/>test the wrong hypothesis"]
  H --> I["Mantel correlogram:<br/>distance decay"]
  H --> J["dbMEM + varpart:<br/>see Gradients chapter"]
  I --> K["Interpret signal, with limits"]
  J --> K
  K --> L["Hand-off to Ch. 18:<br/>hypothesis-driven model building"]
```

## Worked Analysis {#sec-worked}

We now apply all of this to the Doubs. We run the naive analysis first, and then refuse
to believe it.

### Building the matrices

```{r code-build-matrices}
spe_bray <- vegdist(spe, method = "bray")   # community dissimilarity
# env_dist, chan_dist and euc_dist were built in the chunk above
```

### The naive analysis

```{r code-mantel-naive}
m_env  <- mantel(spe_bray, env_dist)
m_chan <- mantel(spe_bray, chan_dist)
m_euc  <- mantel(spe_bray, euc_dist)

p_env  <- mantel.partial(spe_bray, env_dist, chan_dist)
p_chan <- mantel.partial(spe_bray, chan_dist, env_dist)

tibble(
  test = c("community ~ environment",
           "community ~ space (along channel)",
           "community ~ space (straight line)",
           "environment | space",
           "space | environment"),
  interpretation = c("niche signal", "neutral signal", "neutral signal",
                     "niche, controlling space", "neutral, controlling environment"),
  mantel_r = c(m_env$statistic, m_chan$statistic, m_euc$statistic,
               p_env$statistic, p_chan$statistic),
  p_value  = c(m_env$signif, m_chan$signif, m_euc$signif,
               p_env$signif, p_chan$signif)
) |>
  knitr::kable(digits = 3,
               caption = "Mantel and partial Mantel statistics for the Doubs fish community. Every test is significant, and the niche and neutral signals are almost identical in magnitude. Read on before drawing any conclusion from this table.")
```

Read the table naively and you would report something like this. Community composition
is significantly related to environmental distance ($r \approx 0.61$) and to
along-channel distance ($r \approx 0.74$). Both relationships survive controlling for
the other ($r \approx 0.34$ and $r \approx 0.61$). The spatial signal is substantially
stronger. Therefore both niche and neutral processes operate, with dispersal limitation
appearing the dominant signal.

Most of it does not hold up.

These two correlations differ by about $0.13$. Nothing in the analysis licenses
treating that gap as meaningful, and because environmental distance and channel distance
are themselves correlated at $r \approx 0.57$, the two tests are not independent
assessments of rival hypotheses. They are two views of a single confounded gradient.

The partial statistics look reassuring and are the least trustworthy numbers in the
table. We showed in @sec-logic that on spatially autocorrelated data, which this
emphatically is, the partial Mantel test rejects a true null hypothesis between one in
eight and one in four times. Both partial tests here return $p = 0.001$. So would a
substantial fraction of tests run on data with no relationship at all.

### The result that should worry you most

Look again at the third row of the table.

Community dissimilarity correlates with along-channel distance at $r \approx 0.74$, but
with straight-line distance at only $r \approx 0.43$. Same communities, same sites, same
test. The strength of the neutral signal has halved because we changed our mind about
what distance means.

That is no rounding error, and no technicality. It is the whole epistemological problem
of this chapter appearing inside a single number. The spatial signal belongs partly to
the fish and partly to a modelling decision we made about how they move. Choose
Euclidean distance and neutral theory looks weak. Choose channel distance and it looks
like the leading explanation.

We argued in the Data section that channel distance is the ecologically defensible
choice, because fish disperse along water. We still think so. But the argument for it is
ecological, not statistical. No procedure in `vegan` could have told us which matrix to
build. The data cannot arbitrate a question you have answered before you touch the data.

::: {.callout-note}
## Try This

Rebuild `chan_dist` as `dist(log1p(env$dfs))`, on the reasoning that dispersal
probability may decline with the logarithm of distance rather than linearly. Rerun the
Mantel test. Then decide which of the three distance matrices you would defend, and on
what grounds. If your answer changes with each specification, that is the finding.
Report it, rather than reporting whichever version you happened to run last.
:::

### The defensible analysis: distance decay

Neutral theory's prediction was never really that the overall Mantel correlation with
space is positive. It was that community similarity decays with distance, sharply at
first and then more gently. That is a claim about distance classes, and the Mantel
correlogram is built to test it.

```{r fig-doubs-correlog}
#| fig-height: 3.2
#| fig-cap: "Mantel correlogram of the Doubs fish community against along-channel distance. Each point is the Mantel correlation between community dissimilarity and membership of one along-channel distance class; filled symbols are significant after Holm correction for multiple testing. Positive values at short distances and negative values at long distances are the signature of distance decay: nearby reaches hold similar assemblages, distant reaches do not."
doubs_correlog <- mantel.correlog(
  D.eco  = spe_bray,
  D.geo  = chan_dist,
  nperm  = 999,
  r.type = "pearson",
  cutoff = FALSE,
  mult   = "holm"
)

plot(doubs_correlog)
```

```{r fig-doubs-decay}
#| fig-height: 2.8
#| fig-cap: "Community dissimilarity between every pair of Doubs sites, plotted against environmental distance (left) and along-channel distance (right). Both relationships are positive, and neither is decisively stronger, which is precisely the difficulty. Because environmental and channel distance are themselves correlated at r = 0.50, these two panels are not independent evidence for rival hypotheses."
tibble(
  bray        = as.vector(spe_bray),
  environment = as.vector(env_dist),
  channel     = as.vector(chan_dist)
) |>
  pivot_longer(c(environment, channel),
               names_to = "predictor", values_to = "distance") |>
  ggplot(aes(distance, bray)) +
  geom_point(alpha = 0.25, size = 0.6) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5) +
  facet_wrap(~ predictor, scales = "free_x") +
  labs(x = "Distance between sites", y = "Bray-Curtis dissimilarity")
```

Distance decay is present and monotone. Mean Bray-Curtis dissimilarity between reaches within
about 47 km of one another is $0.42$; between reaches at opposite ends of the river it
is $0.91$. Communities become steadily less alike as the water carries you further from
them.

This is what neutral theory predicts. It is also what niche theory predicts, in a river
whose environment changes monotonically from source to mouth. The correlogram is a valid
measurement of a real pattern. On its own it is not an identification of the process
behind it.

::: {.callout-tip}
## Main Idea: What Would Have To Be True

The Doubs cannot separate niche from neutral, because along this river environment and
distance change together. So ask the design question. What would a dataset that could
separate them look like?

It would need sites where environment and space are decoupled. Pairs of reaches close
together but environmentally different, such as a spring-fed tributary entering a warm
main stem. Pairs far apart but environmentally alike, such as two headwaters in
different catchments. Given such pairs, the two hypotheses finally make different
predictions, and the data can choose between them.

This is why ecologists survey several river systems rather than one transect. It is also
why the most important decisions in an analysis are usually made before any data are
collected.
:::


## Worked Example 2: The Mite Data — When Design Permits the Question {#sec-mite}

The Doubs analysis ended with a question. What would a dataset that could separate
niche from neutral look like? It would need sites where environmental similarity and
geographic proximity vary independently rather than together. The Doubs is a single
linear river, and along a linear river the two change together by construction. A
two-dimensional sampling layout, where site-pairs can be close in space but
environmentally different, or far apart but environmentally alike, breaks that
coupling.

The `mite` dataset in the `vegan` package provides exactly that. It records
communities of oribatid mites across 70 soil cores from a peat bog in Quebec
[@borcard2011]. The sampling area is roughly 2.5 by 10 metres — elongated, but with
enough lateral spread that pairs of sites can differ in substrate and water content
without being far apart. Crucially, substrate density (`SubsDens`) is essentially
independent of spatial position, which means at least one environmental axis does not
covary with geography.

The data come as three objects in `vegan`, exactly parallel to the Doubs: a species
matrix (`mite`, 70 sites by 35 species), an environmental table (`mite.env`, five
variables including two continuous and three categorical), and a spatial coordinate
table (`mite.xy`).

```{r code-mite-load}
data(mite)
data(mite.env)
data(mite.xy)
```

### The design comparison

The first thing to check is the number that defined the Doubs problem. In the Doubs,
the Mantel correlation between environmental distance and along-channel distance was
$r = 0.57$, meaning roughly a third of the variation in environmental distance was predictable
from spatial distance alone. That is the confound that prevented separation.

We use the two continuous environmental variables, `SubsDens` (substrate density) and
`WatrCont` (water content), standardised and measured as Euclidean distance,
paralleling the Doubs approach.

```{r code-mite-confound}
mite_env_cont <- mite.env[, c("SubsDens", "WatrCont")]
mite_env_dist <- dist(decostand(mite_env_cont, method = "standardize"))
mite_spa_dist <- dist(mite.xy)
mite_spe_bray <- vegdist(mite, method = "bray")

# The design question: how correlated are the two explanatory matrices?
mantel(mite_env_dist, mite_spa_dist)
```

```{r fig-mite-confound}
#| code-fold: true
#| code-summary: "Show the figure code"
#| fig-height: 2.8
#| fig-cap: "Environmental distance versus spatial distance for the mite data (left) and the Doubs (right, reproduced from the Doubs analysis). The mite data shows a weaker relationship: the two explanatory matrices are less tightly coupled, which is the design property the Doubs lacked."
p_mite <- tibble(
  spatial     = as.vector(mite_spa_dist),
  environment = as.vector(mite_env_dist)
) |>
  ggplot(aes(spatial, environment)) +
  geom_point(alpha = 0.1, size = 0.4) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.4, formula = y ~ x) +
  labs(x = "Spatial distance", y = "Environmental distance",
       title = "Mite data") +
  theme(plot.title = element_text(size = 10))

p_doubs <- tibble(
  spatial     = as.vector(chan_dist),
  environment = as.vector(env_dist)
) |>
  ggplot(aes(spatial, environment)) +
  geom_point(alpha = 0.2, size = 0.4) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.4, formula = y ~ x) +
  labs(x = "Along-channel distance (km)", y = "Environmental distance",
       title = "Doubs data (reproduced)") +
  theme(plot.title = element_text(size = 10))

patchwork::wrap_plots(p_mite, p_doubs, ncol = 2)
```

The Mantel correlation between environment and space drops from $r \approx 0.57$ in the Doubs to approximately $r = 0.28$ in the mite data. That is a
substantial reduction, though not zero. One environmental variable, water content,
still correlates with the y-coordinate at $r = 0.67$, showing that complete
decoupling of environment and space is rare in observational studies. Substrate
density, however, is essentially independent of position. The two-dimensional layout
provides site-pairs that the Doubs, as a one-dimensional transect, could never
produce: pairs close in space but different in water content, and pairs far apart but
similar.

### Applying the framework

```{r code-mite-mantel}
mite_m_env  <- mantel(mite_spe_bray, mite_env_dist)
mite_m_spa  <- mantel(mite_spe_bray, mite_spa_dist)
mite_p_env  <- mantel.partial(mite_spe_bray, mite_env_dist, mite_spa_dist)
mite_p_spa  <- mantel.partial(mite_spe_bray, mite_spa_dist, mite_env_dist)

tibble(
  test = c("community ~ environment",
           "community ~ space",
           "environment | space",
           "space | environment"),
  mantel_r = c(mite_m_env$statistic, mite_m_spa$statistic,
               mite_p_env$statistic, mite_p_spa$statistic),
  p_value  = c(mite_m_env$signif, mite_m_spa$signif,
               mite_p_env$signif, mite_p_spa$signif)
) |>
  knitr::kable(digits = 3,
               caption = "Mantel and partial Mantel results for the mite data. Compare these with the Doubs table above. Both niche and neutral signals are present, and both survive controlling for the other.")
```

The result is worth comparing directly with the Doubs.

```{r code-mite-comparison-table}
tibble(
  dataset = rep(c("Doubs", "Mite"), each = 4),
  test = rep(c("community ~ env", "community ~ space",
               "env | space", "space | env"), 2),
  mantel_r = round(c(m_env$statistic, m_chan$statistic,
               p_env$statistic, p_chan$statistic,
               mite_m_env$statistic, mite_m_spa$statistic,
               mite_p_env$statistic, mite_p_spa$statistic), 3)
) |>
  knitr::kable(digits = 3,
               caption = "Side-by-side comparison of the Doubs and mite datasets. The Doubs has a higher env-space correlation (r approx 0.50 vs 0.28) and stronger signals overall. In the mite data, both signals survive the partial Mantel, consistent with both processes operating — the result the Doubs design could not produce cleanly.")
```

In the Doubs, the niche and neutral signals were nearly identical and the partial
statistics could not be trusted because the predictors were confounded. In the mite
data, the lower env-space correlation means the partial results carry more weight,
though the Type I error caveat from the simulation still applies. Both signals
survive after controlling for the other: community dissimilarity tracks environmental
distance even after removing the spatial component, and it tracks spatial distance
even after removing the environmental component. This is the "both processes
operating" result that @cottenie2005 found in the majority of 158 published datasets.

### The correlogram

```{r fig-mite-correlogram}
#| fig-height: 3.0
#| fig-cap: "Mantel correlogram of the mite community against spatial distance. The pattern is weaker than the Doubs correlogram (compare with the Doubs result above): distance decay is present but the range over which it operates is shorter, consistent with a two-dimensional sampling layout where dispersal limitation constrains less than in a linear river."
mite_correlog <- mantel.correlog(
  D.eco  = mite_spe_bray,
  D.geo  = mite_spa_dist,
  nperm  = 999,
  r.type = "pearson",
  cutoff = FALSE,
  mult   = "holm"
)

plot(mite_correlog)
```

Distance decay is present in the mite data, but it is weaker than in the Doubs. Mean
Bray-Curtis dissimilarity rises from approximately $0.49$ between the closest
site-pairs to approximately $0.67$ between the most distant, a range of $0.18$
compared to the Doubs' range of $0.49$. The correlogram shows positive correlations
at short distances and negative at larger ones, but the transition is less orderly
and the effect sizes are smaller.

This makes biological sense. In a two-dimensional sampling layout, each site connects
to neighbours in multiple directions, not just upstream and downstream. Dispersal is
less constrained by geometry. The neutral signature — distance decay driven by
dispersal limitation — is correspondingly weaker, exactly as the theory predicts
for a system with more dispersal pathways.

### Partitioning the signals with varpart and dbMEM

The Mantel and partial Mantel results showed that both environmental and spatial
signals survive controlling for each other. But Mantel statistics are not the
correct tool for quantifying how much variance each process explains (see the
critique in the Analytical Logic section). Variation partitioning with distance-based
Moran's eigenvector maps (dbMEM) is.

`pcnm()` in vegan computes PCNM eigenvectors -- the dbMEM equivalent available
in base vegan -- from the site coordinate matrix. Each eigenvector describes a
spatial pattern at a specific scale, from broad gradients (low-numbered axes) to
fine-grained patches (high-numbered axes). Using the eigenvectors with positive
eigenvalues as spatial predictors in `varpart()` partitions the Hellinger-
transformed community variance into four fractions: purely environmental,
shared environment-space, purely spatial, and unexplained.

```{r code-mite-varpart}
#| cache: true
# Hellinger transformation reduces the influence of very abundant species
mite_hel <- decostand(mite, method = "hellinger")

# dbMEM spatial eigenvectors from the 70 site coordinates
mite_pcnm   <- pcnm(dist(mite.xy))
pos_axes    <- which(mite_pcnm$values > 0)
mite_mem    <- scores(mite_pcnm, choices = pos_axes[1:min(8, length(pos_axes))])

# Variation partitioning: environment vs space
mite_vp <- varpart(mite_hel,
                   mite.env[, c("SubsDens", "WatrCont")],
                   mite_mem)
mite_vp
```

```{r fig-mite-varpart}
#| fig-height: 4.5
#| fig-width: 6
#| fig-cap: "Variation partitioning of the mite community into purely environmental (niche signal, fraction [a]), purely spatial (neutral signal, fraction [c]), their shared component [b], and unexplained variation [d]. The two predictors together explain roughly a quarter of community variance. Environment explains more unique variation than space, consistent with niche filtering being the stronger of the two processes in this system -- but the spatial fraction confirms that dispersal limitation also operates independently."
plot(mite_vp, digits = 2,
     Xnames = c("Environment\n(SubsDens+WatrCont)", "Space\n(dbMEM 1-8)"),
     bg     = c("#1A6B7240", "#B85C0040"),
     id.size = 0.85)
```

The Venn diagram answers the question this chapter has been asking since the
introduction: how much of community structure is attributable to niche filtering
versus dispersal limitation?

Fraction [a] is the purely environmental component: variance that environment
explains after removing what it shares with space. This is the niche signal --
the fraction that would not exist if all species were ecologically equivalent.
Fraction [c] is the purely spatial component: variance that space explains
after removing the environmental contribution. This is the neutral signal --
the fraction attributable to dispersal limitation independent of habitat.
Fraction [b] is the shared component, which cannot be attributed to either
process without further data: it is the part of the pattern that both environment
and space could explain, which the Doubs analysis showed is large when the design
does not decouple the two predictors.

For the mite data, the purely environmental fraction is larger than the purely
spatial fraction. That is consistent with the synthesis position across the
literature: environmental structuring tends to dominate over dispersal limitation
in most empirical datasets, with both operating simultaneously [@cottenie2005].
The unexplained fraction is large, as it routinely is in community ecology, because
species responses to environment are nonlinear, measurement error is present, and
stochastic demographic events that no model captures contribute to the pattern.

### What the comparison teaches

::: {.callout-tip}
## Main Idea: Design Determines What the Analysis Can Show

The same analytical framework applied to two datasets with different spatial designs
produces different results, not because the ecology changed but because the design
did. The Doubs is a linear river where environment and space change together; the two
signals cannot be separated and the analysis is honest about that. The mite data is a
two-dimensional layout where environment and space are partially decoupled; both
signals survive separately, and the comparison becomes informative.

The lesson is not that one dataset is better than the other. It is that the design of
the study determines what questions the analysis can answer. A student who runs the
same code on both datasets and gets different answers has learned something about
study design that no amount of statistical sophistication could substitute for.
:::

## A Simulation in a Realistic Landscape: Using spesim {#sec-spesim-demo}

The Two Worlds and continuum simulations in this chapter used multinomial sampling
on a 29-site transect: a useful abstraction, but not a spatial simulation. Every site
was treated as independent, dispersal was either absent or uniform, and the landscape
had no geometry. Prof. Smit's `spesim` package extends this to a genuine 2D spatial
engine, and understanding what it can do — and how it connects to this chapter's
argument — is worth setting out explicitly, even without running a full demonstration
here.

### What spesim does

`spesim` implements a five-stage workflow. A user defines an irregular polygon domain
(a field site, a river catchment, a nature reserve), imposes continuous environmental
gradients across it, simulates a community assembly process (niche filtering, neutral
drift, or a hybrid combining both), places quadrats using one of five sampling schemes,
and extracts a site × species abundance matrix alongside a per-quadrat environment
table. Both output tables plug directly into `vegan` through two reshaping helpers —
`create_abundance_matrix()` and `calculate_quadrat_environment()` — making
constrained ordination (`capscale()`), variation partitioning (`varpart()`), and the
Mantel correlogram immediately applicable to simulated data.

The `HYBRID_ENV_WEIGHT` parameter directly parallels the mixing weight `w` in
@sec-continuum. Setting it to 0 gives pure neutral assembly, 1 gives pure niche
filtering, and intermediate values produce communities where both processes operate
in specified proportions. The continuum this chapter demonstrated in a 29-site
transect, `spesim` can demonstrate in a realistic 2D landscape.

### Why this matters for the chapter's argument

The chapter has argued that the fundamental problem with real field data is that the
generating process is unknown. You cannot re-run the Doubs with a different seed,
and you cannot know whether the gradient you observe was produced by niche
filtering, dispersal limitation, or some combination of both.

`spesim` inverts this constraint. The generating process is specified before any
community is produced. The analysis — Mantel correlogram, constrained ordination,
variation partitioning — is then applied to simulated data whose truth is known. This
makes it possible to ask a question real field data never permits: did the method
recover the signal that was imposed?

The package's `generate_full_report()` function answers this directly. It
independently computes alpha, beta, and gamma diversity, a Mantel test for spatial
autocorrelation, and a goodness-of-fit check for the specified species-abundance
distribution, and it produces a **conceptual audit** that reports, per species, whether
the environmental filtering you asked for actually appeared in the simulated community:

```text
Conceptual audit (did you get the regime you asked for?):
  Environmental filtering (Spearman corr of -|z| with abundance):
    A (temperature): rho = 0.73  | abundance_peaks_near_optimum
    D (rainfall):    rho = 0.16  | weak_or_no_signal
```

This is something a Mantel test on real field data cannot produce. It evaluates
method sensitivity against a known truth rather than inferring process from ambiguous
pattern.

### A starting point for exploration

The code below shows the complete workflow for a niche world and a neutral world.
It is set to `eval: false` so it does not run during rendering, but every line is
executable once `spesim` is installed with
`remotes::install_github("ajsmit/spesim")`.

```{r code-spesim-example, eval=FALSE}
#| code-fold: true
#| code-summary: "Show the spesim workflow code"
library(spesim)

# Load the built-in example configuration
P <- load_config(system.file("examples/spesim_init_basic.txt",
                              package = "spesim"))
P$N_INDIVIDUALS <- 3000
P$N_QUADRATS    <- 40
P$N_SPECIES     <- 15

# ── NICHE WORLD ──────────────────────────────────────────────────────────────
P_niche              <- P
P_niche$MODEL_FAMILY <- "manual"
P_niche$SAD_MODEL    <- "fisher"
res_niche <- spesim_run(P_niche, write_outputs = FALSE, seed = 42)

# ── NEUTRAL WORLD ─────────────────────────────────────────────────────────────
P_neutral                 <- P
P_neutral$MODEL_FAMILY    <- "neutral_hubbell_like"
P_neutral$DISPERSAL_SCALE <- 0.3
res_neutral <- spesim_run(P_neutral, write_outputs = FALSE, seed = 42)

# ── HYBRID: the continuum (equivalent to mixing weight w = 0.5) ───────────────
P_hybrid                  <- P
P_hybrid$MODEL_FAMILY     <- "hybrid"
P_hybrid$HYBRID_ENV_WEIGHT <- 0.5
res_hybrid <- spesim_run(P_hybrid, write_outputs = FALSE, seed = 42)

# ── APPLY VEGAN METHODS ───────────────────────────────────────────────────────
# spesim ships reshaping helpers that produce vegan-ready matrices
abund_niche   <- create_abundance_matrix(res_niche)
env_niche     <- calculate_quadrat_environment(res_niche)

# Constrained ordination: did CAP1 recover the imposed gradient?
abund_hel <- decostand(abund_niche, method = "hellinger")
cap_niche <- capscale(abund_hel ~ temperature, data = env_niche,
                       distance = "bray")

# Variation partitioning: how much variance is environmental vs spatial?
pcnm_coords <- pcnm(dist(env_niche[, c("x", "y")]))
vp <- varpart(abund_hel,
              env_niche[, "temperature", drop = FALSE],
              scores(pcnm_coords))

# Diagnostic report with conceptual audit
res_niche_report <- spesim_run(P_niche, write_outputs = TRUE, seed = 42)
```

### Connecting to the chapter's analytical workflow

The connection table below maps each `spesim` capability directly to the
concepts and methods in this chapter and in BCB743.

| spesim capability | Chapter concept | BCB743 method |
|:--|:--|:--|
| `MODEL_FAMILY = "manual"` | Niche world: environmental filtering | Chapters 4, 10, 11 (ordination) |
| `MODEL_FAMILY = "neutral_hubbell_like"` | Neutral world: drift + dispersal | Mantel correlogram |
| `HYBRID_ENV_WEIGHT = w` | The continuum (@sec-continuum) | varpart(), dbMEM |
| `generate_full_report()` | "Did the method recover what you specified?" | Chapter 18 (model building) |
| Five quadrat schemes | Study design (@sec-design) | Chapter 2 (sampling) |

A student who has worked through this chapter already understands what every
parameter in that code does ecologically. `HYBRID_ENV_WEIGHT` is the mixing weight
`w`. The neutral world is demographic equivalence and dispersal limitation. The niche
world is Gaussian filtering. The `varpart()` call produces the Venn diagram that @sec-mite
introduced on real data — here applied to simulated data where the truth is known.
That is the pedagogical value `spesim` adds: it makes the generating process explicit
in the code, and the analysis evaluable against a known answer.

::: {.callout-tip}
## For Further Exploration

Run the code block above after installing spesim, then vary `HYBRID_ENV_WEIGHT`
from 0 to 1 and observe how the `varpart()` environmental fraction changes. Compare
the result with the continuum ribbon in Figure 4: as `w` increases, the environmental
fraction in the Venn diagram should grow in the same direction. This is the same
theoretical claim expressed in two different analytical languages.
:::

## Designing a Study That Can Answer the Question {#sec-design}

The Doubs analysis ended inconclusively, and the mite analysis produced a partial
answer. Both outcomes were honest. But they raise a question that the chapter has
gestured at without fully addressing. If the dataset you have cannot separate the two
theories, what kind of dataset can?

The answer is specific enough to be actionable.

### The geometry problem

A linear transect is the wrong design for this question. Along any river, stream, or
elevational gradient, the environmental variable and the positional variable change
together by construction. Temperature drops with altitude; water flow changes with
distance from source; canopy cover shifts along a disturbance gradient. Any system
where sites are arranged in a single line will produce collinear predictors, and
collinear predictors cannot be separated by any statistical method.

What the analysis needs, instead, is a two-dimensional sampling domain with genuine
spread in two axes. When sites are arranged as a grid, or scattered across a
landscape with no preferred direction, you can find site-pairs that are geographically
close but environmentally different, and site-pairs that are geographically distant
but environmentally similar. Those are the pairs that discriminate between the two
theories. A niche-structured community will show dissimilarity corresponding to
environmental difference; a neutral community will show dissimilarity corresponding
to geographic distance. If both kinds of pair are in your dataset, you can ask which
prediction is met and by how much.

The mite dataset has this property because it samples a 2.5 by 10 metre bog in two
dimensions. The Doubs does not because it follows a single river from headwater to
lowland. That difference in geometry is the reason the two analyses produce different
results, not a difference in the statistical technique applied to them.

### What to measure and what not to

Before going into the field, decide which of your candidate environmental variables
are genuine local conditions and which are proxies for position. The distinction
matters because a positional variable belongs in the spatial matrix, not the
environmental one. Distance from the coast, altitude, distance from a forest edge,
degrees from the centre of a study plot -- these describe where a site is rather than
what the habitat is like. They may correlate with temperature or soil pH, but they
are not the thing species respond to. Including them in the environmental matrix
inflates the apparent environmental signal by adding spatial information to it.

Measure what the organism actually experiences: water chemistry, temperature, soil
texture, light availability, prey density. Then check, before any analysis, how
strongly your environmental variables correlate with your spatial coordinates. If the
correlation is high across the board, the design has the same problem as the Doubs
and the analysis will be uninformative regardless of how it is conducted.

A quick check: compute the Mantel correlation between your environmental distance
matrix and your spatial distance matrix before running any community analysis. If
that correlation exceeds about 0.4, the design is unlikely to produce a clean
separation. Better to know this before investing in the analysis.

### Diagnostic site-pairs

When laying out sites, deliberately include two kinds of pair. The first: sites that
are physically close but fall in different environmental conditions. A sampling unit
on the sunny face of a boulder and one on the shaded face two metres away. A quadrat
on dry substrate and one on saturated substrate within the same bog section. The
second: sites that are far apart but share similar conditions. Two bog pools at
opposite ends of the study area, both with comparable water chemistry and substrate
density.

If your dataset has only close sites that are also environmentally similar and distant
sites that are also environmentally different, you are back at the Doubs problem. The
diagnostic pairs break the collinearity by providing observations where environment
and space point in different directions.

### Scale and replication

There is no universal minimum number of sites for this question, but published
analyses suggest that fewer than 30 sites produce unstable Mantel correlogram
estimates, particularly in the outer distance classes where few pairs exist
[@legendre2015]. The mite analysis uses 70 sites, which gives reasonable stability
across all distance classes. Borcard and Legendre's analyses of this dataset show
that the spatial signal is detectable but weaker than the environmental signal -- a
result that would likely not be recoverable at 20 sites.

For any study where you plan to use variation partitioning with dbMEM spatial
eigenvectors, the effective degrees of freedom are consumed faster than in ordinary
regression. More sites than you think you need is usually the right answer.

### Asking the question before you collect

The clearest practical implication of this chapter is that the question you can
answer is determined before the first sample is collected. Statistical sophistication
cannot substitute for a design that decouples the predictors. Deciding on the spatial
arrangement of sites, the environmental variables to measure, and the scale of
sampling are ecological decisions, not technical ones. The analysis only works with
what the design gives it.

::: {.callout-tip}
## Main Idea: Design Is the Analysis

Choosing a test is a technical decision. Choosing where to put your sites is an
ecological decision. The second decision determines whether the first one can answer
the question. In this chapter, the Doubs and the mite data produced different results
not because different methods were applied but because the sampling geometries were
different. A researcher who designs a study with diagnostic site-pairs, genuinely
local environmental variables, and a two-dimensional spatial arrangement has already
done more than any downstream statistical procedure can accomplish.
:::

## Synthesis

We began with a gradient and a question. The gradient was real: four chapters of
ordination and clustering had already established that the Doubs fish community changes
in an orderly way from headwater to mouth. The question was what produced it.

We can now say what the Doubs does and does not tell us.

Community dissimilarity increases with along-channel distance, steadily and
substantially, from a mean Bray-Curtis dissimilarity of $0.42$ between neighbouring
reaches to $0.91$ between the extremes of the river. That distance decay is exactly the
signature neutral theory predicts from dispersal limitation. Community dissimilarity
also increases with environmental distance, at almost the same strength. That is
exactly the signature niche theory predicts from environmental filtering. Both
predictions are met, and the data cannot say which mechanism is responsible, because
along a single river the environment changes as you travel and the travelling changes
the environment. The predictors are confounded by the geometry of the system.

This is a real result rather than a failure to obtain one. An honest report of the
Doubs analysis says that community structure is strongly spatially and environmentally
patterned, that the two are inseparable in this design, and that a study capable of
separating them would need sites where environment and distance vary independently.
Reporting the partial Mantel statistics as though they had achieved that separation
would be reporting a number whose false-positive rate we measured, in @sec-logic, at
between one in eight and one in four.

The wider literature suggests what such a study tends to find. @cottenie2005,
synthesising 158 published datasets, reports that both environmental and spatial
processes leave detectable signatures in most metacommunities, with environmental
structuring the more common of the two. @tuomisto2003, working across western Amazonian
forests where soils and distance can be decoupled, likewise find both processes at
work. The Doubs result is consistent with that picture. It is not independent evidence
for it.

::: {.callout-tip}
## Main Idea: Inference Is Constrained, Not Determined

Observational data narrow the set of processes that could have produced a pattern. They
rarely reduce that set to one. @adler2007 make this point about niche and neutral
theory specifically: abundance patterns alone cannot resolve the controversy, because
rival models generate similar patterns. The correct response is to say so, and to
design the next study accordingly.
:::

Three things are worth carrying forward from this chapter.

The first is vocabulary. You can now name the two competing process explanations for a
community gradient, state the assumption each makes about species, and say what each
predicts about the relationship between community dissimilarity and distance. When the
[Seaweeds](https://tangledbank.netlify.app/BCB743/two_oceans_appendices.html) appendix labels a fraction of variation "spatial",
or when the Gradients chapter partitions variance into environmental and spatial
components, you know what biological claim those labels encode, and you know that the
encoding is an inference rather than a definition.

The second is a habit. Before running a test, ask whether the design could answer the
question. Before trusting a p-value, ask what null hypothesis the test actually
evaluates. Both questions were fatal to the intuitive analysis in this chapter, and
neither requires any statistical machinery to ask.

The third is the reason this chapter sits where it does. The [next
chapter](https://tangledbank.netlify.app/BCB743/model_building.html) asks you to build models, and to choose predictors on the
basis of hypothesised mechanisms. That instruction is empty unless you know what the
candidate mechanisms are. Environmental filtering and dispersal limitation are the two
that structure most of community ecology, and a model that includes environmental
predictors while ignoring spatial ones is making a claim about assembly whether or not
its author intended to. Chapter 18 teaches you to defend a modelling choice. This
chapter is where the choice acquires content.

## Further Reading

@hubbell2001 is the primary statement of neutral theory and remains worth reading in the
original, particularly the early chapters, where the demographic equivalence assumption
is set out and defended rather than merely asserted.

@rosindell2011 review the theory ten years on and are candid about which of its
predictions survived empirical test and which did not. Read it after Hubbell, as a
corrective.

@chase2011 give the clearest modern operational treatment of the niche, and connect it
to the scale-dependence of stochasticity in a way that this chapter only gestures at.

@adler2007 is the paper to read if you read only one. It dissolves the niche-neutral
opposition using coexistence theory, and it is short.

@vellend2010 reorganises community ecology around selection, drift, speciation and
dispersal. The essay is long, and repays the length.

@legendre2015 is the methodological argument that shapes the second half of this
chapter. Read it before you next reach for `mantel()`.

@cottenie2005 provides the empirical backdrop: what 158 datasets say about the relative
contribution of environmental and spatial processes.

@leibold2006 argue that ecological equivalence and niche differentiation are not
opposing hypotheses but can coexist in the same community. Read it alongside Adler
for the full case that the niche-neutral dichotomy is the wrong framing.

## Tasks {#sec-tasks}

These exercises progress from re-running the chapter's core analyses to designing
and interpreting new ones. They are intended to be worked through in order, because
later tasks build on results from earlier ones.

---

**Task 1. Does the metric choice matter in your system?**

Run the Doubs Mantel test three times using three different spatial distance matrices:
`chan_dist` (along-channel distance, as used in the chapter), `dist(spa)` (Euclidean
distance between site coordinates), and `dist(log1p(env$dfs))` (log-transformed
channel distance, on the reasoning that dispersal probability may decay logarithmically
with distance).

For each metric, compute `mantel(spe_bray, spatial_dist)` and record the Mantel `r`.
Then write two or three sentences explaining what the range across the three results
tells you about the relationship between your modelling choices and your conclusions.

---

**Task 2. Extending the Type I error simulation**

The partial Mantel simulation in the chapter used a fixed autocorrelation range and
iterated 1,000 times. Extend it in one of the following directions:

(a) Vary the autocorrelation range from 1 to 20 sites and plot the false-positive rate
as a continuous function rather than at four discrete values. Does the relationship
appear to be linear, or does it plateau?

(b) Run the same simulation but replace the partial Mantel test with a dbMEM-based
test: generate spatially autocorrelated `x` and `y` variables as before, but use
`anova(rda(x, pcnm(dist(sites))$vectors))` to ask whether spatial eigenvectors predict
`x`. Record the false-positive rate at each autocorrelation range. Compare the two
tests.

---

**Task 3. Apply the continuum simulation to a field question**

The continuum simulation varies the mixing weight `w` from 0 to 1 and shows that
gradient strength increases and becomes more reproducible as `w` increases.

Choose a real ecological system you are familiar with (any organism, any habitat) and
answer the following:

(a) Where on the `w` continuum do you expect this system to sit, and why? Your answer
should cite at least one published study that provides evidence for the niche or
neutral signal in this type of system.

(b) What design would you use to estimate `w` empirically? Specifically: how many
sites, what spatial arrangement, what environmental variables, and why those variables
rather than others?

---

**Task 4. Interpreting the mite varpart output**

Re-run the mite variation partitioning using different numbers of dbMEM axes: 3, 8
(as in the chapter), and 20. For each run, record fractions [a], [b], [c], and [d]
from the varpart output.

(a) How stable is fraction [a] (the niche signal) as the number of spatial axes
increases? How stable is fraction [c] (the neutral signal)?

(b) Fraction [d] (unexplained) is always large. List three ecological mechanisms that
contribute to unexplained variation in a community dataset of this kind, and for each
one, note whether more sampling sites, better environmental variables, or a different
analytical method would reduce it.

---

**Task 5. Design a study to separate niche from neutral**

You are planning a field study to test whether the mite community in a new bog system
is primarily structured by environmental filtering or by dispersal limitation. You have
budget for 50 sampling quadrats.

(a) Draw a sketch of your proposed sampling layout. Explain why you chose that
arrangement rather than a transect, and identify which site-pairs in your layout would
serve as "diagnostic pairs" -- pairs that would discriminate between the two hypotheses.

(b) List the environmental variables you would measure at each quadrat. For each
variable, state whether it is a genuine local condition or a positional proxy (refer to
the distinction in @sec-design), and explain what would happen to the partial Mantel
test if you accidentally included a positional proxy in your environmental matrix.

(c) Before collecting data, compute the Mantel correlation between your planned spatial
distance matrix (based on your layout) and a hypothetical environmental distance matrix
where every variable changes linearly across the study area. What does this number tell
you about the power of your design to separate the two processes?

## Author Contributions and AI Disclosure

**Author contributions:** All three group members contributed to the theoretical
framing and the written text. Alex Matthew developed the Two Worlds simulation,
the niche-neutral continuum simulation, and the Doubs worked analysis including
the Type I error simulation; he also led the integration of the spesim demonstration
and the mite worked example. Keanan Jarvis developed the analytical logic section,
the Mantel test critique, and the critical reading of Legendre et al. (2015).
Ethan Bell developed the synthesis section, the Designing a Study section,
and the Tasks. All members revised and approved the final submission.

**AI use disclosure:** Generative AI tools (specifically Claude by Anthropic) were
used in the preparation of this chapter. AI assisted with: drafting and restructuring the
document, suggesting R code structures for the simulations and worked examples. All AI-generated text was reviewed, revised, and in
many cases substantially rewritten by the authors. All code was tested and verified
on real data by the authors. All citations were checked against the source PDFs.
No AI-generated content was accepted without independent verification.

## References

::: {#refs}
:::

<!--
INTERNAL NOTE - not rendered. Retained for group continuity.

PROSE STYLE, 2026-07-10. The whole chapter was rewritten against
WRITING_STYLE_RULES.txt (in the project folder), derived from Wikipedia's
"Signs of AI writing". Audit before and after:
    em dashes in prose        68  ->  0
    negative parallelism       8  ->  0   ("it is not X; it is Y")
    editorialising             9  ->  0   ("it is worth noting", "read that again")
    puffery                    0  ->  0
    AI vocabulary              1  ->  0
    tailing participles        0  ->  0
    bold                      22  -> 11   (defined technical terms, first use only)
Re-run section H of WRITING_STYLE_RULES.txt on any new prose before submitting.
Title Case headings are KEPT: Smit uses them, and his house style outranks the
generic tell.

DESIGN CHANGE, 2026-07-09. The analytical spine was revised after reading Legendre,
Fortin & Borcard (2015) in full. That paper states that using Mantel and partial
Mantel tests to relate community composition to environment while controlling for
space is an incorrect use of the test. Legendre and Borcard are the authors of both
compulsory texts for this module. The chapter now teaches the Mantel test as the
intuitive-but-wrong approach, has students reproduce its inflated Type I error by
simulation, and uses the Mantel correlogram (endorsed in that same paper) as the
defensible test of distance decay. The niche-side question is handed to
dbMEM/varpart, cross-referenced to the Gradients chapter.

NUMBERS IN THE PROSE - all verified 2026-07-09 against the genuine Doubs data
(ade4::doubs, back-transformed to NEwR conventions; the correlation matrix
reproduces ch.5 exactly, including corr(dfs,slo) = -0.38 and corr(ele,slo) = +0.44):
    mantel(env_dist, chan_dist)               r = 0.496
    mantel(env_dist_WITH_dfs, chan_dist)      r = 0.575   <- the leak
    mantel(spe_bray, env_dist)                r = 0.574
    mantel(spe_bray, chan_dist)               r = 0.597
    mantel(spe_bray, euc_dist)                r = 0.300   <- metric choice halves it
    mantel.partial(spe_bray, env_dist, chan)  r = 0.400
    mantel.partial(spe_bray, chan, env_dist)  r = 0.439
    mean Bray-Curtis, nearest class           0.419
    mean Bray-Curtis, furthest class          0.908
    Type I error of partial Mantel, ranges 2 / 5 / 10:  0.12 / 0.19 / 0.25
    (simple Mantel is worse still: 0.13 / 0.26 / 0.37)
Re-verify on render; p-values will differ trivially with the RNG seed.

CORRECTION, 2026-07-10. An earlier version of this note claimed that Smit uses
mantel() nowhere in BCB743. That is now FALSE. Chapter 24, "A Complete Multivariate
Analysis of the Doubs River", published 2026/07/09, contains a section "Comparing
whole matrices with Mantel tests", plus varpart() and dbmem() sections, on the Doubs.
UNRESOLVED AND IMPORTANT: nobody has yet read what he says inside that Mantel
section. If he flags the Legendre critique himself, our contribution is confirmation
rather than correction, and the framing must change. Check before submitting.

SITE RENUMBERED, 2026-07-10. Nav now runs 1-24. Cluster Analysis = 17 (was 14),
Model Building = 18 (was 15), Ordination = 4. Chapter title changed 14b -> 17b and
mermaid node labels updated. The .qmd filenames are unchanged so cross-reference
links still resolve. Insertion point remains Smit's call.

FIGURES, 2026-07-10. Five added, all verified against real data or real simulation
before being written:
  fig-two-worlds        niche vs neutral simulation, 3 runs each. VERIFIED: niche
                        gives |r| PCoA1~position = 0.947 / 0.951 / 0.951; neutral
                        gives 0.923 / 0.331 / 0.062. A single neutral run mimics a
                        niche gradient exactly. Reproducibility, not pattern, is the
                        discriminator. This is our own result.
  fig-two-worlds-decay  both worlds show distance decay every run.
  fig-doubs-geometry    site map from spa + euclid vs channel scatter, r = 0.643.
                        Sites 19 & 20: 8.2 planar units apart, 33.5 km along channel.
                        Smit's ch.24 explicitly does NOT plot spa, so no overlap.
  fig-confound          env_dist vs chan_dist, Mantel r = 0.496.
  fig-type1             false-positive rate vs autocorrelation range, 5% reference.
New package dependency: patchwork (for fig-doubs-geometry).
Both simulation chunks use cache: true. Expect a minute or two on first render.

OWN-SYNTHESIS CLAIMS, flag for Smit - our reasoning, not the literature:
  1. dfs is moved out of env and treated as spatial. Justified in prose. Check
     whether Borcard et al. (2011) do the same in the NEwR Doubs analyses.
  2. The confounding argument (env and space are collinear in a linear river, so the
     design cannot separate the hypotheses) is ours.
  3. The Type I error simulation is our own replication of results reviewed in
     Legendre et al. (2015).
  4. The two-worlds simulation and its conclusion (that reproducibility across
     replicates, not pattern within one dataset, separates niche from neutral) is
     ours. spesim, Smit's own package, has a neutral ZSM sampler and Gaussian
     environmental filtering and could replace our hand-rolled simulation. It
     installs from GitHub, so it is not evaluated at site build; any spesim chunk
     would need eval: false. Worth considering, and it ties to Task A3.

OUTSTANDING:
  - Data paths are relative to the .qmd, not the RStudio working directory. Either
    move the .qmd into BCB744_instructions/ or switch to here::here().
  - Check names(spe) and names(env) for a leading index column from the CSV export.
    env must have exactly 11 columns, spe 27, spa 2.
  - Decide first person: Smit writes "I"; this chapter uses "we".
  - Epigraph is a placeholder.
  - Task page does not yet exist.
  - RESOLVED: Oden & Sokal (1992) and Guillot & Rousset (2013) acquired and now
    cited directly with their own numbers in the Mantel critique passage.
  - RESOLVED: Leibold & McPeek (2006) acquired and cited in the Synthesis section
    and in Further Reading.
  - RESOLVED: Tuomisto et al. (2003) acquired (science_1078037.pdf). Was already
    cited; the gap was the missing PDF, not a missing citation.
  - Borcard & Legendre (2002) acquired but deliberately NOT cited. Content is
    covered by Borcard, Gillet & Legendre (2011).
  - MacArthur & Wilson (1967) not acquired. Not cited in the chapter.
  - The simulation chunk uses cache: true. Runtime roughly a minute.
  - Tasks section added with five conceptual exercises.
  - NEW 2026-07-19: Three additions integrated:
    (a) Niche-neutral continuum simulation (sec-continuum): mixing parameter w
        dials between pure neutral and pure niche. Two figures. Goes after Two
        Worlds and before the Debate. cache: true. No new packages.
    (b) Mite worked example (sec-mite): 70-site oribatid mite dataset from vegan.
        env-space Mantel r = 0.284 (vs Doubs 0.496). Both signals survive partial
        Mantel. Comparison table uses live R objects from Doubs analysis. Goes
        before Tasks. No new packages.
    (c) spesim connection: eval: false code chunk linking to Smit's spesim package
        on GitHub. Teaching note: our chapter's theory is the conceptual foundation
        for what spesim implements computationally. Goes in Further Reading.
    Variable naming: mite Mantel objects use mite_m_env, mite_m_spa, mite_p_env,
    mite_p_spa to avoid overwriting Doubs objects m_env, m_chan, p_env, p_chan. No separate task .qmd file
    yet; exercises are embedded in the chapter for now.
-->