---
title: "4. Distributions, Sampling, and Uncertainty"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_grey(base_size = 8)
)
```

```{r code-repro-seed, echo=FALSE}
set.seed(74404)
```

```{r code-libraries, echo=FALSE}
library(tidyverse)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   Why distributions shape model choice
-   Discrete and continuous random variables
-   Common distributions encountered in biology
-   Sampling variation and the central limit theorem
-   Confidence intervals as measures of uncertainty
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   None
:::

# Introduction

A good grasp of data distributions is a prerequisite for any statistical analysis. Distributions help us describe variation in data, choose plausible models, and judge whether inferential conclusions are defensible.

My job here is to take you through three linked ideas:

1. the **distribution of observed data**,
2. the **distribution of estimates across repeated samples**, and
3. the **uncertainty** attached to those estimates.

# Data-Level Distributions

## What a Distribution Tells Us

A distribution describes the pattern of possible values for a variable and how often or how probably those values occur. In earlier chapters we used distributions to describe centre, spread, skewness, and outliers. Here we will see that those same features also determine which probability models are plausible and which inferential procedures are defensible.

The important question is therefore not only "what do the data look like?" but also "what kind of stochastic process could have generated them?" In other words, the biological process that's responsible for generating the data affects the data's distribution.

## Discrete and Continuous Variables

Probability distributions are usually divided into two broad classes.

### Discrete Variables

Discrete variables take countable values such as `0`, `1`, `2`, and so on. They arise when we count events, individuals, or successes.

Common biological examples include:

-   the number of offspring in a brood,
-   the number of infected individuals in a sample,
-   the number of seeds that germinate,
-   the number of birds detected in a transect.

### Continuous Variables

Continuous variables can take any value within a range. They arise when we measure quantities such as length, mass, temperature, or nutrient concentration.

Common examples include:

-   plant height,
-   chlorophyll-*a* concentration,
-   growth rate,
-   body temperature.

This distinction is descriptive and it determines which probability models are appropriate for the data. Counts often lead to binomial, Poisson, or negative-binomial models, whereas continuous responses are more often modelled with normal, gamma, or log-normal distributions.

## Common Distributions in Biology

When is a distribution sensible for a particular model?

### Normal Distribution

Use the normal distribution when the response is continuous, roughly symmetric, and not bounded in a way that strongly distorts the shape. It is often a reasonable model for traits influenced by many small contributing factors. Many basic statistical tests rely on normality either in the response itself or, more often, in model residuals.

### Binomial Distribution

The binomial distribution is typically used when the outcome is the number of successes in a fixed number of independent trials with constant success probability. This is the natural model for germination, occupancy, infection status, and other success/failure data.

### Poisson Distribution

The Poisson distribution is generally used for counts of events in a fixed interval of time, space, or effort when the mean and variance are of similar magnitude. It is often the starting point for count data such as individuals per quadrat or events per hour.

### Negative Binomial Distribution

We will frequently find the negative binomial distribution in count data that are more variable than a Poisson model allows. This is common in ecological field data where aggregation, clustering, or unmeasured heterogeneity inflate the variance.

### Log-Normal and Gamma Distributions

Use these distributions for positive, right-skewed continuous variables such as biomass, waiting times, and concentration data. The gamma distribution is often used when the response is strictly positive and variance changes with the mean. Log-normal models are often useful when multiplicative processes generate strong right skew.

::: {.callout-note appearance="simple"}
## The Process Matters

Ask what process plausibly generated the data. The distribution is often a clue about the process, and the process should inform the model.
:::

# Sampling-Level Reasoning

So far I have described variation in observed data. I now shift to variation in estimates across samples.

## Samples, Populations, and Sampling Variation

Statistical inference depends on the fact that we usually observe a **sample**, not the full **population**. Different samples from the same population will differ. This is **sampling variation**.

If we repeatedly sampled the same population and calculated the sample mean each time, those means would themselves form a distribution. That distribution is the **sampling distribution of the mean**.

Figure @fig-clt shows exactly that shift: the raw population is strongly skewed, but the distribution of sample means changes as the sample size increases.

This distinction is important because:

-   the **data distribution** describes the values of observations,
-   the **sampling distribution** describes the values of a statistic across repeated samples.

Many misunderstandings in introductory statistics come from confusing these two ideas.

## The Central Limit Theorem

One of the most important principles in statistics is the **central limit theorem (CLT)**. In practical terms, it says that the sampling distribution of the sample mean tends toward normality as sample size increases, even when the underlying data are skewed.

This is important because many inferential procedures rely on the behaviour of estimates such as sample means or model coefficients rather than on the raw data alone.

In @fig-clt, panels with small sample sizes still show irregular sampling distributions, while the larger sample sizes produce much more nearly normal distributions of the sample mean.

The CLT leaves poor design, extreme skewness in small samples, and non-independence as real problems. Its main contribution is to explain why normal-based approximations often work surprisingly well for estimates.

```{r fig-clt}
#| fig-cap: "A simulation showing the central limit theorem. The population is strongly right-skewed, but the sampling distribution of the sample mean becomes more nearly normal as sample size increases."
#| fig-width: 5.5
#| fig-height: 4

population <- tibble(x = rgamma(100000, shape = 2, rate = 1))

simulate_means <- function(sample_n, reps = 4000) {
  tibble(
    n = paste("n =", sample_n),
    mean_x = replicate(reps, mean(sample(population$x, size = sample_n)))
  )
}

bind_rows(
  simulate_means(2),
  simulate_means(5),
  simulate_means(25),
  simulate_means(99)
) |>
  mutate(n = factor(n, levels = c("n = 2", "n = 5", "n = 25", "n = 99"))) |>
  ggplot(aes(x = mean_x)) +
  geom_histogram(bins = 30, fill = "salmon", colour = "white") +
  facet_wrap(~n, scales = "free_y") +
  labs(x = "Sample mean", y = "Count")
```

# Quantifying Uncertainty

Once we recognise that estimates vary from sample to sample, we need measures that describe that uncertainty directly.

## Standard Error

The **standard deviation** describes variability among observations in a sample. The **standard error** describes variability in an estimate across repeated samples.

Standard error therefore tells us how much an estimate, such as the sample mean, would vary if we repeated the sampling procedure many times.

For the sample mean:

$$
SE = \frac{SD}{\sqrt{n}}
$$

As sample size increases, the standard error decreases. So, larger samples tend to produce more precise estimates of population parameters.

## Confidence Intervals

A **confidence interval (CI)** gives a range of plausible values for an unknown population parameter. For a 95% CI, the interpretation is tied to repeated sampling: if we repeated the same sampling procedure many times and calculated a CI each time, about 95% of those intervals would contain the true parameter.

Confidence intervals are useful because they show both the estimated value and its precision. A narrow interval suggests greater precision; a wide interval suggests greater uncertainty.

They are often more informative than a bare *p*-value because they show:

-   the scale of the effect,
-   the precision of the estimate,
-   whether biologically meaningful values remain plausible.

For group comparisons, the confidence interval is often calculated for the **difference** between groups. If that interval excludes zero, the result is typically consistent with a statistically detectable difference.

I will use confidence intervals in that way for the first time in [Identify Which Groups Differ](08-anova.qmd#sec-identify-groups-differ), where I interpret Tukey intervals for differences among group means after a significant ANOVA.

# A Simple Example in R

```{r}
x <- c(12.1, 11.8, 12.4, 12.0, 11.9, 12.3, 12.2, 11.7)

n <- length(x)
mean_x <- mean(x)
sd_x <- sd(x)
se_x <- sd_x / sqrt(n)
t_crit <- qt(0.975, df = n - 1)
ci_low <- mean_x - t_crit * se_x
ci_high <- mean_x + t_crit * se_x

tibble(
  n = n,
  mean = mean_x,
  sd = sd_x,
  se = se_x,
  ci_low = ci_low,
  ci_high = ci_high
)
```

This example uses the *t* distribution to calculate a confidence interval for the mean of a small sample. Later, I will connect this thinking directly to hypothesis tests and model coefficients.

In practice, we should read the estimate and the interval together: the mean gives the centre of the result, while the confidence interval tells us how precise that estimate is.

# Linking Distributions to Later Chapters

In this chapter, I work at the bridge between description and inference.

-   [Chapter 2](02-summarise-and-describe.qmd) describes data numerically.
-   [Chapter 3](03-visualise.qmd) describes data graphically.
-   This chapter adds probabilistic structure and uncertainty.
-   [Chapter 5](05-inference.qmd) then uses that structure to formalise inference.

Hypothesis tests and regression models rely on these same ideas of sampling variation and uncertainty.

# Summary

-   Distributions describe the pattern of variation in data.
-   The discrete-versus-continuous distinction helps determine which probability models are appropriate.
-   Biological processes often suggest which distributions are plausible.
-   Statistical inference depends on sampling variation, not just raw data values.
-   The central limit theorem explains why sampling distributions of means are often approximately normal.
-   Standard errors and confidence intervals quantify uncertainty in sample estimates.

Understanding these ideas is essential before moving on to formal hypothesis tests and regression models.
