---
date: "2021-01-01"
title: "7. *t*-Tests"
subtitle: "One-Sample, Two-Sample, and Paired Mean Comparisons"
author: "A. J. Smit"
reference-location: margin
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE,
  fig.asp = NULL, # control via width/height
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_grey(base_size = 8)
)
```

```{r code-knitr-opts-chunk-set, echo=FALSE}
library(tidyverse)
library(ggpubr)
library(ggthemes)
```

```{r fn-two-assum-function-x}
#| echo: false
two_assum <- function(x) {
  x_var <- var(x)
  x_norm <- as.numeric(shapiro.test(x)[2])
  result <- c(x_var, x_norm)
  return(result)
}
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   One-sample *t*-tests
-   Two-sample *t*-tests
-   Paired *t*-tests
-   Rank-based alternatives within the *t*-test family
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   Task F
:::

# Introduction

As biologists, we often want to use sample data to make an inference about a larger population. Sometimes we want to know whether the mean of a single sample differs from an expected value. At other times we want to compare the means of two groups, or the mean change within the same individuals measured twice. The family of tests most commonly used for these questions are *t*-tests.

A *t*-test is therefore not only a test for comparing two groups. More generally, it is a statistical hypothesis test used to assess whether a sample mean, or the difference between two means, is likely to be zero under a null hypothesis. The exact form of the test depends on the structure of the data:

-   in a **one-sample *t*-test** we compare one sample mean, $\bar{x}$, against a known or hypothesised value, $\mu_{0}$;
-   in a **two-sample *t*-test** we compare the means of two independent groups, $\bar{A}$ and $\bar{B}$;
-   in a **paired *t*-test** we compare the mean difference, $\bar{d}$, between matched measurements.

These questions arise frequently in ecology and biology. We might ask whether the mean chlorophyll-*a* concentration in an estuary differs from a regulatory threshold, whether the mean size of plants differs between a control and a treatment, or whether the same individuals are heavier after a feeding trial than before it started.

In all cases, we work with sample data but wish to say something about the population from which the sample was drawn. This means that the chapter must begin with the inferential question, the hypotheses, and the nature of the data before we rush to the function in R.

# Key Concepts

These ideas define the logic of the chapter before we move into the worked tests.

- ***t*-test:** A statistical hypothesis test used to determine whether a sample mean, or a difference between means, is consistent with a null hypothesis.
- **One-sample vs. two-sample vs. paired:** The choice of *t*-test depends on the data structure: one sample against a fixed value, two independent groups against one another, or two related measurements on the same sampling units.
- **One-sided vs. two-sided tests:** A two-sided test asks whether a value differs in either direction, while a one-sided test asks whether the difference is specifically greater than or less than zero.
- **Student versus Welch:** Student's *t*-test assumes equal variances between groups; Welch's *t*-test relaxes this assumption and is the default in R.
- **Assumptions are important:** The appropriateness of a *t*-test depends on the nature of the data, especially the scale of measurement, independence, approximate normality, and in the two-sample case the equality or inequality of variances.
- **Non-parametric alternatives:** When the assumptions for a *t*-test are not met, rank-based alternatives such as `wilcox.test()` may be more appropriate.
- **`t.test()` is the main R function:** The same function handles one-sample, two-sample, and paired designs through arguments such as `mu`, `paired`, `var.equal`, and `alternative`.

# Nature of the Data and Assumptions

The appropriateness of a *t*-test hinges on the nature of the response data and on the study design.

-   **Continuous response variable** The response variable should be measured on a continuous scale, such as length, mass, concentration, or temperature.

-   **Correct data structure** The design determines which *t*-test is appropriate:

    -   **One-sample *t*-test** when one sample is compared to a hypothesised mean, $\mu_{0}$;
    -   **Independent two-sample *t*-test** when two groups are made up of different individuals or sampling units;
    -   **Paired *t*-test** when measurements are matched, such as before-and-after measurements on the same individuals or matched pairs.

-   **Independence** The observations used in the test must be independent in the way assumed by the design. In a two-sample test, observations in one group should not determine those in the other group. In a paired design, the pairs are deliberate, but the pairs themselves should still be independent of one another.

-   **Approximate normality** The data, or more precisely the relevant differences or sampling distribution, should be approximately normally distributed for Student's or Welch's *t*-tests. If this assumption is not met, especially in small samples, you should consider a non-parametric alternative such as the Wilcoxon signed-rank test or the Mann-Whitney test via `wilcox.test()`. This assumption is relaxed in larger samples because of the [Central Limit Theorem](04-distributions-sampling-uncertainty.qmd).

-   **Homoscedasticity for Student's two-sample test** In a classical Student's two-sample *t*-test, the variances of the two groups should be approximately equal. If this assumption is not met, you should use Welch's *t*-test instead. In R, Welch's test is the default, and a Student's test is obtained by setting `var.equal = TRUE`.

-   **Random sampling or random assignment** The data should come from a defensible sampling design so that the observations are representative of the population or experimental process of interest.

Before carrying out a *t*-test, you should examine the data visually and statistically. Histograms, boxplots, Q-Q plots, and tests such as `shapiro.test()` or Levene's test help you decide whether the parametric route is appropriate. When assumptions fail badly, please refer back to [Chapter 6](06-assumptions-and-transformations.qmd) for remedies and alternatives.

# The Core Equations

The purpose of a *t*-test is to compare an observed mean difference with the amount of variation expected from sampling. The precise formula changes with the design, but the teaching approach does not.

For a one-sample *t*-test, the test statistic is:

$$t = \frac{\bar{x} - \mu_0}{s / \sqrt{n}}$$ {#eq-onesample}

In @eq-onesample, $\bar{x}$ is the sample mean, $\mu_0$ is the hypothesised mean under the null hypothesis, $s$ is the sample standard deviation, and $n$ is the sample size. The numerator is the difference we want to test, while the denominator scales that difference by the expected sampling variability.

For Student's two-sample *t*-test, where a common variance is assumed, the same approach becomes:

$$t = \frac{\bar{A} - \bar{B}}{s_p\sqrt{\frac{1}{n_A} + \frac{1}{n_B}}}$$ {#eq-student-t}

The pooled standard deviation in @eq-student-t is:

$$s_p^2 = \frac{(n_A - 1)s_A^2 + (n_B - 1)s_B^2}{n_A + n_B - 2}$$ {#eq-pooledvar}

Here $\bar{A}$ and $\bar{B}$ are the sample means, $s_A^2$ and $s_B^2$ are the sample variances, and $n_A$ and $n_B$ are the sample sizes of the two groups. Student's formulation is useful when the two groups are reasonably similar in variance.

For Welch's two-sample *t*-test, which does not assume equal variances, the test statistic is:

$$t = \frac{\bar{A} - \bar{B}}{\sqrt{\frac{s_A^2}{n_A} + \frac{s_B^2}{n_B}}}$$ {#eq-welch}

The associated degrees of freedom are approximated by the Welch-Satterthwaite expression:

$$\text{d.f.} = \frac{\left(\frac{s_A^2}{n_A} + \frac{s_B^2}{n_B}\right)^2}{\frac{(s_A^2 / n_A)^2}{n_A - 1} + \frac{(s_B^2 / n_B)^2}{n_B - 1}}$$ {#eq-satterthwaite}

For a paired *t*-test, the data are first reduced to within-pair differences, $d_i$, and the test is then carried out as a one-sample *t*-test on those differences:

$$t = \frac{\bar{d}}{s_d/\sqrt{n}}$$ {#eq-paired-t}

Here $\bar{d}$ is the mean of the paired differences, $s_d$ is the standard deviation of those differences, and $n$ is the number of pairs. This is why paired designs should always be understood as tests of the mean difference, not as tests of two independent sample means.

# R Functions

The main function used throughout this chapter is `t.test()`. It can be used for:

-   one-sample *t*-tests by specifying `mu = ...`,
-   independent two-sample *t*-tests using either `x, y` or a formula such as `response ~ group`,
-   paired *t*-tests by specifying `paired = TRUE`.

The most important arguments are:

-   `mu` for the hypothesised mean in a one-sample test,
-   `alternative` to choose `"two.sided"`, `"less"`, or `"greater"`,
-   `paired` to indicate matched observations,
-   `var.equal` to request Student's two-sample *t*-test rather than Welch's.

When the assumptions for a *t*-test are not met, the main alternative used in this chapter is `wilcox.test()`. For one-sample and paired settings it serves as a rank-based alternative to the *t*-test, and for two independent groups it provides the Mann-Whitney or Wilcoxon rank-sum approach, depending on the argument structure.

# A Simple Experiment

::: callout-important
## Do This Now!
**Task F.1.** Work with two friends and devise a real 'experiment' (or designed sampling to determine some real-world phenomenon, more likely) in which you demonstrate the following two principles:

a. the effect of sample size on the estimate of the mean and variance (use SD),
b. the effect of repeated sampling on the value of the mean and variance. In both cases, apply the correct *t*-test (later in this Chapter) to test whether sample size and repeat sampling has a statistically significant effect. What conclusion do you reach?

Report back comprehensively with Task F.
:::

The simplest experiment that makes any scientific sense is one where data are collected to represent two groups --- maybe a treatment compared to a control. For example, "does adding manure to some plants have a measurable effect on their growth compared to the situation where nothing was added that might affect the plants' growth?" To answer this very basic question one often uses a ***t*-test**. 

The most commonly used *t*-test is **Student's *t*-test**[^1], a statistical hypothesis test used to determine whether two samples of data are significantly different from each other. It was introduced by William Gosset, who wrote under the pseudonym 'Student,' in 1908. This is a parametric test, and as already indicated, it assumes the data are normally distributed, that the variances of the two samples being compared are equal.

[^1]: Whereas the Student's *t*-test applies to data whose variances are equal among the two groups, the **Welch's *t*-test** is used when the variances are unequal. The Welch's *t*-test is a more robust test, is preferred when the sample sizes are unequal or the variances are unequal. The Welch's *t*-test is the default setting in the`t.test()` function in R (through the argument `var.equal = FALSE`, which is the default).

Now it becomes interesting. There are several flavours of *t*-tests, depending on the nature of our samples, the type of question being asked but all of them involve comparing two samples. Specifically, *the comparison involves the sample means*, and as such, they test whether or not the central tendencies of two samples differ. Here are the variations:

-   **One-sample *t*-tests:** only one sample set of data that we wish to compare against a known population mean:

    -   one-sided one-sample *t*-tests
    -   two-sided one-sample *t*-tests

-   **Two-sample (or independent sample) *t*-tests:** the means of two groups are compared against each other:

    -   one-sided two-sample *t*-tests
    -   two-sided two-sample *t*-tests

-   **Paired sample *t*-tests:** two samples that are 'matched' or as the name indicates, *paired*, *i.e.* they are not independent of one another:

    -   one-sided
    -   two-sided

Let us go into detail into each of these variations of *t*-tests.

# One-Sample *t*-Tests

Sometimes we have only one sample, but we still want to make an inference about a population. In that case the question is not whether two groups differ from one another, but whether the mean of our sample differs from some hypothesised value, $\mu_{0}$. That hypothesised value may come from a regulatory threshold, a historical mean, a theoretical expectation, or a biologically meaningful reference point.

The one-sample *t*-test asks whether the difference between the sample mean, $\bar{x}$, and the hypothesised mean, $\mu_{0}$, is large relative to the variation in the sample, as formalised in @eq-onesample. As in the one-sample hypothesis chapter in `BCB_Stats`, the logic should always proceed in the same order: first decide whether the one-sample design is appropriate, then state the hypotheses, then look carefully at the data, and only then run `t.test()`.

## When to Use a One-Sample *t*-Test

Use a one-sample *t*-test when:

-   you have one sample of observations;
-   the response variable is continuous;
-   the observations are independent;
-   you want to compare the sample mean against a known or hypothesised value, $\mu_{0}$;
-   the data are approximately normally distributed, or the sample size is large enough that the test is reasonably robust.

If you already know the direction of the expected effect before seeing the data, you may use a one-sided test. If you simply want to know whether the mean differs from $\mu_{0}$ in either direction, use a two-sided test.

## Two-Sided One-Sample *t*-Test

For a two-sided one-sample *t*-test, the hypotheses are:

$$H_{0}: \bar{x} = \mu_{0}$$
$$H_{a}: \bar{x} \ne \mu_{0}$$

This is the same as:

$$H_{0}: \bar{x} - \mu_{0} = 0$$
$$H_{a}: \bar{x} - \mu_{0} \ne 0$$

Here $\bar{x}$ is the sample mean and $\mu_{0}$ is the hypothesised population mean against which the sample is being compared.

Because this is a two-sided test, the alternative hypothesis does not specify direction. We are asking whether $\bar{x}$ differs from $\mu_{0}$, not whether it is specifically less than or greater than $\mu_{0}$.

In R, a one-sample *t*-test is done with `t.test(x, mu = ...)`. The default is a two-sided test, so unless you specify otherwise, `t.test()` evaluates whether the sample mean differs from $\mu_{0}$ in either direction.

# Example 1: A One-Sample Comparison Against a Reference Mean

```{r code-set-seed}
# create a single sample of random normal data
set.seed(666)
r_one <- data.frame(dat = rnorm(n = 20, mean = 20, sd = 5),
                    sample = "A")
``` 

We will work with one sample of random normal data and compare its mean against two possible reference values. The purpose of the example is to run the test and to see how the result depends on the hypothesised mean, $\mu_{0}$.

## Do an Exploratory Data Analysis (EDA)

Before fitting the test, we should inspect the data numerically and graphically.

```{r code-r-one-summary}
summary(r_one$dat)
sd(r_one$dat)
shapiro.test(r_one$dat)
```

The sample is small, so it is useful to check both the summary statistics and the approximate normality of the observations. We should also look at a plot.

```{r fig-boxplot1}
#| fig-cap: "Boxplot of random normal data with. A hypothetical (reference) population mean of 20 is shown as a blue line, with the red line showing another reference mean of 30."
#| fig-width: 4
#| fig-height: 3

ggplot(data = r_one, aes(y = dat, x = sample)) +
  geom_boxplot(fill = "indianred", notch = TRUE,
               alpha = 0.3, colour = "black") +
  # population  mean (mu) = 20
  geom_hline(yintercept = 20, colour = "dodgerblue2", 
             size = 0.9) +
  # population  mean (mu) = 30
  geom_hline(yintercept = 30, colour = "indianred2", 
             size = 0.9) +
  labs(y = "Value", x = NULL) +
  coord_flip()
```

The boxplot shows the sample distribution relative to two possible hypothesised means. Visually, the sample appears to be centred much closer to 20 than to 30. That immediately suggests that a test against 20 may not be significant, whereas a test against 30 may be.

## State the Hypotheses

Suppose we first ask whether this sample could reasonably have come from a population with mean 20. The hypotheses are:

$$H_{0}: \mu = 20$$
$$H_{a}: \mu \ne 20$$

We can then ask the same question for a hypothesised population mean of 30:

$$H_{0}: \mu = 30$$
$$H_{a}: \mu \ne 30$$

## Apply the Test

```{r code-t-test-r-one-dat}
# compare random data against a population mean of 20
t.test(r_one$dat, mu = 20)

# compare random data against a population mean of 30
t.test(r_one$dat, mu = 30)
```

## Interpret the Results

The test against $\mu_{0} = 20$ asks whether the observed sample mean is far enough from 20, relative to the sample variation, to reject the null hypothesis. The test against $\mu_{0} = 30$ asks the same question, but now the hypothesised mean is much farther from the centre of the observed data.

This is the key inferential logic of the one-sample *t*-test: the same data can support different conclusions depending on the null value being tested. The output should therefore always be interpreted in relation to the biological or experimental meaning of $\mu_{0}$.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

A one-sample *t*-test was used to compare the sample mean with a hypothesised reference value. Separate tests were considered for two biologically distinct null values, $\mu_0 = 20$ and $\mu_0 = 30$, to illustrate how inference depends on the chosen reference mean.

**Results**

The mean of the sample ($\bar{x} = 20.01$, SD = 6.61, $n = 20$) did not differ from the hypothesised value of $\mu_0 = 20$ (one-sample *t*-test: $t_{19} = 0.005$, $p > 0.05$). The 95% confidence interval for the mean (16.91 to 23.10) included the reference value, indicating that the observed data are consistent with a population mean of 20. In contrast, the sample mean differed from a hypothesised value of $\mu_0 = 30$ ($t_{19} = -6.76$, $p < 0.001$). The same confidence interval did not include 30, indicating that the observed values are not compatible with a population mean of 30.

**Discussion**

The inference depends on the biological meaning of the null value. These data support a mean near 20, but they provide strong evidence against a mean as large as 30.
:::



# One-Sided One-Sample *t*-Tests

For example, when we are concerned that our sample mean, $\bar{x}$, should be less than the *a priori* established value, $\mu_{0}$, the hypotheses are:

$$H_{0}: \bar{x} \ge \mu_{0}$$
$$H_{a}: \bar{x} \lt \mu_{0}$$

Only one of the two directional options is shown here, but the same logic applies if the alternative is that the mean should be greater than $\mu_{0}$.

One-sided tests are used when the direction of the alternative hypothesis is specified before the data are analysed. For example, we may wish to know whether the sample mean is specifically less than $\mu_{0}$, or specifically greater than $\mu_{0}$. This must be justified by the scientific question.

In R, one-sided one-sample tests are requested with the `alternative` argument:

```{r code-t-test-r-one-dat-2}
# check against the trailing tail
t.test(r_one$dat, mu = 30, alternative = "less")

# check against the leading tail
t.test(r_one$dat, mu = 30, alternative = "greater")
```

Are these the results we would have expected? Why does the second test not return a significant result? The answer lies in the direction of the alternative hypothesis: the first test asks whether the mean is less than 30, which is compatible with the data, whereas the second asks whether the mean is greater than 30, which is not.

# If One-Sample Assumptions Fail

If the observations are strongly non-normal, especially in small samples, or if the data contain severe outliers that undermine the mean-based logic of the test, a one-sample *t*-test may not be appropriate. In such cases, a rank-based alternative such as `wilcox.test()` may be more defensible. The broader reasoning behind the checking of assumptions and deciding on alternatives is discussed in [Chapter 6](06-assumptions-and-transformations.qmd).

::: callout-important
## Do This Now!
Create a visualisation to graphically demonstrate the outcome of this *t*-test.
:::

# Two-Sample *t*-Tests

Very often, however, the inferential question is not whether one sample differs from a fixed value, but whether the means of two groups differ from one another. This happens when we compare a control against a treatment, one habitat against another, or one species against another. The response variable is still continuous, but now the grouping variable has two levels.

In this case we use a **two-sample *t*-test**, also called an **independent-samples *t*-test**, because the observations in the two groups are assumed to be independent. As with the one-sample test, the logic must still proceed in the same order: decide whether the design is appropriate, state the hypotheses, examine the data, choose the correct version of the test, and then interpret the result in biological terms.

## When to Use a Two-Sample *t*-Test

Use a two-sample *t*-test when:

-   you have a continuous response variable;
-   you have two groups to compare;
-   the observations in one group are independent of those in the other group;
-   the data in each group are approximately normal, or the sample sizes are large enough for the test to be robust;
-   you are interested in whether the difference between the group means is zero.

At this point we must also decide which two-sample *t*-test is appropriate. If the two groups have approximately equal variances, we may use **Student's *t*-test**. If the variances differ, we use **Welch's *t*-test**, which is more robust and is the default in R.

## Two-Sided Two-Sample *t*-Test

For a two-sided two-sample *t*-test, the hypotheses are:

$$H_{0}: \bar{A} = \bar{B}$$
$$H_{a}: \bar{A} \ne \bar{B}$$

where $\bar{A}$ is the population mean of the first sample and $\bar{B}$ the population mean of the second sample. In this example we have a two-sided two-sample *t*-test, which is the default in R's `t.test()` function.

This is the same as saying:

$$H_{0}: \bar{A} - \bar{B} = 0$$
$$H_{a}: \bar{A} - \bar{B} \ne 0$$

Again, $H_{0}$ does not distinguish between whether the difference is greater than or less than 0. We are simply asking whether the two means differ.

This is the most common *t*-test in experimental biology. We use it when our two samples come from different individuals or sampling units, and when there is no natural pairing between the observations.

## Student's Versus Welch's *t*-Test

There are two main varieties of the two-sample *t*-test. In the case of samples whose variances do not differ appreciably, we may perform a **Student's *t*-test**, which uses the pooled-variance formulation in @eq-student-t and @eq-pooledvar. If the variances differ between the two groups, we instead perform **Welch's *t*-test** as written in @eq-welch and @eq-satterthwaite. Welch's *t*-test is more robust and is therefore the default in R's `t.test()` function.

In the olden days, we had to calculate the *t*-statistic and degrees of freedom by hand and then read probabilities from a printed table. R now does this for us automatically, but the formulas still matter because they explain why Student's and Welch's tests differ.

# Example 2: Two Independent Groups

Let us generate some new random normal data and ask whether the values in two independent groups differ from one another.

```{r code-set-seed-2}
# random normal data
set.seed(666)
r_two <- data.frame(dat = c(rnorm(n = 20, mean = 4, sd = 1),
                            rnorm(n = 20, mean = 5, sd = 1)),
                    sample = c(rep("A", 20), rep("B", 20)))
```

## Do an Exploratory Data Analysis (EDA)

As before, we begin by exploring the data before fitting a model.

```{r code-r-two-summary}
r_two |>
  group_by(sample) |>
  summarise(
    n = n(),
    mean = mean(dat),
    sd = sd(dat),
    .groups = "drop"
  )

by(r_two$dat, r_two$sample, shapiro.test)
```

The summaries tell us whether the two groups differ in their sample means and whether their spread appears similar. We should also inspect the groups visually.

```{r fig-boxplot2}
#| fig-cap: "Boxplots of two independent samples. Group B has a slightly higher mean than Group A, but the spread is similar."
#| fig-width: 4
#| fig-height: 3

ggplot(data = r_two, aes(x = sample, y = dat, fill = sample)) +
  geom_boxplot(alpha = 0.35, colour = "black", outlier.shape = NA) +
  geom_jitter(width = 0.08, alpha = 0.7, size = 1.8) +
  labs(x = "Group", y = "Value") +
  theme(legend.position = "none")
```

The boxplots suggest that the two groups are roughly similar in spread, but that Group B may have a higher mean than Group A. Because these data were simulated with the same standard deviation, a Student's *t*-test is appropriate here.

## State the Hypotheses

For a two-sided comparison, the hypotheses are:

$$H_{0}: \mu_{A} = \mu_{B}$$
$$H_{a}: \mu_{A} \ne \mu_{B}$$

Equivalently:

$$H_{0}: \mu_{A} - \mu_{B} = 0$$
$$H_{a}: \mu_{A} - \mu_{B} \ne 0$$

## Apply the Test

Now we apply the *t*-test. Note how we set the `var.equal` argument to `TRUE` because we know these simulated data have the same standard deviation.

```{r code-t-test-r-two}
# perform t-test
# note how we set the `var.equal` argument to TRUE because we know
# our data has the same SD (they are simulated as such!)
t.test(dat ~ sample, data = r_two, var.equal = TRUE)

# if the variances are not equal, simply set `var.equal` to false
# and a Welch's t-test will be performed
```

The first argument we see in `t.test()` is `dat ~ sample`. Usually in R when we see a `~` (tilde), we are creating what is known as a formula. A formula tells R how it should look for relationships between a response and a grouping or explanatory variable. In our code above, `dat ~ sample` means that the values in the `dat` column are being compared across the levels of the `sample` column. We will see this same formula notation repeatedly later in ANOVAs and linear models.

## Interpret the Results

The two-sample *t*-test asks whether the observed difference between the group means is large relative to the variation within the groups. If the *p*-value is small, we reject the null hypothesis that the mean difference is zero. If the *p*-value is not small, we conclude that the data do not provide strong evidence for a difference in means.

The confidence interval is also important. It gives an estimated range for the mean difference between the two groups. If this interval excludes zero, the result is consistent with a statistically significant difference.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

Group means were compared with a two-sample Student's *t*-test because the simulated data were generated with similar variances and equal sample sizes. Group means, standard deviations, and the confidence interval for the mean difference were used to support interpretation of the test outcome.

**Results**

The mean value in Group A was $4.00 \pm 1.32$ (SD, $n = 20$), while the mean value in Group B was $4.75 \pm 1.08$ (SD, $n = 20$). A two-sample Student's *t*-test indicated that this difference was not statistically significant at the 0.05 level ($t_{38} = -1.95$, $p > 0.05$). The 95% confidence interval for the mean difference between the groups ranged from $-1.52$ to $0.03$, which included zero.

**Discussion**

Although Group B tended to have the higher mean, the evidence was not quite strong enough to conclude that the population means differ. In a manuscript, the observed pattern and the inferential conclusion should be kept distinct.
:::

::: callout-important
## Do This Now!
Create a visualisation to graphically demonstrate the outcome of this *t*-test.
:::

# One-Sided Two-Sample *t*-Test

For example, when we are concerned that the sample mean of the first population, $\bar{A}$, should be greater than that of the second, $\bar{B}$, the hypotheses are:

$$H_{0}: \bar{A} \le \bar{B}$$
$$H_{a}: \bar{A} \gt \bar{B}$$

Only one of the two directional formulations is shown here.

Just as with the one-sample *t*-tests above, we may also specify which tail of the distribution we are interested in when we compare the means of two samples. This is a **one-sided two-sample *t*-test**. Here too we may use either Student's or Welch's form, depending on the variances. The direction of the test must be justified before looking at the results.

In R we do this with the `alternative` argument:

```{r code-compare-means-dat-sample-data}
# is the mean of sample B smaller than that of sample A?
compare_means(dat ~ sample, data = r_two,
              method = "t.test", var.equal = TRUE,
              alternative = "less")

# is the mean of sample B greater than that of sample A?
compare_means(dat ~ sample, data = r_two,
              method = "t.test", var.equal = TRUE,
              alternative = "greater")
```

What do these results show? Is this surprising? It should not be. The data were generated so that Group B tends to be larger than Group A, so a directional test framed that way is more likely to support the alternative than one framed in the opposite direction.

# If Two-Sample Assumptions Fail

If the group variances differ, use Welch's *t*-test rather than Student's *t*-test. If the data are badly non-normal, strongly skewed, or dominated by outliers, a rank-based alternative such as `wilcox.test()` may be preferable. The broader workflow for checking assumptions and choosing alternatives is dealt with in [Chapter 6](06-assumptions-and-transformations.qmd).

# Paired *t*-Tests

Sometimes the two measurements we want to compare do not come from different individuals or sampling units, but from the same units measured twice, or from units that have been deliberately matched. In such cases the observations are not independent across the two groups, and a standard two-sample *t*-test is no longer appropriate.

The logic of the paired *t*-test is therefore slightly different. We do not compare two independent means directly. Instead, we calculate the difference within each pair, and then ask whether the mean of those differences is zero. In other words, the paired *t*-test is really a one-sample *t*-test applied to the set of paired differences.

## When to Use a Paired *t*-Test

Use a paired *t*-test when:

-   the response variable is continuous;
-   each observation in one group is naturally matched to one observation in the other group;
-   the pairing is meaningful, such as before-and-after measurements on the same individuals, left-versus-right measurements on the same organism, or deliberately matched experimental units;
-   the paired differences are approximately normally distributed.

The crucial point is that the analysis focuses on the **differences within pairs**, not on the raw values considered as though they came from independent groups.

For a two-sided paired *t*-test, the hypotheses are:

$$H_{0}: \bar{d} = 0$$
$$H_{a}: \bar{d} \ne 0$$

where $\bar{d}$ is the mean population difference based on the matched pairs.

Again, here $H_{0}$ does not consider if differences are greater than or less than 0 --- a difference in either direction is significant. One-sided versions of the hypothesis can be stated as well.

Paired *t*-tests are therefore used when the dependence between observations is not a problem to be ignored, but the central feature of the design. This test is also sometimes called a dependent-samples *t*-test or a repeated-measures *t*-test.

For example, we might want to determine whether the right arms of a group of people differ in length from their left arms. For each individual, the right arm measurement is matched to the left arm measurement from the same person. Or, we may wish to determine whether piglets weigh more after a feeding trial than they did at the start of the trial. Each individual piglet then contributes a before-and-after pair of measurements.

# Example 3: Before-and-After Measurements on Piglets

Let us use a small example in which piglets are weighed before and after a three-week feeding period. The inferential question is whether body mass increases over time within the same individuals.

```{r code-paired-piglets}
set.seed(666)
piglets <- data.frame(
  id = 1:12,
  before = round(rnorm(12, mean = 8.2, sd = 0.7), 2),
  after = round(rnorm(12, mean = 9.0, sd = 0.7), 2)
)

piglets$diff <- piglets$after - piglets$before
piglets
```

## Do an Exploratory Data Analysis (EDA)

Before running the paired *t*-test, we should inspect the paired differences because these are the values that the test uses directly.

```{r code-paired-summary}
summary(piglets$diff)
sd(piglets$diff)
shapiro.test(piglets$diff)
```

```{r fig-paired-differences}
#| fig-cap: "Distribution of paired differences in piglet body mass after minus before the feeding period."
#| fig-width: 4
#| fig-height: 3

ggplot(data = piglets, aes(x = diff)) +
  geom_histogram(binwidth = 0.5, fill = "indianred", colour = "black", alpha = 0.5) +
  geom_vline(xintercept = 0, colour = "dodgerblue2", linewidth = 0.9) +
  labs(x = "Paired difference in mass (after - before)", y = "Frequency")
```

The distribution of paired differences is centred above zero, which already suggests that the piglets tended to be heavier after the feeding period. The Shapiro-Wilk test also indicates that the differences are reasonably consistent with normality, so a paired *t*-test is appropriate.

## State the Hypotheses

For a two-sided paired *t*-test, the hypotheses are:

$$H_{0}: \mu_{d} = 0$$
$$H_{a}: \mu_{d} \ne 0$$

where $\mu_{d}$ is the population mean of the paired differences. If we had a strictly directional biological question, such as whether piglets are heavier after feeding than before, we could also frame this as a one-sided test.

## Apply the Test

```{r code-paired-test}
t.test(piglets$after, piglets$before, paired = TRUE)
```

We can also use the formula interface with paired data if the data are arranged appropriately, but in practice the paired vectors are often the clearest way to express the analysis.

## Interpret the Results

The paired *t*-test asks whether the mean of the within-individual differences is zero. In this example the mean paired difference is positive, which indicates that the piglets tended to gain mass over the feeding period. Because the confidence interval for the mean difference lies just above zero and the *p*-value is near 0.05, the evidence suggests a modest but statistically detectable increase in mass.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

Piglet body mass was measured before and after a three-week feeding period, and the resulting paired observations were analysed with a paired *t*-test. The test was based on the within-individual differences in mass, with normality assessed on those paired differences.

**Results**

Piglet body mass increased over the three-week feeding period. A paired *t*-test showed that the mean within-individual change in body mass differed significantly from zero ($t_{11} = 2.21$, $p < 0.05$). The mean paired difference was 0.815 kg, with a 95% confidence interval from 0.004 to 1.626 kg, indicating that the piglets were, on average, heavier after the feeding period than at the start.

**Discussion**

The paired design strengthens the inference because each piglet acts as its own control. In a Discussion section, the emphasis therefore falls on biological change within individuals rather than on a comparison of two independent group means.
:::

# If Paired-Test Assumptions Fail

If the paired differences are strongly non-normal, especially in a small sample, the paired *t*-test may not be appropriate. In that case a Wilcoxon signed-rank test via `wilcox.test(..., paired = TRUE)` may be more defensible. As before, the broader workflow for evaluating assumptions is discussed in [Chapter 6](06-assumptions-and-transformations.qmd).

::: callout-important
## Do This Now!
**Task F.3.** Divide the class into two groups. In each group, collect your own data that are suitable for analysis with a paired *t*-test. State your hypothesis. Test it. Discuss. Report back comprehensively in Task F.
:::

# Rank-Based Alternatives Within the *t*-Test Family

The parametric *t*-tests in this chapter are mean-based procedures. When the response is badly non-normal in a small sample, strongly skewed, or dominated by influential outliers, a rank-based alternative may be more defensible. The important point is that these methods belong to the same **inferential family** as the *t*-tests. They answer closely related questions under the same one-sample, two-sample, or paired designs, but they do so through ranks rather than raw values.

The three main rank-based counterparts are:

-   **one-sample Wilcoxon signed-rank test** for a single sample compared with a reference value;
-   **Wilcoxon rank-sum test** for two independent groups;
-   **paired Wilcoxon signed-rank test** for matched or before-after observations.

All three are handled by `wilcox.test()` in R. The design still determines the correct analysis:

-   `wilcox.test(x, mu = ...)` for a one-sample problem;
-   `wilcox.test(x, y)` or `wilcox.test(response ~ group, data = ...)` for two independent groups;
-   `wilcox.test(x, y, paired = TRUE)` for paired data.

These tests are useful when the assumptions for the corresponding *t*-test fail badly and when the biological question can still be answered sensibly from a rank-based comparison. They also change the interpretation from a focus on means to the median in the same direct way as a *t*-test.

::: {.callout-note appearance="simple"}
## Use the Design, Not Just the Function Name

The one-sample and paired Wilcoxon procedures are often confused because they use the same function. The distinction is exactly the same as for *t*-tests:

-   one sample against a reference value → one-sample signed-rank test;
-   two related measurements on the same units → paired signed-rank test;
-   two independent groups → rank-sum test.
:::

# A *t*-Test Workflow

Now that we have seen how to compare the means of two sample sets of data, let us see what that complete workflow would look like in R. For this example we will use the `ecklonia` data from [Intro R Workshop: Data Manipulation, Analysis, Graphing](https://robwschlegel.github.io/Intro_R_Workshop/).

## Loading Data

Before we can run any analyses we will need to load our data. We are also going to convert these data from their wide format into a long format because this is more useful for the rest of our workflow.

```{r code-ecklonia-read-csv-here-here}
ecklonia <- read_csv(file.path("..", "..", "data", "BCB744", "ecklonia.csv")) %>% 
  gather(key = "variable", value = "value", -species, -site, -ID)
```

## Visualising Data

With our data loaded, let us visualise them in order to ensure that these are indeed the data we are after (@fig-ecklonia1). Visualising the data will also help us to formulate a hypothesis.

```{r fig-ecklonia1}
#| fig-cap: "Boxplots showing differences in morphometric properties of the kelp _Ecklonia maxima_ at two sites in False Bay."

ggplot(data = ecklonia, aes(x = variable, y = value, fill = site)) +
  geom_boxplot(colour = "black", fill = "dodgerblue4", alpha = 0.4) +
  coord_flip()
```

The first thing we should notice from the figure above is that our different measurements are on very different scales. This makes comparing all of our data visually rather challenging. Even given this complication, one should readily be able to make out that the measurement values at Batsata Rock appear to be greater than at Boulders Beach. Within the framework of the scientific process, that is what we would call an 'observation', and is the first step towards formulating a hypothesis. The next step is to refine our observation into a hypothesis. By what measurement are the kelps greater at one site than the other?

## Formulating a Hypothesis

Looking at the figure above it appears that for almost all measurements of length, Batsata Rock far exceeds that of Boulders Beach however, the stipe masses between the two sites appear to be more similar. Let us pull out just this variable, create a new boxplot (@fig-ecklonia2).

```{r fig-ecklonia2}
#| fig-cap: "Boxplots showing the difference in stipe mass (kg) of the kelp _Ecklonia maxima_ at two sites in False Bay."

# filter the data
ecklonia_sub <- ecklonia %>% 
  filter(variable == "stipe_mass")

# then create a new figure
ggplot(data = ecklonia_sub, aes(x = variable, y = value, fill = site)) +
  geom_boxplot(colour = "black", alpha = 0.4) +
  coord_flip() +
  labs(y = "Stipe mass (kg)", x = "") +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank())
```

Now we have a more interesting comparison at hand. The question I think of when I look at these data is "Are the stipe masses at Batsata Rock greater than at Boulders Beach?". The hypothesis necessary to answer this question would look like this:

$$H_{0}: \bar{A} \leq \bar{B}$$
$$H_{a}: \bar{A} > \bar{B}$$

In words, the null hypothesis says that stipe mass at Batsata Rock is not greater than at Boulders Beach, while the alternative says that it is.

Which test must we use for this hypothesis?

## Choosing a Test

Before we can pick the correct statistical test for our hypothesis, we need to be clear on what it is we are asking. Starting with the data being used is usually a good first step. As we may see in the above figure, we have two sample sets that we are comparing. Therefore, unsurprisingly, we will likely be using a *t*-test. But we are not done yet. How is it that we are comparing these two sample sets? Remember from the examples above that there are multiple different ways to compare two sets of data. For our hypothesis we want to see if the stipe mass at Batsata Rock is **greater than** the stipe mass at Boulders Beach, not just that they are different. Because of this we will need a one-sided *t*-test. But wait, there is more! We have zeroed in on which sort of test would be appropriate for our hypothesis, but before we run it we need to check our assumptions.

## Checking Assumptions

In case we forgot, here are the assumptions for a *t*-test:

-   the dependent variable must be **continuous**,
-   the observations in the groups being compared are **independent** of each other,
-   the data are **normally distributed**,
-   that the data are **homoscedastic**, and in particular, that there are no outliers.

We know that the first two assumptions are met because our data are measurements of mass at two different sites. Before we can run our one-sided *t*-test we must meet the last two assumptions. Lucky us, we have a function tat will do that automagically.

Please refer to [Chapter 6](06-assumptions-and-transformations.qmd) to see what to do if the assumptions fail.

```{r code-ecklonia-sub}
ecklonia_sub %>% 
  group_by(site) %>% 
  summarise(stipe_mass_var = two_assum(value)[1],
            stipe_mass_norm = two_assum(value)[2])
```

Lovely. The variances are equal and the data are normal. On to the next step.

## Running an Analysis

With our assumptions checked, we may now analyse our data. We will see below how to do this with both of the functions we have learned in this chapter for comparing means of two sample sets.

```{r code-t-test-value-site}
t.test(value ~ site, data = ecklonia_sub,
       var.equal = TRUE, alternative = "greater")
```

## Interpreting the Results

We may reject the null hypothesis that the stipe mass of kelps at Batsata Rock is not greater than at Boulders Beach if our *t*-test returns a *p*-value $\leq$ 0.05. We must also pay attention to some of the other results from our *t*-test, specifically the *t*-value (`t`) and the degrees of freedom (`df`) as these are also needed when we are writing up our results. From all of the information above, we may accept the alternative hypothesis. But how do we write that up?

## Drawing Conclusions

There are many ways to present one's findings. Style, without *too much* flourish, is encouraged as long as certain necessary pieces of information are provided. In a scientific paper, the results section would normally report the sample means, the test used, the test statistic, degrees of freedom, the *p*-value, and the biological interpretation in one compact paragraph.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

Stipe mass of *Ecklonia maxima* was compared between Batsata Rock and Boulders Beach using a one-sided two-sample Student's *t*-test. The analysis treated site as the grouping factor and stipe mass as the continuous response variable, after checking that the sample variances were similar and that the data were approximately normally distributed.

**Results**

Mean stipe mass of *Ecklonia maxima* was higher at Batsata Rock ($6.12 \pm 1.41$ kg SD, $n = 13$) than at Boulders Beach ($5.00 \pm 1.63$ kg SD, $n = 13$). A one-sided two-sample Student's *t*-test indicated that stipe mass at Batsata Rock was significantly greater than at Boulders Beach ($t_{24} = 1.87$, $p < 0.05$). The 95% confidence interval for the mean difference was $0.10$ kg to $\infty$, supporting the conclusion that kelp at Batsata Rock had greater stipe mass.

**Discussion**

The biological conclusion is that kelps at Batsata Rock had heavier stipes than those at Boulders Beach in this comparison. A Discussion section would then move beyond the statistical result itself and ask what local ecological or hydrodynamic factors might account for that difference.
:::

## Going Further

But why though? As is often the case in life, and science is no exception, answers to our questions just create even more questions! Why would the mass of kelp stipes at one locations in the same body of water, only a kilometre or so apart be significantly different? It looks like we are going to need to design a new experiment... Masters thesis anyone?

# Conclusion

*t*-tests are one of the first classes of inferential statistics students typically lerarn in a module on biostatistics. They are suitable for situations where you have two groups of data to compare and to establish if the difference between their means is zero. Lesson learned here are generalisable to more complex analysis.
