---
title: "9. Correlation and Association"
subtitle: "Quantifying Relationships Without Imposing a Response Model"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
params:
  hide_answers: true
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_grey(base_size = 8)
)
```

```{r code-libraries, echo=FALSE}
library(tidyverse)
library(ggpubr)
library(corrplot)
library(gt)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   correlation as a measure of association
-   choosing Pearson, Spearman, or Kendall
-   correlation matrices and heatmaps
-   correlation and non-independence
-   correlation versus regression
::: 
::: {.callout-note appearance="simple"}
## Cheatsheet

Find here a [Cheatsheet](../../docs/Methods_cheatsheet_v1.pdf) on statistical methods.
::: 

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   Self-Assessment Task 9-1
::: 

# Introduction

In the previous chapters, I asked whether means differ among groups. Correlation answers, do two variables vary together?

Here we concern ourselves with **association** rather than group comparison. It is the last of the staple inferential tools and prepares the transition into regression. Correlation does not impose a response model and it simply measures the strength and direction of association between two variables. As we shall see later, the purpose of regression is to have additional expectations about the *roles* of the variables: one variable is the response and the other predicts.

Correlation coefficients are effect sizes. Their sign shows direction and their magnitude shows how tightly the variables vary together. Coefficients vary from `-1.0` (perfect inverse correlation), to `0` (no association), to `1.0` (perfect positive correlation). 

# Choosing the Appropriate Correlation

The main decision we will face is the form of the relationship.

-   Use **Pearson's correlation** when the relationship is approximately linear and the main question concerns linear co-variation.
-   Use **Spearman's rank correlation** when the relationship is monotonic but not especially linear, or when ranked order is more informative than raw spacing.
-   Use **Kendall's rank correlation** when the emphasis is on concordance in rank order, especially with ordinal data, many tied ranks, or a direct question about agreement between rankings.

Start with a scatterplot and then decide whether the pattern is:

-   roughly linear;
-   monotonic but curved or unevenly spaced;
-   clustered, outlier-driven, or structured by site, time, or some other grouping.

Association does not imply causation. Correlation can reveal co-variation, but it cannot identify the mechanism behind it.

## How to Read Magnitude

Rough verbal labels can help, but they are only a starting point:

-   around `0.1` to `0.3`: weak association;
-   around `0.3` to `0.7`: moderate association;
-   above `0.7`: strong association.

Those labels are only heuristics. A coefficient of `0.4` may be unremarkable in one system and biologically substantial in another, so it is highly context specific. A moderate correlation may be biologically useful if the variables are noisy or hard to measure. A strong correlation may still be uninformative if it is driven by site structure, repeated measurements, or a lurking third variable. Sample size also affects stability because correlation estimates are less stable in small datasets. 

# Data Structure and Diagnostics

Three questions should be settled before calculating a coefficient.

## What is the Data Structure?

-   Each observation in one variable must correspond to the same observation in the other variable.
-   The sampling units should be independent.
-   Correlation is calculated for one variable pair at a time, even if many pairs are later assembled into a matrix.

## What is the Relationship Form?

-   Pearson focuses on linear association.
-   Spearman and Kendall focus on ordered association.
-   A scatterplot is the main diagnostic because it shows whether the pattern is linear, monotonic, clustered, or broken by outliers.

## What's the Data Quality?

-   Outliers can distort any coefficient, but Pearson is especially sensitive to them.
-   Site, time, transect, quadrat, or repeated-measures structure can create correlations that do not reflect the biological question of interest.

## Secondary Considerations

Distributional shape is less important than relationship form, outliers, and independence. Pearson is most sensitive to non-linearity and influential points. Normality is not the first issue to inspect.

# R Functions

The main function in this chapter is `cor.test()`. Use:

-   `method = "pearson"` for Pearson's product-moment correlation;
-   `method = "spearman"` for Spearman's rank correlation;
-   `method = "kendall"` for Kendall's rank correlation.

The function `cor()` calculates the coefficient itself and is useful for pairwise matrices. The inferential version, `cor.test()`, is usually more useful in worked examples because it returns the coefficient, test statistic, and *p*-value.

::: {.callout-note collapse="true" appearance="simple"}
## Mathematical Detail

For Pearson correlation:

$$r_{xy} = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n}(x_i - \bar{x})^2}\sqrt{\sum_{i=1}^{n}(y_i - \bar{y})^2}}$$ {#eq-pearson-r}

For Spearman rank correlation:

$$\rho = 1 - \frac{6 \sum_{i=1}^n d_i^2}{n(n^2-1)}$$ {#eq-spearman-rho}

For Kendall rank correlation:

$$\tau = \frac{n_c - n_d}{\binom{n}{2}}$$ {#eq-kendall-tau}

These formulas show the structure of the coefficients. In practice, first decide which coefficient matches the pattern in the data.
:::

# Pearson Correlation

Pearson's $r$ measures the strength and direction of linear association. It is appropriate when the relationship is reasonably well described by a straight line and no small set of outliers dominates the pattern.

# Example 1: Pearson Correlation in *Ecklonia maxima*

I begin by asking whether the length of *Ecklonia maxima* stipes tends to increase together with frond length.

```{r code-read-ecklonia}
ecklonia <- read_csv(file.path("..", "..", "data", "BCB744", "ecklonia.csv"))
```

## Do an Exploratory Data Analysis (EDA)

```{r code-ecklonia-summary}
ecklonia |>
  summarise(
    n = n(),
    mean_stipe_length = mean(stipe_length),
    sd_stipe_length = sd(stipe_length),
    mean_frond_length = mean(frond_length),
    sd_frond_length = sd(frond_length)
  )
```

```{r fig-corr1}
#| fig-cap: "Scatterplot showing the relationship between _Ecklonia maxima_ stipe length and frond length. The fitted line is included only to show the overall linear tendency."
#| fig-width: 4
#| fig-height: 3

r_print <- paste0(
  "r = ",
  round(cor(ecklonia$stipe_length, ecklonia$frond_length), 2)
)

ggplot(data = ecklonia, aes(x = stipe_length, y = frond_length)) +
  geom_smooth(method = "lm", colour = "blue3", se = FALSE, linewidth = 1) +
  geom_point(size = 2.7, colour = "red3", shape = 16) +
  geom_label(x = 300, y = 240, label = r_print) +
  labs(x = "Stipe length (cm)", y = "Frond length (cm)")
```

The scatterplot shows a clear positive linear trend. The points are not tightly packed around the fitted line, but the relationship is straight enough that Pearson's coefficient is appropriate. The main diagnostic here is the linear form of the line and univariate normality is secondary.

## State the Hypotheses

$$H_{0}: \rho = 0$$
$$H_{a}: \rho \ne 0$$

Here $\rho$ is the population Pearson correlation coefficient.

## Apply the Test

```{r code-cor-test-x-ecklonia}
cor.test(
  x = ecklonia$stipe_length,
  y = ecklonia$frond_length,
  use = "everything",
  method = "pearson"
)
```

## Interpret the Results

The correlation is positive and moderately strong. Longer stipes tend to be associated with longer fronds. The coefficient is large enough to be biologically informative, and the *p*-value is well below 0.001, so we reject $H_0$.

The effect size is the coefficient itself. Here $r = 0.65$ indicates a fairly strong linear association for a biological dataset of this kind. That does not imply causation. It shows co-variation.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

The linear association between stipe length and frond length in *Ecklonia maxima* was assessed with a Pearson product-moment correlation.

**Results**

Stipe length and frond length in *Ecklonia maxima* were positively correlated (Pearson correlation: $r = 0.65$, $n = 26$, $p < 0.001$), indicating that kelps with longer stipes also tended to have longer fronds.

**Discussion**

The two morphological variables co-vary strongly in these data. The result shows association only. It does not identify a causal mechanism.
:::

# Spearman Rank Correlation

When the relationship is ordered but not especially linear, Spearman's $\rho$ is often the better choice. It replaces the raw values with ranks and asks whether the two variables tend to increase together in rank order.

# Example 2: Spearman's $\rho$ When the Rank Pattern Is Clearer Than the Linear One

In the `ecklonia` data, the relationship between stipe diameter and primary blade length is not especially well described by a straight line, but there is still a biological suggestion that thicker stipes tend to occur with longer primary blades.

## Do an Exploratory Data Analysis (EDA)

As I always do, I start with a scatterplot.

```{r code-ecklonia-spearman-summary}
ecklonia |>
  summarise(
    n = n(),
    mean_stipe_diameter = mean(stipe_diameter),
    sd_stipe_diameter = sd(stipe_diameter),
    mean_primary_blade_length = mean(primary_blade_length),
    sd_primary_blade_length = sd(primary_blade_length)
  )
```

```{r fig-ecklonia-spearman}
#| fig-cap: "Relationship between stipe diameter and primary blade length in _Ecklonia maxima_. The fitted line is shown only as a visual guide to the overall trend."
#| fig-width: 4
#| fig-height: 3

ggplot(ecklonia, aes(x = stipe_diameter, y = primary_blade_length)) +
  geom_point(shape = 1, colour = "dodgerblue4") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Stipe diameter (mm)", y = "Primary blade length (cm)")
```

The scatterplot suggests an increasing pattern, but the spacing around a straight line is uneven and a few observations influence the fitted line strongly. The biological question is about ordered increase rather than precise linear scaling. That points to Spearman rather than Pearson.

## State the Hypotheses

$$H_{0}: \rho_{s} = 0$$
$$H_{a}: \rho_{s} \ne 0$$

Here $\rho_s$ is the population Spearman rank-correlation coefficient.

## Apply the Test

```{r code-cor-test-ecklonia-spearman}
cor.test(
  ecklonia$stipe_diameter,
  ecklonia$primary_blade_length,
  method = "spearman",
  exact = FALSE
)
```

It is useful to compare this with Pearson's correlation on the same variable pair:

```{r code-cor-test-ecklonia-spearman-compare}
cor.test(
  ecklonia$stipe_diameter,
  ecklonia$primary_blade_length,
  method = "pearson"
)
```

## Interpret the Results

Spearman's $\rho$ shows a moderate positive rank association. In these data, the rank-based signal is clearer than the strictly linear one. Pearson understates the pattern because it is asking a different question.

This is the main reason to use Spearman: the biological pattern is ordered, but the raw spacing around a line is not especially stable.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

A Spearman rank correlation was used to assess the monotonic association between stipe diameter and primary blade length in *Ecklonia maxima*. A rank-based method was preferred because the visual pattern suggested ordered increase without a clean linear form.

**Results**

Stipe diameter and primary blade length showed a moderate positive rank association (Spearman correlation: $\rho = 0.51$, $n = 26$, $p < 0.01$), indicating that kelps with thicker stipes also tended to have longer primary blades. On the same variable pair, the Pearson correlation was weaker and not statistically convincing at the 5% level ($r = 0.32$, $p > 0.05$).

**Discussion**

Spearman's $\rho$ is more informative here because the biological conclusion concerns ordered increase rather than exact linear scaling.
:::

# Kendall Rank Correlation

Kendall's $\tau$ measures concordance in rank ordering. It asks whether pairs of observations tend to agree in how they are ordered on the two variables.

# Example 3: Kendall's $\tau$ for Concordance in Rank Order

Suppose my question is whether longer primary blades also tend to be wider in the same rank order. This is a direct concordance question, which is the natural setting for Kendall's $\tau$.

## Do an Exploratory Data Analysis (EDA)

```{r fig-ecklonia-kendall}
#| fig-cap: "Relationship between primary blade length and primary blade width in _Ecklonia maxima_. The question here is whether longer blades also tend to rank as wider blades."
#| fig-width: 4
#| fig-height: 3

ggplot(ecklonia, aes(x = primary_blade_length, y = primary_blade_width)) +
  geom_point(shape = 1, colour = "seagreen4") +
  labs(x = "Primary blade length (cm)", y = "Primary blade width (cm)")
```

The scatterplot shows a positive tendency, but the main point is not exact linear spacing. The question is whether larger values in one variable tend to be matched by larger values in the other. That is a concordance question, so Kendall's $\tau$ is appropriate.

## State the Hypotheses

$$H_{0}: \tau = 0$$
$$H_{a}: \tau \ne 0$$

Here $\tau$ is the population Kendall rank-correlation coefficient.

## Apply the Test

```{r code-cor-test-ecklonia-primary-blade-length}
cor.test(
  ecklonia$primary_blade_length,
  ecklonia$primary_blade_width,
  method = "kendall"
)
```

## Interpret the Results

Kendall's $\tau$ is positive, so the rank ordering is broadly consistent: longer blades also tend to be wider. The effect is moderate rather than strong, and the *p*-value is below 0.05, so we reject $H_0$.

Kendall is often less familiar than Spearman, but its interpretation is very direct when the scientific question is about agreement in rank order.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

Kendall's $\tau$ was used to assess concordance between primary blade length and primary blade width in *Ecklonia maxima*.

**Results**

Primary blade length and primary blade width showed a positive association in rank order (Kendall correlation: $\tau = 0.34$, $n = 26$, $p < 0.05$), indicating that longer blades also tended to be wider.

**Discussion**

Kendall's $\tau$ is useful when the scientific message concerns agreement in rank order rather than precise linear scaling.
:::

# Correlation Matrices and Heatmaps

Once a single pairwise relationship is understood, the same idea can be scaled to many continuous variables at once.

```{r code-ecklonia-sub}
ecklonia_sub <- ecklonia |>
  select(-species, -site, -ID)

ecklonia_sub <- ecklonia_sub[, order(colnames(ecklonia_sub))]
```

## Create the Correlation Matrix

```{r tbl-ecklonia-pearson}
#| column: page
#| tbl-cap: "A pairwise Pearson correlation matrix of the _Ecklonia_ dataset."
ecklonia_pearson <- round(cor(ecklonia_sub), 2)

ecklonia_pearson |>
  as.data.frame() |>
  tibble::rownames_to_column("variable") |>
  gt() |>
  fmt_number(columns = -variable, decimals = 2) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

By producing many pairwise comparisons at once, these matrices offer a useful exploratory tool. They identify strong positive and negative relationships amongst many variables. Some of these correlations will appear by chance and other will be real, so you will have to apply your expert judgement. A matrix should therefore guide further inspection and not be used as a finished inferential result.

## Visualise the Matrix

```{r fig-corr2}
#| fig-cap: "Pairwise correlations showing the strength of all Pearson correlations between variables as a scale from red (negative) to blue (positive)."
#| fig-width: 6
#| fig-height: 4

ecklonia_pearson[upper.tri(ecklonia_pearson)] <- NA
corrplot(ecklonia_pearson, method = "circle", na.label.col = "white")
```

```{r fig-corr3}
#| fig-cap: "Pairwise correlations of the _Ecklonia_ dataset visualised as a heatmap in **ggplot2**."
#| fig-width: 6
#| fig-height: 4

ecklonia_pearson |>
  as.data.frame() |>
  mutate(x = rownames(ecklonia_pearson)) |>
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "r"
  ) |>
  ggplot(aes(x, y, fill = r)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1, 1),
    na.value = "grey95",
    space = "Lab",
    name = "r"
  ) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  coord_fixed()
```

The matrix and heatmap are useful for scanning the dataset quickly, but they should be treated with caution. They do not control the false-positive rate across all displayed comparisons, and they do not explain why a correlation exists.

# Correlation and Non-Independent Data

Even the correct coefficient can mislead when the observations are not independent. This is common in ecology because measurements are often grouped within sites, transects, quadrats, times, or individuals.

# Example 4: A Correlation Can Be Modified by Site Structure

I again use the kelp data.

```{r code-ecklonia-site-correlation}
ecklonia |>
  group_by(site) |>
  summarise(
    n = n(),
    r_site = cor(stipe_length, epiphyte_length)
  )
```

```{r fig-ecklonia-site-structure}
#| fig-cap: "Relationship between stipe length and epiphyte length in _Ecklonia maxima_, coloured by site. The black dashed line is the naive overall fit, while the coloured lines show the within-site trends."
#| fig-width: 4
#| fig-height: 3

ggplot(ecklonia, aes(x = stipe_length, y = epiphyte_length, colour = site)) +
  geom_point(shape = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  geom_smooth(aes(group = 1),
              method = "lm",
              se = FALSE,
              colour = "black",
              linetype = "dashed",
              linewidth = 0.6) +
  labs(x = "Stipe length (cm)", y = "Epiphyte length (cm)", colour = "Site")
```

The pooled correlation across all observations is clearly positive, but the plot also shows clustering by site. Part of the overall association reflects site differences rather than individual-level co-variation.

This is why independence must be taken seriously. A naive pooled coefficient mixes at least two sources of structure:

-   variation among individuals within sites;
-   variation between sites.

If the biological question is about site-level differences, then site should be modelled explicitly. If the question is about individual-level association, a grouped or mixed-effects model is more defensible than a pooled correlation.

# Correlation Versus Regression

Correlation and regression are related, but they do different jobs.

-   **Correlation** quantifies how strongly two variables vary together.
-   **Regression** estimates how the expected value of a response changes with a predictor.

Both are rooted in covariance, but Pearson's correlation standardises covariance into a unitless coefficient while simple linear regression uses the same covariance structure to estimate a slope. See the next chapter.

If the question is "do these variables co-vary?", correlation is appropriate. If the question is "how much does the response change when the predictor changes?", the analysis has already become a regression.

# If Assumptions Fail

If the relationship is not linear, the variables are ordinal, or the data contain influential outliers, Pearson correlation may not be appropriate. In such cases:

-   use Spearman or Kendall for rank-based association;
-   inspect the scatterplot before trusting any coefficient;
-   ask whether site, time, or repeated-measures structure is inflating the coefficient;
-   treat matrices as exploratory summaries;
-   move to regression if the real question is one of response and predictor, or if confounding and grouped structure need to be handled directly.

Assumption checking itself is discussed in [Chapter 6](06-assumptions-and-transformations.qmd).

:::: {.callout-important}
## Self-Assessment Task 9-1

Find **two datasets** of your own and do a full correlation analysis on each. Briefly describe the data and why they exist. State the hypotheses, do an EDA, make exploratory figures, choose and justify the appropriate correlation method, assess assumptions, and write up the results in publication style.

`r if (params$hide_answers) "::: {.content-hidden}"`

**Answer**

A strong answer should include:

1. **Dataset and variable choice**  
   Identify two biologically meaningful variable pairs. Explain why the question is one of association rather than response modelling.
2. **Hypotheses**  
   For Pearson correlation, for example:

$$H_0: \rho = 0$$
$$H_a: \rho \ne 0$$

   For rank-based methods, state the equivalent null for $\rho_s$ or $\tau$.
3. **EDA and visualisation**  
   Show scatterplots first. Comment on whether the relationship is linear, monotonic, clustered, or outlier-driven. Add summary statistics where useful.
4. **Choice of method**  
   Justify Pearson, Spearman, or Kendall according to the observed pattern. Do not choose mechanically.
5. **Assumption checks**  
   Discuss the relevant assumptions: independence, relationship form, and outliers. Normality is secondary for Pearson and not central for rank-based methods.
6. **Reporting**  
   Report the coefficient, sample size, p-value, and biological interpretation. Distinguish statistical support from causal interpretation.

A strong submission shows that the student understands why correlation is appropriate, why a particular coefficient was chosen, and what the result does and does not mean biologically.

`r if (params$hide_answers) ":::"`
::::

# Summary

Correlation quantifies association without fitting a response model. The working sequence is:

1. plot the data;
2. decide whether the question is really about association;
3. choose Pearson, Spearman, or Kendall according to the pattern in the data;
4. check independence before trusting the coefficient;
5. interpret sign, magnitude, and uncertainty together;
6. treat pairwise matrices as exploratory, not definitive.
