---
title: "9. Correlation and Association"
subtitle: "Quantifying Relationships Without Imposing a Response Model"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
number-sections: true
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_grey(base_size = 8)
)
```

```{r code-libraries, echo=FALSE}
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(corrplot)
library(kableExtra)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   Correlation as a measure of association
-   Pearson's product-moment correlation
-   Spearman's rank correlation
-   Kendall's rank correlation
-   Pairwise correlation matrices and heatmaps
-   Correlation versus regression
:::

::: {.callout-note appearance="simple"}
## Cheatsheet

Find here a [Cheatsheet](../../docs/Methods_cheatsheet_v1.pdf) on statistical methods.
:::

::: {.callout-note appearance="simple"}
## Slides with Additional Information About Correlations

-   [Slide deck 1](../../slides/BCB744_Correlations_slides--1-.html)
-   [Slide deck 2](../../slides/BCB744_Correlations_slides--2-.html)
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   Task H
:::

# Introduction

In the previous chapters we focused mainly on comparing means across groups. Sometimes, however, our question is not whether groups differ, but whether two continuous variables tend to vary together. For example, do longer kelp stipes also tend to have longer fronds? Do warmer waters also contain less dissolved oxygen? Do larger animals tend to have larger home ranges?

These are questions about **association**. The most familiar tool for this purpose is **correlation**, which quantifies the strength and direction of the relationship between two variables. Unlike regression, correlation does not formally assign one variable the role of response and the other the role of predictor. That distinction matters, because correlation is descriptive while regression is explicitly model-based.

Correlation analysis is therefore used to quantify the *strength and direction of the linear relationship between two continuous variables*. In some cases rank-based versions of correlation can also be used with ordinal data or with data that violate the assumptions of Pearson's correlation.

This chapter therefore acts as a bridge. It closes the sequence of core inferential tools by showing how association can be quantified, but it also prepares the way for the regression chapters that follow.

# Key Concepts

These ideas define the logic of the chapter before we move into the worked examples.

-   **Correlation** quantifies the strength and direction of association between variables.
-   **Association is not causation** because correlation alone does not show that one variable causes changes in the other.
-   **Correlation is not regression** because correlation does not specify a formal response and predictor.
-   **Pearson, Spearman, and Kendall** answer related but not identical questions.
-   **Strength, sign, and uncertainty** all matter when interpreting a correlation coefficient.
-   **Pairwise correlation matrices** are useful for exploring patterns in multivariate datasets.

# Nature of the Data and Assumptions

Correlation analysis has its own set of expectations about the data.

-   **Continuous variables** Pearson correlation is most appropriate when both variables are measured on a continuous scale.
-   **Pairwise data** Each value of one variable must correspond to the same observation in the other variable.
-   **Bivariate relationship** Correlation is used to assess the relationship between two variables at a time, even though we can inspect many pairwise correlations in a matrix.
-   **Linearity or monotonicity** Pearson correlation assumes an approximately linear relationship. Spearman's $\rho$ and Kendall's $\tau$ are more appropriate when the relationship is monotonic or when rank-based inference is preferred.
-   **No influential outliers** Outliers can have a strong influence on a correlation coefficient and can easily produce misleading results.
-   **Normality** Pearson correlation is most comfortable when the variables are reasonably well behaved and the bivariate relationship is approximately normal.
-   **Independence of sampling units** The sampling units themselves should be independent. Clustered, temporal, or repeated-measures data can violate this assumption.
-   **Random sampling** As with the other inferential methods, the strongest inference comes from an appropriate sampling design.

Before calculating a correlation coefficient, it is therefore necessary to make a scatterplot and inspect the overall pattern in the data.

# R Functions

The main function used in this chapter is `cor.test()`. It can be applied with different methods:

-   `method = "pearson"` for Pearson's product-moment correlation;
-   `method = "spearman"` for Spearman's rank correlation;
-   `method = "kendall"` for Kendall's rank correlation.

The base function `cor()` calculates the coefficient itself, but `cor.test()` is more useful for inference because it also returns a test statistic, a confidence interval where appropriate, and a *p*-value. When we want to inspect many pairwise correlations at once, `cor()` is useful for generating a correlation matrix.

# Pearson Correlation

Pearson's $r$ is a measure of the linear relationship between two variables. It assumes that the relationship between the variables is approximately linear, and is calculated as the ratio of the covariance between the variables to the product of their standard deviations:

$$r_{xy} = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n}(x_i - \bar{x})^2}\sqrt{\sum_{i=1}^{n}(y_i - \bar{y})^2}}$$ {#eq-pearson-r}

In @eq-pearson-r, $r_{xy}$ is the Pearson correlation coefficient, $x_i$ and $y_i$ are the observed values of the two variables for each observation $i$, $\bar{x}$ and $\bar{y}$ are the sample means of the two variables, and $n$ is the sample size.

The value of $r$ varies from `-1` to `1`, with `0` indicating no linear association, `1` showing a perfect positive association, and `-1` a perfect negative association.

## A Worked Pearson Example

In order to investigate correlations in biological data let us load the `ecklonia` dataset.

```{r code-read-ecklonia}
ecklonia <- read_csv(file.path("..", "..", "data", "BCB744", "ecklonia.csv"))
```

### Do an exploratory data analysis (EDA)

We will begin by asking whether the length of *Ecklonia maxima* stipes tends to increase together with frond length.

```{r code-ecklonia-summary}
ecklonia %>%
  summarise(
    n = n(),
    mean_stipe_length = mean(stipe_length),
    sd_stipe_length = sd(stipe_length),
    mean_frond_length = mean(frond_length),
    sd_frond_length = sd(frond_length)
  )
```

```{r fig-corr1}
#| fig-cap: "Scatterplot showing the relationship between _Ecklonia maxima_ stipe length and frond length. The line is from a linear model and helps show the overall linear tendency."
#| fig-width: 4
#| fig-height: 3

r_print <- paste0(
  "r = ",
  round(cor(ecklonia$stipe_length, ecklonia$frond_length), 2)
)

ggplot(data = ecklonia, aes(x = stipe_length, y = frond_length)) +
  geom_smooth(method = "lm", colour = "blue3", se = FALSE, linewidth = 1) +
  geom_point(size = 2.7, colour = "red3", shape = 16) +
  geom_label(x = 300, y = 240, label = r_print) +
  labs(x = "Stipe length (cm)", y = "Frond length (cm)") +
  theme_pubclean()
```

The scatterplot suggests that these data increase together at a roughly similar rate. The relationship is not perfect, but it is clearly positive. This makes a Pearson correlation sensible.

### State the hypotheses

For a Pearson correlation, the hypotheses are:

$$H_{0}: \rho = 0$$
$$H_{a}: \rho \ne 0$$

where $\rho$ is the population correlation coefficient.

### Apply the test

```{r code-cor-test-x-ecklonia}
cor.test(
  x = ecklonia$stipe_length,
  y = ecklonia$frond_length,
  use = "everything",
  method = "pearson"
)
```

### Interpret the results

The correlation is positive and reasonably strong. This indicates that longer stipes also tend to be associated with longer fronds. The *p*-value is small (`p < 0.001`), so the data provide strong evidence against the null hypothesis of no linear association.

The result does not imply that stipe length causes frond length or vice versa. It simply indicates that the two variables co-vary positively in these data.

### Reporting

::: {.callout-note appearance="simple"}
## Marine Biology Write-Up

**Methods**

> The linear association between stipe length and frond length in *Ecklonia maxima* was assessed with a Pearson product-moment correlation. The analysis focused on whether the two continuous variables co-varied positively across individuals.

**Results**

> Stipe length and frond length in *Ecklonia maxima* were positively correlated (Pearson correlation: $r = 0.65$, $n = 26$, $p < 0.001$), indicating that kelps with longer stipes also tended to have longer fronds.

**Discussion**

> The main discussion point is that the two morphological variables co-vary strongly. At the same time, the interpretation should remain disciplined: correlation alone does not establish a causal relationship between them.
:::

# Pairwise Correlation Matrices

If our dataset contains multiple continuous variables, as `ecklonia` does, we may investigate many pairwise correlations simultaneously. In order to do so we want a subset that contains only the continuous variables appropriate for Pearson correlation.

```{r code-ecklonia-sub}
ecklonia_sub <- ecklonia %>%
  select(-species, -site, -ID)

ecklonia_sub <- ecklonia_sub[, order(colnames(ecklonia_sub))]
```

## Create the Correlation Matrix

```{r tbl-ecklonia-pearson}
#| column: page
ecklonia_pearson <- round(cor(ecklonia_sub), 2)

ecklonia_pearson |>
  kbl(caption = "A pairwise correlation matrix of the _Ecklonia_ dataset.") |>
  kable_classic(full_width = FALSE)
```

Pairwise matrices are useful for identifying patterns and relationships between variables that may be hidden in the overall structure of a dataset. This is particularly useful when several variables have been measured on the same individuals.

## Visualise the Matrix

One way to visualise a pairwise matrix is with `corrplot()`:

```{r fig-corr2}
#| fig-cap: "Pairwise correlations showing the strength of all correlations between all variables as a scale from red (negative) to blue (positive)."
#| fig-width: 5
#| fig-height: 5

ecklonia_pearson[upper.tri(ecklonia_pearson)] <- NA
corrplot(ecklonia_pearson, method = "circle", na.label.col = "white")
```

We may also do this in **ggplot2** as a heatmap:

```{r fig-corr3}
#| fig-cap: "Pairwise correlations of the _Ecklonia_ dataset visualised as a heatmap in **ggplot2**."
#| fig-width: 5
#| fig-height: 5

ecklonia_pearson |>
  as.data.frame() |>
  mutate(x = rownames(ecklonia_pearson)) |>
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "r"
  ) |>
  ggplot(aes(x, y, fill = r)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1, 1),
    na.value = "grey95",
    space = "Lab",
    name = "r"
  ) +
  labs(x = NULL, y = NULL) +
  theme_grey() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  coord_fixed()
```

These graphs are useful for quickly identifying subsets of variables that are strongly related to one another, which can then point the path to further analysis and modelling.

# Spearman Rank Correlation

Spearman correlation is used to measure the strength and direction of the relationship between two variables based on their rank order. Unlike Pearson correlation, which assumes that the relationship between the variables is linear, Spearman correlation can be used when the relationship is monotonic but not necessarily linear. It is also useful when the data are not normally distributed, or when outliers make Pearson's correlation less reliable. The usual textbook formula is:

$$\rho = 1 - \frac{6 \sum_{i=1}^n d_i^2}{n(n^2-1)}$$ {#eq-spearman-rho}

In @eq-spearman-rho, $\rho$ is the Spearman correlation, $d_i$ is the difference between the ranks of the two variables for the $i^{th}$ observation, and $n$ is the sample size.

## Worked Spearman Example

To create an ordinal variable for demonstration, we will add a simple ranked body-length variable to `ecklonia`:

```{r code-ecklonia-length-as-numeric}
ecklonia$length <- as.numeric(
  cut(ecklonia$stipe_length + ecklonia$frond_length, breaks = 3)
)

head(select(ecklonia, species, site, stipe_length, frond_length, length), 10)
```

Now we correlate this ordinal `length` variable with `digits`:

```{r code-cor-test-ecklonia-length}
cor.test(ecklonia$length, ecklonia$digits, method = "spearman")
```

Spearman's $\rho$ asks whether the ranks of the two variables tend to increase together. This makes it a good option when the values themselves are not especially meaningful on a continuous scale, but their order is.

## Reporting

::: {.callout-note appearance="simple"}
## Marine Biology Write-Up

**Methods**

> A Spearman rank correlation was used to assess the monotonic association between ordinal body-length class and the number of digits. A rank-based method was preferred here because the analysis concerned ordered values rather than strict linear covariation on a continuous scale.

**Results**

> In a journal Results section, the Spearman coefficient, sample size, and *p*-value would be reported directly from the `cor.test()` output, together with a short statement on the direction of the rank association between body-length class and digit number.

**Discussion**

> Spearman's $\rho$ describes the tendency of ranks to increase together. It is therefore useful when relative order matters more than exact distance between measurements.
:::

# Kendall Rank Correlation

Kendall's $\tau$ is another non-parametric correlation method for assessing the strength and direction of the relationship between two variables. It is similar to Spearman's correlation, but it is calculated from the number of concordant and discordant pairs of observations:

$$\tau = \frac{n_c - n_d}{\binom{n}{2}}$$ {#eq-kendall-tau}

In @eq-kendall-tau, $\tau$ is Kendall's correlation coefficient, $n_c$ is the number of concordant pairs, $n_d$ is the number of discordant pairs, and $\binom{n}{2}$ is the total number of possible pairs.

Kendall's $\tau$ is useful for ordinal or non-parametric data and is robust to outliers and non-normal distributions.

## Worked Kendall Example

Let us inspect the normality of the `ecklonia_sub` variables so that we can identify a case where a rank-based correlation may be sensible:

```{r code-ecklonia-norm-ecklonia-sub}
ecklonia_norm <- ecklonia_sub %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(variable_norm = as.numeric(shapiro.test(value)[["p.value"]]))

ecklonia_norm
```

From this analysis we can see that some variables are not especially well behaved. Let us therefore examine the association between `primary_blade_length` and `primary_blade_width` using Kendall's $\tau$:

```{r code-cor-test-ecklonia-primary-blade-length}
cor.test(
  ecklonia$primary_blade_length,
  ecklonia$primary_blade_width,
  method = "kendall"
)
```

Here the coefficient is called Kendall's $\tau$, but the interpretation is similar to the other correlation coefficients: positive values indicate a positive association, negative values indicate a negative association, and values near zero indicate little or no association.

## Reporting

::: {.callout-note appearance="simple"}
## Marine Biology Write-Up

**Methods**

> Kendall's $\tau$ was used to assess the association between primary blade length and primary blade width because the variables were not especially well behaved under normal-theory assumptions and a rank-based measure of concordance was therefore more appropriate.

**Results**

> In a Marine Biology-style Results section, the Kendall coefficient, sample size, and *p*-value would be reported together with a short statement on whether longer blades also tended to be wider. The emphasis would be on concordance in rank order rather than on a linear effect size.

**Discussion**

> Kendall's $\tau$ is especially helpful when the scientific message concerns ordered agreement rather than precise linear scaling. That distinction should be made explicit so the result is not over-interpreted as if it were a regression slope.
:::

# Correlation Versus Regression

It is important not to confuse correlation with regression.

-   **Correlation** asks whether two variables vary together.
-   **Regression** asks whether a response variable changes systematically as a function of one or more predictors.

This distinction is not just technical. If your question is "do these variables co-vary?", correlation is appropriate. If your question is "how does the response change with the predictor?", then you are already in the territory of regression. That is why this chapter naturally leads into the next one.

# If Assumptions Fail

If the relationship is not linear, the variables are ordinal, or the data contain influential outliers, Pearson correlation may not be appropriate. In such cases:

-   use Spearman or Kendall correlation for rank-based association;
-   inspect the scatterplot carefully before interpreting any coefficient;
-   use pairwise matrices as exploratory tools rather than as substitutes for biological reasoning;
-   move to a regression model if the real question is one of response and predictor rather than simple association.

Assumption checking itself is discussed in [06-assumptions-and-transformations.qmd](06-assumptions-and-transformations.qmd).

# Summary

Correlation is a measure of association, not a model of response. The important workflow is therefore:

1. Plot the data.
2. Decide whether the question is really about association.
3. Choose Pearson, Spearman, or Kendall according to the data and relationship.
4. Interpret the coefficient, uncertainty, and *p*-value together.
5. Use pairwise correlation matrices as exploratory summaries when several continuous variables are available.
6. Do not confuse association with causation, or correlation with regression.
