---
title: "9. Correlation and Association"
subtitle: "Quantifying Relationships Without Imposing a Response Model"
author: "A. J. Smit"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_grey(base_size = 8)
)
```

```{r code-libraries, echo=FALSE}
library(tidyverse)
library(ggpubr)
library(corrplot)
library(gt)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   Correlation as a measure of association
-   Pearson's product-moment correlation
-   Spearman's rank correlation
-   Kendall's rank correlation
-   Pairwise correlation matrices and heatmaps
-   Why non-independent data can mislead a correlation analysis
-   Correlation versus regression
:::

::: {.callout-note appearance="simple"}
## Cheatsheet

Find here a [Cheatsheet](../../docs/Methods_cheatsheet_v1.pdf) on statistical methods.
:::

::: {.callout-note appearance="simple"}
## Slides with Additional Information About Correlations

-   [Slide deck 1](../../slides/BCB744_Correlations_slides--1-.html)
-   [Slide deck 2](../../slides/BCB744_Correlations_slides--2-.html)
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   Task H
:::

# Introduction

In the previous chapters we focused mainly on comparing means across groups. Sometimes, however, our question is not whether groups differ, but whether two continuous variables tend to vary together. For example, do longer kelp stipes also tend to have longer fronds? Do warmer waters also contain less dissolved oxygen? Do larger animals tend to have larger home ranges?

These are questions about **association**. The most familiar tool for this purpose is **correlation**, which quantifies the strength and direction of the relationship between two variables. Unlike regression, correlation does not formally assign one variable the role of response and the other the role of predictor. That distinction matters, because correlation is descriptive while regression is explicitly model-based.

Correlation analysis is therefore used to quantify the *strength and direction of the linear relationship between two continuous variables*. In some cases rank-based versions of correlation can also be used with ordinal data or with data that violate the assumptions of Pearson's correlation.

This chapter therefore acts as a bridge. It closes the sequence of core inferential tools by showing how association can be quantified, but it also prepares the way for the regression chapters that follow.

# Key Concepts

These ideas define the logic of the chapter before we move into the worked examples.

-   **Correlation** quantifies the strength and direction of association between variables.
-   **Association is not causation** because correlation alone does not show that one variable causes changes in the other.
-   **Correlation is not regression** because correlation does not specify a formal response and predictor.
-   **Pearson, Spearman, and Kendall** answer related but not identical questions.
-   **Strength, sign, and uncertainty** all matter when interpreting a correlation coefficient.
-   **Pairwise correlation matrices** are useful for exploring patterns in multivariate datasets.

# Nature of the Data and Assumptions

Correlation analysis has its own set of expectations about the data.

-   **Continuous variables** Pearson correlation is most appropriate when both variables are measured on a continuous scale.
-   **Pairwise data** Each value of one variable must correspond to the same observation in the other variable.
-   **Bivariate relationship** Correlation is used to assess the relationship between two variables at a time, even though we can inspect many pairwise correlations in a matrix.
-   **Linearity or monotonicity** Pearson correlation assumes an approximately linear relationship. Spearman's $\rho$ and Kendall's $\tau$ are more appropriate when the relationship is monotonic or when rank-based inference is preferred.
-   **No influential outliers** Outliers can have a strong influence on a correlation coefficient and can easily produce misleading results.
-   **Normality** Pearson correlation is most comfortable when the variables are reasonably well behaved and the bivariate relationship is approximately normal.
-   **Independence of sampling units** The sampling units themselves should be independent. Clustered, temporal, or repeated-measures data can violate this assumption.
-   **Random sampling** As with the other inferential methods, the strongest inference comes from an appropriate sampling design.

Before calculating a correlation coefficient, it is therefore necessary to make a scatterplot and inspect the overall pattern in the data.

# The Core Equations

Different correlation coefficients answer slightly different questions, but they all attempt to summarise how two variables vary together.

For Pearson correlation, the coefficient is:

$$r_{xy} = \frac{\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i=1}^{n}(x_i - \bar{x})^2}\sqrt{\sum_{i=1}^{n}(y_i - \bar{y})^2}}$$ {#eq-pearson-r}

In @eq-pearson-r, $r_{xy}$ is the Pearson correlation coefficient, $x_i$ and $y_i$ are the observed values of the two variables for observation $i$, $\bar{x}$ and $\bar{y}$ are the sample means, and $n$ is the sample size. The equation expresses the covariance between the variables scaled by their standard deviations, which is why $r$ is unitless and ranges from `-1` to `1`.

For Spearman rank correlation, the textbook version is:

$$\rho = 1 - \frac{6 \sum_{i=1}^n d_i^2}{n(n^2-1)}$$ {#eq-spearman-rho}

Here $\rho$ is Spearman's rank correlation, $d_i$ is the difference between the ranks of the two variables for the $i^{th}$ observation, and $n$ is the sample size.

For Kendall rank correlation, the coefficient is:

$$\tau = \frac{n_c - n_d}{\binom{n}{2}}$$ {#eq-kendall-tau}

In @eq-kendall-tau, $n_c$ is the number of concordant pairs, $n_d$ is the number of discordant pairs, and $\binom{n}{2}$ is the total number of possible pairs. Kendall's formulation therefore focuses directly on agreement in rank order.

# R Functions

The main function used in this chapter is `cor.test()`. It can be applied with different methods:

-   `method = "pearson"` for Pearson's product-moment correlation;
-   `method = "spearman"` for Spearman's rank correlation;
-   `method = "kendall"` for Kendall's rank correlation.

The base function `cor()` calculates the coefficient itself, but `cor.test()` is more useful for inference because it also returns a test statistic, a confidence interval where appropriate, and a *p*-value. When we want to inspect many pairwise correlations at once, `cor()` is useful for generating a correlation matrix.

# Pearson Correlation

Pearson's $r$ is a measure of the linear relationship between two variables, as defined in @eq-pearson-r. The value of $r$ varies from `-1` to `1`, with `0` indicating no linear association, `1` showing a perfect positive association, and `-1` a perfect negative association.

## Example 1: Pearson correlation in *Ecklonia maxima*

In order to investigate correlations in biological data let us load the `ecklonia` dataset.

```{r code-read-ecklonia}
ecklonia <- read_csv(file.path("..", "..", "data", "BCB744", "ecklonia.csv"))
```

### Do an exploratory data analysis (EDA)

We will begin by asking whether the length of *Ecklonia maxima* stipes tends to increase together with frond length.

```{r code-ecklonia-summary}
ecklonia %>%
  summarise(
    n = n(),
    mean_stipe_length = mean(stipe_length),
    sd_stipe_length = sd(stipe_length),
    mean_frond_length = mean(frond_length),
    sd_frond_length = sd(frond_length)
  )
```

```{r fig-corr1}
#| fig-cap: "Scatterplot showing the relationship between _Ecklonia maxima_ stipe length and frond length. The line is from a linear model and helps show the overall linear tendency."
#| fig-width: 4
#| fig-height: 3

r_print <- paste0(
  "r = ",
  round(cor(ecklonia$stipe_length, ecklonia$frond_length), 2)
)

ggplot(data = ecklonia, aes(x = stipe_length, y = frond_length)) +
  geom_smooth(method = "lm", colour = "blue3", se = FALSE, linewidth = 1) +
  geom_point(size = 2.7, colour = "red3", shape = 16) +
  geom_label(x = 300, y = 240, label = r_print) +
  labs(x = "Stipe length (cm)", y = "Frond length (cm)")
```

The scatterplot suggests that these data increase together at a roughly similar rate. The relationship is not perfect, but it is clearly positive. This makes a Pearson correlation sensible.

### State the hypotheses

For a Pearson correlation, the hypotheses are:

$$H_{0}: \rho = 0$$
$$H_{a}: \rho \ne 0$$

where $\rho$ is the population correlation coefficient.

### Apply the test

```{r code-cor-test-x-ecklonia}
cor.test(
  x = ecklonia$stipe_length,
  y = ecklonia$frond_length,
  use = "everything",
  method = "pearson"
)
```

### Interpret the results

The correlation is positive and reasonably strong. This indicates that longer stipes also tend to be associated with longer fronds. The *p*-value is small (`p < 0.001`), so the data provide strong evidence against the null hypothesis of no linear association.

The result does not imply that stipe length causes frond length or vice versa. It simply indicates that the two variables co-vary positively in these data.

### Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

The linear association between stipe length and frond length in *Ecklonia maxima* was assessed with a Pearson product-moment correlation. The analysis focused on whether the two continuous variables co-varied positively across individuals.

**Results**

Stipe length and frond length in *Ecklonia maxima* were positively correlated (Pearson correlation: $r = 0.65$, $n = 26$, $p < 0.001$), indicating that kelps with longer stipes also tended to have longer fronds.

**Discussion**

The main discussion point is that the two morphological variables co-vary strongly. At the same time, the interpretation should remain disciplined: correlation alone does not establish a causal relationship between them.
:::

# Pairwise Correlation Matrices

If our dataset contains multiple continuous variables, as `ecklonia` does, we may investigate many pairwise correlations simultaneously. In order to do so we want a subset that contains only the continuous variables appropriate for Pearson correlation.

```{r code-ecklonia-sub}
ecklonia_sub <- ecklonia %>%
  select(-species, -site, -ID)

ecklonia_sub <- ecklonia_sub[, order(colnames(ecklonia_sub))]
```

## Create the Correlation Matrix

```{r tbl-ecklonia-pearson}
#| column: page
#| tbl-cap: "A pairwise correlation matrix of the _Ecklonia_ dataset."
ecklonia_pearson <- round(cor(ecklonia_sub), 2)

ecklonia_pearson |>
  as.data.frame() |>
  tibble::rownames_to_column("variable") |>
  gt() |>
  fmt_number(columns = -variable, decimals = 2) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())
  )
```

Pairwise matrices are useful for identifying patterns and relationships between variables that may be hidden in the overall structure of a dataset. This is particularly useful when several variables have been measured on the same individuals. The numerical values in @tbl-ecklonia-pearson make it easier to identify especially strong positive or negative relationships before moving on to a visual summary.

## Visualise the Matrix

One way to visualise a pairwise matrix is with `corrplot()`:

```{r fig-corr2}
#| fig-cap: "Pairwise correlations showing the strength of all correlations between all variables as a scale from red (negative) to blue (positive)."
#| fig-width: 6
#| fig-height: 4

ecklonia_pearson[upper.tri(ecklonia_pearson)] <- NA
corrplot(ecklonia_pearson, method = "circle", na.label.col = "white")
```

We may also do this in **ggplot2** as a heatmap:

```{r fig-corr3}
#| fig-cap: "Pairwise correlations of the _Ecklonia_ dataset visualised as a heatmap in **ggplot2**."
#| fig-width: 6
#| fig-height: 4

ecklonia_pearson |>
  as.data.frame() |>
  mutate(x = rownames(ecklonia_pearson)) |>
  pivot_longer(
    cols = -x,
    names_to = "y",
    values_to = "r"
  ) |>
  ggplot(aes(x, y, fill = r)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(
    low = "blue",
    high = "red",
    mid = "white",
    midpoint = 0,
    limit = c(-1, 1),
    na.value = "grey95",
    space = "Lab",
    name = "r"
  ) +
  labs(x = NULL, y = NULL) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  coord_fixed()
```

These graphs are useful for quickly identifying subsets of variables that are strongly related to one another, which can then point the path to further analysis and modelling.

# Spearman Rank Correlation

Spearman correlation is used to measure the strength and direction of the relationship between two variables based on their rank order. Unlike Pearson correlation, which assumes that the relationship between the variables is linear, Spearman correlation can be used when the relationship is monotonic but not necessarily linear. It is also useful when the data are not normally distributed, or when outliers make Pearson's correlation less reliable. The usual textbook formula is given in @eq-spearman-rho.

## Example 2: Spearman's $\rho$ when the rank pattern is clearer than the linear one

## Do an Exploratory Data Analysis (EDA)

In the `ecklonia` data, the relationship between stipe diameter and primary blade length is not especially well described by a straight line, but there is still a biological suggestion that thicker stipes may tend to occur with longer primary blades. This is a good setting in which to compare Pearson's and Spearman's outcomes.

```{r code-ecklonia-spearman-summary}
ecklonia |>
  summarise(
    n = n(),
    mean_stipe_diameter = mean(stipe_diameter),
    sd_stipe_diameter = sd(stipe_diameter),
    mean_primary_blade_length = mean(primary_blade_length),
    sd_primary_blade_length = sd(primary_blade_length)
  )
```

```{r fig-ecklonia-spearman}
#| fig-cap: "Relationship between stipe diameter and primary blade length in _Ecklonia maxima_. The fitted line is shown only as a visual guide to the overall trend."
#| fig-width: 4
#| fig-height: 3

ggplot(ecklonia, aes(x = stipe_diameter, y = primary_blade_length)) +
  geom_point(shape = 1, colour = "dodgerblue4") +
  geom_smooth(method = "lm", se = FALSE, colour = "firebrick") +
  labs(x = "Stipe diameter (mm)", y = "Primary blade length (cm)")
```

The scatterplot suggests a generally increasing trend, but not a particularly tidy linear one. A few observations have disproportionate influence on the straight-line fit, and the biological question is more about ordered increase than about exact linear scaling, which is the kind of situation in which a rank-based correlation can be more informative.

## State the Hypotheses

The biological question is whether larger stipe diameters also tend to be associated with longer primary blades in rank order.

$$H_{0}: \rho_{s} = 0$$
$$H_{a}: \rho_{s} \ne 0$$

Here $\rho_{s}$ is the population Spearman rank-correlation coefficient. A value of zero means there is no monotonic association in rank between the two variables.

## Apply the Test

```{r code-cor-test-ecklonia-spearman}
cor.test(ecklonia$stipe_diameter,
         ecklonia$primary_blade_length,
         method = "spearman",
         exact = FALSE)
```

It is useful to compare this directly with Pearson's correlation on the same pair of variables:

```{r code-cor-test-ecklonia-spearman-compare}
cor.test(ecklonia$stipe_diameter,
         ecklonia$primary_blade_length,
         method = "pearson")
```

This contrast is exactly why Spearman's $\rho$ exists. Pearson's $r$ asks about linear covariation and is weak here, whereas Spearman's $\rho$ asks whether the variables tend to increase together in rank order. In these data, that rank-based question is better aligned with the pattern visible in @fig-ecklonia-spearman.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

A Spearman rank correlation was used to assess the monotonic association between stipe diameter and primary blade length in *Ecklonia maxima*. A rank-based method was preferred because the visual pattern suggested an ordered increase that was not especially well captured by a single straight-line relationship.

**Results**

Stipe diameter and primary blade length showed a moderate positive rank association (Spearman correlation: $\rho = 0.51$, $n = 26$, $p < 0.01$), indicating that kelps with thicker stipes also tended to have longer primary blades. On the same variable pair, the Pearson correlation was weaker and not statistically convincing at the 5% level ($r = 0.32$, $p > 0.05$), which shows that the main signal in these data is monotonic rather than cleanly linear.

**Discussion**

Spearman's $\rho$ is useful when the biological interpretation concerns ordered increase rather than a precise linear scaling. In this example, that distinction changes the practical conclusion of the analysis.
:::

# Kendall Rank Correlation

Kendall's $\tau$ is another non-parametric correlation method for assessing the strength and direction of the relationship between two variables. It is similar to Spearman's correlation, but it is calculated from the number of concordant and discordant pairs of observations, as shown in @eq-kendall-tau.

Kendall's $\tau$ is useful for ordinal or non-parametric data and is robust to outliers and non-normal distributions.

## Example 3: Kendall's $\tau$ for concordance in rank order

Let us inspect the normality of the `ecklonia_sub` variables so that we can identify a case where a rank-based correlation may be sensible:

```{r code-ecklonia-norm-ecklonia-sub}
ecklonia_norm <- ecklonia_sub %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(variable_norm = as.numeric(shapiro.test(value)[["p.value"]]))

ecklonia_norm
```

From this analysis we can see that some variables are not especially well behaved. Let us therefore examine the association between `primary_blade_length` and `primary_blade_width` using Kendall's $\tau$:

## State the Hypotheses

The biological question is whether longer primary blades also tend to be wider in rank order.

$$H_{0}: \tau = 0$$
$$H_{a}: \tau \ne 0$$

Here $\tau$ is the population Kendall rank-correlation coefficient. A value of zero means there is no concordance in the rank ordering of blade length and blade width.

## Apply the Test

```{r code-cor-test-ecklonia-primary-blade-length}
cor.test(
  ecklonia$primary_blade_length,
  ecklonia$primary_blade_width,
  method = "kendall"
)
```

Here the coefficient is called Kendall's $\tau$, but the interpretation is similar to the other correlation coefficients: positive values indicate a positive association, negative values indicate a negative association, and values near zero indicate little or no association.

## Reporting

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

Kendall's $\tau$ was used to assess the association between primary blade length and primary blade width because the variables were not especially well behaved under normal-theory assumptions and a rank-based measure of concordance was therefore more appropriate.

**Results**

Primary blade length and primary blade width showed a positive association in rank order (Kendall correlation: $\tau = 0.34$, $n = 26$, $p = 0.018$), indicating that longer blades also tended to be wider. The concordance was moderate rather than strong, but it was sufficiently consistent to support a positive rank-based relationship in these data.

**Discussion**

Kendall's $\tau$ is especially helpful when the scientific message concerns ordered agreement rather than precise linear scaling. That distinction should be made explicit so the result is not over-interpreted as if it were a regression slope.
:::

# Correlation Versus Regression

It is important not to confuse correlation with regression.

-   **Correlation** asks whether two variables vary together.
-   **Regression** asks whether a response variable changes systematically as a function of one or more predictors.

This distinction is not just technical. If your question is "do these variables co-vary?", correlation is appropriate. If your question is "how does the response change with the predictor?", then you are already in the territory of regression. That is why this chapter naturally leads into the next one.

# Correlation and Non-Independent Data

Even when the right coefficient has been chosen, correlation can still mislead if the observations are not independent. This matters a great deal in ecology because measurements are often grouped within sites, transects, quadrats, times, or individuals.

In the `ecklonia` data, the observations come from two sites. If individuals within a site resemble one another, then a naive overall correlation may partly reflect site structure rather than an individual-level biological relationship.

## Example 4: A correlation can be inflated by site structure

### Do an Exploratory Data Analysis (EDA)

```{r code-ecklonia-site-correlation}
ecklonia |>
  group_by(site) |>
  summarise(
    n = n(),
    r_site = cor(stipe_length, epiphyte_length)
  )
```

```{r fig-ecklonia-site-structure}
#| fig-cap: "Relationship between stipe length and epiphyte length in _Ecklonia maxima_, coloured by site. The black dashed line is the naive overall fit, while the coloured lines show the within-site trends."
#| fig-width: 4
#| fig-height: 3

ggplot(ecklonia, aes(x = stipe_length, y = epiphyte_length, colour = site)) +
  geom_point(shape = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  geom_smooth(aes(group = 1),
              method = "lm",
              se = FALSE,
              colour = "black",
              linetype = "dashed",
              linewidth = 0.6) +
  labs(x = "Stipe length (cm)", y = "Epiphyte length (cm)", colour = "Site")
```

The pooled correlation across all 26 observations is clearly positive, but @fig-ecklonia-site-structure shows that the data are also clustered by site. The within-site relationships are not identical, and part of the overall association comes from the sites differing in their average sizes.

### Interpret the results

This does not mean the pooled correlation is "wrong". It means its meaning is limited. The naive overall coefficient mixes at least two sources of structure:

-   how individuals vary within a site; and
-   how sites differ from one another.

If the biological question is about site-level differences, then averaging or modelling site explicitly may be appropriate. If the question is about individual-level association, then a grouped or mixed-effects framework is more defensible than a simple pooled correlation.

This is one reason a correlation coefficient should never be interpreted without first understanding the sampling design.

# If Assumptions Fail

If the relationship is not linear, the variables are ordinal, or the data contain influential outliers, Pearson correlation may not be appropriate. In such cases:

-   use Spearman or Kendall correlation for rank-based association;
-   ask whether the apparent pattern is partly driven by site, time, or repeated-measures clustering;
-   inspect the scatterplot carefully before interpreting any coefficient;
-   use pairwise matrices as exploratory tools rather than as substitutes for biological reasoning;
-   move to a regression model if the real question is one of response and predictor rather than simple association.

Assumption checking itself is discussed in [06-assumptions-and-transformations.qmd](06-assumptions-and-transformations.qmd).

# Summary

Correlation is a measure of association, not a model of response. The important workflow is therefore:

1. Plot the data.
2. Decide whether the question is really about association.
3. Choose Pearson, Spearman, or Kendall according to the data and relationship.
4. Check whether the observations are genuinely independent before trusting the coefficient.
5. Interpret the coefficient, uncertainty, and *p*-value together.
6. Use pairwise correlation matrices as exploratory summaries when several continuous variables are available.
7. Do not confuse association with causation, or correlation with regression.
