---
title: "14. Collinearity, Confounding, and Measurement Error"
subtitle: "Three Threats to Interpretable Regression Models"
author: "A. J. Smit"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6.5,
  fig.height = 4.5,
  out.width = "88%",
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

```{r code-knitr-opts-chunk-set, echo=FALSE}
library(tidyverse)
library(car)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   what collinearity is and why it matters;
-   how confounding differs from collinearity;
-   why measurement error weakens inference;
-   how proxy variables complicate interpretation;
-   what a practical response looks like when these problems appear.
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   None
:::

# Introduction

Regression models are sometimes taught as though the difficult part is fitting the correct formula and reading the summary table. In practice, some of the hardest problems sit below that surface. Predictors may overlap strongly, causal roles may be confused, and the variables we fit may be noisy measurements or only rough proxies for the processes we care about.

This chapter brings together three closely related but distinct problems:

1. **collinearity**, where predictors share too much information;
2. **confounding**, where a third variable distorts interpretation; and
3. **measurement error**, where the variables in the model are noisy versions of the quantities we would really like to observe.

All three problems matter because they threaten interpretation. A model may still run and produce a neat table of coefficients, but those coefficients may no longer mean what we would like them to mean.

# Key Concepts

Keep the following distinctions clear.

-   **Collinearity** is overlap among predictors.
-   **Confounding** is a problem of attribution, not merely correlation.
-   **Measurement error** weakens and can bias estimated effects.
-   **Proxy variables** are often useful but come with interpretive limits.
-   **Good design and good variable choice** solve more of these problems than software does.

# Collinearity

## What it is

Collinearity occurs when two or more predictors are strongly correlated with one another. In ecology this is common because environmental variables often co-vary through shared physical processes.

Examples include:

-   altitude and temperature;
-   nitrate and phosphate;
-   temperature and dissolved oxygen;
-   rainfall and river flow.

## Why it is a problem

Collinearity is mainly a problem of interpretation. A model can still fit the response well while the individual coefficients become:

-   unstable;
-   imprecise;
-   assigned the wrong sign;
-   difficult to interpret biologically.

The model struggles to partition shared explanatory information among overlapping predictors.

## A worked collinearity check

Using the seaweed example from the previous chapter, we can inspect the candidate climate predictors for overlap.

```{r code-collinearity-data}
sw <- read.csv("../../data/BCB743/seaweed/spp_df2.csv")
sw_ectz <- sw |>
  filter(bio == "ECTZ") |>
  select(Y, augMean, febRange, febSD, augSD, annMean)

cor(sw_ectz[, -1])
```

```{r code-collinearity-vif}
col_mod <- lm(Y ~ augMean + febRange + febSD + augSD + annMean,
              data = sw_ectz)

vif(col_mod)
```

The VIF values show that some predictors overlap strongly. This does not mean the model is mathematically invalid, but it does mean that the separate interpretation of those coefficients becomes hard to defend.

## Responding to collinearity

Possible responses include:

-   removing a redundant predictor;
-   choosing one variable over another based on biological reasoning;
-   combining correlated predictors into an index or ordination axis;
-   accepting some overlap if prediction is the main goal but avoiding causal over-interpretation.

There is no magical threshold that solves the problem automatically. The question is whether the overlap is small enough that the coefficients still mean something defensible.

# Confounding

Confounding is a different problem. It is not simply that two predictors overlap. It is that a third variable distorts the apparent relationship between a predictor of interest and the response.

A confounder is a variable that influences both:

-   the predictor of interest; and
-   the response.

This can create a spurious association or distort a real one. For example, if you relate abundance to temperature but nutrient supply is associated with both temperature and abundance, the temperature coefficient may partly absorb nutrient effects.

## Confounding is not the same as collinearity

These ideas are related but not identical.

-   **Collinearity** is about overlap among predictors in the data.
-   **Confounding** is about mistaken attribution of an effect.

Two variables can be highly collinear without one being a confounder. Equally, a confounder can matter even when the correlation structure does not look especially dramatic.

# Measurement Error

Standard linear regression assumes that predictors are measured without serious error. That assumption is rarely exactly true in field biology.

When a predictor is noisy, its estimated effect is often biased towards zero. This is called **attenuation bias**. In practical terms, a real effect may appear weaker and less stable than it truly is.

This matters because:

-   field instruments are imperfect;
-   environmental conditions fluctuate;
-   biological measurements can be imprecise;
-   some important variables are difficult or expensive to measure directly.

# Proxy Variables

Many biological models use proxy variables because the true mechanistic variables are unavailable.

Examples include:

-   altitude as a proxy for temperature or radiation;
-   distance from shore as a proxy for exposure;
-   chlorophyll as a proxy for productivity;
-   body size as a proxy for age or condition.

Proxy variables are often reasonable, but they limit interpretation. A proxy coefficient should not be read as if it identified one clean mechanism.

::: {.callout-note appearance="simple"}
## Mechanism Versus Proxy

If the question is physiological, use physiological drivers where possible. If the question is broad-scale pattern, a proxy may be perfectly acceptable. The important thing is that the variable should match the scale and aim of the question.
:::

# A Practical Workflow

When building and interpreting a regression model:

1. state the biological hypothesis first;
2. identify which predictors are mechanistic and which are proxies;
3. ask which variables could confound the relationship of interest;
4. inspect the predictor correlations and calculate VIF where appropriate;
5. remove or combine redundant predictors when interpretation is the goal;
6. be honest about causal limits when important variables are missing or poorly measured.

# Summary

-   Collinearity makes coefficients unstable because predictors share information.
-   Confounding is a problem of attribution rather than simple overlap.
-   Measurement error weakens inference and can bias estimated effects.
-   Proxy variables are often useful, but they limit what a coefficient can be said to mean.
-   These problems are handled best through strong biological reasoning and careful design before model fitting.

The next chapter turns from model construction to the question of whether a fitted model is actually behaving well: diagnostics, comparison, and evaluation.
