---
title: "15. Model Checking and Evaluation"
subtitle: "Diagnostics, Comparison, and Generalisation"
author: "A. J. Smit"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6.5,
  fig.height = 4.5,
  out.width = "88%",
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_minimal(base_size = 8)
)
ggplot2::theme_set(
  ggplot2::theme_bw(base_size = 8)
)
```

```{r code-knitr-opts-chunk-set, echo=FALSE}
library(tidyverse)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   Why model fitting is not the end of the analysis
-   Diagnostic checks for linear models
-   Influence, leverage, and residual structure
-   Comparing candidate models with information criteria
-   Overfitting and generalisation
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   None
:::

# Introduction

Once a model has been fitted, two further questions immediately follow:

1. **Is the model adequate for the data?**
2. **How does this model compare with plausible alternatives?**

These questions belong to model checking and model evaluation. A statistically significant model may still be poorly specified, violate assumptions, overfit the data, or generalise badly to new observations.

# Key Concepts

The following concepts structure the chapter.

-   **Model checking** asks whether a fitted model is adequate for the data and assumptions.
-   **Residuals** are central diagnostic tools for linear-model behaviour.
-   **Influence and leverage** identify observations that matter disproportionately.
-   **Model comparison** weighs plausible alternatives rather than treating one fitted model as final.
-   **Generalisation** matters because good in-sample fit does not guarantee useful prediction.

# Model Checking

Model checking asks whether the fitted model behaves in a way consistent with its assumptions and intended interpretation.

For ordinary linear models, the most useful checks usually involve **residuals**.

## Residual patterns

Residuals should not show strong systematic structure.

Problems to look for include:

-   curvature, suggesting non-linearity,
-   funnel shapes, suggesting heteroscedasticity,
-   unusual points with strong leverage,
-   heavy tails or severe asymmetry.

## Standard diagnostic plots

The standard four-plot diagnostic display in R remains a good starting point:

```{r}
#| eval: false
plot(fitted_model)
```

These typically show:

-   residuals vs fitted values,
-   normal Q-Q plot,
-   scale-location plot,
-   residuals vs leverage.

## Influence and leverage

Not all unusual points are equally important. Some observations have:

-   high **residuals**: poorly fitted values,
-   high **leverage**: unusual predictor values,
-   high **influence**: strong impact on the fitted coefficients.

An influential point deserves attention, but not automatic deletion. It may indicate:

-   a data entry problem,
-   measurement error,
-   an important biological extreme, or
-   a model that is too simple.

# Assumption Checks and Their Limits

Formal tests such as Shapiro-Wilk or Breusch-Pagan can be useful, but they should not replace graphical assessment and biological judgement.

In practice:

-   tiny deviations from normality are often unimportant,
-   large datasets can make trivial deviations look “significant,” and
-   visual residual structure is often more informative than a single test.

The real question is not whether a model is perfect, but whether it is **adequate for the inferential goal**.

# Model Comparison

Biological analysis often involves several plausible candidate models, each representing a different hypothesis.

Model comparison should therefore be framed as comparing **competing explanations**, not merely hunting for the best numerical fit.

## Information criteria

The most common tools are:

-   **AIC**,
-   **AICc** for small samples,
-   sometimes **BIC** for a stronger penalty on complexity.

These criteria balance:

-   model fit, and
-   model complexity.

Lower values indicate better relative support among the models being compared.

## What not to do

Automated stepwise procedures often produce unstable and hard-to-interpret models. They can also encourage post hoc storytelling around whichever terms happen to survive selection.

Candidate models should ideally be motivated by theory first and compared second.

# Overfitting and Generalisation

A model can fit the present data very well while performing poorly on new data. This is **overfitting**.

Overfitting occurs when the model captures noise as if it were signal. It is especially common when:

-   sample size is modest,
-   many predictors are included,
-   interactions are added freely,
-   model choice is heavily data-driven.

The key distinction is between:

-   **training performance**: fit to the data used to estimate the model,
-   **generalisation performance**: fit to new or unseen data.

## Cross-validation

Cross-validation provides one way to estimate predictive performance on unseen data.

In principle, the workflow is:

1. split the data into parts,
2. fit the model on one part,
3. test it on the held-out part,
4. repeat.

This is especially valuable when the goal of modelling is prediction rather than explanation.

# A Practical Workflow

After fitting a model:

1. Inspect residual plots.
2. Check for influential points.
3. Ask whether any detected pattern suggests a missing variable or wrong functional form.
4. Compare only biologically sensible candidate models.
5. Use information criteria or cross-validation according to the modelling goal.
6. Report uncertainty and limitations clearly.

# Summary

-   Fitting a model is only the middle of the analysis.
-   Residual diagnostics are essential for checking adequacy.
-   Influence and leverage can distort fitted results and must be inspected.
-   Model comparison should evaluate competing biological hypotheses, not just maximise fit.
-   Overfitting is a major risk when complexity outruns signal.

Good modelling requires both a plausible model and evidence that the model is behaving well.
