---
title: "19. Generalised Linear Models"
subtitle: "Extending Regression Beyond Normal Responses"
author: "A. J. Smit"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6.5,
  fig.height = 4.5,
  out.width = "88%",
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   Why ordinary linear models are not enough for many responses
-   The logic of families and link functions
-   Logistic regression for binary data
-   Poisson and related models for counts
-   What overdispersion tells us
:::

# Introduction

The linear model assumes a normally distributed response with constant variance. Many biological data do not have that structure. Counts, proportions, presences and absences, and success-failure outcomes require a broader framework. Generalised linear models (GLMs) provide that framework while retaining the central logic of regression.

A GLM combines three parts:

1. a response distribution;
2. a linear predictor;
3. a link function that connects the linear predictor to the expected response scale.

The important point is that GLMs do not replace the regression logic you have already learned. They extend it to different kinds of response variables.

# Key Concepts

-   **GLMs** extend regression to non-normal response variables.
-   **The family** should match the data-generating structure, such as binomial data for binary outcomes or Poisson data for counts.
-   **The link function** connects the linear predictor to the response scale.
-   **Interpretation still depends on the biological question**, not only on the model family.
-   **Overdispersion** is a warning sign that the simplest GLM may be inadequate.

# When This Method Is Appropriate

You should consider a GLM when:

-   the response is binary, such as alive/dead or present/absent;
-   the response is a proportion based on counts of successes and failures;
-   the response is a count and cannot sensibly be modelled with a normal distribution;
-   the variance changes with the mean in a way that a standard linear model does not handle well.

This chapter therefore picks up directly from the earlier proportion-testing work and places it in the broader modelling framework where binary and count responses can be analysed with predictors.

# The GLM Structure

A generalised linear model combines a response distribution with a linear predictor and a link function. A compact way to write this is:

$$g(\mu_i) = \eta_i = \alpha + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_p X_{pi}$$ {#eq-glm}

In @eq-glm, $\mu_i$ is the expected response for observation $i$, $g(\cdot)$ is the link function, and $\eta_i$ is the linear predictor.

# Common GLM Families

## Logistic regression

Use a **binomial GLM** when the response is binary or a proportion.

The most common link in introductory logistic regression is the logit link:

$$\log\left(\frac{p_i}{1-p_i}\right) = \alpha + \beta_1 X_{1i} + \cdots + \beta_p X_{pi}$$ {#eq-logit}

Equation @eq-logit says that the predictors act linearly on the log-odds scale, not directly on the probability scale.

Examples:

-   infection status as a function of temperature and host size;
-   settlement success as a function of habitat and season;
-   survival as a function of treatment.

## Poisson regression

Use a **Poisson GLM** when the response is a count.

For a Poisson GLM, the most common link is the log link:

$$\log(\mu_i) = \alpha + \beta_1 X_{1i} + \cdots + \beta_p X_{pi}$$ {#eq-poisson}

Equation @eq-poisson means that the coefficients are interpreted on the log-count scale unless they are transformed back to the response scale.

Examples:

-   number of individuals per quadrat;
-   number of flowers per plant;
-   number of parasite eggs per host.

## Overdispersed count models

If the count variance is much larger than the mean, a simple Poisson model may be too restrictive. In such cases, a **negative binomial model** or another overdispersion-aware model is often more appropriate.

# R Functions

The main function is `glm()`:

```{r}
#| eval: false
glm(response ~ predictors, family = ..., data = df)
```

A logistic regression might look like:

```{r}
#| eval: false
glm(cbind(successes, failures) ~ temperature + treatment,
    family = binomial,
    data = df)
```

or, for a binary response:

```{r}
#| eval: false
glm(presence ~ salinity + habitat,
    family = binomial,
    data = df)
```

A count model might look like:

```{r}
#| eval: false
glm(count ~ reef + depth,
    family = poisson,
    data = df)
```

# A Practical Workflow

The teaching workflow remains familiar:

1. identify the biological question;
2. identify the response structure;
3. choose a family that matches the response;
4. fit the model;
5. inspect model adequacy, especially for overdispersion and fit;
6. interpret coefficients on the correct scale.

The new difficulty is that the coefficients are often estimated on the **link scale**, so interpretation requires more care than in an ordinary linear model.

# Interpretation and Reporting

In GLMs, interpretation often needs to distinguish between:

-   the coefficient on the link scale;
-   the effect on the response scale;
-   the biological meaning of that effect.

For example, in logistic regression a positive coefficient means the log-odds of success increase with the predictor. In practice, it is often clearer to translate that into statements about the probability of success increasing or decreasing rather than reporting only the raw coefficient.

# Reporting

Even when the fitted model is not developed as a full worked example in this chapter, the reporting logic should follow the same pattern used elsewhere in the sequence: the model structure belongs in Methods, the fitted effect belongs in Results, and the biological meaning belongs in Discussion.

::: {.callout-note appearance="simple"}
## Marine Biology Write-Up

**Methods**

> Presence-absence data were analysed using a generalised linear model with a binomial error structure and logit link function. Predictor variables were selected to reflect the hypothesised effects of habitat and salinity on the probability of occurrence.

**Results**

> The fitted model indicated that the probability of occurrence changed systematically with the predictors included in the analysis. On the logit scale, the coefficients described the direction and relative strength of these effects, while back-transformation to the response scale showed how the estimated probability of occurrence increased or decreased across the observed predictor range.

**Discussion**

> In a biological interpretation, the emphasis should fall on the ecological meaning of the change in probability rather than on the link-scale coefficients alone. The value of the GLM framework is that it allows a response with non-normal error structure to be analysed in a way that remains statistically coherent and biologically interpretable.
:::

# Common Mistakes

Common mistakes include:

-   choosing a family because it is familiar rather than because it matches the response;
-   fitting Poisson models to strongly overdispersed counts;
-   interpreting link-scale coefficients as though they were ordinary linear slopes;
-   forgetting that diagnostics still matter in GLMs.

# Summary

-   GLMs extend regression to binary, proportional, and count responses.
-   The model family should match the response structure.
-   Logistic and Poisson models are the most common introductory GLMs.
-   Overdispersion is an important practical warning sign.
-   The regression logic is unchanged, but interpretation becomes more careful because of the link function.

This chapter broadens the modelling family. The next chapter turns from non-normal responses to relationships that are not well described by straight lines.
