---
title: "19. Generalised Linear Models"
subtitle: "Extending Regression Beyond Normal Responses"
author: "A. J. Smit"
date: last-modified
date-format: "YYYY/MM/DD"
reference-location: margin
---

```{r code-brewing-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>",
  warning = FALSE,
  message = FALSE,
  fig.asp = NULL,
  fig.align = "center",
  fig.retina = 2,
  dpi = 300
)

ggplot2::theme_set(
  ggplot2::theme_grey(base_size = 8)
)
```

::: {.callout-note appearance="simple"}
## In This Chapter

-   Why ordinary linear models are not enough for many responses
-   The logic of families and link functions
-   Logistic regression for binary data
-   Poisson and related models for counts
-   What overdispersion tells us
:::

::: {.callout-important appearance="simple"}
## Tasks to Complete in This Chapter

-   None
:::

# Introduction

The linear model assumes a normally distributed response with constant variance. Many biological data do not have that structure. Counts, proportions, presences and absences, and success-failure outcomes require a broader framework. Generalised linear models (GLMs) provide that framework while retaining the central logic of regression.

A GLM combines three parts:

1. a response distribution;
2. a linear predictor;
3. a link function that connects the linear predictor to the expected response scale.

The important point is that GLMs do not replace the regression logic you have already learned. They extend it to different kinds of response variables.

# Key Concepts

-   **GLMs** extend regression to non-normal response variables.
-   **The family** should match the data-generating structure, such as binomial data for binary outcomes or Poisson data for counts.
-   **The link function** connects the linear predictor to the response scale.
-   **Interpretation still depends on the biological question**, not only on the model family.
-   **Overdispersion** is a warning sign that the simplest GLM may be inadequate.

# When This Method Is Appropriate

You should consider a GLM when:

-   the response is binary, such as alive/dead or present/absent;
-   the response is a proportion based on counts of successes and failures;
-   the response is a count and cannot sensibly be modelled with a normal distribution;
-   the variance changes with the mean in a way that a standard linear model does not handle well.

This chapter therefore picks up directly from the earlier proportion-testing work and places it in the broader modelling framework where binary and count responses can be analysed with predictors.

# The Core Equation

A generalised linear model combines a response distribution with a linear predictor and a link function. A compact way to write this is:

$$g(\mu_i) = \eta_i = \alpha + \beta_1 X_{1i} + \beta_2 X_{2i} + \cdots + \beta_p X_{pi}$$ {#eq-glm}

In @eq-glm, $\mu_i$ is the expected response for observation $i$, $g(\cdot)$ is the link function, and $\eta_i$ is the linear predictor. This equation is the common template. Logistic and Poisson regression are then just special cases with different response distributions and different link functions.

# Common GLM Families

## Logistic regression

Use a **binomial GLM** when the response is binary or a proportion.

The most common link in introductory logistic regression is the logit link:

$$\log\left(\frac{p_i}{1-p_i}\right) = \alpha + \beta_1 X_{1i} + \cdots + \beta_p X_{pi}$$ {#eq-logit}

Equation @eq-logit says that the predictors act linearly on the log-odds scale, not directly on the probability scale.

Examples:

-   infection status as a function of temperature and host size;
-   settlement success as a function of habitat and season;
-   survival as a function of treatment.

## Poisson regression

Use a **Poisson GLM** when the response is a count.

For a Poisson GLM, the most common link is the log link:

$$\log(\mu_i) = \alpha + \beta_1 X_{1i} + \cdots + \beta_p X_{pi}$$ {#eq-poisson}

Equation @eq-poisson means that the coefficients are interpreted on the log-count scale unless they are transformed back to the response scale.

Examples:

-   number of individuals per quadrat;
-   number of flowers per plant;
-   number of parasite eggs per host.

## Overdispersed count models

If the count variance is much larger than the mean, a simple Poisson model may be too restrictive. In such cases, a **negative binomial model** or another overdispersion-aware model is often more appropriate.

# R Functions

The main function is `glm()`:

```{r}
#| eval: false
glm(response ~ predictors, family = ..., data = df)
```

A logistic regression might look like:

```{r}
#| eval: false
glm(cbind(successes, failures) ~ temperature + treatment,
    family = binomial,
    data = df)
```

or, for a binary response:

```{r}
#| eval: false
glm(presence ~ salinity + habitat,
    family = binomial,
    data = df)
```

A count model might look like:

```{r}
#| eval: false
glm(count ~ reef + depth,
    family = poisson,
    data = df)
```

# A Practical Workflow

The teaching workflow remains familiar:

1. identify the biological question;
2. identify the response structure;
3. choose a family that matches the response;
4. fit the model;
5. inspect model adequacy, especially for overdispersion and fit;
6. interpret coefficients on the correct scale.

The new difficulty is that the coefficients are often estimated on the **link scale**, so interpretation requires more care than in an ordinary linear model.

# Interpretation and Reporting

In GLMs, interpretation often needs to distinguish between:

-   the coefficient on the link scale;
-   the effect on the response scale;
-   the biological meaning of that effect.

For example, in logistic regression a positive coefficient means the log-odds of success increase with the predictor. In practice, it is often clearer to translate that into statements about the probability of success increasing or decreasing rather than reporting only the raw coefficient.

# A Compact Logistic Example

To make the reporting logic concrete, consider a small illustrative logistic-regression example in which occurrence depends on salinity and habitat.

```{r code-glm-demo-data, echo=FALSE}
set.seed(74419)

glm_demo <- data.frame(
  salinity = runif(120, 28, 38),
  habitat = factor(sample(c("reef", "sand"), 120, replace = TRUE))
  )

glm_demo$eta <- -7.6983417 +
  0.2532772 * glm_demo$salinity +
  ifelse(glm_demo$habitat == "sand", -1.0056404, 0)
glm_demo$probability <- plogis(glm_demo$eta)
glm_demo$presence <- rbinom(nrow(glm_demo), 1, glm_demo$probability)

glm_demo_mod <- glm(presence ~ salinity + habitat,
                    family = binomial,
                    data = glm_demo)
```

```{r code-glm-demo-summary}
summary(glm_demo_mod)
```

The fitted coefficients show that occurrence becomes more likely as salinity increases, while sand habitat has a lower probability of occurrence than reef habitat at the same salinity.

# Reporting

Even when a GLM chapter is mostly conceptual, the reporting logic should still follow the same pattern used elsewhere in the sequence: the model structure belongs in Methods, the fitted effect belongs in Results, and the biological meaning belongs in Discussion.

::: {.callout-note appearance="simple"}
## Write-Up

**Methods**

Presence-absence data were analysed using a generalised linear model with a binomial error structure and logit link function. Salinity and habitat were included as predictors to test whether occurrence probability changed across the salinity gradient and between reef and sand habitats.

**Results**

Occurrence probability increased significantly with salinity (logit coefficient $= 0.25 \pm 0.07$ SE, $z = 3.41$, $p < 0.001$) and was lower on sand than on reef habitat (logit coefficient $= -1.01 \pm 0.40$ SE, $z = -2.50$, $p = 0.012$). Across the observed salinity range, the fitted probability of occurrence for sand habitat rose from about 0.18 to 0.71, indicating a substantial but habitat-dependent increase in occurrence with salinity.

**Discussion**

In a biological interpretation, the emphasis should fall on the ecological meaning of the change in probability rather than on the link-scale coefficients alone. In this illustrative example, higher salinity increased the likelihood of occurrence, but occurrence remained consistently less likely on sand than on reef habitat at comparable salinity.
:::

# Common Mistakes

Common mistakes include:

-   choosing a family because it is familiar rather than because it matches the response;
-   fitting Poisson models to strongly overdispersed counts;
-   interpreting link-scale coefficients as though they were ordinary linear slopes;
-   forgetting that diagnostics still matter in GLMs.

# Summary

-   GLMs extend regression to binary, proportional, and count responses.
-   The model family should match the response structure.
-   Logistic and Poisson models are the most common introductory GLMs.
-   Overdispersion is an important practical warning sign.
-   The regression logic is unchanged, but interpretation becomes more careful because of the link function.

This chapter broadens the modelling family. The next chapter turns from non-normal responses to relationships that are not well described by straight lines.
