\documentclass[12pt]{article}

\usepackage[english]{babel}
\usepackage[portrait,a5paper,hmargin=2.5cm,vmargin=4cm]{geometry}
% \usepackage[babel,activate={true,nocompatibility},protrusion=true,tracking=true,expansion=false]{microtype}
\usepackage[babel,activate={true,nocompatibility},protrusion=true,tracking=true,expansion=true,factor=750,stretch=5,shrink=5]{microtype}   

\usepackage{fontspec}
    \defaultfontfeatures{SmallCapsFeatures = {
        Letters = SmallCaps,
        LetterSpace = 2.25,
        Numbers = Lining}}

    \setmainfont{Cochineal-Roman}[
        Path = /usr/local/texlive/2025/texmf-dist/fonts/opentype/public/cochineal/,
        Extension = .otf,
        BoldFont = Cochineal-Bold,
        ItalicFont = Cochineal-Italic,
        BoldItalicFont = Cochineal-BoldItalic,
        Numbers={Proportional,OldStyle}]
    \newfontfamily\dropcaps{TypographerWoodcutInitialsOne}

\usepackage{xcolor}
\definecolor{MyCrimson}{HTML}{880000}
\definecolor{MyGrey}{HTML}{080808}
\definecolor{MyBackground}{HTML}{FFFFEB}

\usepackage{pagecolor}
\pagecolor{MyBackground}

\usepackage{lettrine}
\renewcommand{\LettrineFontHook}{\dropcaps}

\pagestyle{empty}

\begin{document}
    \begin{quote}
        \begin{flushright}
            \itshape\large\textcolor{MyCrimson}
            {{\LARGE Everything~}{\Large written symbols can say has already passed \rlap{by\,…}}}
        \end{flushright}
    \end{quote}

    \lettrine[lines=4,findent=.25em,nindent=0em,lhang=.04,loversize=-.01]
    {\textcolor{MyCrimson}{T}}{\textcolor{MyGrey}{\kern .25pt hey are like tracks left by
    animals.} }\\[2pt]{\textcolor{MyGrey}{That is why the masters of
    meditation re\-fuse to accept that writings are final. The aim is to reach
    true being by means of those tracks, those letters, those signs; but
    reality itself is not a sign, and it leaves no tracks. It doesn't come to
    us by way of letters or words. We can go toward it, by following those
    words and letters back to whence they came from.}

\vspace{1ex}

    \noindent {\textcolor{MyGrey}{ But so long as we are preoccupied with symbols, theories and opinions, we will fail to reach the principle.}

\vspace{1ex}

    \noindent{}\kern .5pt—~But when we give up symbols and opinions, aren't we left in the utter nothingness of being?

\vspace{-3.5ex}

    \begin{quote}
        \hfill\itshape\large\textcolor{MyCrimson}{\noindent{}\kern .5pt—~Yes.}
    \end{quote}

\vspace{2\baselineskip}

    \begin{flushleft}
        {\textcolor{MyGrey}{
        \small
        \noindent\textsc{Kimura Kyūho} — \emph{Kenjutsu Fushigi Hen} \newline
        [On the Mysteries of Swordsmanship], 1768}}
    \end{flushleft}
\end{document}
