---
title: "Ecoevidence Doubs River Data Set"
subtitle: "Package Test"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  warning: true
  message: false
---

```{r}
#| label: setup

# install.packages(c("ade4", "vegan", "adespatial", "betapart"))

library(ade4)
library(vegan)
library(adespatial)
library(betapart)

set.seed(2026)
```

```{r}
#| label: inspect-doubs

data(doubs, package = "ade4")

names(doubs)

str(doubs, max.level = 1)

?doubs
```
This vignette demonstrates a typical ecological analysis workflow using the **ecoevidence** package. The Doubs River dataset, distributed with the **ade4** package, contains fish community data, environmental variables, and spatial coordinates collected from 30 sampling sites along the Doubs River in France.

```{r}
# Doubs River Workflow
## Load required packages
 warning=FALSE
library(ecoevidence)
library(ade4)
library(vegan)
library(adespatial)
library(betapart)

set.seed(2026)
```

## Load the Doubs River dataset

The Doubs River dataset is included in the **ade4** package and provides four data objects:

- **fish** – fish species abundance matrix
- **env** – environmental variables
- **xy** – spatial coordinates
- **species** – taxonomic information for each fish species

```{r load-data}
data(doubs, package = "ade4")
```

## Inspect the dataset

Before beginning an analysis, it is good practice to inspect the structure of the dataset.

```{r inspect-data}
names(doubs)
str(doubs, max.level = 1)
```

## Extract the data components

For clarity, extract each component into a separate object.

```{r extract-data}
fish <- doubs$fish
environment <- doubs$env
coordinates <- doubs$xy
species_info <- doubs$species
```

Inspect their dimensions.

```{r dimensions}
dim(fish)
dim(environment)
dim(coordinates)
dim(species_info)
```

Preview the first few observations.

```{r preview}
head(fish)
head(environment)
head(coordinates)
head(species_info)
```

## Create an EcoStudy object

The `observe()` function creates an **EcoStudy** object that stores the community, environmental, and spatial data together for subsequent analyses.

```{r create-study}
study <- ecoevidence::observe(
  species = doubs$fish,
  environment = doubs$env
)
```
## Summarise the study

The `eco_summary()` function provides a concise overview of the ecological dataset, including key characteristics of the species, environmental, and spatial data stored within the `EcoStudy` object.

```{r}
eco_summary(study)

```
```{r}
eco_nmds(study)
```
Hypothesis Test
```{r}
hyp1 <- eco_hypothesis(

    name = "Altitude shapes fish communities",

    response = "species",

    predictor = "alt"

)

hyp1
```
```{r}
test <- eco_test(study, hyp1)

test
```
```{r}
eco_test
```

```{r}
test1 <- eco_test(study, hyp1)

str(test1)
```
```{r}
test1$fitted

test1$statistics
```

Does altitude influence fish community composition?

```{r}
hyp1 <- eco_hypothesis(
  name = "Altitude shapes fish communities",
  response = "species",
  predictor = "alt"
)

test1 <- eco_test(study, hyp1)

test1
```

Does water pH change with altitude?

```{r}
hyp2 <- eco_hypothesis(
  name = "Altitude influences pH",
  response = "pH",
  predictor = "alt"
)

test2 <- eco_test(study, hyp2)

test2
```

Does dissolved oxygen vary with altitude?

```{r}
hyp3 <- eco_hypothesis(
  name = "Altitude influences dissolved oxygen",
  response = "oxy",
  predictor = "alt"
)

test3 <- eco_test(study, hyp3)

test3
```
ECOSCORE Test
```{r}
score1 <- eco_score(test1)

score1
```


